---
title: "Tables"
format:
  docx:
    reference-doc: ../../04.references/style-formal.docx
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| output: false

## Journal of Applied Entomology asks for tables as a separate, editable file. These are
## read back from the CSVs the manuscript wrote as it rendered, not recomputed, so this
## file cannot report a different number from the article it accompanies. If a CSV is
## missing, render the manuscript first.
library(knitr)
TBL <- file.path("..", "..", "03.outputs", "TBL")
get <- function(n) {
  f <- file.path(TBL, paste0(n, ".csv"))
  if (!file.exists(f))
    stop("missing ", f, ". Render beetle-topography-wind-study-short.qmd first.")
  read.csv(f, check.names = FALSE)
}
```

<!-- Footnote symbols follow the journal's rule: dagger, double dagger, section and
     pilcrow in that order, with asterisks reserved for P-values. Statistical measures
     are named in the column headings, as the guidelines require. -->

# Table 1

Landscape attributes entered in this study, the direction expected of each, and the
reasoning behind that expectation.

```{r}
kable(get("table-1-hypotheses"), booktabs = TRUE, align = "lll")
```

# Table 2

Moderate-to-high beetle disturbance by year inside the study perimeter.

```{r}
kable(get("table-2-prevalence"), booktabs = TRUE, align = "lrr")
```

# Table 3

Stand structure across the study perimeter, from the Vegetation Resources Inventory.

```{r}
kable(get("table-3-stand-structure"), booktabs = TRUE, align = "lrrrr")
```

# Table 4

Model comparison. Each model adds one pathway to the previous.

```{r}
kable(get("table-4-model-comparison"), booktabs = TRUE, align = "lrrr")
```

# Table 5

Full model, continuous terms. Coefficients are log-odds per standard deviation on a
class-balanced sample, so the intercept is not landscape prevalence. SE is the standard
error of the coefficient.

```{r}
kable(get("table-5-full-model"), booktabs = TRUE, align = "lrrrr")
```

# Table 6

The wind terms under two specifications, both with previous-year pressure in the model.
M4 carries station wind, which is flat within a year and identified only across years.
M5 carries the terrain-resolved MicroMet field, which varies within the year and is
identified in space.

```{r}
kable(get("table-6-wind-terms"), booktabs = TRUE, align = "lllrr")
```

# Table 7

Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced
sample. The 25 cm boundary is the source-sink threshold of the species' bionomics.

```{r}
kable(get("table-7-diameter-class"), booktabs = TRUE, align = "lrr")
```
