---
title: "Tables"
format:
  docx:
    reference-doc: ../../04.references/style-formal.docx
    number-sections: false
bibliography: ../../04.references/references-beetle.bib
csl: ../../04.references/apa-6th-edition.csl
## The bibliography is declared so the citations inside Table 2's rationale column resolve.
## It is suppressed because the reference list belongs in the main document, not here.
suppress-bibliography: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| output: false

## Journal of Applied Entomology asks for tables as a separate, editable file. These are
## read back from the CSVs the manuscript wrote as it rendered, not recomputed, so this
## file cannot report a different number from the article it accompanies. If a CSV is
## missing, render 01.manuscript/beetle-topography-wind-study-vri-timeseries.qmd first.
##
## Audited 2026-08-28. Three faults were fixed here at once, all of which shipped to the
## package on that date:
##   1. The numbering was one behind the main document, because the data-inventory table
##      was added as Table 1 after this file was written and everything else kept its old
##      number. The CSV names have been renumbered in the manuscript to match.
##   2. Tables 1 and 9 were missing entirely, so the editor received seven of nine.
##   3. No bibliography or CSL was declared, so Table 2 printed its raw citation keys,
##      "[@krawchuk2020; @cartwright2018; @powell2014]", into a submission document.
## The captions below are the manuscript's own, in full. The abridged versions used here
## previously dropped the abbreviation definitions the journal requires: it states that
## "the table, legend, and footnotes must be understandable without reference to the text"
## and that "All abbreviations must be defined in footnotes".
library(knitr)
TBL <- file.path("..", "..", "03.outputs", "TBL")
get <- function(n) {
  f <- file.path(TBL, paste0(n, ".csv"))
  if (!file.exists(f))
    stop("missing ", f,
         ". Render 01.manuscript/beetle-topography-wind-study-vri-timeseries.qmd first.")
  read.csv(f, check.names = FALSE)
}
```

<!-- Footnote symbols follow the journal's rule: dagger, double dagger, section and
     pilcrow in that order, with asterisks reserved for P-values. Statistical measures
     are named in the column headings, as the guidelines require. -->

# Table 1

The datasets this study combines, with the structure and resolution of each. Every count is
read from the files at render time. Spatial resolution is the grid the variable is analysed
on; temporal resolution is the interval at which it varies. Note that the response varies
every 16 days, the inventory once a year and the terrain not at all, which is the asymmetry
the wind analysis turns on.

```{r}
kable(get("table-1-inventory"), booktabs = TRUE)
```

# Table 2

Landscape attributes entered in this study, the direction expected of each, and the
reasoning behind that expectation.

```{r}
kable(get("table-2-hypotheses"), booktabs = TRUE, align = "lll")
```

# Table 3

Moderate-to-high beetle disturbance by year inside the study perimeter.

```{r}
kable(get("table-3-prevalence"), booktabs = TRUE, align = "lrr")
```

# Table 4

Stand structure across the study perimeter, from the Vegetation Resources Inventory. n is
cell-years. SD is the standard deviation of the landscape; SE is the standard error of the
mean and is small by construction at this n, so it should not be read as precision about any
one cell. Skewness and kurtosis are the bias-corrected third and fourth standardised moments;
kurtosis is excess, so 0 is Gaussian and positive is heavy-tailed.

```{r}
kable(get("table-4-stand-structure"), booktabs = TRUE, align = "lrrrr")
```

# Table 5

Model comparison, panel A of the model table. Each model adds one pathway to the previous.
AIC ranks on likelihood and a parameter penalty; the remaining columns are predictive error
on the fitted probabilities. RMSE is the root mean squared error and is the square root of
the Brier score; RMSE (%) expresses it against the prevalence of the response. MAE is the
mean absolute error. AUC is the area under the receiver operating characteristic curve.
Brier skill is the improvement over predicting the prevalence for every cell, where 0 is no
better than the base rate. MAPE and Theil's U are not reported: both divide by the observed
value, which is zero for the majority class of a binary response, so both are undefined
without discarding that class.

```{r}
kable(get("table-5-model-comparison"), booktabs = TRUE, align = "lrrr")
```

# Table 6

Full model M3, panel B of the model table: continuous terms, ordered by absolute effect.
Coefficients are log-odds per standard deviation on a class-balanced sample, so the intercept
is not landscape prevalence. SE is the standard error of the coefficient; z is the Wald
statistic. The geomorphon landform classes are also in this model and are not reported here.

```{r}
kable(get("table-6-full-model"), booktabs = TRUE, align = "lrrrr")
```

# Table 7

The wind terms under two specifications, both with previous-year pressure in the model.
M4 carries station wind, which is flat within a year and identified only across years.
M5 carries the terrain-resolved MicroMet field, which varies within the year and is
identified in space. z is the Wald statistic.

```{r}
kable(get("table-7-wind-terms"), booktabs = TRUE, align = "lllrr")
```

# Table 8

Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced
sample. QMD is quadratic mean diameter over stems 12.5 cm and up. The 25 cm boundary is the
source-sink threshold of the species' bionomics. Intervals are Wilson score intervals on the
class proportion, which is why they are asymmetric in the smallest class; the normal
approximation is not used because it misbehaves at the small class sizes at both ends of the
range. Note that 30 m cells in a spreading outbreak are not independent, so the tests
reported beneath this table in the main text are anti-conservative.

```{r}
kable(get("table-8-diameter-class"), booktabs = TRUE, align = "lrr")
```

# Table 9

The 16-day model refitted under four definitions of the wind window. The epoch wind regime,
every hour, is the definition used throughout this paper. Coefficients are log-odds per
standard deviation. AUC is the area under the receiver operating characteristic curve. The
thermal-gate row is fitted on fewer cell-epochs because two epochs carried too few qualifying
hours, so its AIC is not comparable with the other three and AIC is therefore not shown.

```{r}
kable(get("table-9-flight-window"), booktabs = TRUE)
```
