---
title: "Testing the wind disruption hypothesis for beetle refugia"
subtitle: "Host distribution, thermal load, and terrain-wind effects on Mountain pine beetle attack (Dendroctonus ponderosae) across 1,740 m of relief in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
abstract: >
  Mass attack by mountain pine beetle (Dendroctonus ponderosae) is coordinated by an airborne pheromone plume.
  Refugia are therefore hypothesised to form where stands are thin enough for wind to disrupt that plume. The
  mechanism has never been tested spatially. We test it across 168 km2 of inventoried forest and 1,740 m of relief
  in British Columbia, combining an aerial survey of beetle-caused mortality (1999-2015) with inventory host
  attributes and Global Wind Atlas speed.
  Attack is unimodal, peaking near 1,430 m. Below 1,900 m that peak is the distribution of lodgepole pine.
  Above 1,900 m it is not: attack halves while cover holds. Coding host as cover alone overstates that result.
  Attack rises monotonically with stand diameter, from 15 to 41 per cent across the range, with the steepest step
  at the 25 cm source-sink threshold the species' bionomics predicts, and the share of stands above that threshold
  falls four-fold at the top of the gradient. Conditioning on diameter and the remaining host terms removes about a
  quarter of the high-elevation suppression and leaves three quarters standing.
  We find no wind effect. Orthogonalised against a linear elevation term, wind is the strongest negative predictor
  in host-bearing stands and clears the interval test at every severity threshold above trace. But a unimodal
  response leaves elevation's curvature inside that term, and refitting elevation as a quadratic or spline collapses
  the coefficient everywhere. The signal is a specification artefact. Nor is the predictor the quantity the insect
  responds to: the atlas reports an annual mean at 10 m over open ground, while flight is arrested by instantaneous
  below-canopy speeds near 2 m s-1.
  Thermal alternatives survive. Deep-winter air temperature within 100 km never approaches the lethal range for
  overwintering late instars, but our coldest reading lies inside the range for early instars, 41 per cent of
  minima fall in December before hardening completes, and the record excludes the shoulder months entirely. Loss
  of univoltine development at elevation predicts the same pattern without invoking wind.
  Elevation explains 0.63 of the wind surface's variance landscape-wide but 0.80 within a 2 by 4 km clip, where the
  estimate reverses sign; grain scarcely matters. At the extent disturbance ecology usually adopts this study would
  have reported the opposite; with severity collapsed to presence, no result at all; and without a host layer,
  absence of pine misread as refugia.

keywords:
  - Beetle refugia
  - Mountain pine beetle
  - Aerial overview survey
  - Global Wind Atlas
  - Scale dependence
  - Collinearity

date: today
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/springer-basic-author-date.csl
df-print: kable
---

```{r}
#| label: setup
#| echo: true
#| output: false

## Every quantity in this manuscript is computed here at render time.
## patchwork is loaded for its `+` operator on ggplots (@fig-gradient), which is
## why no patchwork function appears by name anywhere below.
pkgs <- c("terra","knitr","kableExtra","ggplot2","dplyr","patchwork","tibble","kernlab",
          "sf","ggspatial","tidyterra","ggrepel","rnaturalearth","scales","parallel")
invisible(lapply(pkgs, function(p) library(p, character.only = TRUE)))
## dpi 300 is the Elsevier minimum for a combination line/halftone figure.
## error = FALSE on purpose: in a manuscript whose every number is computed at
## render time, a chunk that fails quietly while downstream inline expressions
## carry stale values is the one failure this design exists to prevent.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      comment = NA, dpi = 500, fig.align = "center",
                      cache.path = "_cache/", autodep = TRUE, cache.lazy = FALSE)
## Caching is opt-in per chunk and is applied only to the model-fitting chunks, which
## between them refit the block bootstrap roughly four thousand times per render.
## `autodep = TRUE` makes knitr read each chunk's globals and invalidate a cached
## chunk when any chunk that created those globals changes, so a cached result
## cannot outlive the data it was computed from. That guarantee is the only reason
## caching is acceptable in a manuscript whose whole design is that no number can go
## stale. It is not absolute: knitr tracks R objects, not files on disk. If an input
## raster or CSV changes without its reading chunk changing, delete `_cache/` and
## render again. `make clean-cache` does that, and the submission render should
## always be run from an empty cache.
knitr::dep_auto()
options(scipen = 999)
theme_set(theme_minimal(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                plot.title = element_text(face = "bold", size = 10)))

B_BOOT   <- 150L    # spatial block bootstrap replicates
BLOCK    <- 40L     # block side in cells (40 x 25.1 m ~= 1 km)
OUTBREAK <- 1999:2015
## The elevation above which conditioning on host stops removing the decline.
## It is a band edge of the 200 m banding in @tbl-gradient and is quoted in the
## prose, so it is defined once here rather than typed into each comparison.
ALPINE   <- 1900
```

```{r}
#| label: paths
#| echo: true

DATA_ROOT <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
dpath <- function(...) file.path(DATA_ROOT, ...)
here  <- function(...) file.path("..", ...)     # repo-relative, from 01.manuscript/
stopifnot(dir.exists(dpath("terrain_environment")))

## The elevation surface is the analysis grid: everything else is resampled onto
## it. Loaded here so the Study area section can quote the relief from the raster
## rather than from memory.
G <- rast(dpath("terrain_environment", "Elevation.utm.tif"))
GRID_RELIEF <- diff(as.vector(minmax(G)))
```

```{r}
#| label: map-furniture
#| echo: true

## Cartography shared by the four map figures, defined once here so each figure can
## sit in the section it documents instead of being grouped at the end of Materials
## for want of its furniture. Graticule, EPSG, north arrow, bar scale and a
## representative fraction quoted against a stated printed width are set once and
## reused, so the panels are directly comparable.

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
clip_sf  <- st_as_sf(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)))
grid_sf  <- st_as_sf(as.polygons(ext(G), crs = crs(G)))
ev <- as.vector(ext(G)); gw <- ev["xmax"] - ev["xmin"]; gh <- ev["ymax"] - ev["ymin"]
## Graticule and axis ticks are drawn in the same coordinate reference system the
## panel declares, so the CRS named beneath each title is the CRS the reader is
## reading off the frame. Ticks are metres divided by 1,000; the subtitle says so.
km_lab  <- function(x) sprintf("%.0f", x / 1000)
sub_utm <- "EPSG:32611 WGS 84 / UTM 11N; ticks in km"
sub_alb <- "EPSG:3005 NAD83 / BC Albers; ticks in km"

map_theme <- theme_bw(base_size = 8) +
  theme(panel.grid = element_line(colour = "grey70", linewidth = .25, linetype = "dotted"),
        axis.title = element_blank(), axis.text = element_text(size = 6, colour = "grey25"),
        plot.title = element_text(face = "bold", size = 8.5, margin = margin(b = 1)),
        plot.subtitle = element_text(size = 6, colour = "grey35", margin = margin(b = 3)),
        legend.key.width = unit(3.2, "mm"), legend.key.height = unit(7, "mm"),
        legend.title = element_text(size = 6.5), legend.text = element_text(size = 6))

north <- annotation_north_arrow(location = "tr", which_north = "true",
  height = unit(6, "mm"), width = unit(5, "mm"),
  pad_x = unit(1.2, "mm"), pad_y = unit(1.2, "mm"),
  style = north_arrow_orienteering(text_size = 5, line_width = .6,
    fill = c("white", "grey20"), text_col = "grey20"))

## White plates behind the north arrow and the clip label. Without them the arrow is
## unreadable over the dark end of the wind ramp and the label collides with the
## severity polygons.
plate <- annotate("rect", xmin = ev["xmin"] + .885*gw, xmax = ev["xmax"],
                  ymin = ev["ymax"] - .115*gh, ymax = ev["ymax"],
                  fill = "white", alpha = .72, colour = NA)
clip_layer <- geom_sf(data = clip_sf, fill = NA, colour = "#C0392B", linewidth = .45)
study <- list(coord_sf(crs = 32611, datum = st_crs(32611), expand = FALSE,
                       xlim = ev[c("xmin","xmax")], ylim = ev[c("ymin","ymax")]),
              scale_x_continuous(breaks = seq(520000, 530000, 5000), labels = km_lab),
              scale_y_continuous(breaks = seq(5440000, 5450000, 5000), labels = km_lab),
              plate, north, map_theme)

## Colour limits from robust quantiles. The raw topographic position surface runs
## past 1,000 m at the grid edge, where the 625 m window is truncated, and on a
## full-range ramp that artefact flattens the whole interior to one colour.
panel <- function(r, title, legend, pal = "viridis", dir = 1, q = c(.02, .98)) {
  lim <- unname(quantile(values(r, mat = FALSE), q, na.rm = TRUE))
  ggplot() + geom_spatraster(data = r) +
    scale_fill_viridis_c(option = pal, direction = dir, na.value = "grey88",
                         name = legend, limits = lim, oob = scales::squish) +
    clip_layer + study + labs(title = title, subtitle = sub_utm)
}
```

**Keywords:** beetle refugia; mountain pine beetle; aerial overview survey; Global Wind Atlas; scale dependence; collinearity

# Literature review {#sec-review}

The review below is written against this study's four objectives rather than as a survey of the
species, and its findings are numbered R1 to R5 and cited by number in every section that follows.
The protocol, the screening decisions and the per-study attribute extraction are recorded in
`02.inputs/literature/review-protocol.md`; every count reported here is computed at render time from
`02.inputs/literature/review-screening.csv` and none is typed into the prose.

```{r}
#| label: review-screen
#| echo: true
#| output: false

## The screening table is built by 02.inputs/literature/build-review-screening.py against the
## central store, and read here. Counts in the prose below are inline expressions over this
## object, so a change to the screening cannot leave the text behind.
rev  <- read.csv(here("02.inputs/literature/review-screening.csv"), stringsAsFactors = FALSE)
inc  <- rev[rev$stage2_include == "yes", ]

N_CAND    <- nrow(rev)
N_INC     <- nrow(inc)
N_FULL    <- sum(inc$full_text_checked == "yes")
N_WIND    <- sum(inc$wind_predictor == "yes")
N_NOWIND  <- N_INC - N_WIND
WIND_WHO  <- inc$key[inc$wind_predictor == "yes"]
YR_RANGE  <- range(suppressWarnings(as.integer(inc$year)), na.rm = TRUE)

## Every included study was read in full text, because the review's central claim is an
## absence and an absence established from titles is not evidence.
stopifnot(N_FULL == N_INC)
```

## Search and screening

The corpus is the central store described in @sec-review-limits, holding 579 distinct papers at the
date of the search. Matching titles and filenames on *beetle*, *Dendroctonus*, *MPB*,
*Ips typographus*, *Scolytinae* and *refugi* returned `r N_CAND` candidates. A study was retained if
it models bark-beetle-caused tree mortality spatially against terrain, landscape or climate
predictors, which is the only design that can address objectives 1, 3 and 4. `r N_INC` studies
qualified, spanning `r YR_RANGE[1]` to `r YR_RANGE[2]`, and all `r N_FULL` were read in full text.

Excluded studies are not discarded. Papers on mechanism and physiology carry the life history the
Introduction rests on, and papers on detection carry the error properties of the response variable;
both are cited there. They are absent from the counts because they cannot answer the objectives.

```{r}
#| label: tbl-review
#| echo: true
#| tbl-cap: "The studies retained by the review, with the attributes the objectives turn on, extracted from full text. The wind column records whether a wind term entered the model, not whether the word appears: every occurrence of `wind` in all eleven texts was read in context, and the per-study result is recorded in the review protocol."

inc |>
  transform(Study = ifelse(nzchar(key), key, "Zabihi 2021 (unverified)")) |>
  (\(d) d[order(suppressWarnings(as.integer(d$year))), ])() |>
  transform(Year = year,
            `Wind in model` = ifelse(wind_predictor == "yes", "yes", "no"),
            `Occurrences of \"wind\"` = wind_mentions,
            `Terrain terms` = terrain_terms,
            Pages = n_pages) |>
  subset(select = c("Study", "Year", "Terrain terms", "Wind in model",
                    "Occurrences of \"wind\"", "Pages")) |>
  kable(booktabs = TRUE, row.names = FALSE, align = "llccrr")
```

## Findings

**R1. The wind-disruption mechanism has been proposed but never fitted.** Of the `r N_INC` studies
that model bark beetle attack against terrain, `r N_WIND` enters a wind term, and that one is the
companion study on this same landscape whose response is conifer regeneration rather than beetle
attack [@murphy2026]. In the remaining `r N_NOWIND`, wind is either absent from the text entirely or
occurs only in a reference-list entry or a discussion sentence. @cartwright2018 is the clearest case:
the predictor set is slope, topographic position, heat load, compound topographic index and soil bulk
density, and wind appears three times, all in the discussion, invoked to explain an observed
thinner-stand association. @krawchuk2020 states the hypothesis and fits no model. The mechanism has
therefore been carried in this literature for at least `r YR_RANGE[2] - 2018` years as an
explanation, without once being entered as a predictor. Objective 3 addresses this directly.

**R2. The host-density half of the hypothesis already has support, and it is confounded.** The one
included study designed to identify insect refugia and model their controls found them associated
with thinner stands, shaded slopes and convergent topographic positions [@cartwright2018]. Thin
stands are exactly where a wind effect would be expected, and also where host is scarce, so the
association is consistent with the hypothesis and with its principal alternative. Separating them
requires an independent host layer, which is objective 1.

**R3. Host is never coded as the attribute the life history says matters.** None of the `r N_INC`
included studies mentions the 25 cm source-sink threshold or quadratic mean diameter, though the
threshold has been in the standard bionomics review since 2003 [@carroll2004bionomics]. Host enters
as cover, as basal area, or as a susceptibility class. Since brood production is governed by stem
size rather than by how much host is present, a host layer coded as cover cannot distinguish a stand
that produces beetles from one that consumes them. Objective 2 tests what changes when it is coded
as diameter instead.

**R4. Scale is recognised in this literature but conclusions are not tested against it.** Scale
vocabulary appears in six of the `r N_INC` included studies, so the issue is not unacknowledged, and
the included extents run from a single landscape to the whole of British Columbia. What the review
did not find is a study that refits one model across extent, grain and response coding on one dataset
and reports whether its own conclusion survives. Objective 4 does that, and @sec-extent, @sec-grain
and @sec-ordinal report the result.

**R5. No available wind product reports the quantity the hypothesis is about.** The hypothesis
concerns an instantaneous speed, below the canopy, during the July and August flight period, at the
height a beetle flies. Gridded products report a long-run mean at 10 m over open ground, and they are
produced by downscaling a coarse field over a digital elevation model [@badger2014; @davis2023],
which makes them partly a transform of the terrain they are offered alongside. The ceiling the
hypothesis leans on, about 2 m s⁻¹, is itself an inference across species
[@carroll2004bionomics; @jones2019]. Any test of objective 3 is therefore a test of a proxy, and the
review's conclusion is that the proxy's dependence on elevation has to be measured before its
coefficient can be read at all. R6 takes the same question along the time axis and reaches a
harder answer: not only does no product report the right quantity, no study says over what
interval the quantity should be reported.

```{r}
#| label: review-wind-temporal
#| echo: true
#| output: false

## The temporal resolution at which each included study treats wind, extracted from
## full text by 02.inputs/literature/extract-wind-temporal.py. Every wind sentence and
## its classification is written to wind-temporal-sentences.csv so any count below can
## be disputed against the sentence it came from. "window" and "windthrow" are excluded
## from the match, as they are in the screening table: a first run without that
## exclusion returned six wind mentions in @haukema2008movement that were all the
## phrase "temporal window".
wt <- read.csv(here("02.inputs/literature/wind-temporal-summary.csv"),
               stringsAsFactors = FALSE)

WT_N       <- nrow(wt)
WT_SILENT  <- sum(wt$n_wind == 0)
WT_SCALED  <- sum(nzchar(wt$finest))
WT_UNSCALED<- WT_N - WT_SCALED
WT_MENTION <- WT_N - WT_SILENT          # studies that use the word at all
WT_M_UNSC  <- WT_MENTION - WT_SCALED    # of those, how many say nothing about time
WT_KRAW    <- wt$n_wind[grepl("krawchuk", wt$study)]   # the study that states the hypothesis
WT_WHO     <- wt$study[nzchar(wt$finest)]
WT_MAX     <- wt$study[which.max(wt$n_wind)]
WT_MAXN    <- max(wt$n_wind)
stopifnot(WT_N == N_INC)
## the single study carrying temporal vocabulary must be the companion paper; if a
## future corpus adds another, R6's wording has to change rather than silently mislead
stopifnot(WT_SCALED == 1, grepl("murphy", WT_MAX))
```

**R6. The literature specifies no temporal resolution for wind, so the choice cannot be
deferred to it.** The analysis in this study has to decide what wind means in a model: a
long-run climatology, a growing-season mean, a monthly mean, a flight-window statistic
or an hourly series are five different covariates and they do not agree with each other.
Reading every wind sentence in the full text of all `r WT_N` included studies settles
where the literature stands, and the answer is that it does not stand anywhere.
`r WT_SILENT` of the `r WT_N` do not contain the word at all once "window" and
"windthrow" are excluded. Of the `r WT_MENTION` that do use it, `r WT_M_UNSC` attach no
temporal vocabulary to it whatsoever: no averaging period, no season, no time of day.
Across the whole included set that is `r WT_UNSCALED` of `r WT_N` studies silent on the
question. The single exception is @murphy2026, with `r WT_MAXN` mentions carrying monthly and annual
vocabulary, and that is the companion study on this same landscape, which took mean
monthly speeds at 10 m for 2015 to 2020 from the Global Wind Atlas [@murphy2026] and so
described a period five to fifteen years after the outbreak it was modelling.
@krawchuk2020, who state the hypothesis, mention wind `r WT_KRAW` times and never once say over
what interval it should be measured.

The consequence for this study is a positive one and it is the reason R6 exists. Because
no prior specification can be cited, the temporal resolution has to be argued from the
mechanism rather than adopted from precedent, and the mechanism is explicit about it:
dispersal and pheromone-mediated mass attack happen during a narrow midsummer flight
period, and within it they depend on whether the air is still enough to hold a plume.
That argues for the finest resolution the observations support, reduced only at the last
step into the flight window itself. @sec-krawchuk reports what is used and what it costs.

## Limits of this review {#sec-review-limits}

The corpus is one curated library rather than a systematic search of an abstracting database, so
this is a review of what was read, made checkable, rather than a claim about the complete
literature. A study that entered a wind predictor and is absent from the store would not have been
found, and R1 carries that scope wherever it is used. The screening table records the decision and
the extracted attribute for every candidate so that a reader can dispute either.

# Introduction

Disturbance refugia, defined by @krawchuk2020 as locations "disturbed less severely or less frequently than other areas within the surrounding landscape", have become an organising concept in forest conservation under a changing climate. The framework identifies where biological legacies persist through a disturbance and why those legacies matter to the recovery of the surrounding stand. Its empirical base is predominantly a wildfire literature [@krawchuk2020; @littlefield2019]. For insect outbreaks the base is thinner: @krawchuk2020 observe that few studies have explicitly identified refugia from insect outbreaks, even though the stand and landscape controls on outbreak severity are themselves well documented for the better-studied pests.

For mountain pine beetle (*Dendroctonus ponderosae* Hopkins), @krawchuk2020 advance three hypotheses. Refugia might occur, first, in areas with cooler temperatures, for example from topographic shading, that protect trees from water stress; second, in areas of lower host density, "allowing for greater wind disruption of beetle pheromone communication and more vigorous tree growth and chemical defenses"; and third, in areas with fewer large-diameter host trees, or among trees investing more heavily in resistance traits such as resin ducts. These are not mutually exclusive. All three predict reduced attack on high, cool, sparsely stocked ground, and they are separable only by the mechanism each invokes. The present study tests the second hypothesis and treats the first and third as the competing explanations against which it must be judged.

@cartwright2018 provides the closest existing test. Working at 30 m resolution in lodgepole and whitebark pine in southern Oregon, that study identified refugia from Landsat moisture anomalies during a concurrent 2009 drought and beetle outbreak, and modelled their landscape controls. Refugia were associated with shaded slopes, convergent topographic positions, low soil bulk density and thinner stands. The thinner-stand association is consistent with the host-density component of the second hypothesis, but no wind surface was available and the disruption mechanism itself was not evaluated. The review in @sec-review makes that observation countable rather than rhetorical: of the studies that model bark beetle attack against terrain, one enters a wind term and it is the companion study on the present landscape (R1).

## Beetle life history

Interpreting any spatial pattern of attack requires the population biology that generates it. Mountain pine beetle populations occupy one of four phases defined by population size relative to available host: endemic, incipient epidemic, epidemic and post-epidemic [@safranyik2006chap1]. Endemic populations persist in weakened, generally small-diameter trees. Incipient-epidemic populations develop when larger-diameter hosts become colonisable, through a local decline in host resistance, immigration, or favourable breeding conditions. Epidemics operate at landscape scale and arise mainly from extensive, contiguous areas of susceptible host under favourable weather [@safranyik2006chap1]. The transition between phases is a density-dependent threshold rather than a gradient: @howe2022 show that stand-level spatial aggregation, behavioural shifts and the quality of attainable hosts converge to define a critical threshold beyond which population growth becomes self-driving. Outbreak history in British Columbia has been governed jointly by the availability of susceptible age classes and by climate [@taylor2003; @sambaraju2021].

Colonisation proceeds by mass attack, which is a density phenomenon rather than a behaviour. A lodgepole pine resists a single beetle through constitutive resin flow and an induced response that saturates the phloem with terpenes, and the constitutive terpene profile itself varies geographically in ways that correlate with historical attack [@clark2010]. Overwhelming that defence requires the near-simultaneous arrival of many individuals, and the arrival is coordinated chemically. Pioneer females oxidise the host monoterpene α-pinene to *trans*-verbenol, which is preferentially attractive to males; arriving males release *exo*-brevicomin, which at low concentrations attracts mainly females; and the host kairomones α-pinene and myrcene synergise both, so that colonisation of a tree is normally completed within one to two days [@safranyik2006chap1]. At approximately 60 attacks per square metre of bark the signal reverses. High concentrations of *exo*-brevicomin and frontalin, together with the anti-aggregation pheromone verbenone, terminate colonisation and redirect arrivals to neighbouring trees [@safranyik2006chap1]. Because this signal is airborne, it is in principle susceptible to disruption by wind, and that susceptibility is the premise of the hypothesis under test.

Host quality sets a second constraint, and it is a constraint on stem size rather than on host abundance. Bark roughness, bark thickness, bark surface area, phloem thickness and sapwood moisture retention all increase with tree diameter, and together they determine how many beetles a colonised tree can produce. @carroll2004bionomics summarise the consequence as a threshold: on average, lodgepole pine trees of 25 cm diameter or less are beetle sinks, in that more beetles attack than emerge, whereas trees above 25 cm are beetle sources. The threshold is stated for individual trees. We apply it to a stand-level attribute, quadratic mean diameter, and therefore treat it as an approximate ordering of stand suitability rather than as a per-stand criterion transferred intact from the source.

Temperature imposes the third and fourth constraints, and they act on different parts of the life cycle. Dispersal is bounded by a flight envelope. @carroll2004bionomics summarise it as lower and upper limits of 19 and 41 °C after @mccambridge1971, with most beetles flying between 22 and 32 °C after Safranyik, negative phototaxis above 35 °C and severely restricted flight above 38 °C. The primary measurement behind the outer bounds was made on beetles reared from field-infested ponderosa pine and tested at 50 per cent relative humidity without direct insolation: @mccambridge1971 places the threshold at approximately 67 to 68 °F, or 19 to 20 °C, and extrapolates an upper limit near 107 °F, or 42 °C. That study also records field flight below its own laboratory threshold where sunlight warms emerged beetles or the bark around them, so the laboratory envelope is a conservative bound on ambient air temperature rather than a hard limit.

Development is bounded separately. Stage-specific development thresholds maintain synchrony within and among seasons, and sub-population-specific development rates maintain univoltinism across much of the species range [@safranyik2006chap1]. Where heat accumulation is insufficient, as at high elevation and near the northern range margin, populations become semivoltine. Stretching the life cycle across two years imposes severe mortality, because the insect must overwinter twice, often in cold-susceptible stages, and remains exposed to natural enemies for longer [@safranyik2006chap1]. In semivoltine populations, climate becomes the dominant control on distribution and abundance.

Overwinter survival is bounded by cold tolerance, which is acquired rather than fixed. Glycerol accumulates in the hemolymph as temperatures decline through autumn, so cold-hardiness is greatest from December to February and is incomplete on either side of that window [@safranyik2006chap1]. Tolerance also varies by stage. Lethal temperature ranges have been estimated at -18 °C for eggs, -18 to -34 °C for pupae, -23 to -34 °C for adults, and, for larvae, -23 to -29 °C for first instars, -23 to -34 °C for second instars and -29 to -40 °C for third and fourth instars [@safranyik2006chap1]. Late larval instars are the normal overwintering stage and withstand temperatures near -40 °C for extended midwinter periods, which is the origin of the conventionally quoted lethal figure. The same source is explicit that a midwinter minimum of -30 °C would be expected to cause little mortality, and that significant mortality instead follows cold arriving early, before sufficient glycerol has accumulated, or late, after it has begun to be metabolised. Any assessment of thermal limitation on a landscape therefore depends on the date of the minimum as much as on its depth, a point we return to in @sec-cold.

## Wind and flight

The wind-disruption hypothesis requires wind to interfere with the pheromone plume. A second and simpler route from wind to reduced attack acts on the insect directly, by suppressing flight, and the two cannot be separated by the design used here. The standard figure is a wind ceiling of about 2 m s⁻¹. @carroll2004bionomics state it directly: bark beetles in general do not fly in winds exceeding their maximum flight speed, and for a large-bodied species such as mountain pine beetle the maximum wind speed for flight, and so the probable maximum flight velocity, is approximately 2 m s⁻¹. The provenance of that number is worth stating, because it is an inference across species rather than a measurement on this one: the primary citation is a study of *Dendroctonus pseudotsugae* Hopkins. The comparative evidence assembled by @jones2019 is consistent but no firmer. *Dendroctonus frontalis* Zimmermann cannot fly above 2 m s⁻¹, *Ips pini* Say and *Scolytus multistriatus* Marsham arrest flight at 4.8 and 5 km h⁻¹, and in the one laboratory wind tunnel trial on *D. ponderosae* the beetles flew for longer at 0, 0.5 and 1 m s⁻¹ than at 2 m s⁻¹, which is a reduction in flight duration rather than arrest. Flight-mill velocities for *D. ponderosae* are reported in the same review as 1.55 to 1.93 km h⁻¹, or roughly 0.43 to 0.54 m s⁻¹, with mean flight distances of 2.12 to 5.95 km, so the ceiling is several times the sustained cruising speed. Dispersal is also density-dependent, switching between short-range movement below the canopy and long-range transport above it as population density rises [@powell2014; @johnson2026density; @johnson2026stratified]. At low wind speeds the dispersal direction of *D. ponderosae* is not influenced by wind direction [@jones2019].

## Measurement problems

Testing the hypothesis requires a spatial wind surface, and the available surfaces carry three properties that constrain what can be inferred from them.

The first is construction. Gridded wind products, including the Global Wind Atlas, are produced by driving a mesoscale model from reanalysis, generalising the output to remove local terrain and roughness effects, and then downscaling it over a digital elevation model and a land-cover roughness map [@badger2014; @davis2023]. Elevation is an input to the product rather than a covariate measured independently of it. A wind surface entered into a model alongside elevation is therefore not an independent measurement, and the degree of dependence is a function of study extent, because a coarse driving field is close to flat over a few kilometres and almost all within-product variation at small extents derives from terrain. The extents at which disturbance ecology commonly works, a burn perimeter or a cluster of field plots, are the extents at which this dependence is most severe and least visible.

The second is correspondence to the biological quantity. The hypothesis concerns an instantaneous speed, below the canopy, during the July and August flight period, at the height at which beetles fly. The atlas reports an annual mean at 10 m above open ground. The two quantities differ in height, season, averaging period and canopy position, and no downscaled product currently available at landscape scale reports the former.

The third is a confound specific to refugia work. A cell recorded as unattacked may simply contain no host. In the terms of @krawchuk2020 this is the distinction between what a location is a refugium from and what it is a refugium for, and it cannot be resolved without an independent map of host abundance. The precedent is direct: @hadley1994 predicted heavier beetle-caused mortality on hot, dry, south-facing slopes and observed the reverse, because host distribution rather than moisture stress governed where attack succeeded. We extend the argument by noting that a map of host cover is also insufficient, because the attribute governing whether a stand produces beetles is stem diameter rather than cover.

## Objectives

Each question below follows from a numbered finding of the review in @sec-review, and the finding it answers is named with it. We address four questions on a single landscape spanning `r format(round(GRID_RELIEF), big.mark = ",")` m of relief.

1.  **From R2.** Does terrain condition beetle-caused mortality once host abundance is accounted for, and at what point on the elevation gradient does host distribution cease to explain the observed pattern? The review found the host-density half of the hypothesis supported but confounded with the wind half, and separable only with an independent host layer.
2.  **From R3.** Does that result survive coding host as stem diameter rather than as cover, given the 25 cm source-sink threshold reported for individual trees by @carroll2004bionomics? No study retained by the review codes host as the attribute the life history says governs brood production.
3.  **From R1 and R5.** Does the component of the wind surface that is independent of elevation predict reduced attack where host is present, as hypothesised by @krawchuk2020? The review found the mechanism proposed repeatedly and fitted once, in a companion study with a different response, and found that no available product reports the wind quantity the hypothesis concerns.
4.  **From R4.** To what extent does the answer to question 3 depend on analytical decisions that are not fixed by the data: the extent of the study area, the grain of the analysis cells, and whether damage is coded as a binary outcome or as the ordinal severity class the survey records? Scale is acknowledged in this literature without any retained study refitting one model across all three axes and reporting whether its conclusion survives.

The fourth question is answered by refitting the same response, predictors and model across each axis in turn: across the full landscape and within a 2 by 4 km clip drawn from it, at 25 m, 50 m and 100 m grain, and at each severity threshold. Two of the three axes alter the conclusion.

# Materials

## Study area

The Darkwoods Conservation Area lies in the South Selkirk Mountains of southeastern British Columbia. The terrain grid spans 16 by 17.5 km and `r format(round(GRID_RELIEF), big.mark = ",")` m of relief, from valley bottom to alpine. The modelled area is a subset, because cells carrying no inventory basal area cannot enter a model conditioning on stand structure; @tbl-relief reports its extent and relief. The two extents are distinct and are not interchangeable. The range crosses the full biogeoclimatic sequence of the southern interior wet belt, from the dry warm Interior Cedar-Hemlock subzones through the moist warm subzones to the Engelmann Spruce-Subalpine Fir zone and its parkland variant [@ketcheson1991; @braumandl2005]. Lodgepole pine (*Pinus contorta* var. *latifolia*), the beetle's principal host here, occupies the middle of that sequence and is scarce at both ends.

The site description and terrain surfaces are inherited from a companion analysis of conifer regeneration on the same ground [@murphy2026], which used them to predict seedling intensity after a 2015 wildfire. Disturbance was a predictor there; here it is the response. @fig-site locates the landscape and shows the three spatial inputs.

## Beetle-caused mortality

```{r}
#| label: read-response
#| echo: true

aos_box <- ext(project(as.polygons(ext(G), crs = crs(G)), "EPSG:3005"))
aos <- vect(here("02.inputs/aos/pest_infestation_poly.gdb"),
            layer = "pest_infestation_poly", extent = aos_box)
mpb <- project(aos[aos$PEST_SPECIES_CODE == "IBM", ], crs(G))   # IBM = mountain pine beetle
m   <- mpb[mpb$CAPTURE_YEAR %in% OUTBREAK, ]

## BC severity classes are ordinal: trace, light, moderate, severe, very severe
sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)
m$sev <- unname(sev_rank[m$PEST_SEVERITY_CODE]); m$sev[is.na(m$sev)] <- 0

severity  <- rasterize(m, G, field = "sev", fun = "max", background = 0)

c(polygons = nrow(m), years = length(unique(m$CAPTURE_YEAR)),
  hectares = round(sum(m$AREA_HA)))
```

The response is the British Columbia Aerial Overview Survey, the operational provincial record of forest health, in which observers sketch-map damage annually from aircraft and label each polygon by year, causal agent, severity class and host species. We take polygons coded `IBM`, mountain pine beetle, for the outbreak window 1999 to 2015, and rasterise the maximum severity class reached by each cell.

The survey was preferred to a spectral classification for a specific reason. Red attack has been mapped from Landsat with improving accuracy [@white2005; @meddens2013; @ye2021], but every such product is a model whose errors are spatially structured and not independent of terrain, because illumination, shadow and view angle vary with slope and aspect. For a question about terrain, a response from a terrain-sensitive classifier risks recovering the classifier's own terrain sensitivity. Sketch-mapping has cruder errors, but they are observer errors, not errors generated by the predictors under test.

The severity scale is anchored to ground measurement. Twenty-eight field plots established in beetle-affected stands in 2020 recorded the basal area of pine killed, spanning 0.6 to 47.4 m2/ha, or 1.3 to 94.7 per cent of plot basal area, and calibrated the spectral mapping used in the companion study [@murphy2026]. Plot coordinates were not recoverable, so the plots calibrate the severity scale but cannot verify the spatial placement of individual polygons. No validation of polygon placement is therefore claimed.

@fig-site shows the landscape and the response together. Panel (b) is the elevation surface that defines the grid and panel (d) the survey record this paper models: attack concentrated in a mid-elevation band, sparse in the valley and again along the alpine crest. Panel (c) is the wind surface whose relationship to that terrain the second half of the paper is about.

```{r}
#| label: fig-site
#| echo: true
#| fig-width: 7.1
#| fig-height: 6.8
#| fig-cap: "The study area and the three spatial inputs. (a) Location within British Columbia; the marker is a 22 km symbol, not the footprint. (b) The elevation surface, which is the analysis grid. (c) Global Wind Atlas mean speed at 10 m; the shaded southern band lies outside the wind product's extent. (d) Beetle-caused mortality from the Aerial Overview Survey, 1999 to 2015, as the maximum severity class in any year. The outlined rectangle in (b) to (d) is the 2 by 4 km clip of @sec-extent. Each panel's graticule, axis ticks and stated coordinate reference system are the same; the system is named beneath the panel title and ticks are in kilometres."

## The wind raster begins 715 m north of the grid's southern edge, so resampling onto G
## turns that strip into NA cells and na.value draws it grey. Left transparent it reads
## as a rendering fault rather than as missing data.
Wind_r <- resample(rast(dpath("terrain_environment", "Wind.utm.tif")), G)

prov <- st_transform(ne_states(country = c("Canada","United States of America"),
                               returnclass = "sf"), 3005)
site <- st_transform(st_centroid(grid_sf), 3005)
lab  <- transform(as.data.frame(st_coordinates(site)), lab = "Darkwoods\nstudy area")
## BC is framed to the analysis grid's aspect ratio so the two rows of the figure are
## the same height and no band of white opens between them.
b0 <- st_bbox(prov[prov$name == "British Columbia", ]); tgt <- unname(gw/gh)
w0 <- unname(b0["xmax"]-b0["xmin"]); h0 <- unname(b0["ymax"]-b0["ymin"])
cx <- unname(b0["xmax"]+b0["xmin"])/2; cy <- unname(b0["ymax"]+b0["ymin"])/2
if (w0/h0 > tgt) h0 <- w0/tgt else w0 <- h0*tgt
pd <- .04*w0
bb <- c(xmin = cx-w0/2-pd, xmax = cx+w0/2+pd, ymin = cy-h0/2-pd, ymax = cy+h0/2+pd)

p_a <- ggplot() +
  geom_sf(data = prov[prov$name != "British Columbia", ], fill = "grey93",
          colour = "grey65", linewidth = .2) +
  geom_sf(data = prov[prov$name == "British Columbia", ], fill = "grey82",
          colour = "grey30", linewidth = .4) +
  geom_sf(data = st_buffer(site, 22000), fill = "#C0392B", colour = "grey15",
          linewidth = .3) +
  geom_label_repel(data = lab, aes(X, Y, label = lab), size = 2.1, lineheight = .95,
                   colour = "grey10", nudge_x = -330000, nudge_y = 300000,
                   segment.colour = "grey25", segment.size = .3, min.segment.length = 0,
                   box.padding = .4, seed = 1, fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm")) +
  annotate("text", x = bb["xmin"] + .34*(bb["xmax"]-bb["xmin"]),
           y = bb["ymin"] + .70*(bb["ymax"]-bb["ymin"]), label = "British\nColumbia",
           size = 2.3, colour = "grey30", fontface = "italic", lineheight = .95) +
  coord_sf(crs = 3005, xlim = bb[c("xmin","xmax")], ylim = bb[c("ymin","ymax")],
           datum = st_crs(3005), expand = FALSE) +
  scale_x_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  scale_y_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  north + map_theme +
  labs(title = "(a) Location", subtitle = sub_alb)

p_b <- ggplot() + geom_spatraster(data = G) +
  scale_fill_hypso_c(palette = "dem_screen", na.value = "transparent",
                     name = "Elevation\n(m)") +
  clip_layer + study + labs(title = "(b) Elevation, the analysis grid", subtitle = sub_utm)

p_c <- ggplot() + geom_spatraster(data = Wind_r) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey88",
                       name = "Wind speed\n(m s⁻¹)") +
  clip_layer + study +
  labs(title = "(c) Global Wind Atlas mean speed at 10 m", subtitle = sub_utm)

sev_f <- as.factor(severity)
levels(sev_f) <- data.frame(ID = 0:5, class = c("none","trace","light","moderate",
                                                "severe","very severe"))
p_d <- ggplot() + geom_spatraster(data = sev_f) +
  scale_fill_manual(values = c(none = "grey92", trace = "#FEE0B6", light = "#FDB863",
                    moderate = "#E08214", severe = "#B35806", `very severe` = "#7F3B08"),
                    na.value = "transparent", name = "Severity\nclass",
                    na.translate = FALSE) +
  clip_layer + study +
  labs(title = "(d) Beetle-caused mortality, 1999 to 2015", subtitle = sub_utm)

fig_site <- (p_a | p_b) / (p_c | p_d)
fig_site
ggsave("../03.outputs/PNG/fig-site-maps.png", fig_site,
       width = 180, height = 172, units = "mm", dpi = 500, bg = "white")
```

## Red-stage classification

The aerial overview survey resolves damage at the scale an observer can sketch
from an aircraft, and the refugia question is about wind and terrain acting at
tens of metres. So we also classified red-stage attack annually from Landsat
across the whole grid, to see whether a finer response layer would support the
test that the polygons cannot. The pipeline below is that classification. Its
result is reported here rather than in Results because it decides which response
layer the paper uses, and the answer is that it does not replace the survey.

Every stage runs in Google Earth Engine through `rgee`. The code is gated on the
exported rasters: where they are absent it builds and downloads them, which takes
roughly half an hour, and where they are present the render reads them and needs
no network. Function definitions sit outside the gate so they are readable
whether or not they run.

```{r}
#| label: ee-pipeline-setup
#| echo: true

RED_DIR   <- here("02.inputs/beetle-classification/red-stage")
BASE_YEAR <- 2003          # differencing baseline, see below
CLASS_YRS <- 2005:2014
dec_files <- file.path(RED_DIR, sprintf("dec_%d.tif", CLASS_YRS))

## Sensor supplying each year, taken from Landsat scene metadata rather than
## from the parent study's Table 2, which lists Landsat 5 for a 2013 acquisition.
## Landsat 5 stopped acquiring in November 2011 and Landsat 8 did not begin until
## 2013, so 2012 is the one year only Landsat 7 covers. The parent study omitted
## 2012 for that reason and not for cloud: six Landsat 7 scenes met its own
## selection rule of under 20 per cent cloud between 1 June and 30 August, the
## cleanest at 1 per cent. Landsat 7 has flown scan-line-corrector-off since May
## 2003, so that year alone carries striping. Inventory in
## 02.inputs/beetle-classification/landsat-scene-inventory.csv.
SENSOR <- c(`2003`="LT05",`2005`="LT05",`2006`="LT05",`2007`="LT05",`2008`="LT05",
            `2009`="LT05",`2010`="LT05",`2011`="LT05",`2012`="LE07",
            `2013`="LC08",`2014`="LC08")
COLL   <- c(LT05="LANDSAT/LT05/C02/T1_L2", LE07="LANDSAT/LE07/C02/T1_L2",
            LC08="LANDSAT/LC08/C02/T1_L2")
SB     <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")
BANDS  <- c(SB, "NDMI", "dNDMI")
```

```{r}
#| label: ee-pipeline-functions
#| echo: true

## Stage 1, imagery. Collection 2 Level-2 surface reflectance, scaled by the
## published factors, with cloud, cloud shadow, cirrus and snow masked from
## QA_PIXEL bits 1 to 5. Thematic Mapper and Enhanced Thematic Mapper carry the
## reflective bands as SR_B1-5 and SR_B7; the Operational Land Imager shifts them
## to SR_B2-7 because it adds a coastal aerosol band. Both are renamed to common
## names so nothing downstream has to know which sensor it is reading.
ee_prep <- function(sensor) {
  bn <- if (sensor == "LC08") paste0("SR_B", 2:7) else paste0("SR_B", c(1:5, 7))
  rgee::ee_utils_pyfunc(function(im) {
    qa   <- im$select("QA_PIXEL")
    keep <- qa$bitwiseAnd(2L)$eq(0)$And(qa$bitwiseAnd(4L)$eq(0))$
      And(qa$bitwiseAnd(8L)$eq(0))$And(qa$bitwiseAnd(16L)$eq(0))$
      And(qa$bitwiseAnd(32L)$eq(0))
    im$select(bn, SB)$multiply(0.0000275)$add(-0.2)$updateMask(keep)$
      copyProperties(im, list("SUN_ELEVATION","SUN_AZIMUTH","system:time_start"))
  })
}

ee_year <- function(y, AOI) {
  s <- SENSOR[[as.character(y)]]
  ee$ImageCollection(COLL[[s]])$filterBounds(AOI)$
    filterDate(sprintf("%d-06-01", y), sprintf("%d-08-31", y))$
    filter(ee$Filter$lt("CLOUD_COVER", 60))$map(ee_prep(s))
}

## Stage 2, terrain correction. Mandatory here and not at the 7.78 km2 extent the
## parent study worked at. Across 1,776 m of relief, illumination varies enough
## with slope and aspect to put a terrain signal into reflectance, and this study
## tests terrain effects, so an uncorrected artefact would be indistinguishable
## from the result. SCS+C after Soenen et al. 2005: the C term is the ratio of
## intercept to slope in a regression of reflectance on the illumination term,
## fitted per band per scene. Where a scene is wholly masked over the grid the
## regression returns null; C is then set very large, which drives the correction
## factor to one, so a scene with nothing to fit is passed through unchanged
## rather than destroyed.
ee_illum <- function(im, slope, asp) {
  sz <- ee$Image$constant(90)$subtract(ee$Number(im$get("SUN_ELEVATION")))$multiply(pi/180)
  sa <- ee$Image$constant(ee$Number(im$get("SUN_AZIMUTH")))$multiply(pi/180)
  slope$cos()$multiply(sz$cos())$
    add(slope$sin()$multiply(sz$sin())$multiply(sa$subtract(asp)$cos()))$rename("cos_i")
}

ee_scs_c <- function(im, slope, asp, AOI) {
  ci <- ee_illum(im, slope, asp)
  sz <- ee$Image$constant(90)$subtract(ee$Number(im$get("SUN_ELEVATION")))$multiply(pi/180)
  n0 <- slope$cos()$multiply(sz$cos())
  one <- function(b) {
    f  <- ci$addBands(im$select(b))$reduceRegion(ee$Reducer$linearFit(), AOI, 90,
                                                 maxPixels = 1e9, bestEffort = TRUE)
    sc <- ee$Number(ee$Algorithms$If(f$get("scale"),  f$get("scale"),  0))
    of <- ee$Number(ee$Algorithms$If(f$get("offset"), f$get("offset"), 0))
    C  <- ee$Number(ee$Algorithms$If(sc$abs()$gt(1e-8), of$divide(sc), 1e6))$max(0)$min(1e6)
    im$select(b)$multiply(n0$add(C))$divide(ci$add(C))$rename(b)
  }
  bs <- one(SB[1]); for (b in SB[-1]) bs <- bs$addBands(one(b))
  bs$copyProperties(im, list("SUN_ELEVATION","SUN_AZIMUTH","system:time_start"))
}

## Stage 3, index and differencing. NDMI is (NIR - SWIR1) / (NIR + SWIR1) and each
## annual composite is differenced against the pre-outbreak baseline. The baseline
## is 2003, not the 2005 the parent study's text names twice. Its own Table 2
## labels 2003 pre-outbreak and 2005 early-outbreak, and its 2005 raster carries
## 534 cells coded red attack, 9.73 per cent of the cells it maps. A scene
## differenced against itself is zero everywhere, so 2005 cannot be its own
## baseline. See 02.inputs/beetle-classification/02-parent-raster-audit.R.
ee_stack <- function(y, BASE, slope, asp, AOI) {
  m  <- ee_year(y, AOI)$map(rgee::ee_utils_pyfunc(
          function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()
  nd <- m$normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")
  m$addBands(nd)$addBands(nd$subtract(BASE)$rename("dNDMI"))
}
```

The classifier is trained on the aerial overview survey itself. The parent
study's 310-row spectral sample cannot be reused: every band column in the source
workbook is prefixed 2020, the plots were established in July and August 2020,
and the values sit inside the Collection 2 Level-1 range of the 2020 Landsat 8
scene, reading as negative reflectance in four bands if taken as Level-2. A model
built from one uncorrected 2020 scene was applied to map attack from 2005 to
2014, eight of those years from a different satellite. The plot coordinates are
also unrecoverable, so the spectra cannot be re-extracted on a matching footing.
Training on the survey instead keeps labels and imagery in the same year and the
same radiometry.

```{r}
#| label: ee-classify
#| echo: true

SAMP_FILE <- file.path(RED_DIR, "training-sample.csv")

## Stage 4, labels. Positives are cells the survey mapped at moderate severity or
## worse in that year, negatives cells it mapped as undamaged, 250 of each per
## year with a quarter held out. 2014 takes the pooled model because the survey
## records no cell above trace over this grid that year.
##
## The sample is written to disk and the classifier is then trained from the
## table rather than from the sampling graph. That is not tidiness. An Earth
## Engine classifier carries its training computation into every image it is
## later applied to, so classifying an annual composite from a classifier trained
## directly on nine years of sampled composites re-runs all of it and the
## download fails with a 400. Trained from stored numbers, the same download
## succeeds.
if (!file.exists(SAMP_FILE)) {
  library(rgee); ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  dir.create(RED_DIR, recursive = TRUE, showWarnings = FALSE)
  AOI   <- ee$Geometry$Rectangle(c(-116.7861, 49.0884, -116.5649, 49.2460))
  DEM   <- ee$Image("USGS/SRTMGL1_003")
  slope <- ee$Terrain$slope(DEM)$multiply(pi/180)
  asp   <- ee$Terrain$aspect(DEM)$multiply(pi/180)
  BASE  <- ee_year(BASE_YEAR, AOI)$map(rgee::ee_utils_pyfunc(
             function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()$
             normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")

  set.seed(42); parts <- list()
  for (y in 2005:2013) {
    ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev",
                    fun = "max", background = 0)
    if (global(ry >= 3, "sum", na.rm = TRUE)[[1]] < 30) next
    pick <- function(mk, lab) {
      p <- spatSample(mk, 250, "random", na.rm = TRUE, xy = TRUE, values = FALSE)
      data.frame(x = p[,1], y = p[,2], lab = lab,
                 elev = extract(G, p[,1:2])[,1], yr = y)
    }
    d <- rbind(pick(ifel(ry >= 3, 1, NA), 1L), pick(ifel(ry == 0, 1, NA), 0L))
    d <- d[!is.na(d$elev), ]
    d$split <- as.integer(runif(nrow(d)) < 0.25)
    pts <- sf::st_transform(sf::st_as_sf(d, coords = c("x","y"), crs = crs(G)), 4326)
    parts[[length(parts) + 1]] <- ee_stack(y, BASE, slope, asp, AOI)$sampleRegions(
      collection = sf_as_ee(pts), properties = list("lab","elev","yr","split"),
      scale = 30, tileScale = 4, geometries = TRUE)
  }
  got <- ee_as_sf(ee$FeatureCollection(parts)$flatten(), quiet = TRUE)
  xy  <- sf::st_coordinates(got)
  write.csv(cbind(sf::st_drop_geometry(got), lon = xy[,1], lat = xy[,2]),
            SAMP_FILE, row.names = FALSE)
}
samp <- read.csv(SAMP_FILE)
stopifnot(all(c(BANDS, "lab", "elev", "yr", "split", "lon", "lat") %in% names(samp)))

## Stage 4b, the classifier. Fitted here rather than by ee$Classifier$libsvm,
## for a mechanical reason. An Earth Engine classifier carries its training
## features into every request that uses it, so each annual download embedded the
## whole labelled table and returned HTTP 400; the same eight-band composite
## downloads without complaint on its own. A linear kernel is a plane, so the
## fitted model transfers exactly as a band expression and Earth Engine is left
## doing arithmetic. The check below is that transfer, and it holds to machine
## precision. Predictors are standardised on the training rows only.
SVM_TR <- samp[samp$split == 0, ]; SVM_TE <- samp[samp$split == 1, ]
SVM_MU <- sapply(SVM_TR[BANDS], mean); SVM_SD <- sapply(SVM_TR[BANDS], sd)
zscore <- function(d) as.matrix(sweep(sweep(d[BANDS], 2, SVM_MU, "-"), 2, SVM_SD, "/"))
set.seed(1)
SVM_FIT <- kernlab::ksvm(zscore(SVM_TR), factor(SVM_TR$lab), type = "C-svc",
                         kernel = "vanilladot", C = 1, scaled = FALSE)
SVM_W <- colSums(kernlab::coef(SVM_FIT)[[1]] * kernlab::xmatrix(SVM_FIT)[[1]])
SVM_B <- -SVM_FIT@b
stopifnot(identical(
  as.integer(as.vector(zscore(SVM_TE) %*% SVM_W + SVM_B) > 0),
  as.integer(as.character(predict(SVM_FIT, zscore(SVM_TE))))))

## Collapsed into raw reflectance units, which is the form Earth Engine applies.
SVM_A <- SVM_W / SVM_SD
SVM_C <- SVM_B - sum(SVM_W * SVM_MU / SVM_SD)
stopifnot(max(abs((as.matrix(SVM_TE[BANDS]) %*% SVM_A + SVM_C) -
                  (zscore(SVM_TE) %*% SVM_W + SVM_B))) < 1e-8)

## Stage 5, validation. Computed here from the held-out quarter, so the agreement
## table below and the exported rasters come from one fitted object.
SVM_TE$pred <- as.integer(as.vector(as.matrix(SVM_TE[BANDS]) %*% SVM_A + SVM_C) > 0)

if (!all(file.exists(dec_files))) {
  library(rgee); ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  AOI   <- ee$Geometry$Rectangle(c(-116.7861, 49.0884, -116.5649, 49.2460))
  DEM   <- ee$Image("USGS/SRTMGL1_003")
  slope <- ee$Terrain$slope(DEM)$multiply(pi/180)
  asp   <- ee$Terrain$aspect(DEM)$multiply(pi/180)
  BASE  <- ee_year(BASE_YEAR, AOI)$map(rgee::ee_utils_pyfunc(
             function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()$
             normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")

  ## Stage 6, export. The decision plane is applied band by band, thresholded at
  ## zero, classified at the 30 m native grid, then resampled onto the analysis
  ## grid by nearest neighbour, which is the correct rule for a class raster and
  ## makes the alignment exact rather than nearly so.
  gbox <- ee$Geometry$Rectangle(c(xmin(G), ymin(G), xmax(G), ymax(G)),
                                proj = "EPSG:32611", geodesic = FALSE)
  for (y in CLASS_YRS) {
    st <- ee_stack(y, BASE, slope, asp, AOI)
    d  <- ee$Image$constant(SVM_C)
    for (bn in BANDS) d <- d$add(st$select(bn)$multiply(unname(SVM_A[bn])))
    ## The continuous decision value is exported, not a thresholded map, so the
    ## threshold stays a visible choice made below rather than one baked into a
    ## download. Scaled by 1,000 and offset by 10,000 because a GeoTIFF written
    ## from an integer image takes zero as its nodata value: exporting a signed
    ## surface directly would silently delete every cell sitting on the decision
    ## boundary, and exporting a 0/1 map deleted every unattacked cell outright
    ## on the first attempt, returning 100 per cent attacked in all ten years.
    u <- d$multiply(1000)$add(10000)$round()$toInt16()$rename("dec")$
      getDownloadURL(list(name = sprintf("dec_%d", y), crs = "EPSG:32611",
                          scale = 30, region = gbox, format = "GEO_TIFF"))
    tmp <- tempfile(fileext = ".tif"); download.file(u, tmp, mode = "wb", quiet = TRUE)
    ## Kept in the scaled integer form Earth Engine sent, deflated, because the
    ## same surface as 32-bit float is 2.2 MB a year and this is 0.2 MB. The
    ## offset is removed at read time below, not here.
    writeRaster(resample(rast(tmp), G, method = "bilinear"),
                file.path(RED_DIR, sprintf("dec_%d.tif", y)),
                overwrite = TRUE, datatype = "INT2S",
                gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2"))
  }
}

dec <- (rast(dec_files) - 10000) / 1000     # undo the export scaling and offset
names(dec) <- CLASS_YRS

## The model's own threshold is zero, which is what a balanced training sample
## implies. It is the wrong prior for this landscape. Training drew 250 attacked
## and 250 unattacked cells per year, so the implied prevalence is one half,
## while the survey maps attack on a few per cent of the grid. Thresholded at
## zero the classification calls a third to two thirds of the landscape attacked.
## Both maps are therefore carried: the model's own, and one whose threshold is
## calibrated per year so the classified area matches the area the survey mapped.
SURV_AREA <- sapply(CLASS_YRS, function(y) {
  ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev", fun = "max", background = 0)
  global(ry >= 3, "mean", na.rm = TRUE)[[1]]
})
names(SURV_AREA) <- CLASS_YRS
RAW_AREA <- sapply(CLASS_YRS, function(y)
  global(dec[[as.character(y)]] > 0, "mean", na.rm = TRUE)[[1]])
names(RAW_AREA) <- CLASS_YRS

red <- rast(lapply(CLASS_YRS, function(y) {
  d <- dec[[as.character(y)]]
  thr <- if (SURV_AREA[[as.character(y)]] > 0)
    global(d, function(v, ...) quantile(v, 1 - SURV_AREA[[as.character(y)]], na.rm = TRUE))[[1]]
    else Inf
  d > thr
}))
names(red) <- CLASS_YRS
## A classification that is entirely one class, or entirely masked, is a
## pipeline failure rather than a result, and must not reach the prose.
stopifnot(nlyr(red) == length(CLASS_YRS),
          all(dim(red)[1:2] == dim(G)[1:2]),
          isTRUE(all.equal(as.vector(ext(red)), as.vector(ext(G)), tolerance = 1e-6)),
          all(RAW_AREA > 0.001), all(RAW_AREA < 0.999),
          all(global(!is.na(red), "mean")[[1]] > 0.5))
```

```{r}
#| label: tbl-redstage
#| echo: true
#| tbl-cap: "Agreement between the Landsat red-stage classification and the aerial overview survey, by 300 m elevation band. The held-out quarter of the labelled sample, pooled over 2005 to 2013. This is agreement with the survey, not accuracy against ground truth: the survey is also the training source, and no independent reference exists at this extent."

v <- SVM_TE
v$classification <- v$pred
v$band <- cut(v$elev, seq(500, 2400, by = 300), dig.lab = 5)

kap <- function(o, p) {
  cm <- table(factor(o, 0:1), factor(p, 0:1))
  oa <- sum(diag(cm)) / sum(cm)
  pe <- sum(rowSums(cm) * colSums(cm)) / sum(cm)^2
  c(OA = oa, kappa = (oa - pe) / (1 - pe))
}
POOLED <- kap(v$lab, v$classification)

RED_TBL <- do.call(rbind, lapply(levels(v$band), function(b) {
  s <- v[!is.na(v$band) & v$band == b, ]
  if (nrow(s) < 10)
    return(data.frame(Band = b, n = nrow(s), Attacked = sum(s$lab == 1),
                      Agreement = NA_real_, Kappa = NA_real_,
                      Note = "too few reference points, uncertain"))
  k <- kap(s$lab, s$classification)
  data.frame(Band = b, n = nrow(s), Attacked = sum(s$lab == 1),
             Agreement = round(k["OA"], 3), Kappa = round(k["kappa"], 3), Note = "")
}))
kable(RED_TBL, row.names = FALSE,
      col.names = c("Elevation band (m)", "n", "Attacked", "Agreement", "Kappa", ""))
```

```{r}
#| label: tbl-redstage-area
#| echo: true
#| tbl-cap: "Where the Landsat classification and the aerial overview survey disagree, by year. Survey area is the share of the grid the survey mapped at moderate severity or worse. Model area is what the classifier calls attacked at its own threshold of zero. Hit rate is the share of calibrated classified cells that fall inside a survey polygon, and lift is that rate divided by the survey's own prevalence, so 1.00 is chance."

RED_CMP <- do.call(rbind, lapply(CLASS_YRS, function(y) {
  yc <- as.character(y)
  ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev", fun = "max", background = 0)
  S  <- ry >= 3
  p  <- SURV_AREA[[yc]]
  if (p == 0) return(data.frame(Year = y, Survey = 0, Model = round(100*RAW_AREA[[yc]], 1),
                                Hit = NA_real_, Lift = NA_real_))
  hits <- global(S & red[[yc]], "sum", na.rm = TRUE)[[1]]
  nC   <- global(red[[yc]], "sum", na.rm = TRUE)[[1]]
  data.frame(Year = y, Survey = round(100*p, 3), Model = round(100*RAW_AREA[[yc]], 1),
             Hit = round(100*hits/nC, 2), Lift = round((hits/nC)/p, 2))
}))
kable(RED_CMP, row.names = FALSE,
      col.names = c("Year", "Survey area (%)", "Model area (%)",
                    "Hit rate (%)", "Lift"))
```

The classification does not replace the survey, and the reason is not a marginal
shortfall in accuracy. Pooled over the held-out quarter it agrees with the survey
on `r sprintf("%.1f", 100 * POOLED["OA"])` per cent of cells, kappa
`r sprintf("%.2f", POOLED["kappa"])`, against the kappa of 0.750 and overall
accuracy of 0.824 the parent study reports for its 7.78 km2 window. But
@tbl-redstage-area shows that figure is flattered by how the sample was drawn.
Training took 250 attacked and 250 unattacked cells per year, so the model's
implied prior is one half, while the survey maps attack on
`r sprintf("%.2f", 100 * min(SURV_AREA[SURV_AREA > 0]))` to
`r sprintf("%.2f", 100 * max(SURV_AREA))` per cent of the grid. At its own
threshold the classifier therefore calls
`r sprintf("%.0f", 100 * min(RAW_AREA))` to `r sprintf("%.0f", 100 * max(RAW_AREA))`
per cent of the landscape attacked, one to two orders of magnitude more ground
than the survey maps.

Matching the area does not rescue it. Calibrating the threshold each year so the
classified area equals the surveyed area, the classification lands on
`r sprintf("%.1f", min(RED_CMP$Hit, na.rm = TRUE))` to
`r sprintf("%.1f", max(RED_CMP$Hit, na.rm = TRUE))` per cent of the cells the
survey mapped. Expressed as lift against the survey's own prevalence, where 1.00
is chance, the years run from `r sprintf("%.2f", min(RED_CMP$Lift, na.rm = TRUE))`
to `r sprintf("%.2f", max(RED_CMP$Lift, na.rm = TRUE))`, and
`r sum(RED_CMP$Lift == 0, na.rm = TRUE)` of `r sum(!is.na(RED_CMP$Lift))` years
share no cells with the survey at all. Once area is held equal the map is close to
uninformative about where attack was.

The year that most exposes the mechanism is 2012. It is the only year covered by
Landsat 7 alone, whose scan line corrector has been broken since May 2003, and it
is the year the classifier calls the most attacked at
`r sprintf("%.0f", 100 * RAW_AREA[["2012"]])` per cent while the survey maps the
least at `r sprintf("%.2f", 100 * SURV_AREA[["2012"]])` per cent. The classifier
is reading striping as change.

Two readings of the pooled agreement figure were available before these tables
existed. The classifier might have been wrong, or the polygons might have been,
and a one-sided error towards mapping more attack than an observer drew is what a
finer instrument would also produce. @tbl-redstage-area settles it against the
classifier: an instrument genuinely resolving attack the surveyor missed would
still overlap the attack the surveyor found, and this one does not. We cannot
say the survey is right, only that the classification is not measuring the same
thing.

We therefore keep the aerial overview survey as the response and report the
classification as a comparison layer. That preserves the argument the rest of the
paper rests on, which is that a response whose errors are observer errors is
safer here than one whose errors are generated by the predictors under test. A
classification trained on terrain-corrected imagery does not escape that problem,
because terrain correction is itself a model of the terrain whose effect is being
tested. It also means the timing analysis a per-cell year of first detection would
allow is not available: an annual series this weak cannot support a survival
model, and fitting one would give the wind test a precision the response layer
cannot justify.

## Host abundance

```{r}
#| label: read-host
#| echo: true

vri <- project(vect(here("02.inputs/vri/vri_darkwoods.geojson")), crs(G))
vd  <- as.data.frame(vri)
num <- function(x) suppressWarnings(as.numeric(x))

## VRI records up to three species per polygon with percentage cover
accumulate <- function(pattern) {
  out <- rep(0, nrow(vd))
  for (i in 1:3) {
    cd <- vd[[paste0("SPECIES_CD_", i)]]
    pc <- num(vd[[paste0("SPECIES_PCT_", i)]]); pc[is.na(pc)] <- 0
    out <- out + ifelse(!is.na(cd) & grepl(pattern, cd), pc, 0)
  }
  out
}
vri$PinePct <- accumulate("^PL")               # lodgepole, the primary host
vri$Basal   <- num(vd$BASAL_AREA)

## Quadratic mean diameter of stems 12.5 cm and over. Carried because stem size,
## not cover, is what decides whether a stand produces more beetles than it
## consumes: below about 25 cm a lodgepole stand is a sink, above it a source
## [@carroll2004bionomics]. Cover and basal area cannot express that.
vri$QMD <- num(vd$QUAD_DIAM_125)

host <- rast(lapply(c("PinePct","Basal","QMD"), function(f) rasterize(vri, G, field = f)))
names(host) <- c("PinePct","Basal","QMD")
c(vri_polygons = nrow(vri), with_lodgepole = sum(vri$PinePct > 0),
  with_diameter = sum(!is.na(vri$QMD)))
```

Host abundance comes from the provincial Vegetation Resources Inventory, layer `VEG_COMP_LYR_R1_POLY`, retrieved from the British Columbia Data Catalogue web feature service for the study window. We accumulate the percentage cover of lodgepole pine across the three species slots each polygon records, and carry stand basal area alongside it.

We also carry quadratic mean diameter, which the inventory records for stems 12.5 cm and over. Cover and basal area describe how much host is present. Neither describes whether that host can produce beetles. Brood production per tree rises with bark surface area, bark thickness, phloem thickness and sapwood moisture retention, and all four rise with stem diameter, so a stand of small stems consumes more beetles than it returns. Carroll and Safranyik put the crossover at about 25 cm: below it a lodgepole stand is on average a beetle sink, above it a source [@carroll2004bionomics]. That threshold is the reason a host layer coded only as cover cannot answer the question this paper asks at high elevation, and it is the gap review finding R3 identifies in the retained studies. @sec-size tests it directly. Diameter is recorded for `r format(sum(!is.na(vri$QMD)), big.mark=",")` of the `r format(nrow(vri), big.mark=",")` polygons, which covers the analysis frame almost exactly, because the polygons that lack it are largely those that also lack a basal area.

The inventory's vintage matters. These polygons carry a reference year of 2014 and an interpretation date of August 2014; the projected date of 31 December 2025 is the growth model's roll-forward of age, height and volume, not a re-observation. Species composition therefore comes from photography flown in the fourteenth year of a seventeen-year outbreak window.

This limitation cannot be removed and is stated explicitly. A stand whose pine was killed early will have been interpreted after the fact, and its recorded lodgepole percentage will be lower than the pine it carried when the beetle arrived. The bias runs in a specific direction: host is under-recorded exactly where attack was heaviest and earliest. It is therefore a candidate explanation for the reversal of the lodgepole coefficient at the severe threshold in @sec-ordinal. It does not touch the presence-absence distinction the layer is mainly used for, since a stand that carried enough pine to be killed is not recorded as pine-free.

Without this layer the analysis cannot distinguish a refugium from an absence of pine, and @sec-host shows that the distinction accounts for most of the apparent terrain signal.

@fig-host maps both layers. Lodgepole cover is concentrated in the middle of the elevation range and scarce in the valley, which is the pattern the host-conditioning argument turns on: the valley is not lightly attacked because it is sheltered but because it holds almost no host. Basal area defines the analysis frame, and the blank cells in panel (b) are those dropped for want of it.

```{r}
#| label: fig-host
#| echo: true
#| fig-width: 7.1
#| fig-height: 3.8
#| fig-cap: "The forest inventory host layers, rasterised onto the analysis grid. (a) Lodgepole pine cover, accumulated across the three species slots each polygon records. (b) Stand basal area, which defines the analysis frame: cells with no recorded basal area are dropped. Unshaded ground is unmapped or non-productive. Projection and scale as in @fig-site."

fig_host <-
  (ggplot() + geom_spatraster(data = host$PinePct) +
     scale_fill_viridis_c(option = "viridis", direction = -1, na.value = "grey88",
                          name = "Lodgepole\ncover (%)") +
     clip_layer + study + labs(title = "(a) Lodgepole pine cover", subtitle = sub_utm)) |
  (ggplot() + geom_spatraster(data = host$Basal) +
     scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "grey88",
                          name = "Basal area\n(m² ha⁻¹)") +
     clip_layer + study + labs(title = "(b) Stand basal area", subtitle = sub_utm))
fig_host
ggsave("../03.outputs/PNG/fig-host-maps.png", fig_host,
       width = 180, height = 96, units = "mm", dpi = 500, bg = "white")
```

## Terrain and wind

```{r}
#| label: read-terrain
#| echo: true

to_grid <- function(f) {
  r <- rast(dpath("terrain_environment", f))
  b <- project(as.polygons(ext(G), crs = crs(G)), crs(r))
  project(crop(r, b, snap = "out"), G, method = "bilinear")
}
terr <- rast(list(Elevation = G,               Slope = to_grid("Slope.utm.tif"),
                  TWI       = to_grid("TWI.utm.tif"),
                  RIX       = to_grid("Rutm.tif"),
                  Wind      = to_grid("Wind.utm.tif")))
names(terr) <- c("Elevation","Slope","TWI","RIX","Wind")

## Topographic position: mean deviation from a surrounding annulus.
tpi <- function(r, w) {
  k <- matrix(1, w, w); k[ceiling(w/2), ceiling(w/2)] <- 0
  r - focal(r, k / sum(k), fun = sum, na.policy = "omit", na.rm = TRUE)
}
terr$TPI600 <- tpi(G, 25)     # ~625 m neighbourhood
```

```{r}
#| label: aspect-terms
#| echo: true

## Two surfaces built from aspect, which the inherited set lacks entirely.
## Both are computed directly from the elevation model rather than draped over
## it, which is why @tbl-orthogonality can show them independent of elevation.
d2r <- pi / 180
asp  <- rast(dpath("terrain_environment", "Aspect.utm.tif"))
slp  <- rast(dpath("terrain_environment", "Slope.utm.tif"))
stopifnot(identical(dim(asp), dim(G)), identical(as.vector(ext(asp)), as.vector(ext(G))))

## Heat load after McCune and Keon, aspect folded about the SW-NE axis so that
## southwest is hottest. Flat cells carry a -1 sentinel; the sin(slope) factor
## nulls the aspect term there, so the substitution below is inert.
asp[asp < 0] <- 225
fold <- abs(180 - abs(asp - 225))
LAT  <- mean(crds(project(as.polygons(ext(G), crs = crs(G)), "EPSG:4326"))[, 2])
terr$HLI <- 0.339 + 0.808 * cos(LAT*d2r) * cos(slp*d2r) -
            0.196 * sin(LAT*d2r) * sin(slp*d2r) -
            0.482 * cos(fold*d2r) * sin(slp*d2r)

## Winstral shelter index: the maximum upwind slope angle within SEARCH metres
## along a bearing. Positive means terrain rises upwind, so the cell is
## sheltered; negative means it is exposed. The Global Wind Atlas gives speed
## but no direction, so rather than assume a prevailing bearing we compute all
## eight and average, giving an omnidirectional shelter index.
SEARCH <- 500
shift_mat <- function(m, dr, dc) {
  nr <- nrow(m); nc <- ncol(m); out <- matrix(NA_real_, nr, nc)
  ri <- seq_len(nr); ci <- seq_len(nc); sr <- ri + dr; sc <- ci + dc
  vr <- sr >= 1 & sr <= nr; vc <- sc >= 1 & sc <= nc
  out[ri[vr], ci[vc]] <- m[sr[vr], sc[vc]]
  out
}
zm <- as.matrix(G, wide = TRUE); rstep <- res(G)[1]
sx_dir <- function(az) {
  th <- az * d2r; best <- matrix(-Inf, nrow(zm), ncol(zm))
  for (k in 1:floor(SEARCH / rstep)) {
    dc <- round(k * sin(th)); dr <- round(-k * cos(th))
    if (dc == 0 && dr == 0) next
    best <- pmax(best, atan((shift_mat(zm, dr, dc) - zm) / (k * rstep)) / d2r, na.rm = TRUE)
  }
  best[!is.finite(best)] <- NA_real_
  best
}
sx <- rast(lapply(seq(0, 315, by = 45), function(a) {
  r <- G; values(r) <- as.vector(t(sx_dir(a))); r }))
terr$SxMean <- app(sx, mean, na.rm = TRUE)
```

Terrain surfaces are inherited unchanged from @murphy2026, which derived them from the 1 m Natural Resources Canada High Resolution Digital Elevation Model [@nrcan2017], smoothed, depression-filled by the algorithm of @wang2006 and aggregated to 25 m, with compound indices following @beers1973 and @parker1982. We add three surfaces of our own. The first is a topographic position index at a 625 m neighbourhood, because @cartwright2018 found slope, elevation and heat load among the strongest controls on refugia and reported topographic position alongside them, and the inherited set carried none.

The other two exist because the inherited set carries no aspect information, and so no term that can distinguish a warm slope from a cold one at the same elevation. That matters because the confound this paper cannot resolve is between wind exposure and cold, and aspect bears on the two differently. We compute a heat load index after @mccune2002, in which aspect is folded about the southwest to northeast axis so that southwest is hottest, and a shelter index after @winstral2002, the maximum upwind slope angle within 500 m along a bearing, positive where terrain rises upwind and negative on exposed convexities. The Global Wind Atlas supplies speed but no direction, so rather than assume a prevailing bearing we compute the index on all eight cardinal and intercardinal bearings and average them.

The decisive property of both is how they are built. They are derived from the elevation model by trigonometry rather than draped over it by downscaling, and @tbl-orthogonality shows the consequence: their variance is essentially unrelated to elevation, where the wind product's is largely elevation. A terrain-derived predictor is not automatically an elevation transform; the distinction is in the construction.

The ruggedness surface carries a documentation gap we state rather than paper over. Its exact formulation is not recorded beyond the citations above, and its distribution is inconsistent with a vector ruggedness measure. We treat it as a reproducible but unnamed ruggedness surface and confine claims to its statistical behaviour.

Ambient wind is Global Wind Atlas mean wind speed at 10 m above ground level [@globalwindatlas]. The atlas is built by driving a mesoscale model from reanalysis, generalising the result by the method of @badger2014 to remove the local terrain and roughness signal, then downscaling it over a digital elevation model and a land-cover roughness map [@davis2023]. Elevation and roughness are inputs to the product, not covariates measured independently of it. That is why it cannot be assumed independent of elevation, and testing that assumption is the second half of this paper.

Figure S1 maps the six derived surfaces. Their point is visual rather than statistical: slope, wetness, ruggedness and topographic position all trace the same drainage network and valley wall, whereas heat load and the shelter index divide the landscape by aspect, picking out southwest-facing ground the others do not distinguish. That is the difference @tbl-orthogonality quantifies.

```{r}
#| echo: true
#| output: false
#| fig-width: 7.1
#| fig-height: 4.6

fig_terrain <-
  (panel(terr$Slope,  "(a) Slope", "Degrees", "cividis") |
   panel(terr$TWI,    "(b) Topographic wetness index", "TWI", "mako", -1) |
   panel(terr$RIX,    "(c) Ruggedness", "Index", "rocket", -1)) /
  (panel(terr$TPI600, "(d) Topographic position, 625 m", "TPI (m)", "viridis") |
   panel(terr$HLI,    "(e) Heat load index", "HLI", "inferno") |
   panel(terr$SxMean, "(f) Shelter index, omnidirectional", "Sx (deg)", "cividis", -1))
fig_terrain
ggsave("../03.outputs/PNG/fig-terrain-maps.png", fig_terrain,
       width = 180, height = 118, units = "mm", dpi = 500, bg = "white")
```

## Climate records

Because the interpretation in @sec-krawchuk turns on whether it is cold enough at the top of the gradient to kill beetles, we assembled the observational record. Hourly observations from the British Columbia Wildfire Service network and daily observations from Environment and Climate Change Canada were retrieved for all stations within 100 km for the winters of the outbreak window. The script in `02.inputs/climate/` streams roughly 2.7 GB and writes the per-station-winter summary the manuscript reads. That division is deliberate: the fetch cannot sit inside a document render, but every decision about which records to believe is made in @sec-cold, in code, and is visible there.

Two screens matter. Many older Wildfire Service sensors floor at exactly -20.0 °C, so a station-winter whose minimum sits on that floor has an unknown true minimum and is excluded; the rule is applied and checked in @sec-cold rather than inherited. Isolated spikes, colder than both neighbouring observations by more than 15 °C, are rejected as instrument error at the observation level, the one screen that must happen during retrieval because it needs the sub-daily series. One such value, a -40.0 °C daily minimum on a day whose maximum was -4.5 °C and carrying no quality flag, would otherwise have produced exactly the conclusion the record does not support.

A second script in the same directory, `add-minimum-dates.R`, reads the winter extracts the fetch already cached and records the date on which each station-winter reached its minimum. It runs outside the render for the same reason, and its output is small and is kept with the manuscript. The date is needed because lethal temperature for this insect is conditional on the state of cold-hardening rather than fixed, so a minimum without a date cannot be compared with a lethal range.

The window is December to February, which is what the question about deep-winter minima required. That choice bounds what the record can support, and @sec-cold states the bound rather than working around it. This network cannot furnish a temperature surface. It can establish whether deep-winter air temperature in this region approaches the lethal range, a weaker and, for one purpose, sufficient claim.

```{r}
#| label: assemble
#| echo: true

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
in_clip  <- rasterize(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)),
                      G, field = 1, background = 0)

X <- as.data.frame(c(terr, host, severity, in_clip), na.rm = FALSE)
names(X)[(ncol(X) - 1):ncol(X)] <- c("Severity","InClip")
X$Attacked <- X$Severity > 0
X$HasHost  <- !is.na(X$PinePct) & X$PinePct > 0

## spatial blocks for resampling-based inference
rc <- rowColFromCell(G, seq_len(ncell(G)))
X$block <- paste(rc[,1] %/% BLOCK, rc[,2] %/% BLOCK, sep = "_")

PRED  <- c("Elevation","Slope","TWI","RIX","Wind","TPI600")
EXTRA <- c("HLI","SxMean")          # aspect-derived; enter only the extended model
MVARS <- c(PRED, "PinePct", "Basal")   # defines the analysis frame, deliberately unchanged

## Cells lacking an inventory basal area are unmapped or non-productive and cannot
## enter a model that conditions on stand structure. Dropping them is reported here
## rather than left to silent na.action inside glm.
n_before <- sum(complete.cases(X[, PRED]) & !is.na(X$PinePct))
S <- X[complete.cases(X[, MVARS]), ]
H <- S[S$HasHost, ]
c(with_terrain_and_host = n_before, dropped_no_basal_area = n_before - nrow(S),
  analysis_cells = nrow(S), km2 = round(nrow(S) * prod(res(G)) / 1e6, 1),
  attacked = sum(S$Attacked), host_cells = nrow(H))
```

```{r}
#| label: tbl-relief
#| echo: true
#| tbl-cap: "The analysis landscape, and the 2 by 4 km clip drawn from it that the companion study used. The clip samples under two thirds of the relief and under 4 per cent of the cells."

tibble(
  Extent = c("full landscape", "2 by 4 km clip"),
  Cells  = c(nrow(S), sum(S$InClip == 1)),
  `Area (km²)` = round(c(nrow(S), sum(S$InClip == 1)) * prod(res(G)) / 1e6, 1),
  `Elevation (m)` = c(sprintf("%.0f to %.0f", min(S$Elevation), max(S$Elevation)),
                      sprintf("%.0f to %.0f", min(S$Elevation[S$InClip==1]),
                                              max(S$Elevation[S$InClip==1]))),
  `Relief (m)` = round(c(diff(range(S$Elevation)), diff(range(S$Elevation[S$InClip==1])))),
  `Attacked (%)` = round(100 * c(mean(S$Attacked), mean(S$Attacked[S$InClip==1])), 1)
) |> kable(booktabs = TRUE, align = "llllll", format.args = list(big.mark = ","))
```

# Methods

## Inference

The response is heavily clustered, so cell counts vastly overstate independent information. All coefficients carry percentile intervals from a spatial block bootstrap, meaning the landscape is cut into square tiles and whole tiles are drawn at random with replacement, so neighbouring cells travel together and are never treated as separate evidence. The landscape is partitioned into blocks of `r BLOCK` by `r BLOCK` cells, roughly 1 km on a side, resampled with replacement `r B_BOOT` times. Blocks of this size exceed any plausible correlation range of the survey polygons. We report the point estimate, the block interval and the proportion of resamples reproducing its sign, and treat a term as supported only when the interval excludes zero.

```{r}
#| label: fn-boot
#| echo: true

## The bootstrap refits several hundred times on >100,000 rows, so it works on a
## model matrix built once and calls glm.fit directly. This is arithmetically
## identical to glm() with the same formula, and roughly thirty times faster.
MODEL_VARS <- c("Elevation","WindResid","PinePct","Basal","TPI600","Slope","TWI","RIX")

design_matrix <- function(z, vars = MODEL_VARS)
  cbind(`(Intercept)` = 1, as.matrix(z[, vars, drop = FALSE]))

fit_fast <- function(Xm, y) glm.fit(Xm, y, family = binomial())

## The resamples are drawn first, in one sequence from one seed, and only then
## fitted. That ordering is the whole point: the random number stream is consumed
## in exactly the order the original serial loop consumed it, so every interval in
## this paper is bit-identical to the serial implementation, and the fitting, which
## is the expensive part, can be spread across cores. Seeding inside each worker
## instead would be simpler and would change every reported interval.
##
## Measured on the analysis frame: the matrix subsetting costs about 2 per cent of
## a replicate and glm.fit about 98, so there is nothing to gain from cleverer
## indexing, and two attempts at it were slower than the original. Four cores gives
## about 1.7 times; eight is slower again on this machine, which has four
## performance cores.
BOOT_CORES <- max(1L, min(4L, parallel::detectCores(logical = FALSE) - 1L))

block_boot <- function(z, vars = MODEL_VARS, B = B_BOOT, seed = 42) {
  Xm <- design_matrix(z, vars); y <- z$y
  idx_of <- split(seq_len(nrow(z)), z$block)
  blocks <- names(idx_of)
  set.seed(seed)
  draws <- lapply(seq_len(B), function(b) sample(blocks, length(blocks), replace = TRUE))
  one <- function(k) {
    idx <- unlist(idx_of[k], use.names = FALSE)
    f <- try(fit_fast(Xm[idx, , drop = FALSE], y[idx]), silent = TRUE)
    if (inherits(f, "try-error")) rep(NA_real_, ncol(Xm)) else coef(f)
  }
  ## mclapply forks, which is unavailable on Windows; fall back rather than fail.
  res <- if (BOOT_CORES > 1L && .Platform$OS.type == "unix")
           parallel::mclapply(draws, one, mc.cores = BOOT_CORES) else lapply(draws, one)
  matrix(unlist(res), B, ncol(Xm), byrow = TRUE,
         dimnames = list(NULL, colnames(Xm)))
}

model_table <- function(z, vars = MODEL_VARS, seed = 42) {
  Xm <- design_matrix(z, vars)
  fit <- fit_fast(Xm, z$y)
  cf  <- coef(fit)
  se  <- sqrt(diag(chol2inv(fit$qr$qr[1:fit$rank, 1:fit$rank, drop = FALSE])))
  bs  <- block_boot(z, vars, seed = seed)
  tibble(term = names(cf), beta = as.numeric(cf), naive_z = as.numeric(cf / se),
         lo = apply(bs, 2, quantile, .025, na.rm = TRUE),
         hi = apply(bs, 2, quantile, .975, na.rm = TRUE),
         sign_stab = sapply(names(cf), function(n)
                       mean(sign(bs[, n]) == sign(cf[n]), na.rm = TRUE)))
}

fmt_model <- function(mt) {
  mt |> filter(term != "(Intercept)") |>
    transmute(Term = term, beta = sprintf("%+.3f", beta),
              `naive z` = sprintf("%+.1f", naive_z),
              `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
              `sign stability` = sprintf("%.2f", sign_stab),
              Supported = ifelse(lo * hi > 0, "yes", "no"))
}
```

## Orthogonalisation

Review finding R5 is the reason this step exists. Because the wind surface is built over the elevation surface, its coefficient in a model containing elevation is estimated from whatever small part of the joint distribution lets the two vary separately. We therefore replace wind by the residual of wind regressed on elevation, rescaled to unit variance. Orthogonalisation here means exactly that: keep only the part of the wind surface that a straight line in elevation cannot predict.

The substitution is sometimes presented as a remedy for collinearity, which it is not. The substitution is a linear reparameterisation of the same model: fitted values and deviance are unchanged, and by the Frisch-Waugh-Lovell theorem the residualised coefficient is the raw wind coefficient multiplied by the square root of one minus the R² of wind on elevation, with its standard error scaled identically. Every supported-or-not verdict in this paper is therefore numerically the same in both parameterisations, and orthogonalisation buys no precision whatever. What changes is interpretation. The wind coefficient now refers to the part of the wind surface elevation does not explain, which is the quantity the @krawchuk2020 hypothesis is about, and the elevation coefficient becomes the total elevation association rather than the partial effect at fixed wind. The reparameterisation is therefore justified on interpretative grounds alone.

```{r}
#| label: design
#| echo: true

standardise <- function(dd) {
  vars <- intersect(c(PRED, EXTRA, "PinePct", "Basal"), names(dd))
  z <- as.data.frame(scale(dd[, vars]))
  z$y     <- as.integer(dd$Attacked)
  z$block <- dd$block
  z$Severity <- dd$Severity          # carried for the ordinal fits in @sec-ordinal
  z$WindResid <- as.numeric(scale(residuals(lm(Wind ~ Elevation, data = z))))
  z
}
Z  <- standardise(S)     # whole landscape
ZH <- standardise(H)     # cells containing lodgepole
ZC <- standardise(S[S$InClip == 1, ])   # the 2 by 4 km clip
round(cor(Z[, c("Elevation","Wind","WindResid")]), 3)
```

```{r}
#| label: tbl-orthogonality
#| echo: true
#| tbl-cap: "How much of each candidate predictor is elevation. The wind product is a mesoscale field draped over terrain and is largely an elevation transform. The heat load and shelter indices are computed from the same elevation model by trigonometry and are almost unrelated to it."

tibble(Surface = c("Global Wind Atlas wind speed", "Heat load index (McCune and Keon)",
                   "Shelter index (Winstral, omnidirectional)", "Topographic position, 625 m"),
       `How it is built` = c("mesoscale field downscaled over the DEM",
                             "trigonometric function of DEM aspect and slope",
                             "maximum upwind slope from the DEM within 500 m",
                             "deviation from surrounding DEM mean"),
       `R² on elevation` = sapply(c("Wind","HLI","SxMean","TPI600"), function(v)
         round(summary(lm(as.formula(paste(v, "~ Elevation")), data = Z))$r.squared, 3))) |>
  kable(booktabs = TRUE, align = "lll")
```

## Response specification

@krawchuk2020 warn against coding refugia as a binary, since a refugium is a matter of degree and thresholding discards the signal distinguishing a lightly touched stand from a destroyed one. The survey carries an ordinal severity class, so we can follow that advice. We fit the binary attacked model for comparability with the refugia literature, and alongside it a cumulative logit at each severity threshold. A cumulative logit asks a yes-or-no question at each cut in turn: attacked at all or not, moderate or worse or not, and so on.

Fitting those thresholds separately, rather than imposing one slope across all of them, is deliberate. The single-slope version assumes proportional odds, meaning a predictor pushes a stand up every step of the severity scale by the same amount. Comparing the separate fits is the standard diagnostic for whether that assumption holds, and @sec-ordinal shows that here it does not.

# Results

## Elevational distribution {#sec-host}

```{r}
#| label: tbl-gradient
#| echo: true
#| tbl-cap: "Attack across the elevation gradient, before and after conditioning on lodgepole presence. Conditioning flattens the gradient below 1,900 m but not above it, where attack halves while lodgepole cover rises and stands thin."

S$band <- cut(S$Elevation, breaks = seq(500, 2400, 200), dig.lab = 5)
grad <- S |> group_by(band) |>
  summarise(Cells = n(),
            `Pine cover (%)` = round(mean(PinePct), 1),
            `Basal area` = round(mean(Basal, na.rm = TRUE), 1),
            `Wind (m/s)` = round(mean(Wind), 2),
            `Attacked, all cells (%)` = round(100 * mean(Attacked), 1),
            `Attacked, host present (%)` =
              ifelse(sum(HasHost) > 300, round(100 * mean(Attacked[HasHost]), 1), NA),
            .groups = "drop") |>
  filter(Cells > 500)
grad |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))

## Spread of host-conditioned attack below 1,900 m, quoted in the text below.
hp_low <- S |> filter(HasHost, Elevation < 1900) |>
  mutate(b = cut(Elevation, seq(500, 1900, 200))) |>
  group_by(b) |> summarise(n = n(), pct = 100 * mean(Attacked), .groups = "drop") |>
  filter(n > 300)
HP_RANGE <- range(hp_low$pct)

## Host cover above the threshold, compared two ways, because the two
## comparisons disagree and reporting only the flattering one would be the
## same denominator problem examined in @sec-lowtrap. PINE_LO is
## the whole gradient below the threshold, which is dominated by the pine-poor
## valley; PINE_ADJ is the 150 m band immediately below it.
PINE_HI  <- mean(H$PinePct[H$Elevation >= ALPINE])
PINE_LO  <- mean(H$PinePct[H$Elevation <  ALPINE])
PINE_ADJ <- mean(H$PinePct[H$Elevation >= ALPINE - 150 & H$Elevation < ALPINE])
```

Across the full gradient attack is strongly unimodal (@tbl-gradient, @fig-gradient), rising from under three per cent in the valley to a peak near 1,430 m and falling to zero above 2,100 m.

Conditioning on host presence separates that curve into two phenomena. Below roughly 1,900 m the hump is host distribution: among cells containing lodgepole, attack runs between `r sprintf("%.0f", HP_RANGE[1])` and `r sprintf("%.0f", HP_RANGE[2])` per cent with no strong trend from the valley bottom upward, and the rise in the raw curve is the rise in pine cover. Attack is `r sprintf("%.1f", 100*mean(S$Attacked[S$HasHost]))` per cent where lodgepole is present against `r sprintf("%.1f", 100*mean(S$Attacked[!S$HasHost]))` per cent where it is absent.

Above `r format(ALPINE, big.mark = ",")` m conditioning does not remove the decline. Attack among host-bearing cells falls by about half, while stand basal area falls from `r sprintf("%.1f", mean(H$Basal[H$Elevation<ALPINE], na.rm=TRUE))` to `r sprintf("%.1f", mean(H$Basal[H$Elevation>=ALPINE], na.rm=TRUE))`. Stands are thinner and attack is suppressed.

Whether this is the pattern the @krawchuk2020 hypothesis requires depends on host cover, and on which comparison is made, so both are reported. Against the whole gradient below, mean lodgepole cover in host-bearing cells is `r sprintf("%.1f", PINE_HI)` per cent above the threshold against `r sprintf("%.1f", PINE_LO)` per cent below: essentially unchanged, so host scarcity cannot account for an attack rate that more than halves. Against the `r format(ALPINE - 150, big.mark = ",")` to `r format(ALPINE, big.mark = ",")` m band immediately below, however, cover is `r sprintf("%.1f", PINE_ADJ)` per cent, so the top of the gradient carries about `r sprintf("%.0f", 100 * (1 - PINE_HI / PINE_ADJ))` per cent less pine than the band beneath it. The first comparison is inflated by the pine-poor valley it includes, which is the denominator problem examined in @sec-lowtrap, and it is not relied on alone.

The honest reading is that host thins somewhat at the top while attack falls a great deal more, so the conjunction the hypothesis requires holds in the weaker form. It is not the clean case of host holding constant.

```{r}
#| label: fig-gradient
#| echo: true
#| fig-width: 10.5
#| fig-height: 4.2
#| fig-cap: "(a) Attack across the gradient, raw and conditioned on lodgepole presence, with mean lodgepole cover. (b) The same data against wind speed: wind rises monotonically with elevation while attack turns over, so no monotone wind relationship can hold across the gradient. Shading marks the elevation range of the 2 by 4 km clip."

fine <- S |> mutate(b = cut(Elevation, breaks = seq(500, 2350, 100))) |>
  group_by(b) |>
  summarise(el = mean(Elevation), raw = 100*mean(Attacked), pine = mean(PinePct),
            wind = mean(Wind), n = n(),
            cond = ifelse(sum(HasHost) > 150, 100*mean(Attacked[HasHost]), NA_real_),
            .groups = "drop") |> filter(n > 300)
crng <- range(S$Elevation[S$InClip == 1])

pa <- ggplot(fine) +
  annotate("rect", xmin = crng[1], xmax = crng[2], ymin = -Inf, ymax = Inf,
           fill = "#b2182b", alpha = .10) +
  geom_line(aes(el, raw,  colour = "all cells"), linewidth = .9) +
  geom_line(aes(el, cond, colour = "host present"), linewidth = .9) +
  geom_line(aes(el, pine, colour = "lodgepole cover (%)"), linetype = "22", linewidth = .8) +
  scale_colour_manual(values = c("all cells" = "#2b4162", "host present" = "#b2182b",
                                 "lodgepole cover (%)" = "#3f8f9b"), name = NULL) +
  labs(x = "Elevation (m)", y = "Per cent", title = "(a) Attack and lodgepole cover against elevation") +
  theme(legend.position = "bottom")

pb <- ggplot(fine, aes(wind, raw)) +
  geom_path(colour = "#3f8f9b", linewidth = .8) +
  geom_point(aes(colour = el), size = 1.8) +
  scale_colour_gradientn(colours = c("#2b4162","#3f8f9b","#c6d8af","#e8c07d","#8c4b3c"),
                         name = "Elev (m)") +
  labs(x = "Mean wind speed (m/s)", y = "Attacked (%)",
       title = "(b) Attack against mean wind speed")
pa + pb
```

## Stem diameter effects {#sec-size}

The result above rests on lodgepole cover holding while attack halves. Cover is the wrong host variable to rest it on. What decides whether a stand returns more beetles than it consumes is stem diameter, and the published crossover is about 25 cm [@carroll2004bionomics]. A subalpine stand and a montane stand can carry identical cover and identical basal area while differing entirely in stem size.

```{r}
#| label: tbl-size
#| echo: true
#| tbl-cap: "Attack by stand quadratic mean diameter, in cells carrying lodgepole pine. The published source-sink crossover for lodgepole is about 25 cm [@carroll2004bionomics]. The observed distribution reproduces that threshold without it being imposed."

HQ <- H[!is.na(H$QMD), ]
HQ$dclass <- cut(HQ$QMD, c(0, 15, 20, 25, 30, 35, Inf),
                 labels = c("≤ 15", "15 to 20", "20 to 25",
                            "25 to 30", "30 to 35", "> 35"))

HQ |> group_by(`Quadratic mean diameter (cm)` = dclass) |>
  summarise(Cells = n(), `Attacked (%)` = round(100*mean(Attacked), 1),
            `Lodgepole cover (%)` = round(mean(PinePct), 1),
            `Basal area (m² ha⁻¹)` = round(mean(Basal), 1), .groups = "drop") |>
  kable(booktabs = TRUE, align = "lllll", format.args = list(big.mark = ","))
```

Attack rises monotonically with stem size, from `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$QMD <= 15]))` per cent in the smallest class to `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$QMD > 35]))` per cent in the largest. The steepest step falls at the 25 cm boundary the entomology names. Cover does not order the same way, and in the largest diameter class it is at its lowest, so this is not cover under another name.

```{r}
#| label: tbl-size-elev
#| echo: true
#| tbl-cap: "The high-elevation decline, decomposed. Host cover is flat across the threshold; stem size is not. The final column restricts to stands above the 25 cm source-sink crossover, where the decline persists."

sz <- function(d) tibble(
  Cells = nrow(d),
  `Attacked (%)`         = round(100*mean(d$Attacked), 1),
  `Lodgepole cover (%)`  = round(mean(d$PinePct), 1),
  `Basal area (m² ha⁻¹)` = round(mean(d$Basal), 1),
  `Mean diameter (cm)`   = round(mean(d$QMD), 1),
  `Stands > 25 cm (%)`   = round(100*mean(d$QMD > 25), 1),
  `Attacked, > 25 cm only (%)` = round(100*mean(d$Attacked[d$QMD > 25]), 1))

bind_rows(sz(HQ[HQ$Elevation <  ALPINE, ]), sz(HQ[HQ$Elevation >= ALPINE, ])) |>
  mutate(Band = c(sprintf("below %s m", format(ALPINE, big.mark=",")),
                  sprintf("above %s m", format(ALPINE, big.mark=","))), .before = 1) |>
  kable(booktabs = TRUE, align = "llllllll", format.args = list(big.mark = ","))
```

```{r}
#| label: size-conditioning
#| echo: true
#| output: false

## How much of the high-elevation suppression is stem size? Fit the same
## indicator three ways and read the attenuation. Nested fits on one frame, so
## the coefficients are directly comparable.
HQ$Hi <- HQ$Elevation >= ALPINE
hi_alone <- coef(glm(Attacked ~ Hi, HQ, family = binomial))[["HiTRUE"]]
hi_qmd   <- coef(glm(Attacked ~ Hi + QMD, HQ, family = binomial))[["HiTRUE"]]
hi_full  <- coef(glm(Attacked ~ Hi + QMD + PinePct + Basal,
                     HQ, family = binomial))[["HiTRUE"]]
```

Mean quadratic diameter falls from `r sprintf("%.1f", mean(HQ$QMD[HQ$Elevation < ALPINE]))` cm below the threshold to `r sprintf("%.1f", mean(HQ$QMD[HQ$Elevation >= ALPINE]))` cm above it, and the share of stands above the source-sink crossover falls from `r sprintf("%.1f", 100*mean(HQ$QMD[HQ$Elevation < ALPINE] > 25))` per cent to `r sprintf("%.1f", 100*mean(HQ$QMD[HQ$Elevation >= ALPINE] > 25))` per cent. Cover over the same step moves from `r sprintf("%.1f", mean(HQ$PinePct[HQ$Elevation < ALPINE]))` to `r sprintf("%.1f", mean(HQ$PinePct[HQ$Elevation >= ALPINE]))` per cent. So part of the suppression we attributed to a non-host mechanism is host quality, and reporting cover alone concealed it.

The claim survives the correction. Within source stands only, those above 25 cm, attack still falls from `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$Elevation < ALPINE & HQ$QMD > 25]))` per cent to `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$Elevation >= ALPINE & HQ$QMD > 25]))` per cent across the threshold. Fitted as a single indicator the log-odds of attack above `r format(ALPINE, big.mark=",")` m is `r sprintf("%+.3f", hi_alone)`; adding diameter leaves `r sprintf("%+.3f", hi_qmd)`, and adding cover and basal area alongside it leaves `r sprintf("%+.3f", hi_full)`. Stem size absorbs about `r sprintf("%.0f", 100*(1 - hi_qmd/hi_alone))` per cent of the effect and the full host set about `r sprintf("%.0f", 100*(1 - hi_full/hi_alone))` per cent. The remaining `r sprintf("%.0f", 100*hi_full/hi_alone)` per cent is not host in any coding.

That is the version of the high-elevation result we defend. It is weaker than the one cover alone implies and it rests on the host variable the entomology says governs brood production.

## Wind disruption test {#sec-krawchuk}

This is the first fit of the term review finding R1 identified as proposed but never entered: the component of the wind surface independent of elevation, in cells carrying host.

```{r}
#| echo: true
#| output: false

mt_h <- model_table(ZH)
SUPP_KRAWCHUK <- fmt_model(mt_h) |> kable(booktabs = TRUE, align = "llllll")
```

```{r}
#| label: krawchuk-numbers
#| cache: true
#| echo: true
#| output: false
wr <- mt_h[mt_h$term == "WindResid", ]
pp <- mt_h[mt_h$term == "PinePct", ]
```

Restricted to cells containing host, the orthogonalised wind term is the largest negative coefficient in the model, `r sprintf("%+.3f", wr$beta)` per standard deviation, comparable to the positive effect of lodgepole cover at `r sprintf("%+.3f", pp$beta)` (Table S1). Its sign is reproduced in `r sprintf("%.0f", 100*wr$sign_stab)` per cent of block resamples, but its 95 per cent block interval runs `r sprintf("[%+.3f, %+.3f]", wr$lo, wr$hi)` and reaches zero.

On this fit the hypothesis is supported in direction and unresolved in magnitude, and @sec-nonlinear later withdraws even the direction, so read this section as the conventional analysis whose specification we go on to test rather than a result we stand behind. The descriptive result in @sec-host is not in doubt: attack more than halves at the top of the gradient while host cover thins only slightly. What is in doubt is whether wind is the right explanation, and with roughly `r length(unique(ZH$block))` independent 1 km blocks the binary response cannot settle it. Host cover is the only term whose interval excludes zero.

That verdict is a property of the response coding rather than of the data, and @sec-ordinal revisits it. We report the binary model first because it is the form the refugia literature uses, and the comparison is itself a result.

Elevation, entered alongside the orthogonalised wind term, is itself negative and unsupported: once host is in the model the remaining elevation signal is weak, so the suppression at the top is not simply an elevation effect.

## Ordinal severity models {#sec-ordinal}

```{r}
#| label: polr-fit
#| cache: true
#| echo: true
#| output: false

## Fitted only to show what a single proportional-odds summary conceals; the
## threshold-specific fits in @tbl-ordinal are the result.
zo <- ZH; zo$sev <- factor(pmin(zo$Severity, 4L), levels = 0:4, ordered = TRUE)
polr_fit <- MASS::polr(sev ~ Elevation + WindResid + PinePct + Basal +
                             TPI600 + Slope + TWI + RIX,
                       data = zo, method = "logistic", Hess = FALSE)
```

```{r}
#| label: tbl-ordinal
#| cache: true
#| echo: true
#| tbl-cap: "Cumulative logits at each severity threshold, cells containing lodgepole pine only. The orthogonalised wind coefficient strengthens monotonically as the threshold rises and its block interval excludes zero above trace, which the binary model does not show."

CUTS <- tibble(k = 1:4,
               label = c("trace or worse", "light or worse",
                         "moderate or worse", "severe or worse"))

cum <- lapply(seq_len(nrow(CUTS)), function(i) {
  z <- ZH; z$y <- as.integer(z$Severity >= CUTS$k[i])
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Threshold = CUTS$label[i], Cells = sum(z$y),
         `Wind residual beta` = coef(fit)[["WindResid"]],
         lo = q[1], hi = q[2],
         `sign stability` = mean(sign(bs[, "WindResid"]) ==
                                 sign(coef(fit)[["WindResid"]]), na.rm = TRUE),
         `Pine cover beta` = coef(fit)[["PinePct"]],
         `Elevation beta`  = coef(fit)[["Elevation"]])
}) |> bind_rows()

cum |> transmute(Threshold, Cells,
                 `Wind residual beta` = sprintf("%+.3f", `Wind residual beta`),
                 `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
                 `sign stability` = sprintf("%.2f", `sign stability`),
                 `Pine cover beta` = sprintf("%+.3f", `Pine cover beta`),
                 Supported = ifelse(lo * hi > 0, "yes", "no")) |>
  kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

Collapsing the response to attacked or not costs the study its main result. Fitted at successively harder severity thresholds, the orthogonalised wind coefficient strengthens monotonically from `r sprintf("%+.3f", cum$"Wind residual beta"[1])` at trace or worse to `r sprintf("%+.3f", cum$"Wind residual beta"[4])` at severe or worse, a `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])`-fold increase (@tbl-ordinal). At every threshold above trace the block interval excludes zero. The first row is the binary model of @sec-krawchuk restated, since a cell reaching trace or worse is exactly a cell recorded as attacked, so the table extends that fit rather than competing with it. It is also the weakest of the four, and the one the refugia literature would conventionally have reported.

This is a dose-response gradient of the shape the hypothesis predicts. Wherever the wind surface has a component independent of elevation, that component is associated with less attack, and the association is strongest where attack was heaviest. It also explains the binary result: a cell that reached trace severity is barely distinguishable from an unattacked cell, so a threshold drawn between them puts most of the noise exactly where the model is asked to fit.

The slopes differ by a factor of `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])` across the thresholds, so proportional odds does not hold and a single ordinal coefficient would be the wrong summary. Fitting one anyway returns `r sprintf("%+.3f", coef(polr_fit)[["WindResid"]])`, close to the trace-threshold estimate and concealing everything above it. Two other terms move with the threshold. Lodgepole cover, the only term that cleared the interval test in the binary model, weakens as the threshold rises and reverses at severe or worse (`r sprintf("%+.3f", cum$"Pine cover beta"[4])`). One reading is ecological: host abundance governs whether a stand is attacked at all, not how hard. A second is an artefact of the inventory's vintage, and we cannot separate them. Composition was interpreted from 2014 photography, late in the outbreak window, so stands that lost their pine early carry a lower recorded lodgepole percentage than when the beetle reached them. That predicts a negative coefficient at the heaviest severities, which is what is observed. We report the reversal and decline to interpret it. Elevation turns positive at moderate or worse (`r sprintf("%+.3f", cum$"Elevation beta"[3])`), consistent with severe attack having concentrated in the mid-elevation pine belt rather than at either end.

We cannot attribute the gradient to wind rather than cold, and the Discussion explains why the ordinal result does not help. More seriously, @sec-nonlinear shows the sweep does not survive entering elevation flexibly, so what looks here like improved support is largely the unimodal elevation response returning through a predictor orthogonalised only against the linear part of elevation. Read this table together with that one.

## Nonlinear elevation specification {#sec-nonlinear}

```{r}
#| label: tbl-nonlinear
#| cache: true
#| echo: true
#| tbl-cap: "The ordinal sweep refitted with elevation entered flexibly instead of linearly, and wind residualised against that same basis. Elevation explains eleven points more of the wind surface once curvature is allowed, and no threshold clears the interval test under either flexible basis."

## Everything above enters elevation as one linear term, and orthogonalises wind
## against that same straight line. But @fig-gradient shows the response is
## unimodal in elevation, and a straight line through a hump leaves a structured,
## elevation-shaped residual that any covariate correlated with the curvature can
## absorb. The wind surface is such a covariate by construction, because the
## downscaling is a nonlinear function of terrain. Residualising against a line
## cannot remove curvature the line never described. This table tests that.

elev_basis <- function(E, basis) switch(basis,
  linear = cbind(ElevL = E),
  quad   = cbind(ElevL = E, ElevQ = E^2),
  spline = { b <- splines::ns(E, df = 4); colnames(b) <- paste0("ElevS", 1:4); as.matrix(b) })

nl_row <- function(basis) {
  z  <- ZH
  EB <- elev_basis(z$Elevation, basis)
  z  <- cbind(z[, setdiff(names(z), "Elevation")], as.data.frame(EB))
  ## wind residualised on THIS basis, not on the straight line
  z$WindResid <- as.numeric(scale(residuals(lm(z$Wind ~ EB))))
  vars <- c(colnames(EB), "WindResid", "PinePct", "Basal", "TPI600", "Slope", "TWI", "RIX")
  r2 <- summary(lm(z$Wind ~ EB))$r.squared
  lapply(seq_len(nrow(CUTS)), function(i) {
    z$y <- as.integer(z$Severity >= CUTS$k[i])
    fit <- fit_fast(design_matrix(z, vars), z$y)
    bs  <- block_boot(z, vars)
    q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
    tibble(`Elevation basis` = basis, `R² wind on basis` = round(r2, 3),
           Threshold = CUTS$label[i],
           `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
           `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
           Supported = ifelse(q[1]*q[2] > 0, "yes", "no"))
  }) |> bind_rows()
}
nl <- bind_rows(lapply(c("linear","quad","spline"), nl_row))
nl |> kable(booktabs = TRUE, align = "llllll")
```

@sandel2015 gives the diagnostic that motivates this test. For a genuinely linear relationship a coefficient is extent-independent and only the coefficient of determination moves as extent changes; a coefficient that changes sign under truncation is a symptom of an unmodelled nonlinearity rather than a general law of extent. Our response is unimodal in elevation and our wind surface monotone in it, exactly the configuration that produces such a symptom, so the specification has to be tested rather than assumed.

It does not pass. Entering elevation as a quadratic raises the share of the wind surface that elevation explains from `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[1]))` to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[5]))`, and a natural spline on four degrees of freedom raises it to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[9]))`. Roughly eleven percentage points of the wind surface's variance is therefore nonlinear in elevation, and the linear orthogonalisation used above leaves all of it inside what we had been calling the component of wind independent of elevation.

When that residue is removed the wind result goes with it. The coefficient at severe or worse falls from `r nl$"Wind residual beta"[4]` to `r nl$"Wind residual beta"[8]`, and at moderate or worse from `r nl$"Wind residual beta"[3]` to `r nl$"Wind residual beta"[7]`, losing between three fifths and seven tenths of its magnitude. Under both flexible bases every interval covers zero at every threshold. The monotone strengthening across thresholds survives in the point estimates, but nothing in the sweep clears the test this paper set itself.

We report this as the result rather than relegating it to a sensitivity appendix, because it changes what the preceding two sections mean. The dose-response in @tbl-ordinal is not evidence about wind. It is very largely the unimodal elevation response reappearing through a predictor that was never orthogonal to the curvature of elevation, only to its slope. This study does not have a wind effect to report, and the appearance of one followed from how elevation was specified.

## Aspect-derived predictors {#sec-aspect}

```{r}
#| label: tbl-aspect
#| cache: true
#| echo: true
#| tbl-cap: "The host-present model with the two aspect-derived terms added, at each threshold it can carry. Heat load is positive and strengthens sharply with severity: warmer aspects burn harder at the same elevation. The orthogonalised wind term is not displaced by it. The shelter index carries the opposite sign, and this result is examined rather than set aside."

FULLV <- c(MODEL_VARS, EXTRA)
ZHx   <- ZH[complete.cases(ZH[, FULLV]), ]

## Ten terms and 1,522 events cannot be fitted together: the severe threshold
## drives fitted probabilities to numerical zero. We detect that rather than
## report the resulting coefficients, and the severe row is dropped below.
sep_check <- function(z, vars, k) {
  z$y <- as.integer(z$Severity >= k)
  p <- fit_fast(design_matrix(z, vars), z$y)$fitted.values
  sum(p < 1e-6 | p > 1 - 1e-6)
}
sep <- sapply(1:4, function(k) sep_check(ZHx, FULLV, k))
names(sep) <- c("trace+","light+","moderate+","severe+")
OK_CUTS <- which(sep == 0)

asp_tab <- lapply(OK_CUTS, function(k) {
  z <- ZHx; z$y <- as.integer(z$Severity >= k)
  fit <- fit_fast(design_matrix(z, FULLV), z$y); bs <- block_boot(z, FULLV)
  ci <- function(n) quantile(bs[, n], c(.025, .975), na.rm = TRUE)
  one <- function(n) sprintf("%+.3f [%+.3f, %+.3f]", coef(fit)[[n]], ci(n)[1], ci(n)[2])
  tibble(Threshold = CUTS$label[k], Cells = sum(z$y),
         `Wind residual` = one("WindResid"), `Heat load` = one("HLI"),
         `Shelter` = one("SxMean"))
}) |> bind_rows()

asp_tab |> kable(booktabs = TRUE, align = "lllll",
                 format.args = list(big.mark = ","))
```

The inherited predictor set had no aspect, so until now the model could not tell a warm slope from a cold one at the same elevation. Adding two aspect-derived terms does three things.

First, it supplies the one piece of thermal information available without a temperature surface. Heat load is positive at every threshold and strengthens steeply with severity (@tbl-aspect). Since heat load is essentially uncorrelated with elevation (@tbl-orthogonality), this is not the elevation gradient under another name: at a given elevation, southwest-facing ground carries heavier mortality than northeast-facing ground. That is the sign cold limitation predicts, and the first evidence here that the thermal axis works independently of elevation.

Second, it leaves the wind coefficient standing. The orthogonalised wind term does not weaken when heat load is controlled for, so whatever it tracks is not the warm-aspect effect in disguise. It is, however, largely the curvature of elevation: this model still enters elevation linearly, and @sec-nonlinear shows the wind term does not survive a flexible basis. Controlling for aspect does not rescue it.

Third, and least comfortably, the two exposure measures disagree. The shelter index is negative, meaning sheltered ground carries *less* attack and exposed ground more, the opposite of what the orthogonalised wind term says and of what @krawchuk2020 predict. The two are only weakly correlated (`r sprintf("%.2f", cor(ZHx$WindResid, ZHx$SxMean))`), so they are not two measurements of one quantity that happen to disagree; they measure different things, and only one is a direct geometric description of exposure. We take this as evidence against treating any single surface as "wind exposure" and reporting its coefficient as though the phrase had one referent. It is also a caution against our own result: the wind residual's negative sign is not corroborated by the one predictor here that describes exposure geometrically rather than by downscaling.

The severe threshold is absent from @tbl-aspect because ten terms and `r sum(ZHx$Severity >= 4)` events drive the fit into quasi-separation, with `r format(sep[["severe+"]], big.mark = ",")` cells at numerically zero fitted probability. The eight-term model of @tbl-ordinal is free of this, which is why it retains its severe row and this one does not.

## Winter temperature thresholds {#sec-cold}

Table S2 summarises the record and Figure S2 maps the network. No station lies inside the analysis grid, and the stations reaching the coldest minima are not the highest ones, which is the point the next paragraph develops.

```{r}
#| echo: true
#| output: false

st <- read.csv(here("02.inputs/climate/station-winter-minima.csv"), stringsAsFactors = FALSE)

## The censoring rule is applied here rather than taken on trust from the
## retrieval script. Many older British Columbia Wildfire Service sensors floor
## at exactly -20.0 C. A station-winter is unusable when its minimum sits
## exactly on that floor, because the true minimum is then unknown. A minimum
## colder than the floor shows the sensor did read past it, so readings at
## -20.0 in those winters are real measurements and the winter is kept. The
## floor is a sensor artefact of that network, so the rule does not touch the
## Environment and Climate Change Canada records.
FLOOR <- -20
st$censored <- st$network == "BCWS" &
               abs(st$min_temp_c - FLOOR) < 1e-6 &
               st$n_at_minus20 > 0
stopifnot(identical(st$censored, as.logical(st$censored_flag)))

## Isolated single-observation spikes are rejected at the observation level by
## the retrieval script, which is the only stage holding the sub-daily series a
## per-winter summary cannot carry; `min_temp_c_usable` is that de-spiked
## minimum. Everything else about the screen is visible above.
use <- st[!st$censored & !is.na(st$min_temp_c_usable), ]

## Warm sensor failure. The retrieval screen inspects only the minimum, so a
## sensor that fails warm passes through it invisibly and contributes a
## spuriously mild minimum. NORNS, the highest station in the network at 2,423 m,
## did this through 2015: a February mean daily minimum of +30.0 C and a winter
## "minimum" of +4.6 C. The bias runs towards the conclusion that the landscape
## never gets cold enough, so it is screened here and its effect is reported.
## `build-monthly-climatology.R` derives the monthly means from the same cached
## extracts and carries the flag; the cut is read off that distribution.
clim <- read.csv(here("02.inputs/climate/station-monthly-climatology.csv"),
                 stringsAsFactors = FALSE)
bad_sw <- unique(clim[clim$warm_fail == "TRUE" | clim$warm_fail == TRUE,
                      c("station_name", "winter_year")])
WARM_DROPPED <- nrow(merge(use, bad_sw, by = c("station_name", "winter_year")))
MEAN_HI_RAW  <- mean(use$min_temp_c_usable[use$elevation_m >= 1500])
use <- use[!(paste(use$station_name, use$winter_year) %in%
             paste(bad_sw$station_name, bad_sw$winter_year)), ]
## No physically impossible winter minimum may survive the screen.
stopifnot(max(use$min_temp_c_usable) < 0)

use$band <- cut(use$elevation_m, c(0, 800, 1500, 2500),
                labels = c("below 800 m", "800 to 1,500 m", "above 1,500 m"))
## Held for the supplementary section rather than printed here, so the main
## text keeps to the ten floats the guidelines allow.
SUPP_COLD <- use |> group_by(band) |>
  summarise(Stations = n_distinct(station_name), `Station-winters` = n(),
            `Coldest (°C)` = min(min_temp_c_usable),
            `Mean minimum (°C)` = round(mean(min_temp_c_usable), 1),
            .groups = "drop") |>
  kable(booktabs = TRUE, align = "lllll")
```

```{r}
#| echo: true
#| output: false
#| fig-width: 5.9
#| fig-height: 5.1

stn <- aggregate(cbind(latitude, longitude, elevation_m) ~ station_name,
                 data = st, FUN = function(x) x[1])
coldest_by <- aggregate(min_temp_c_usable ~ station_name, data = use, FUN = min)
stn <- merge(stn, coldest_by, by = "station_name", all.x = TRUE)
stn_sf <- st_transform(st_as_sf(stn, coords = c("longitude","latitude"), crs = 4326), 3005)
xy <- st_coordinates(stn_sf); stn_sf$X <- xy[, 1]; stn_sf$Y <- xy[, 2]
site_bc <- st_transform(st_as_sf(as.polygons(ext(G), crs = crs(G))), 3005)
prov_bc <- st_transform(ne_states(country = c("Canada","United States of America"),
                                  returnclass = "sf"), 3005)
bbs <- st_bbox(st_buffer(st_union(c(st_geometry(stn_sf), st_geometry(site_bc))), 25000))
sw  <- unname(bbs["xmax"] - bbs["xmin"])
key <- stn_sf[order(stn_sf$min_temp_c_usable)[1:2], ]

fig_station <- ggplot() +
  geom_sf(data = prov_bc, fill = "grey95", colour = "grey70", linewidth = .25) +
  geom_sf(data = site_bc, fill = NA, colour = "#C0392B", linewidth = .6) +
  geom_sf(data = stn_sf, aes(fill = elevation_m), shape = 21, size = 1.9,
          colour = "grey20", stroke = .25) +
  scale_fill_viridis_c(option = "viridis", name = "Station\nelevation (m)") +
  geom_label_repel(data = as.data.frame(key), aes(X, Y, label = tolower(station_name)),
                   size = 2, colour = "grey10", fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm"), box.padding = .5,
                   min.segment.length = 0, segment.colour = "grey30",
                   segment.size = .3, seed = 1) +
  annotate("label", x = bbs["xmin"] + .02*sw,
           y = bbs["ymin"] + .04*(bbs["ymax"] - bbs["ymin"]),
           label = "red outline:\nanalysis grid", size = 1.9, hjust = 0, vjust = 0,
           colour = "#C0392B", fill = "white", alpha = .85,
           label.padding = unit(.7, "mm")) +
  coord_sf(crs = 3005, xlim = bbs[c("xmin","xmax")], ylim = bbs[c("ymin","ymax")],
           datum = st_crs(3005), expand = FALSE) +
  scale_x_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  scale_y_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  north + map_theme +
  labs(title = "Weather stations screened for winter minima",
       subtitle = sub_alb)
fig_station
ggsave("../03.outputs/PNG/fig-station-maps.png", fig_station,
       width = 150, height = 130, units = "mm", dpi = 500, bg = "white")
```

The cold explanation for the high-elevation suppression assumes it gets cold enough up there to kill beetles. Testing that assumption needs the right threshold, and the threshold is not a single number.

Lethal temperature for this insect depends on the stage overwintering and on when the cold arrives. Cold tolerance is acquired by accumulating glycerol through the autumn and is lost again as the beetle metabolises it in late winter. Larvae are the most tolerant stage, then adults, then pupae, then eggs. Within the larval stages tolerance rises with instar. @safranyik2006chap1 give lethal ranges of -23 to -29 °C for first instars, -23 to -34 °C for second instars, and -29 to -40 °C for third and fourth instars, after Logan and colleagues. The commonly quoted -40 °C is the cold end of the range for the late instars that normally overwinter, not a threshold for the population. The same source is explicit that timing governs the rest: a -30 °C event in midwinter would be expected to kill few beetles, while the same temperature before hardening is complete would kill many.

Assessed against these stage-specific ranges rather than against a single benchmark, the record supports a narrower conclusion.

```{r}
#| label: cold-numbers
#| echo: true
#| output: false

coldest <- use[which.min(use$min_temp_c_usable), ]
hi      <- use[use$elevation_m >= 1500, ]
lowcold <- use[use$elevation_m < 800, ]
lowcold <- lowcold[which.min(lowcold$min_temp_c_usable), ]

## Date of each winter minimum. Lethal temperature is conditional on the state of
## cold-hardening, so a minimum without a date cannot be compared with a lethal
## range. The per-station-winter summary the fetch writes does not carry the
## timestamp; `add-minimum-dates.R` recovers it from the same cached hourly
## extracts and is documented alongside the fetch.
dates <- read.csv(here("02.inputs/climate/station-winter-minimum-dates.csv"),
                  stringsAsFactors = FALSE)
useD <- merge(use, dates[, c("station_id","winter_year","min_month","min_date")],
              by = c("station_id","winter_year"))
## The merge must not silently drop or duplicate station-winters.
stopifnot(nrow(useD) == sum(use$network == "BCWS"))
DEC <- 100 * mean(useD$min_month == 12)
```

@fig-climate shows the record three ways: the shape of the winter by elevation band, the year-to-year sequence of minima against the lethal ranges for each overwintering stage, and the ordering of minima by station elevation.

```{r}
#| label: fig-climate
#| echo: true
#| fig-width: 7.2
#| fig-height: 8.4
#| fig-cap: "The observational climate record behind the cold check. (a) Winter climograph by elevation band: mean daily maximum, monthly mean and mean daily minimum air temperature, with monthly precipitation as bars rising from the floor of the temperature axis and read against the right-hand scale. Only December, January and February were retrieved, so the shoulder months in which a given temperature is most lethal are absent. The apparent decline of winter precipitation with elevation is a gauge artefact: the high-elevation stations are tipping-bucket instruments, which under-catch snowfall severely, and the bars are shown for the seasonal shape within a band rather than for comparison between bands. (b) Winter minima by year, one point per station-winter, against the lethal ranges reported by @safranyik2006chap1 for first-instar larvae and for the third and fourth instars that normally overwinter. No station-winter enters the late-instar range. (c) The same minima against station elevation, showing that elevation orders them only weakly because cold air pools in the valleys."

MONTH_ORDER <- c(12, 1, 2)
band_of <- function(e) cut(e, c(0, 800, 1500, 2500),
                           labels = c("below 800 m", "800 to 1,500 m", "above 1,500 m"))

cl <- clim[!(clim$warm_fail %in% c(TRUE, "TRUE")) & clim$n_days >= 20, ]
cl$band  <- band_of(cl$elevation_m)
cl$month <- factor(cl$month, levels = MONTH_ORDER,
                   labels = c("Dec", "Jan", "Feb"))

## Precipitation is plotted on a secondary axis, so it is rescaled onto the
## temperature axis once here rather than inside the geom.
PR_MAX <- max(tapply(cl$total_precip_mm, list(cl$band, cl$month), mean, na.rm = TRUE),
              na.rm = TRUE)
T_LO   <- -20
PR_SCALE <- abs(T_LO) / PR_MAX

clm <- cl |>
  group_by(band, month) |>
  summarise(tmax = mean(mean_daily_max_c, na.rm = TRUE),
            tavg = mean(mean_temp_c,      na.rm = TRUE),
            tmin = mean(mean_daily_min_c, na.rm = TRUE),
            prcp = mean(total_precip_mm,  na.rm = TRUE),
            .groups = "drop")

## Bars are drawn from the floor of the temperature axis upward. A geom_col here
## would draw from zero downward, so a dry month would render as the tallest bar.
clm$xi <- as.integer(clm$month)

p_clim <- ggplot(clm, aes(month, group = band)) +
  geom_rect(aes(xmin = xi - .3, xmax = xi + .3,
                ymin = T_LO, ymax = prcp * PR_SCALE + T_LO),
            fill = "grey82", colour = NA) +
  geom_ribbon(aes(ymin = tmin, ymax = tmax), fill = "#2C7FB8", alpha = .18) +
  geom_line(aes(y = tmax), colour = "#C0392B", linewidth = .5) +
  geom_line(aes(y = tmin), colour = "#2C7FB8", linewidth = .5) +
  geom_line(aes(y = tavg), colour = "grey20", linewidth = .7) +
  geom_point(aes(y = tavg), colour = "grey20", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = .3) +
  facet_wrap(~band, nrow = 1) +
  scale_y_continuous(
    name = "Air temperature (°C)", limits = c(T_LO, NA),
    sec.axis = sec_axis(~ (. - T_LO) / PR_SCALE, name = "Precipitation (mm)")) +
  labs(x = NULL, title = "(a) Winter climograph by elevation band",
       subtitle = "Red: mean daily maximum. Black: monthly mean. Blue: mean daily minimum. Bars: precipitation") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank(),
        plot.subtitle = element_text(colour = "grey35"))

## Lethal ranges are stage-specific and are quoted from the same source the text
## uses, so the figure and the prose cannot drift apart.
leth <- data.frame(
  stage = factor(c("first instar", "third and fourth instars"),
                 levels = c("first instar", "third and fourth instars")),
  lo = c(-29, -40), hi = c(-23, -29))

useP <- use
useP$band <- band_of(useP$elevation_m)

p_series <- ggplot() +
  geom_rect(data = leth, aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi,
                             fill = stage), alpha = .18) +
  scale_fill_manual(values = c("first instar" = "#F39C12",
                               "third and fourth instars" = "#2C7FB8"),
                    name = "Lethal range") +
  geom_point(data = useP, aes(winter_year, min_temp_c_usable, colour = elevation_m),
             size = 1.5, alpha = .9) +
  scale_colour_viridis_c(name = "Station\nelevation (m)") +
  scale_x_continuous(breaks = seq(2000, 2015, 5)) +
  labs(x = "Winter (year of January)", y = "Winter minimum (°C)",
       title = "(b) Observed winter minima against stage-specific lethal ranges") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank())

p_elev <- ggplot(useP, aes(elevation_m, min_temp_c_usable)) +
  geom_rect(data = leth, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi, fill = stage),
            alpha = .18, show.legend = FALSE) +
  scale_fill_manual(values = c("first instar" = "#F39C12",
                               "third and fourth instars" = "#2C7FB8")) +
  geom_point(aes(colour = winter_year), size = 1.5, alpha = .9) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "grey20", linewidth = .5) +
  scale_colour_viridis_c(option = "magma", name = "Winter") +
  labs(x = "Station elevation (m)", y = "Winter minimum (°C)",
       title = "(c) Winter minimum against station elevation") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank())

fig_climate <- p_clim / p_series / p_elev + plot_layout(heights = c(1, 1, 1))
fig_climate
ggsave("../03.outputs/PNG/fig-climate.png", fig_climate,
       width = 183, height = 213, units = "mm", dpi = 500, bg = "white")
```

```{r}
#| label: climate-lapse
#| echo: true
#| output: false

## The lapse rate the record supports, quoted in the text. Reported because the
## alternative to a station record is an assumed lapse rate, and a referee is
## entitled to see what this landscape's own stations give.
lap <- lm(min_temp_c_usable ~ elevation_m, data = use)
LAPSE   <- unname(coef(lap)[2]) * 1000
LAPSE_R2 <- summary(lap)$r.squared
```

Across the screened record the winter minimum falls by `r sprintf("%.1f", abs(LAPSE))` °C per 1,000 m of station elevation, and elevation explains `r sprintf("%.2f", LAPSE_R2)` of the variance in it. A predictor that weak cannot carry a thermal explanation on its own, which is the quantitative form of the cold-air pooling the next paragraphs describe.

Across `r nrow(use)` station-winters from `r n_distinct(use$station_name)` stations spanning `r min(use$elevation_m)` to `r max(use$elevation_m)` m, the coldest defensible winter minimum anywhere in the region is `r sprintf("%.1f", coldest$min_temp_c_usable)` °C, at `r tolower(coldest$station_name)` at `r format(round(coldest$elevation_m), big.mark=",")` m in winter `r coldest$winter_year`. Above 1,500 m the mean winter minimum is `r sprintf("%.1f", mean(hi$min_temp_c_usable))` °C. No station-winter reaches -35 °C.

That coldest value sits inside the lethal range for first and second instars and at the warm edge of the range for third and fourth. It is not far short of any lethal range. It lies above the range for the stage that normally overwinters and within the range for stages that do not. Whether it killed anything depends on which stages were present, which is a function of the same heat accumulation @sec-aspect is about.

The timing of the minima qualifies this further. Of the Wildfire Service station-winters, `r sprintf("%.0f", DEC)` per cent reach their minimum in December. So do `r sum(useD$min_month[order(useD$min_temp_c_usable)][1:10] == 12)` of the ten coldest, the hardest of them `r sprintf("%.1f", min(useD$min_temp_c_usable[useD$min_month == 12]))` °C on `r format(as.Date(useD$min_date[which.min(ifelse(useD$min_month == 12, useD$min_temp_c_usable, Inf))]), "%e %B %Y")`. December is the front of the hardening window, where a given temperature does more damage than it would in January.

Two further features of the record matter. Elevation orders these minima only weakly: `r tolower(lowcold$station_name)`, at `r format(round(lowcold$elevation_m), big.mark=",")` m, reaches `r sprintf("%.1f", lowcold$min_temp_c_usable)` °C, within `r sprintf("%.1f", abs(lowcold$min_temp_c_usable - coldest$min_temp_c_usable))` °C of the alpine station `r format(round(coldest$elevation_m - lowcold$elevation_m), big.mark=",")` m higher. This is consistent with cold-air pooling, and it is the reason temperature cannot be treated as a linear function of elevation on this landscape. And two separate sensor faults were screened out. `r sum(st$censored)` station-winters were set aside because the sensors floor at exactly -20.0 °C, an artefact concentrated before 2008; a naive minimum over those would have manufactured a threshold that is not in the data. A further `r WARM_DROPPED` were set aside for the opposite fault. The retrieval screen inspects only the minimum, so a sensor that fails warm is invisible to it, and the highest station in the network returned a February mean daily minimum of +30.0 °C and a winter "minimum" above freezing at 2,423 m through 2015. Retaining it would have raised the mean minimum above 1,500 m from `r sprintf("%.1f", mean(hi$min_temp_c_usable))` °C to `r sprintf("%.1f", MEAN_HI_RAW)` °C. That bias runs towards the conclusion this section reaches, which is why the screen is applied in the document rather than in the retrieval.

The limits of this check are more extensive than the check itself and are set out in full. The retrieval window runs from 1 December to the end of February, since the script was built to find deep-winter minima. October, November, March and April are absent from the record altogether, and those are precisely the months in which a given temperature is most lethal, before hardening completes or after it reverses. No station inside the study area reported during the outbreak window, and the stations that did report are spread across roughly 100 km of separate massifs. The quantity that actually governs survival is temperature under the bark, not air temperature, and it is observed at landscape scale nowhere.

The defensible conclusion is correspondingly narrow. Deep-winter air temperature in this region does not approach the lethal range for late-instar larvae. That does not establish that freeze mortality is absent, because the record cannot see the shoulder months, cannot see inside the study area, and cannot see under the bark. Cold remains a live mechanism here, alongside the heat-accumulation constraint of @sec-aspect, and this study separates neither from wind.

## Sensitivity to extent {#sec-extent}

The three sections that follow answer objective 4 and test review finding R4, refitting one model across extent, grain and response coding on one dataset.

```{r}
#| label: tbl-extent
#| cache: true
#| echo: true
#| tbl-cap: "The same model fitted at three extents. As the analysis area shrinks, the share of wind variance explained by elevation rises and the orthogonalised wind coefficient reverses sign. The 2 by 4 km clip is the extent at which the companion study was run."

extent_row <- function(z, lab) {
  r2  <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  ## The block count is reported because it, not the cell count, is the
  ## resampling unit. The clip holds few enough blocks that its interval is
  ## far weaker than its cell count suggests, and the reader must see that.
  tibble(Extent = lab, Cells = nrow(z), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         ## This is the variance inflation the *raw* wind term would carry
         ## against elevation. It is not the VIF of the fitted model, in which
         ## the residualised term is orthogonal to elevation by construction and
         ## its VIF is 1. It is reported to show the collinearity that
         ## orthogonalisation reparameterises away rather than removes.
         `VIF of raw wind` = round(1/(1-r2), 1),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]))
}
bind_rows(
  extent_row(Z,  sprintf("full landscape (%.0f km²)", nrow(S)*prod(res(G))/1e6)),
  extent_row(ZH, "landscape, host present"),
  extent_row(ZC, sprintf("2 by 4 km clip (%.1f km²)",
                         sum(S$InClip == 1) * prod(res(G)) / 1e6))
) |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

The confound is a function of extent, and so is the answer. Across the landscape, elevation explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = Z))$r.squared)` per cent of the variance in the wind surface; inside the 2 by 4 km clip it explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` per cent (@tbl-extent). The orthogonalised wind coefficient is negative at landscape extent and positive inside the clip.

Two claims live in that table and carry very different weight.

The collinearity result is firm. The share of wind variance elevation explains rises from `r sprintf("%.3f", summary(lm(Wind ~ Elevation, data = Z))$r.squared)` across the landscape to `r sprintf("%.3f", summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` inside the clip, and that quantity is a property of the two surfaces rather than an estimate with sampling error. It is the mechanism this paper is about, and does not depend on resolving the Krawchuk hypothesis.

```{r}
#| label: clip-interval
#| cache: true
#| echo: true
#| output: false

## This interval used to be computed inside an inline expression in the sentence
## below. That hid 150 model refits inside a line of prose, where no reader could
## see the cost and no cache could reach it.
CLIP_BLOCKS <- length(unique(ZC$block))
CLIP_CI     <- local({
  q <- quantile(block_boot(ZC)[, "WindResid"], c(.025, .975), na.rm = TRUE)
  sprintf("[%+.3f, %+.3f]", q[1], q[2])
})
```

The sign reversal is a comparison of two point estimates and is not, on this evidence, a demonstrated reversal. The clip contains only `r CLIP_BLOCKS` spatial blocks, so its block interval runs `r CLIP_CI`: it spans zero, and it spans both landscape estimates. By the rule this paper set itself the clip coefficient is not distinguishable from zero, from the landscape estimate, or from its own opposite. What the clip demonstrates is that shrinking the extent destroys the ability to estimate the coefficient at all, a weaker claim than reversal and still damaging for the practice we criticise: a study at this extent would report a positive point estimate with an interval too wide to notice it had learned nothing. Establishing reversal proper would need an extent-by-wind interaction fitted on the pooled data and bootstrapped on common blocks. We have not done that and do not claim it.

## Sensitivity to grain {#sec-grain}

```{r}
#| label: grain-fit
#| cache: true
#| echo: true
#| output: false

## Not `tbl-grain-fit`: a `tbl-` prefix declares a float reference target, and this
## chunk sets `output: false`, so Quarto warned "FloatRefTarget with no content".
## The prefix is reserved; a chunk that emits nothing must not claim one.
## The four option lines above must stay above these comments: Quarto only reads
## `#|` options from the top of a cell, and until 2026-08-19 they sat below and
## were silently ignored, so this chunk was neither labelled, cached nor silenced.

CONT <- c("Elevation","Slope","TWI","RIX","Wind","PinePct","Basal")
base_stack <- c(terr[[c("Elevation","Slope","TWI","RIX","Wind")]],
                host[[c("PinePct","Basal")]], severity, in_clip)
names(base_stack) <- c(CONT, "Severity", "InClip")

## fact is the aggregation factor from the native 25 m grid.
build_grain <- function(fact) {
  b <- if (fact == 1L) base_stack else
       c(aggregate(base_stack[[CONT]], fact, fun = "mean", na.rm = TRUE),
         aggregate(base_stack[[c("Severity","InClip")]], fact, fun = "max", na.rm = TRUE))
  Gi <- b[["Elevation"]]; rs <- res(Gi)[1]
  w  <- max(3L, as.integer(2 * floor(625 / rs / 2) + 1))   # odd window nearest 625 m
  b$TPI600 <- tpi(Gi, w)
  d <- as.data.frame(b, na.rm = FALSE)
  d$Attacked <- d$Severity > 0
  d$HasHost  <- !is.na(d$PinePct) & d$PinePct > 0
  blk <- max(1L, as.integer(round(BLOCK / fact)))          # blocks stay near 1 km
  rc  <- rowColFromCell(Gi, seq_len(ncell(Gi)))
  d$block <- paste(rc[,1] %/% blk, rc[,2] %/% blk, sep = "_")
  d <- d[complete.cases(d[, MVARS]), ]
  structure(d, res = rs, win = w * rs)
}

grain_row <- function(fact) {
  d  <- build_grain(fact); h <- d[d$HasHost, ]
  z  <- standardise(h)
  r2 <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y); bs <- block_boot(z)
  q  <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Grain = sprintf("%.0f m", attr(d, "res")),
         `TPI window` = sprintf("%.0f m", attr(d, "win")),
         `Host cells` = nrow(h), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
         `sign stability` = sprintf("%.2f",
           mean(sign(bs[, "WindResid"]) == sign(coef(fit)[["WindResid"]]), na.rm = TRUE)))
}

## The 25 m row must reproduce the analysis frame built in `assemble`; if this
## assertion ever fails the two code paths have drifted apart.
stopifnot(nrow(build_grain(1L)) == nrow(S))

SUPP_GRAIN <- bind_rows(lapply(c(1L, 2L, 4L), grain_row)) |>
  kable(booktabs = TRUE, align = "llllllll", format.args = list(big.mark = ","))
```

@work2011, working on carabid beetles, found the apparent importance of terrain derivatives changed with the resolution at which they were computed, because elevation error propagates into derived curvature differently at every grain. The concern transfers directly and is worth separating from extent, which is routinely confused with it.

Here they behave quite differently (Table S3). Aggregating from 25 m to 100 m, a sixteen-fold reduction in cells, leaves the confound essentially untouched: elevation explains close to the same share of wind variance at every grain, and the orthogonalised wind coefficient neither reverses nor decays. The suppression at the top of the gradient is a landscape-scale feature and survives a coarser lens.

The contrast with @sec-extent is the point, and one we should not overstate. Enlarging the cell fourfold changes almost nothing; shrinking the study area destroys the estimate. Grain governs how finely a pattern is resolved, extent governs which part of the joint distribution of elevation and wind the model ever sees, and here only the second mattered.

That is a result about this landscape and this pair of surfaces, not a general ordering of the two axes. @wiens1989 set out the distinction and warned that correlations evident within one domain may disappear or change sign when the extent is enlarged past it, and @dungan2002 recommended that studies report unit size, shape, spacing and extent together for the same reason. The literature since has not settled which axis matters more, and where it has looked it has often found the opposite of what we find: @mcgarigal2016, reviewing several hundred multi-scale habitat studies, concluded that the emphasis on extent may have left grain under-examined. Nor is reporting the weak point we assumed: @estes2018 found resolution and extent reported at closely comparable rates, near three fifths in both cases.

What our result adds is narrower and more useful. @sandel2015 shows that for a genuinely linear relationship the coefficient is extent-independent and only the coefficient of determination moves with extent. A coefficient that changes sign under truncation is therefore a diagnostic for an unmodelled nonlinearity rather than an instance of a general law. That is very likely what is happening here: @fig-gradient shows attack unimodal in elevation while wind rises monotonically, so no monotone wind term can hold across the whole gradient, and the clip lands on one limb of the hump. Read that way the finding is not "extent reverses inferences" but the sharper "truncating a gradient onto one limb of a unimodal response will reverse the apparent sign of anything monotonically correlated with the gradient." Testing it properly means fitting the nonlinearity rather than assuming it away, which is the first thing we would do next.

# Discussion

Read against the review, the result is a negative answer to a question the literature had been answering in the affirmative without measurement. R1 recorded that the wind-disruption mechanism has been invoked as an explanation for years while entering no model; fitted here, on a landscape chosen for its relief, it does not survive a specification that allows the response its observed curvature. R4 recorded that scale is acknowledged but conclusions are not tested against it; tested here, two of three analytical axes change the answer.

Three of the results above are built and then withdrawn, which is deliberate and can be hard to follow on a first reading. We therefore state at the outset what survives.

Four findings stand. Attack is unimodal in elevation. Below 1,900 m the pattern is the distribution of lodgepole pine. Attack rises monotonically with stem diameter and steps at the published source-sink threshold, so host must be coded as size and not only as cover. And above 1,900 m, `r sprintf("%.0f", 100*hi_full/hi_alone)` per cent of the suppression remains after every host term we have.

Three do not. The orthogonalised wind coefficient does not survive fitting elevation flexibly. The conclusion that freeze mortality can be excluded does not survive the stage-specific and seasonal structure of lethal temperature. And no estimate at all survives truncating the landscape to a 2 by 4 km clip.

One thing was never established. Nothing here identifies the mechanism behind the high-elevation refugia, and three candidates remain live.

The high-elevation suppression is real, and it is only partly host limitation. Among cells carrying lodgepole pine, attack above `r format(ALPINE, big.mark=",")` m runs at `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation>=ALPINE]))` per cent against `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation<ALPINE]))` per cent below. Cover barely moves across that step. Stem size does, and @sec-size shows that the share of stands above the source-sink crossover falls roughly four-fold. Conditioning on diameter and the rest of the host set removes `r sprintf("%.0f", 100*(1 - hi_full/hi_alone))` per cent of the effect on the log-odds scale and leaves the rest standing. Under the @krawchuk2020 definition what remains are refugia: disturbed less severely than the surrounding landscape, with host present, and predictably located by an enduring topographic feature. That matches @cartwright2018 in associating refugia with thinner stands, and it now says something more specific, because thinner here means smaller-stemmed as well as sparser.

We record that the first version of this result rested on cover alone and overstated the case. Reporting cover without diameter concealed a real host effect, and the general lesson is the one this paper keeps arriving at: a host layer is not a single thing, and which attribute of it you use decides the answer.

What we cannot do is attribute the remainder to wind. The ordinal result in @sec-ordinal changes one half of that difficulty and leaves the other where it was.

The statistical half looked, at first, much improved. Once severity is modelled as the ordinal variable it is, the orthogonalised wind term clears the interval test at every threshold above trace and strengthens monotonically with severity. @sec-nonlinear withdraws that reading. Entering elevation flexibly, and residualising wind against the same basis rather than against a straight line, removes between three fifths and seven tenths of the coefficient and leaves no threshold clearing the test. The ordinal coding does recover structure the binary discards, and @krawchuk2020's warning against binary refugia coding stands. But that structure is not attributable to wind: it is very largely the curvature of the elevation response, which the linear specification pushed into the wind term. We therefore do not report a wind effect, in either coding.

The identification half is untouched, and the life history says the competing explanation is more specific than we first allowed.

The upper gradient is where it is coldest, and cold acts on this insect through two separate mechanisms rather than one. The first is overwinter mortality, graded by instar and by the state of cold-hardening [@safranyik2006chap1; @cooke2009mortality]. The second is heat accumulation. A generation must finish in one year, and where summer warmth is insufficient the population becomes semivoltine: it overwinters twice, often in the cold-susceptible early instars, sits longer under natural enemies, and loses the emergence synchrony on which mass attack depends [@safranyik2006chap1; @powell2014].

That second mechanism deserves the emphasis, because it predicts our result without any appeal to wind. Amman's relationship between life-cycle duration, elevation and latitude, reproduced by @safranyik2006chap1, places the ceiling of univoltine development progressively lower as latitude rises, and at the latitude of the Selkirks it falls within the elevation range this study spans. We cannot locate it on this gradient, for reasons the next paragraph gives. But a transition from univoltine to semivoltine ground produces exactly what we observe: a step down in attack, with host present, at an elevation fixed by terrain, and graded in severity because the mechanism acts on how many beetles emerge together.

Cold in either form predicts a dose-response in severity just as wind disruption does. Fewer beetles surviving or emerging in step means both fewer stands attacked and lighter attack in those that are. The monotone strengthening across thresholds is therefore fully consistent with thermal limitation and does nothing to separate it from wind. Orthogonalising against elevation does not orthogonalise against temperature, which is not a linear function of elevation where cold air pools.

We looked for a way out and do not think one exists at this extent. Separating exposure from cold needs a temperature surface whose fine-scale structure is not itself a function of the elevation model, and any surface of the form intercept plus lapse rate times elevation is an affine rescaling of the elevation layer, whatever the lapse rate's source. It is collinear by construction, and only a residual departure from that form would identify anything, which would itself require terrain as the interpolator. The gridded products with genuine sub-kilometre temperature structure all obtain it by regression onto a digital elevation model, so they reproduce the problem rather than solve it.

One result narrows the field, and it narrows it less than we first claimed. @sec-cold shows that deep-winter air temperature in this region does not approach the lethal range for the late-instar larvae that normally overwinter. We had read that as ruling out freeze mortality and leaving heat accumulation as the survivor. It does not go that far. Lethal temperature is stage-specific, and our coldest defensible minimum of `r sprintf("%.1f", coldest$min_temp_c_usable)` °C lies inside the published range for first and second instars. It is also conditional on timing, and `r sprintf("%.0f", DEC)` per cent of our station-winters bottom out in December, at the front of the hardening window. Most decisive, the retrieved record runs only from December to February, so the shoulder months in which a given temperature is most lethal were never observed. Both thermal mechanisms remain live, and we no longer claim to have eliminated either.

Aspect gives partial purchase on one of them. Because heat load is a trigonometric function of the elevation model rather than a field draped over it, it carries thermal information independent of elevation, and @sec-aspect finds it predicts attack severity in the direction warmth-limited development implies. The constraint for which we have an elevation-independent proxy is the heat-accumulation one, and that proxy behaves as the constraint predicts. It says nothing about the overwinter minimum.

The gain is bounded in four ways. The station record is a regional plausibility check, not a temperature surface, since no station inside the study area reported during the outbreak window. Its window excludes the shoulder months. Heat load is a modelled radiation load, not a measured temperature. And none of it touches the possibility that cold-air pooling produces local freeze events the network never sampled, which is precisely the structure @sec-cold shows when a valley station at `r format(round(lowcold$elevation_m), big.mark=",")` m matches an alpine station `r format(round(coldest$elevation_m - lowcold$elevation_m), big.mark=",")` m above it.

The deeper obstacle is that the quantity governing overwinter survival is not air temperature but temperature under the bark, which is not observed at landscape scale anywhere. @goodsman2024, validating the standard Canadian Forest Service winter mortality model against observed mortality in Banff National Park, found its spatial predictions poor and attributed the failure to precisely this. The confound we report unresolved is unresolved in this terrain generally, and is not an artefact of our data being thin.

What would resolve it is a degree-day surface for this landscape, and we did not build one. The station archive was retrieved for winter months only, because the question it was assembled to answer was about winter minima. Growing-season accumulation would need the same archive re-streamed across the summer months of eighteen years, and the thresholds to apply are established: emergence scales with degree-days above 14.4 °C, and about 118 degree-days above 4.4 °C carry half a clutch of eggs to hatch in southern British Columbia [@safranyik2006chap1]. That is the first thing we would do next, and it is a data-collection task rather than a design problem.

There is also a class of mechanism that would produce our result without pheromone plumes at all, and the life history makes it more concrete than we first stated. Wind acts on the beetle before it acts on the plume. Bark beetles do not fly in winds exceeding their own flight speed, and *D. ponderosae* flies for shorter periods at 2 m s⁻¹ than at lower speeds in a wind tunnel [@carroll2004bionomics; @jones2019]. Dispersal within a stand is itself wind-structured: beetles drift downwind below the crowns until they meet an attractive plume, then turn and fly upwind to its source [@carroll2004bionomics]. If exposed ground receives fewer arriving beetles, rather than the same number arriving to a disrupted signal, the observable consequence in a static model is identical, and the Allee effect in colonisation means a modest reduction in arrivals can produce a large reduction in successful attack [@powell2014]. We cannot compare our wind values against the flight threshold, because the atlas reports an annual mean at 10 m over open ground and the threshold is an instantaneous below-canopy speed. That mismatch is itself the point: the study's predictor and the insect's constraint are not the same quantity.

The obvious precedent does not settle the direction. @smithmckenna2013, working on white pine blister rust in whitebark pine treeline communities, established that topographic position governs delivery of a wind-borne propagule, but the direction they report is the opposite of the one this alternative needs, with exposed positions favouring *interception* rather than escape. Their agent is a fungus dispersed passively, so neither direction transfers cleanly to a flying insect that selects where it lands. What their result does establish is that the dispersal pathway is a live confound in terrain of this kind. Nothing in our design distinguishes it from pheromone disruption. Doing so would need the dispersal process observed rather than inferred, through attack years, distance from the previous year's attack, or trapping.

One further boundary belongs here, and it is about the insect rather than the terrain. The 1999 to 2015 window is a single epidemic passage. Mountain pine beetle populations move through endemic, incipient-epidemic, epidemic and post-epidemic phases, and host selection and the spatial configuration of attack are themselves density-dependent [@safranyik2006chap1; @howe2022]. Endemic populations occupy weakened, small-diameter stems, which are distributed quite differently from the large-stemmed stands an epidemic takes. Everything estimated here is therefore beetle behaviour at epidemic density. @krawchuk2020 distinguish refugia by severity from refugia by frequency, and a single outbreak window can speak only to the first.

The honest summary is that we have located candidate refugia, confirmed the host half of the hypothesis in a stronger form than cover alone supports, and cannot support the wind half once elevation is fitted flexibly. The mechanism remains confounded with overwinter mortality, with heat-accumulation limits on voltinism, and with dispersal interception, on the same gradient.

## Host-absence confounding {#sec-lowtrap}

The result that should travel furthest is at the bottom of the gradient. Attack there is near zero, and any refugia mapping exercise lacking a host layer would have identified the valley as a large, spatially coherent, topographically predictable refugium. It is not a refugium. It is a place with almost no pine. Conditioning on host raises attack in the lowest band from `r sprintf("%.1f", grad$"Attacked, all cells (%)"[1])` per cent to `r sprintf("%.1f", grad$"Attacked, host present (%)"[1])` per cent, which is within the range observed at mid elevations.

@krawchuk2020 anticipate this in framing, asking of any candidate refugium both what it is a refugium from and what it is a refugium for, but the framework does not force the check and the data required to perform it are not always at hand. On this landscape the difference between performing it and not is the difference between reporting a roughly `r sprintf("%.0f", sum(S$Elevation < 900) * prod(res(G)) / 1e6)` km2 valley refugium, the modelled ground below 900 m that @tbl-gradient shows carries almost no pine, and reporting none. We would put the requirement more strongly than the framework currently does: a refugium claim without an independent host or habitat layer is not interpretable.

## Terrain limitations

The collinearity result in @sec-extent generalises beyond wind. A fine-resolution climate surface is produced by imposing terrain on a coarse field, so wherever the coarse field is nearly constant the product approaches a function of the terrain. That covers most gridded climate products used in landscape ecology. Three properties govern the severity, and all are checkable before any model is fitted: the ratio of analysis extent to the correlation length of the driving field, the relief of the study area, and which terrain variable does the work in the downscaling, which for wind is elevation and roughness but for radiation would be slope and aspect.

The consequence for synthesis is uncomfortable. If a confound scales with study extent, effect sizes for derived climate predictors are not comparable across studies of different extents, and small studies will systematically report the strongest apparent effects, a pattern easily mistaken for publication bias.

We would ask three things of any study entering such a surface into a disturbance model: report its regression on elevation, report the same diagnostic at more than one extent, and interpret only the orthogonalised coefficient. None requires new data, and here each changed the answer.

Three limitations belong to the host layer specifically. Quadratic mean diameter is a stand-level attribute applied uniformly across a polygon, so it cannot express the within-stand size distribution that decides which individual stems are sources. It is recorded for stems 12.5 cm and over, which is the right domain for this question but means the smallest stands are described by a truncated distribution. And it carries the same 2014 vintage as the species composition, so it too is a post-hoc description of stands the beetle had already reached. The 25 cm crossover we test against is itself an average across studies rather than a constant of this landscape [@carroll2004bionomics], and we use it as a reference point rather than as a fitted threshold.

A fourth limitation is about the insect and not the data. Everything estimated here describes an epidemic population. Endemic populations occupy weakened, small-diameter stems and select hosts differently [@safranyik2006chap1; @howe2022], so the diameter relationship in @sec-size should not be read as a constant of the species.

The plot record calibrates severity but not location, because coordinates for the 28 ground plots were not recoverable. The survey response is observer sketch-mapping with known positional and severity error, and rasterising polygons imposes a grain coarser than the 25 m cell; severity classes are assigned by eye from an aircraft, so the ordinal gradient in @sec-ordinal rests on judgement of proportion killed rather than measurement. The inventory is a modelled operational product whose species percentages are polygon-level attributes applied uniformly. The grain test in @sec-grain spans 25 m to 100 m, which brackets the resolutions in common use and the 900 m2 Landsat grain at which @krawchuk2020 note refugia have typically been described. Finally the analysis is correlational and the outbreak is one epidemic passage; frequency refugia in the sense of @krawchuk2020 cannot be addressed with a single window, and the static design cannot separate a wind effect on pheromone communication from one on dispersal.

The identification gained in @sec-aspect is partial. Heat load bears on the degree-day half of cold limitation, the half governing whether a generation completes in one year. It says almost nothing about the overwinter minimum half, the lethal events that kill larvae outright, which are driven by synoptic air mass intrusions and cold-air pooling rather than by radiation on a slope. Separating warm aspects from cool ones at the same elevation therefore addresses one thermal mechanism and leaves the other untouched, and rests on an assumed functional form, the @mccune2002 equation, rather than an independent measurement.

The shelter index is also imperfect as a test of the mechanism. It is omnidirectional because the Global Wind Atlas supplies no direction, and an omnidirectional index cannot represent a process that depends on where the wind comes from. A directional index on a known prevailing bearing would be sharper, and its disagreement with the wind residual might resolve either way under one.

# Conclusions

Each conclusion below is tied to the review finding it answers. R1 asked whether a mechanism carried in the literature survives being fitted, R3 whether host coded as stem size changes the terrain result, R4 whether the answer is stable across the analyst's free choices, and R5 whether the available wind product measures the quantity the hypothesis concerns.

Beetle-caused mortality at Darkwoods is unimodal across the `r format(round(diff(range(S$Elevation))), big.mark = ",")` m of relief the modelled area spans. Below 1,900 m that pattern is the distribution of lodgepole pine, and conditioning on host removes almost all of it. Above 1,900 m it is not, but the margin is narrower than cover alone implies.

Stem size is the host attribute that matters, and it had not been used. Attack rises monotonically with stand quadratic mean diameter, from `r sprintf("%.0f", 100*mean(HQ$Attacked[HQ$QMD <= 15]))` per cent in the smallest stands to `r sprintf("%.0f", 100*mean(HQ$Attacked[HQ$QMD > 35]))` per cent in the largest, with the steepest step at the 25 cm crossover between beetle sink and beetle source that the species' bionomics predicts [@carroll2004bionomics]. Across the 1,900 m threshold cover is flat while the share of stands above 25 cm falls roughly four-fold. Conditioning on diameter and the remaining host terms removes about `r sprintf("%.0f", 100*(1 - hi_full/hi_alone))` per cent of the high-elevation suppression. The rest stands, and that is where the Krawchuk hypothesis predicts refugia should be.

We do not find a wind effect. The component of the Global Wind Atlas surface orthogonal to a linear elevation term is the strongest negative predictor of attack in host-bearing stands, and modelled as the ordinal severity class it clears the interval test at every threshold above trace. But the response is unimodal, and orthogonalising against a straight line leaves the curvature of elevation inside the wind term. Entering elevation as a quadratic or a spline, and residualising wind against that same basis, collapses the coefficient at every threshold. The apparent wind signal is a specification artefact.

Even had it survived, it would not have been a measurement of the mechanism. Wind arrests bark beetle flight near 2 m s⁻¹, an instantaneous below-canopy speed during a July and August flight window [@carroll2004bionomics; @jones2019]. The atlas supplies an annual mean at 10 m above open ground. Studies that enter such a surface into an insect disturbance model should say which quantity they mean, and we could not.

The thermal alternatives are not eliminated, and we withdraw a stronger claim made earlier. Deep-winter air temperature at every station within 100 km stays clear of the lethal range for the late instars that normally overwinter. It does not stay clear of the range for first and second instars, `r sprintf("%.0f", DEC)` per cent of station-winter minima arrive in December before cold-hardening completes, and the retrieved record excludes October, November and March altogether. Separately, loss of univoltine development at elevation predicts a step down in attack, with host present, at a terrain-fixed elevation, and graded in severity, which is what we observe [@safranyik2006chap1]. A heat load index, being trigonometric in the elevation model and therefore nearly unrelated to elevation, predicts attack at the moderate severity threshold in the direction that constraint implies. But a geometric shelter index computed from the same model carries the opposite sign to the wind residual while being nearly uncorrelated with it, a warning that "wind exposure" is not one measurable thing and that our wind result lacks corroboration from the one predictor describing exposure directly.

The methodological findings are firmer, and all three are failures of design rather than data. Fitted inside a 2 by 4 km clip of the same landscape, elevation explains four fifths of the variance in the wind surface rather than two thirds, the wind coefficient's point estimate turns positive, and its interval widens until it excludes nothing; a four-fold change of grain moves none of this. Extent is the axis that governs whether the coefficient can be estimated here. We do not generalise that ordering: following @sandel2015, the likeliest reading is that truncating a gradient onto one limb of a unimodal response reverses the apparent sign of anything monotonically correlated with it, which is a statement about functional form rather than extent as such. Collapsing severity to attacked or not converts the study's strongest apparent relationship into a null. And without a host layer, the pine-poor valley floor would have been mapped as the largest refugium on the mountain.

# CRediT authorship declaration

Seamus Murphy: Conceptualization, Methodology, Software, Formal analysis, Investigation, Data curation, Writing - original draft, Writing - review & editing, Visualization.

# Competing interest declaration

The author declares no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Funding declaration

This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.

# Declaration of Generative AI

All prose, writing, results in the analysis and reporting contained within this document was derived and verified by the author. No generative tool was used to create, alter or manipulate research data or results. The author is responsible for the accuracy of all content.

# Data availability

Every dataset used here is public and none was collected by the author. Beetle-caused mortality is the British Columbia Aerial Overview Survey pest infestation polygon layer, filtered to `IBM` for 1999 to 2015. Host cover and stand basal area are the provincial Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved from the British Columbia Data Catalogue web feature service. Ambient wind is Global Wind Atlas mean speed at 10 m [@globalwindatlas]. Terrain is derived from the Natural Resources Canada High Resolution Digital Elevation Model [@nrcan2017]. Winter minima are British Columbia Wildfire Service hourly and Environment and Climate Change Canada daily records for stations within 100 km, retrieved by the script named in the Materials section, which also writes the cached per-station-winter summary the manuscript reads.

The analysis is the manuscript. Every table, figure and quoted number is produced by a code chunk in the source document, so rendering it re-runs the pipeline and regenerates the results. The source document and the retrieval scripts are available from the author on request.

# Ethics declarations

This study analysed publicly available spatial and meteorological records only. It involved no human participants, no animal subjects and no sampling of protected species, so ethics approval and informed consent were not required.

# Word counts

Total including references, {{WC_TOTAL}}. Total excluding references, {{WC_NOREFS}}. Abstract, {{WC_ABSTRACT}}. Supplementary information, {{WC_SUPP}}.

# References

::: {#refs}
:::

# Supplementary information

The five items below are cited in the main text as Tables S1 to S3 and Figures
S1 and S2. They are supplied as supplementary material and are not counted
toward the ten figures and tables the main text carries. Each is produced by
code already run above, so the values are identical to those the main text
quotes.

**Table S1.** The binary hypothesis test, cells containing lodgepole pine only. Coefficients are per standard deviation; wind is entered as the component orthogonal to elevation. Intervals are percentile intervals from resampled 1 km spatial blocks. This is the same fit as the first row of the ordinal sweep in the main text, retained because it is the form the refugia literature conventionally reports.

```{r}
#| echo: true
SUPP_KRAWCHUK
```

**Table S2.** Observed winter minimum air temperatures from stations within 100 km, by elevation band, after excluding station-winters censored at the -20.0 °C sensor floor. Nothing approaches the temperatures conventionally cited as lethal for this insect, and elevation orders the minima only weakly.

```{r}
#| echo: true
SUPP_COLD
```

**Table S3.** The host-present model refitted at three grains. Predictors are aggregated by mean, severity class by maximum, and the bootstrap blocks are rescaled to stay near 1 km. Unlike extent, grain changes nothing: collinearity, sign and magnitude are all stable across a four-fold change in cell size.

```{r}
#| echo: true
SUPP_GRAIN
```

**Figure S1.** The six derived terrain predictors, all computed from the same 25 m elevation model. (a) Slope. (b) Topographic wetness index. (c) Ruggedness. (d) Topographic position at a 625 m neighbourhood. (e) Heat load index after McCune and Keon. (f) Omnidirectional shelter index after Winstral, the maximum upwind slope angle within 500 m averaged over eight bearings. Ranges are clipped to the 2nd and 98th percentiles.

```{r}
#| echo: true
#| fig-width: 7.1
#| fig-height: 4.6
fig_terrain
```

**Figure S2.** The weather station network behind the cold argument. All stations within 100 km reporting a usable winter minimum between 1999 and 2015, shaded by station elevation, with the analysis grid outlined. Only the two stations named in the text are labelled. The alpine station and the valley stations near the international boundary reach comparable minima, which is the cold-air pooling the text describes.

```{r}
#| echo: true
#| fig-width: 5.9
#| fig-height: 5.1
fig_station
```
