---
title: "Testing the wind disruption hypothesis for beetle refugia"
subtitle: "Host distribution, thermal load, and terrain-wind effects on Mountain pine beetle attack (Dendroctonus ponderosae) across 1,740 m of relief in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
abstract: >
  Refugia from mountain pine beetle (Dendroctonus ponderosae) are hypothesised to form where host is sparse enough
  for wind to disrupt the aggregation pheromone plume mass attack depends on, never tested spatially.
  We test it across 168 km² of inventoried forest and 1,740 m of relief in British Columbia, combining an aerial
  survey of beetle-caused mortality (1999-2015) with inventory host cover and Global Wind Atlas speed. Attack is
  unimodal, peaking near 1,430 m. Below 1,900 m that peak is host, and conditioning on lodgepole presence removes
  it; above 1,900 m it is not, and attack halves while pine cover thins only slightly. We find no wind effect.
  Orthogonalised against a linear elevation term, wind is the strongest negative predictor in host-bearing stands,
  clearing the interval test at every severity threshold above trace; but a unimodal response leaves elevation's
  curvature inside that term, and refitting it as a quadratic or spline collapses the coefficient everywhere. The
  signal is a specification artefact. Winter minima at every station within 100 km fall ten degrees short of lethal,
  so the cold mechanism is degree-day accumulation, not freeze. Elevation explains 0.63 of the wind
  surface's variance landscape-wide but 0.80 within a 2 by 4 km clip, where the estimate reverses sign; grain
  scarcely matters. At the extent disturbance ecology usually adopts, this study would have reported the
  opposite; with severity collapsed to presence, no result at all; and without a host layer, absence of pine
  misread as refugia.

keywords:
  - Beetle refugia
  - Mountain pine beetle
  - Aerial overview survey
  - Global Wind Atlas
  - Scale dependence
  - Collinearity

date: today
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/springer-basic-author-date.csl
df-print: kable
---

```{r}
#| label: setup
#| echo: false
#| output: false

## Every quantity in this manuscript is computed here at render time.
## patchwork is loaded for its `+` operator on ggplots (@fig-gradient), which is
## why no patchwork function appears by name anywhere below.
pkgs <- c("terra","knitr","kableExtra","ggplot2","dplyr","patchwork","tibble",
          "sf","ggspatial","tidyterra","ggrepel","rnaturalearth","scales")
invisible(lapply(pkgs, function(p) library(p, character.only = TRUE)))
## dpi 300 is the Elsevier minimum for a combination line/halftone figure.
## error = FALSE on purpose: in a manuscript whose every number is computed at
## render time, a chunk that fails quietly while downstream inline expressions
## carry stale values is the one failure this design exists to prevent.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      comment = NA, dpi = 500, fig.align = "center")
options(scipen = 999)
theme_set(theme_minimal(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                plot.title = element_text(face = "bold", size = 10)))

B_BOOT   <- 150L    # spatial block bootstrap replicates
BLOCK    <- 40L     # block side in cells (40 x 25.1 m ~= 1 km)
OUTBREAK <- 1999:2015
## The elevation above which conditioning on host stops removing the decline.
## It is a band edge of the 200 m banding in @tbl-gradient and is quoted in the
## prose, so it is defined once here rather than typed into each comparison.
ALPINE   <- 1900
```

```{r}
#| label: paths
#| echo: false

DATA_ROOT <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
dpath <- function(...) file.path(DATA_ROOT, ...)
here  <- function(...) file.path("..", ...)     # repo-relative, from 01.manuscript/
stopifnot(dir.exists(dpath("terrain_environment")))

## The elevation surface is the analysis grid: everything else is resampled onto
## it. Loaded here so the Study area section can quote the relief from the raster
## rather than from memory.
G <- rast(dpath("terrain_environment", "Elevation.utm.tif"))
GRID_RELIEF <- diff(as.vector(minmax(G)))
```

```{r}
#| label: map-furniture
#| echo: false

## Cartography shared by the four map figures, defined once here so each figure can
## sit in the section it documents instead of being grouped at the end of Materials
## for want of its furniture. Graticule, EPSG, north arrow, bar scale and a
## representative fraction quoted against a stated printed width are set once and
## reused, so the panels are directly comparable.

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
clip_sf  <- st_as_sf(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)))
grid_sf  <- st_as_sf(as.polygons(ext(G), crs = crs(G)))
ev <- as.vector(ext(G)); gw <- ev["xmax"] - ev["xmin"]; gh <- ev["ymax"] - ev["ymin"]
sub_utm <- "EPSG:32611 WGS 84 / UTM 11N"

map_theme <- theme_bw(base_size = 8) +
  theme(panel.grid = element_line(colour = "grey70", linewidth = .25, linetype = "dotted"),
        axis.title = element_blank(), axis.text = element_text(size = 6, colour = "grey25"),
        plot.title = element_text(face = "bold", size = 8.5, margin = margin(b = 1)),
        plot.subtitle = element_text(size = 6, colour = "grey35", margin = margin(b = 3)),
        legend.key.width = unit(3.2, "mm"), legend.key.height = unit(7, "mm"),
        legend.title = element_text(size = 6.5), legend.text = element_text(size = 6))

north <- annotation_north_arrow(location = "tr", which_north = "true",
  height = unit(6, "mm"), width = unit(5, "mm"),
  pad_x = unit(1.2, "mm"), pad_y = unit(1.2, "mm"),
  style = north_arrow_orienteering(text_size = 5, line_width = .6,
    fill = c("white", "grey20"), text_col = "grey20"))

## White plates behind the north arrow and the clip label. Without them the arrow is
## unreadable over the dark end of the wind ramp and the label collides with the
## severity polygons.
plate <- annotate("rect", xmin = ev["xmin"] + .885*gw, xmax = ev["xmax"],
                  ymin = ev["ymax"] - .115*gh, ymax = ev["ymax"],
                  fill = "white", alpha = .72, colour = NA)
clip_layer <- geom_sf(data = clip_sf, fill = NA, colour = "#C0392B", linewidth = .45)
study <- list(coord_sf(crs = 32611, datum = st_crs(4326), expand = FALSE,
                       xlim = ev[c("xmin","xmax")], ylim = ev[c("ymin","ymax")]),
              scale_x_continuous(breaks = seq(-116.75, -116.60, .05)),
              scale_y_continuous(breaks = seq(49.10, 49.24, .05)),
              plate, north, map_theme)

## Colour limits from robust quantiles. The raw topographic position surface runs
## past 1,000 m at the grid edge, where the 625 m window is truncated, and on a
## full-range ramp that artefact flattens the whole interior to one colour.
panel <- function(r, title, legend, pal = "viridis", dir = 1, q = c(.02, .98)) {
  lim <- unname(quantile(values(r, mat = FALSE), q, na.rm = TRUE))
  ggplot() + geom_spatraster(data = r) +
    scale_fill_viridis_c(option = pal, direction = dir, na.value = "grey88",
                         name = legend, limits = lim, oob = scales::squish) +
    clip_layer + study + labs(title = title, subtitle = sub_utm)
}
```

# Introduction

Disturbance refugia, defined by @krawchuk2020 as locations disturbed less severely or less frequently than the surrounding landscape, have become an organising concept for forest conservation under climate change. The concept is well developed for fire and much less so for insects. @krawchuk2020 note that few studies have explicitly identified refugia from insect outbreaks, and propose a testable mechanism for the most consequential insect disturbance in western North America: refugia from mountain pine beetle (*Dendroctonus ponderosae* Hopkins) could occur where host density is lower, thinner stands admitting more wind and so disrupting the aggregation pheromone communication mass attack depends on.

The hypothesis is attractive because it is mechanistic and because each of its parts is independently plausible. Mass attack requires beetles to overwhelm host defences in a coordinated pulse, and the coordination is chemical. Beetle dispersal above the canopy is a flight behaviour sensitive to ambient conditions [@gray1972; @jones2019]. Thinner stands admit more wind. What has been missing is a spatial test.

@cartwright2018 came closest. Working at 30 m in lodgepole and whitebark pine in southern Oregon, she identified refugia from Landsat moisture anomalies during a 2009 drought and beetle outbreak and modelled their landscape controls. Refugia were associated with shaded slopes, convergent positions, low soil bulk density and thinner stands. The thinner-stand result is the host-density half of the hypothesis, confirmed. But she had no wind surface, so the mechanism went untested.

Testing it requires a wind surface, and that is where the problem starts. Gridded wind products, the Global Wind Atlas among them, are produced by downscaling coarse mesoscale fields over a digital elevation model [@davis2023; @badger2014]. A surface built this way is in part a transform of the terrain it was draped over. Offered to a model alongside elevation it is not an independent measurement, and the severity of the dependence is a function of extent: a coarse driving field is nearly flat over a few kilometres, so at small extents almost all the variation in the product is terrain. An analyst working at the scale of a burn perimeter or a cluster of field plots, which is where most disturbance ecology is done, is in the worst position and has the least warning.

A second and more elementary trap attends refugia work. A cell that was not attacked may hold no host. Under the @krawchuk2020 framing this is the difference between what a place is a refugium *from* and what it is a refugium *for*, and it cannot be resolved without an independent map of host abundance. The precedent is old and pointed: @hadley1994 predicted heavier beetle mortality on hot, dry, south-facing slopes and found the reverse, because host density rather than moisture stress governed where attack succeeded.

We take these three problems to one landscape. We ask:

1.  Does terrain condition beetle-caused mortality once host abundance is accounted for, and where along the elevation gradient does host stop explaining the pattern?
2.  Does the component of the wind surface that is independent of elevation predict reduced attack where host is present, as @krawchuk2020 hypothesise?
3.  How much does the answer to question 2 depend on decisions the analyst makes freely: the extent of the study area, the grain of the cells, and whether damage is coded as a binary or as the ordinal severity class the survey records?

The design makes the third question answerable rather than rhetorical. The same response, predictors and model are refitted across each axis: across the full landscape and inside a 2 by 4 km clip drawn from it, at 25 m, 50 m and 100 m, and at each severity threshold in turn. Two of the three change the answer, and they are not the two that get reported.

# Materials

## Study area

The Darkwoods Conservation Area lies in the South Selkirk Mountains of southeastern British Columbia. The terrain grid spans 16 by 17.5 km and `r format(round(GRID_RELIEF), big.mark = ",")` m of relief, from valley bottom to alpine. The modelled area is a subset, because cells carrying no inventory basal area cannot enter a model conditioning on stand structure; @tbl-relief reports its extent and relief. Both numbers are correct and are not interchangeable. The range crosses the full biogeoclimatic sequence of the southern interior wet belt, from the dry warm Interior Cedar-Hemlock subzones through the moist warm subzones to the Engelmann Spruce-Subalpine Fir zone and its parkland variant [@ketcheson1991; @braumandl2005]. Lodgepole pine (*Pinus contorta* var. *latifolia*), the beetle's principal host here, occupies the middle of that sequence and is scarce at both ends.

The site description and terrain surfaces are inherited from a companion analysis of conifer regeneration on the same ground [@murphy2026], which used them to predict seedling intensity after a 2015 wildfire. Disturbance was a predictor there; here it is the response. @fig-site locates the landscape and shows the three spatial inputs.

## Beetle-caused mortality

```{r}
#| label: read-response
#| echo: false

aos_box <- ext(project(as.polygons(ext(G), crs = crs(G)), "EPSG:3005"))
aos <- vect(here("02.inputs/aos/pest_infestation_poly.gdb"),
            layer = "pest_infestation_poly", extent = aos_box)
mpb <- project(aos[aos$PEST_SPECIES_CODE == "IBM", ], crs(G))   # IBM = mountain pine beetle
m   <- mpb[mpb$CAPTURE_YEAR %in% OUTBREAK, ]

## BC severity classes are ordinal: trace, light, moderate, severe, very severe
sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)
m$sev <- unname(sev_rank[m$PEST_SEVERITY_CODE]); m$sev[is.na(m$sev)] <- 0

severity  <- rasterize(m, G, field = "sev", fun = "max", background = 0)

c(polygons = nrow(m), years = length(unique(m$CAPTURE_YEAR)),
  hectares = round(sum(m$AREA_HA)))
```

The response is the British Columbia Aerial Overview Survey, the operational provincial record of forest health, in which observers sketch-map damage annually from aircraft and label each polygon by year, causal agent, severity class and host species. We take polygons coded `IBM`, mountain pine beetle, for the outbreak window 1999 to 2015, and rasterise the maximum severity class reached by each cell.

Using the survey rather than a spectral classification is a deliberate trade. Red attack has been mapped from Landsat with improving accuracy [@white2005; @meddens2013; @ye2021], but every such product is a model whose errors are spatially structured and not independent of terrain, because illumination, shadow and view angle vary with slope and aspect. For a question about terrain, a response from a terrain-sensitive classifier risks recovering the classifier's own terrain sensitivity. Sketch-mapping has cruder errors, but they are observer errors, not errors generated by the predictors under test.

The severity scale is anchored to ground measurement. Twenty-eight field plots established in beetle-affected stands in 2020 recorded the basal area of pine killed, spanning 0.6 to 47.4 m2/ha, or 1.3 to 94.7 per cent of plot basal area, and calibrated the spectral mapping used in the companion study [@murphy2026]. Plot coordinates were not recoverable, so the plots calibrate the severity scale but cannot check the spatial placement of individual polygons. We state that limit rather than implying a validation we did not perform.

@fig-site shows the landscape and the response together. Panel (b) is the elevation surface that defines the grid and panel (d) the survey record this paper models: attack concentrated in a mid-elevation band, sparse in the valley and again along the alpine crest. Panel (c) is the wind surface whose relationship to that terrain the second half of the paper is about.

```{r}
#| label: fig-site
#| echo: false
#| fig-width: 7.1
#| fig-height: 6.8
#| fig-cap: "The study area and the three spatial inputs. (a) Location within British Columbia; the marker is a 22 km symbol, not the footprint. (b) The elevation surface, which is the analysis grid. (c) Global Wind Atlas mean speed at 10 m; the shaded southern band lies outside the wind product's extent. (d) Beetle-caused mortality from the Aerial Overview Survey, 1999 to 2015, as the maximum severity class in any year. The outlined rectangle in (b) to (d) is the 2 by 4 km clip of @sec-extent. Graticules are WGS 84; each panel's projection is beneath its title."

## The wind raster begins 715 m north of the grid's southern edge, so resampling onto G
## turns that strip into NA cells and na.value draws it grey. Left transparent it reads
## as a rendering fault rather than as missing data.
Wind_r <- resample(rast(dpath("terrain_environment", "Wind.utm.tif")), G)

prov <- st_transform(ne_states(country = c("Canada","United States of America"),
                               returnclass = "sf"), 3005)
site <- st_transform(st_centroid(grid_sf), 3005)
lab  <- transform(as.data.frame(st_coordinates(site)), lab = "Darkwoods\nstudy area")
## BC is framed to the analysis grid's aspect ratio so the two rows of the figure are
## the same height and no band of white opens between them.
b0 <- st_bbox(prov[prov$name == "British Columbia", ]); tgt <- unname(gw/gh)
w0 <- unname(b0["xmax"]-b0["xmin"]); h0 <- unname(b0["ymax"]-b0["ymin"])
cx <- unname(b0["xmax"]+b0["xmin"])/2; cy <- unname(b0["ymax"]+b0["ymin"])/2
if (w0/h0 > tgt) h0 <- w0/tgt else w0 <- h0*tgt
pd <- .04*w0
bb <- c(xmin = cx-w0/2-pd, xmax = cx+w0/2+pd, ymin = cy-h0/2-pd, ymax = cy+h0/2+pd)

p_a <- ggplot() +
  geom_sf(data = prov[prov$name != "British Columbia", ], fill = "grey93",
          colour = "grey65", linewidth = .2) +
  geom_sf(data = prov[prov$name == "British Columbia", ], fill = "grey82",
          colour = "grey30", linewidth = .4) +
  geom_sf(data = st_buffer(site, 22000), fill = "#C0392B", colour = "grey15",
          linewidth = .3) +
  geom_label_repel(data = lab, aes(X, Y, label = lab), size = 2.1, lineheight = .95,
                   colour = "grey10", nudge_x = -330000, nudge_y = 300000,
                   segment.colour = "grey25", segment.size = .3, min.segment.length = 0,
                   box.padding = .4, seed = 1, fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm")) +
  annotate("text", x = bb["xmin"] + .34*(bb["xmax"]-bb["xmin"]),
           y = bb["ymin"] + .70*(bb["ymax"]-bb["ymin"]), label = "British\nColumbia",
           size = 2.3, colour = "grey30", fontface = "italic", lineheight = .95) +
  coord_sf(crs = 3005, xlim = bb[c("xmin","xmax")], ylim = bb[c("ymin","ymax")],
           datum = st_crs(4326), expand = FALSE) +
  scale_x_continuous(breaks = seq(-135, -115, 5)) +
  scale_y_continuous(breaks = seq(48, 60, 3)) +
  north + map_theme +
  labs(title = "(a) Location", subtitle = "EPSG:3005 NAD83 / BC Albers")

p_b <- ggplot() + geom_spatraster(data = G) +
  scale_fill_hypso_c(palette = "dem_screen", na.value = "transparent",
                     name = "Elevation\n(m)") +
  clip_layer + study + labs(title = "(b) Elevation, the analysis grid", subtitle = sub_utm)

p_c <- ggplot() + geom_spatraster(data = Wind_r) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey88",
                       name = "Wind speed\n(m s⁻¹)") +
  clip_layer + study +
  labs(title = "(c) Global Wind Atlas mean speed at 10 m", subtitle = sub_utm)

sev_f <- as.factor(severity)
levels(sev_f) <- data.frame(ID = 0:5, class = c("none","trace","light","moderate",
                                                "severe","very severe"))
p_d <- ggplot() + geom_spatraster(data = sev_f) +
  scale_fill_manual(values = c(none = "grey92", trace = "#FEE0B6", light = "#FDB863",
                    moderate = "#E08214", severe = "#B35806", `very severe` = "#7F3B08"),
                    na.value = "transparent", name = "Severity\nclass",
                    na.translate = FALSE) +
  clip_layer + study +
  labs(title = "(d) Beetle-caused mortality, 1999 to 2015", subtitle = sub_utm)

fig_site <- (p_a | p_b) / (p_c | p_d)
fig_site
ggsave("../03.outputs/PNG/fig-site-maps.png", fig_site,
       width = 180, height = 172, units = "mm", dpi = 500, bg = "white")
```

## Host abundance

```{r}
#| label: read-host
#| echo: false

vri <- project(vect(here("02.inputs/vri/vri_darkwoods.geojson")), crs(G))
vd  <- as.data.frame(vri)
num <- function(x) suppressWarnings(as.numeric(x))

## VRI records up to three species per polygon with percentage cover
accumulate <- function(pattern) {
  out <- rep(0, nrow(vd))
  for (i in 1:3) {
    cd <- vd[[paste0("SPECIES_CD_", i)]]
    pc <- num(vd[[paste0("SPECIES_PCT_", i)]]); pc[is.na(pc)] <- 0
    out <- out + ifelse(!is.na(cd) & grepl(pattern, cd), pc, 0)
  }
  out
}
vri$PinePct <- accumulate("^PL")               # lodgepole, the primary host
vri$Basal   <- num(vd$BASAL_AREA)

host <- rast(lapply(c("PinePct","Basal"), function(f) rasterize(vri, G, field = f)))
names(host) <- c("PinePct","Basal")
c(vri_polygons = nrow(vri), with_lodgepole = sum(vri$PinePct > 0))
```

Host abundance comes from the provincial Vegetation Resources Inventory, layer `VEG_COMP_LYR_R1_POLY`, retrieved from the British Columbia Data Catalogue web feature service for the study window. We accumulate the percentage cover of lodgepole pine across the three species slots each polygon records, and carry stand basal area alongside it.

The inventory's vintage matters. These polygons carry a reference year of 2014 and an interpretation date of August 2014; the projected date of 31 December 2025 is the growth model's roll-forward of age, height and volume, not a re-observation. Species composition therefore comes from photography flown in the fourteenth year of a seventeen-year outbreak window.

That is a limitation we cannot remove and should not hide. A stand whose pine was killed early will have been interpreted after the fact, and its recorded lodgepole percentage will be lower than the pine it carried when the beetle arrived. The bias runs in a specific direction: host is under-recorded exactly where attack was heaviest and earliest. It is therefore a candidate explanation for the reversal of the lodgepole coefficient at the severe threshold in @sec-ordinal. It does not touch the presence-absence distinction the layer is mainly used for, since a stand that carried enough pine to be killed is not recorded as pine-free.

Without this layer the analysis cannot distinguish a refugium from an absence of pine, and @sec-host shows that the distinction accounts for most of the apparent terrain signal.

@fig-host maps both layers. Lodgepole cover is concentrated in the middle of the elevation range and scarce in the valley, which is the pattern the host-conditioning argument turns on: the valley is not lightly attacked because it is sheltered but because it holds almost no host. Basal area defines the analysis frame, and the blank cells in panel (b) are those dropped for want of it.

```{r}
#| label: fig-host
#| echo: false
#| fig-width: 7.1
#| fig-height: 3.8
#| fig-cap: "The forest inventory host layers, rasterised onto the analysis grid. (a) Lodgepole pine cover, accumulated across the three species slots each polygon records. (b) Stand basal area, which defines the analysis frame: cells with no recorded basal area are dropped. Unshaded ground is unmapped or non-productive. Projection and scale as in @fig-site."

fig_host <-
  (ggplot() + geom_spatraster(data = host$PinePct) +
     scale_fill_viridis_c(option = "viridis", direction = -1, na.value = "grey88",
                          name = "Lodgepole\ncover (%)") +
     clip_layer + study + labs(title = "(a) Lodgepole pine cover", subtitle = sub_utm)) |
  (ggplot() + geom_spatraster(data = host$Basal) +
     scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "grey88",
                          name = "Basal area\n(m² ha⁻¹)") +
     clip_layer + study + labs(title = "(b) Stand basal area", subtitle = sub_utm))
fig_host
ggsave("../03.outputs/PNG/fig-host-maps.png", fig_host,
       width = 180, height = 96, units = "mm", dpi = 500, bg = "white")
```

## Terrain and wind

```{r}
#| label: read-terrain
#| echo: false

to_grid <- function(f) {
  r <- rast(dpath("terrain_environment", f))
  b <- project(as.polygons(ext(G), crs = crs(G)), crs(r))
  project(crop(r, b, snap = "out"), G, method = "bilinear")
}
terr <- rast(list(Elevation = G,               Slope = to_grid("Slope.utm.tif"),
                  TWI       = to_grid("TWI.utm.tif"),
                  RIX       = to_grid("Rutm.tif"),
                  Wind      = to_grid("Wind.utm.tif")))
names(terr) <- c("Elevation","Slope","TWI","RIX","Wind")

## Topographic position: mean deviation from a surrounding annulus.
tpi <- function(r, w) {
  k <- matrix(1, w, w); k[ceiling(w/2), ceiling(w/2)] <- 0
  r - focal(r, k / sum(k), fun = sum, na.policy = "omit", na.rm = TRUE)
}
terr$TPI600 <- tpi(G, 25)     # ~625 m neighbourhood
```

```{r}
#| label: aspect-terms
#| echo: false

## Two surfaces built from aspect, which the inherited set lacks entirely.
## Both are computed directly from the elevation model rather than draped over
## it, which is why @tbl-orthogonality can show them independent of elevation.
d2r <- pi / 180
asp  <- rast(dpath("terrain_environment", "Aspect.utm.tif"))
slp  <- rast(dpath("terrain_environment", "Slope.utm.tif"))
stopifnot(identical(dim(asp), dim(G)), identical(as.vector(ext(asp)), as.vector(ext(G))))

## Heat load after McCune and Keon, aspect folded about the SW-NE axis so that
## southwest is hottest. Flat cells carry a -1 sentinel; the sin(slope) factor
## nulls the aspect term there, so the substitution below is inert.
asp[asp < 0] <- 225
fold <- abs(180 - abs(asp - 225))
LAT  <- mean(crds(project(as.polygons(ext(G), crs = crs(G)), "EPSG:4326"))[, 2])
terr$HLI <- 0.339 + 0.808 * cos(LAT*d2r) * cos(slp*d2r) -
            0.196 * sin(LAT*d2r) * sin(slp*d2r) -
            0.482 * cos(fold*d2r) * sin(slp*d2r)

## Winstral shelter index: the maximum upwind slope angle within SEARCH metres
## along a bearing. Positive means terrain rises upwind, so the cell is
## sheltered; negative means it is exposed. The Global Wind Atlas gives speed
## but no direction, so rather than assume a prevailing bearing we compute all
## eight and average, giving an omnidirectional shelter index.
SEARCH <- 500
shift_mat <- function(m, dr, dc) {
  nr <- nrow(m); nc <- ncol(m); out <- matrix(NA_real_, nr, nc)
  ri <- seq_len(nr); ci <- seq_len(nc); sr <- ri + dr; sc <- ci + dc
  vr <- sr >= 1 & sr <= nr; vc <- sc >= 1 & sc <= nc
  out[ri[vr], ci[vc]] <- m[sr[vr], sc[vc]]
  out
}
zm <- as.matrix(G, wide = TRUE); rstep <- res(G)[1]
sx_dir <- function(az) {
  th <- az * d2r; best <- matrix(-Inf, nrow(zm), ncol(zm))
  for (k in 1:floor(SEARCH / rstep)) {
    dc <- round(k * sin(th)); dr <- round(-k * cos(th))
    if (dc == 0 && dr == 0) next
    best <- pmax(best, atan((shift_mat(zm, dr, dc) - zm) / (k * rstep)) / d2r, na.rm = TRUE)
  }
  best[!is.finite(best)] <- NA_real_
  best
}
sx <- rast(lapply(seq(0, 315, by = 45), function(a) {
  r <- G; values(r) <- as.vector(t(sx_dir(a))); r }))
terr$SxMean <- app(sx, mean, na.rm = TRUE)
```

Terrain surfaces are inherited unchanged from @murphy2026, which derived them from the 1 m Natural Resources Canada High Resolution Digital Elevation Model [@nrcan2017], smoothed, depression-filled by the algorithm of @wang2006 and aggregated to 25 m, with compound indices following @beers1973 and @parker1982. We add three surfaces of our own. The first is a topographic position index at a 625 m neighbourhood, because @cartwright2018 found slope, elevation and heat load among the strongest controls on refugia and reported topographic position alongside them, and the inherited set carried none.

The other two exist because the inherited set carries no aspect information, and so no term that can distinguish a warm slope from a cold one at the same elevation. That matters because the confound this paper cannot resolve is between wind exposure and cold, and aspect bears on the two differently. We compute a heat load index after @mccune2002, in which aspect is folded about the southwest to northeast axis so that southwest is hottest, and a shelter index after @winstral2002, the maximum upwind slope angle within 500 m along a bearing, positive where terrain rises upwind and negative on exposed convexities. The Global Wind Atlas supplies speed but no direction, so rather than assume a prevailing bearing we compute the index on all eight cardinal and intercardinal bearings and average them.

The decisive property of both is how they are built. They are derived from the elevation model by trigonometry rather than draped over it by downscaling, and @tbl-orthogonality shows the consequence: their variance is essentially unrelated to elevation, where the wind product's is largely elevation. A terrain-derived predictor is not automatically an elevation transform; the distinction is in the construction.

The ruggedness surface carries a documentation gap we state rather than paper over. Its exact formulation is not recorded beyond the citations above, and its distribution is inconsistent with a vector ruggedness measure. We treat it as a reproducible but unnamed ruggedness surface and confine claims to its statistical behaviour.

Ambient wind is Global Wind Atlas mean wind speed at 10 m above ground level [@globalwindatlas]. The atlas is built by driving a mesoscale model from reanalysis, generalising the result by the method of @badger2014 to remove the local terrain and roughness signal, then downscaling it over a digital elevation model and a land-cover roughness map [@davis2023]. Elevation and roughness are inputs to the product, not covariates measured independently of it. That is why it cannot be assumed independent of elevation, and testing that assumption is the second half of this paper.

@fig-terrain maps the six derived surfaces. Their point is visual rather than statistical: slope, wetness, ruggedness and topographic position all trace the same drainage network and valley wall, whereas heat load and the shelter index divide the landscape by aspect, picking out southwest-facing ground the others do not distinguish. That is the difference @tbl-orthogonality quantifies.

```{r}
#| label: fig-terrain
#| echo: false
#| fig-width: 7.1
#| fig-height: 4.6
#| fig-cap: "The six derived terrain predictors, all computed from the same 25 m elevation model. (a) Slope. (b) Topographic wetness index. (c) Ruggedness. (d) Topographic position at a 625 m neighbourhood. (e) Heat load index after McCune and Keon. (f) Omnidirectional shelter index after Winstral, the maximum upwind slope angle within 500 m averaged over eight bearings. Ranges are clipped to the 2nd and 98th percentiles. Projection and scale as in @fig-site."

fig_terrain <-
  (panel(terr$Slope,  "(a) Slope", "Degrees", "cividis") |
   panel(terr$TWI,    "(b) Topographic wetness index", "TWI", "mako", -1) |
   panel(terr$RIX,    "(c) Ruggedness", "Index", "rocket", -1)) /
  (panel(terr$TPI600, "(d) Topographic position, 625 m", "TPI (m)", "viridis") |
   panel(terr$HLI,    "(e) Heat load index", "HLI", "inferno") |
   panel(terr$SxMean, "(f) Shelter index, omnidirectional", "Sx (deg)", "cividis", -1))
fig_terrain
ggsave("../03.outputs/PNG/fig-terrain-maps.png", fig_terrain,
       width = 180, height = 118, units = "mm", dpi = 500, bg = "white")
```

## Climate records

Because the interpretation in @sec-krawchuk turns on whether it is cold enough at the top of the gradient to kill beetles, we assembled the observational record. Hourly observations from the British Columbia Wildfire Service network and daily observations from Environment and Climate Change Canada were retrieved for all stations within 100 km for the winters of the outbreak window. The script in `02.inputs/climate/` streams roughly 2.7 GB and writes the per-station-winter summary the manuscript reads. That division is deliberate: the fetch cannot sit inside a document render, but every decision about which records to believe is made in @sec-cold, in code, and is visible there.

Two screens matter. Many older Wildfire Service sensors floor at exactly -20.0 °C, so a station-winter whose minimum sits on that floor has an unknown true minimum and is excluded; the rule is applied and checked in @sec-cold rather than inherited. Isolated spikes, colder than both neighbouring observations by more than 15 °C, are rejected as instrument error at the observation level, the one screen that must happen during retrieval because it needs the sub-daily series. One such value, a -40.0 °C daily minimum on a day whose maximum was -4.5 °C and carrying no quality flag, would otherwise have produced exactly the conclusion the record does not support.

This network cannot furnish a temperature surface, for the reasons @sec-cold sets out. It can establish whether lethal temperatures occur in this region at all, a weaker and sufficient claim.

```{r}
#| label: assemble
#| echo: false

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
in_clip  <- rasterize(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)),
                      G, field = 1, background = 0)

X <- as.data.frame(c(terr, host, severity, in_clip), na.rm = FALSE)
names(X)[(ncol(X) - 1):ncol(X)] <- c("Severity","InClip")
X$Attacked <- X$Severity > 0
X$HasHost  <- !is.na(X$PinePct) & X$PinePct > 0

## spatial blocks for resampling-based inference
rc <- rowColFromCell(G, seq_len(ncell(G)))
X$block <- paste(rc[,1] %/% BLOCK, rc[,2] %/% BLOCK, sep = "_")

PRED  <- c("Elevation","Slope","TWI","RIX","Wind","TPI600")
EXTRA <- c("HLI","SxMean")          # aspect-derived; enter only the extended model
MVARS <- c(PRED, "PinePct", "Basal")   # defines the analysis frame, deliberately unchanged

## Cells lacking an inventory basal area are unmapped or non-productive and cannot
## enter a model that conditions on stand structure. Dropping them is reported here
## rather than left to silent na.action inside glm.
n_before <- sum(complete.cases(X[, PRED]) & !is.na(X$PinePct))
S <- X[complete.cases(X[, MVARS]), ]
H <- S[S$HasHost, ]
c(with_terrain_and_host = n_before, dropped_no_basal_area = n_before - nrow(S),
  analysis_cells = nrow(S), km2 = round(nrow(S) * prod(res(G)) / 1e6, 1),
  attacked = sum(S$Attacked), host_cells = nrow(H))
```

```{r}
#| label: tbl-relief
#| echo: false
#| tbl-cap: "The analysis landscape, and the 2 by 4 km clip drawn from it that the companion study used. The clip samples under two thirds of the relief and under 4 per cent of the cells."

tibble(
  Extent = c("full landscape", "2 by 4 km clip"),
  Cells  = c(nrow(S), sum(S$InClip == 1)),
  `Area (km²)` = round(c(nrow(S), sum(S$InClip == 1)) * prod(res(G)) / 1e6, 1),
  `Elevation (m)` = c(sprintf("%.0f to %.0f", min(S$Elevation), max(S$Elevation)),
                      sprintf("%.0f to %.0f", min(S$Elevation[S$InClip==1]),
                                              max(S$Elevation[S$InClip==1]))),
  `Relief (m)` = round(c(diff(range(S$Elevation)), diff(range(S$Elevation[S$InClip==1])))),
  `Attacked (%)` = round(100 * c(mean(S$Attacked), mean(S$Attacked[S$InClip==1])), 1)
) |> kable(booktabs = TRUE, align = "llllll", format.args = list(big.mark = ","))
```

# Methods

## Inference

The response is heavily clustered, so cell counts vastly overstate independent information. All coefficients carry percentile intervals from a spatial block bootstrap in which the landscape is partitioned into blocks of `r BLOCK` by `r BLOCK` cells, roughly 1 km on a side, resampled with replacement `r B_BOOT` times. Blocks of this size exceed any plausible correlation range of the survey polygons. We report the point estimate, the block interval and the proportion of resamples reproducing its sign, and treat a term as supported only when the interval excludes zero.

```{r}
#| label: fn-boot
#| echo: false

## The bootstrap refits several hundred times on >100,000 rows, so it works on a
## model matrix built once and calls glm.fit directly. This is arithmetically
## identical to glm() with the same formula, and roughly thirty times faster.
MODEL_VARS <- c("Elevation","WindResid","PinePct","Basal","TPI600","Slope","TWI","RIX")

design_matrix <- function(z, vars = MODEL_VARS)
  cbind(`(Intercept)` = 1, as.matrix(z[, vars, drop = FALSE]))

fit_fast <- function(Xm, y) glm.fit(Xm, y, family = binomial())

block_boot <- function(z, vars = MODEL_VARS, B = B_BOOT, seed = 42) {
  Xm <- design_matrix(z, vars); y <- z$y
  idx_of <- split(seq_len(nrow(z)), z$block)
  blocks <- names(idx_of)
  set.seed(seed)
  out <- matrix(NA_real_, B, ncol(Xm), dimnames = list(NULL, colnames(Xm)))
  for (b in seq_len(B)) {
    idx <- unlist(idx_of[sample(blocks, length(blocks), replace = TRUE)], use.names = FALSE)
    f <- try(fit_fast(Xm[idx, , drop = FALSE], y[idx]), silent = TRUE)
    if (!inherits(f, "try-error")) out[b, ] <- coef(f)
  }
  out
}

model_table <- function(z, vars = MODEL_VARS, seed = 42) {
  Xm <- design_matrix(z, vars)
  fit <- fit_fast(Xm, z$y)
  cf  <- coef(fit)
  se  <- sqrt(diag(chol2inv(fit$qr$qr[1:fit$rank, 1:fit$rank, drop = FALSE])))
  bs  <- block_boot(z, vars, seed = seed)
  tibble(term = names(cf), beta = as.numeric(cf), naive_z = as.numeric(cf / se),
         lo = apply(bs, 2, quantile, .025, na.rm = TRUE),
         hi = apply(bs, 2, quantile, .975, na.rm = TRUE),
         sign_stab = sapply(names(cf), function(n)
                       mean(sign(bs[, n]) == sign(cf[n]), na.rm = TRUE)))
}

fmt_model <- function(mt) {
  mt |> filter(term != "(Intercept)") |>
    transmute(Term = term, beta = sprintf("%+.3f", beta),
              `naive z` = sprintf("%+.1f", naive_z),
              `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
              `sign stability` = sprintf("%.2f", sign_stab),
              Supported = ifelse(lo * hi > 0, "yes", "no"))
}
```

## Orthogonalisation

Because the wind surface is built over the elevation surface, its coefficient in a model containing elevation is estimated from whatever small part of the joint distribution lets the two vary separately. We therefore replace wind by the residual of wind regressed on elevation, rescaled to unit variance.

It is easy to present this as a remedy for collinearity and it is not one. The substitution is a linear reparameterisation of the same model: fitted values and deviance are unchanged, and by the Frisch-Waugh-Lovell theorem the residualised coefficient is the raw wind coefficient multiplied by the square root of one minus the R² of wind on elevation, with its standard error scaled identically. Every supported-or-not verdict in this paper is therefore numerically the same in both parameterisations, and orthogonalisation buys no precision whatever. What changes is interpretation. The wind coefficient now refers to the part of the wind surface elevation does not explain, which is the quantity the @krawchuk2020 hypothesis is about, and the elevation coefficient becomes the total elevation association rather than the partial effect at fixed wind. The reparameterisation is worth making for that reason and no other.

```{r}
#| label: design
#| echo: false

standardise <- function(dd) {
  vars <- intersect(c(PRED, EXTRA, "PinePct", "Basal"), names(dd))
  z <- as.data.frame(scale(dd[, vars]))
  z$y     <- as.integer(dd$Attacked)
  z$block <- dd$block
  z$Severity <- dd$Severity          # carried for the ordinal fits in @sec-ordinal
  z$WindResid <- as.numeric(scale(residuals(lm(Wind ~ Elevation, data = z))))
  z
}
Z  <- standardise(S)     # whole landscape
ZH <- standardise(H)     # cells containing lodgepole
ZC <- standardise(S[S$InClip == 1, ])   # the 2 by 4 km clip
round(cor(Z[, c("Elevation","Wind","WindResid")]), 3)
```

```{r}
#| label: tbl-orthogonality
#| echo: false
#| tbl-cap: "How much of each candidate predictor is elevation. The wind product is a mesoscale field draped over terrain and is largely an elevation transform. The heat load and shelter indices are computed from the same elevation model by trigonometry and are almost unrelated to it."

tibble(Surface = c("Global Wind Atlas wind speed", "Heat load index (McCune and Keon)",
                   "Shelter index (Winstral, omnidirectional)", "Topographic position, 625 m"),
       `How it is built` = c("mesoscale field downscaled over the DEM",
                             "trigonometric function of DEM aspect and slope",
                             "maximum upwind slope from the DEM within 500 m",
                             "deviation from surrounding DEM mean"),
       `R² on elevation` = sapply(c("Wind","HLI","SxMean","TPI600"), function(v)
         round(summary(lm(as.formula(paste(v, "~ Elevation")), data = Z))$r.squared, 3))) |>
  kable(booktabs = TRUE, align = "lll")
```

## Response coding

@krawchuk2020 warn against coding refugia as a binary, since a refugium is a matter of degree and thresholding discards the signal distinguishing a lightly touched stand from a destroyed one. The survey carries an ordinal severity class, so we can follow that advice. We fit the binary attacked model for comparability with the refugia literature, and alongside it a cumulative logit at each severity threshold. Fitting the thresholds separately rather than imposing a single proportional-odds slope is deliberate, and is the standard diagnostic for whether proportional odds holds; @sec-ordinal shows that here it does not.

# Results

## Host effects by elevation {#sec-host}

```{r}
#| label: tbl-gradient
#| echo: false
#| tbl-cap: "Attack across the elevation gradient, before and after conditioning on lodgepole presence. Conditioning flattens the gradient below 1,900 m but not above it, where attack halves while lodgepole cover rises and stands thin."

S$band <- cut(S$Elevation, breaks = seq(500, 2400, 200), dig.lab = 5)
grad <- S |> group_by(band) |>
  summarise(Cells = n(),
            `Pine cover (%)` = round(mean(PinePct), 1),
            `Basal area` = round(mean(Basal, na.rm = TRUE), 1),
            `Wind (m/s)` = round(mean(Wind), 2),
            `Attacked, all cells (%)` = round(100 * mean(Attacked), 1),
            `Attacked, host present (%)` =
              ifelse(sum(HasHost) > 300, round(100 * mean(Attacked[HasHost]), 1), NA),
            .groups = "drop") |>
  filter(Cells > 500)
grad |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))

## Spread of host-conditioned attack below 1,900 m, quoted in the text below.
hp_low <- S |> filter(HasHost, Elevation < 1900) |>
  mutate(b = cut(Elevation, seq(500, 1900, 200))) |>
  group_by(b) |> summarise(n = n(), pct = 100 * mean(Attacked), .groups = "drop") |>
  filter(n > 300)
HP_RANGE <- range(hp_low$pct)

## Host cover above the threshold, compared two ways, because the two
## comparisons disagree and reporting only the flattering one would be the
## same denominator trick this paper criticises in @sec-lowtrap. PINE_LO is
## the whole gradient below the threshold, which is dominated by the pine-poor
## valley; PINE_ADJ is the 150 m band immediately below it.
PINE_HI  <- mean(H$PinePct[H$Elevation >= ALPINE])
PINE_LO  <- mean(H$PinePct[H$Elevation <  ALPINE])
PINE_ADJ <- mean(H$PinePct[H$Elevation >= ALPINE - 150 & H$Elevation < ALPINE])
```

Across the full gradient attack is strongly unimodal (@tbl-gradient, @fig-gradient), rising from under three per cent in the valley to a peak near 1,430 m and falling to zero above 2,100 m.

Conditioning on host presence separates that curve into two phenomena. Below roughly 1,900 m the hump is host distribution: among cells containing lodgepole, attack runs between `r sprintf("%.0f", HP_RANGE[1])` and `r sprintf("%.0f", HP_RANGE[2])` per cent with no strong trend from the valley bottom upward, and the rise in the raw curve is the rise in pine cover. Attack is `r sprintf("%.1f", 100*mean(S$Attacked[S$HasHost]))` per cent where lodgepole is present against `r sprintf("%.1f", 100*mean(S$Attacked[!S$HasHost]))` per cent where it is absent.

Above `r format(ALPINE, big.mark = ",")` m conditioning does not remove the decline. Attack among host-bearing cells falls by about half, while stand basal area falls from `r sprintf("%.1f", mean(H$Basal[H$Elevation<ALPINE], na.rm=TRUE))` to `r sprintf("%.1f", mean(H$Basal[H$Elevation>=ALPINE], na.rm=TRUE))`. Stands are thinner and attack is suppressed.

Host cover decides whether this is the pattern the @krawchuk2020 hypothesis requires, and it depends on the comparison, so we give both. Against the whole gradient below, mean lodgepole cover in host-bearing cells is `r sprintf("%.1f", PINE_HI)` per cent above the threshold against `r sprintf("%.1f", PINE_LO)` per cent below: essentially unchanged, so host scarcity cannot account for an attack rate that more than halves. Against the `r format(ALPINE - 150, big.mark = ",")` to `r format(ALPINE, big.mark = ",")` m band immediately below, however, cover is `r sprintf("%.1f", PINE_ADJ)` per cent, so the top of the gradient carries about `r sprintf("%.0f", 100 * (1 - PINE_HI / PINE_ADJ))` per cent less pine than the band beneath it. The first comparison is flattered by the pine-poor valley it includes, which is the denominator problem @sec-lowtrap is about, and we would be inconsistent to rely on it alone.

The honest reading is that host thins somewhat at the top while attack falls a great deal more, so the conjunction the hypothesis requires holds in the weaker form. It is not the clean case of host holding constant.

```{r}
#| label: fig-gradient
#| echo: false
#| fig-width: 10.5
#| fig-height: 4.2
#| fig-cap: "(a) Attack across the gradient, raw and conditioned on lodgepole presence, with mean lodgepole cover. (b) The same data against wind speed: wind rises monotonically with elevation while attack turns over, so no monotone wind relationship can hold across the gradient. Shading marks the elevation range of the 2 by 4 km clip."

fine <- S |> mutate(b = cut(Elevation, breaks = seq(500, 2350, 100))) |>
  group_by(b) |>
  summarise(el = mean(Elevation), raw = 100*mean(Attacked), pine = mean(PinePct),
            wind = mean(Wind), n = n(),
            cond = ifelse(sum(HasHost) > 150, 100*mean(Attacked[HasHost]), NA_real_),
            .groups = "drop") |> filter(n > 300)
crng <- range(S$Elevation[S$InClip == 1])

pa <- ggplot(fine) +
  annotate("rect", xmin = crng[1], xmax = crng[2], ymin = -Inf, ymax = Inf,
           fill = "#b2182b", alpha = .10) +
  geom_line(aes(el, raw,  colour = "all cells"), linewidth = .9) +
  geom_line(aes(el, cond, colour = "host present"), linewidth = .9) +
  geom_line(aes(el, pine, colour = "lodgepole cover (%)"), linetype = "22", linewidth = .8) +
  scale_colour_manual(values = c("all cells" = "#2b4162", "host present" = "#b2182b",
                                 "lodgepole cover (%)" = "#3f8f9b"), name = NULL) +
  labs(x = "Elevation (m)", y = "Per cent", title = "(a) The hump is host, except at the top") +
  theme(legend.position = "bottom")

pb <- ggplot(fine, aes(wind, raw)) +
  geom_path(colour = "#3f8f9b", linewidth = .8) +
  geom_point(aes(colour = el), size = 1.8) +
  scale_colour_gradientn(colours = c("#2b4162","#3f8f9b","#c6d8af","#e8c07d","#8c4b3c"),
                         name = "Elev (m)") +
  labs(x = "Mean wind speed (m/s)", y = "Attacked (%)",
       title = "(b) Wind is monotone, attack is not")
pa + pb
```

## The Krawchuk test {#sec-krawchuk}

```{r}
#| label: tbl-krawchuk
#| echo: false
#| tbl-cap: "The hypothesis test. Cells containing lodgepole pine only. Coefficients are per standard deviation; wind is entered as the component orthogonal to elevation. Intervals are percentile intervals from resampled 1 km spatial blocks. Only host cover clears the interval test; the orthogonalised wind term is the strongest negative coefficient but its interval reaches zero."

mt_h <- model_table(ZH)
fmt_model(mt_h) |> kable(booktabs = TRUE, align = "llllll")
```

```{r}
#| label: krawchuk-numbers
#| echo: false
#| output: false
wr <- mt_h[mt_h$term == "WindResid", ]
pp <- mt_h[mt_h$term == "PinePct", ]
```

Restricted to cells containing host, the orthogonalised wind term is the largest negative coefficient in the model, `r sprintf("%+.3f", wr$beta)` per standard deviation, comparable to the positive effect of lodgepole cover at `r sprintf("%+.3f", pp$beta)` (@tbl-krawchuk). Its sign is reproduced in `r sprintf("%.0f", 100*wr$sign_stab)` per cent of block resamples, but its 95 per cent block interval runs `r sprintf("[%+.3f, %+.3f]", wr$lo, wr$hi)` and reaches zero.

On this fit the hypothesis is supported in direction and unresolved in magnitude, and @sec-nonlinear later withdraws even the direction, so read this section as the conventional analysis whose specification we go on to test rather than a result we stand behind. The descriptive result in @sec-host is not in doubt: attack more than halves at the top of the gradient while host cover thins only slightly. What is in doubt is whether wind is the right explanation, and with roughly `r length(unique(ZH$block))` independent 1 km blocks the binary response cannot settle it. Host cover is the only term whose interval excludes zero.

That verdict is a property of the response coding rather than of the data, and @sec-ordinal revisits it. We report the binary model first because it is the form the refugia literature uses, and the comparison is itself a result.

Elevation, entered alongside the orthogonalised wind term, is itself negative and unsupported: once host is in the model the remaining elevation signal is weak, so the suppression at the top is not simply an elevation effect.

## Disturbance severity metrics {#sec-ordinal}

```{r}
#| label: polr-fit
#| echo: false
#| output: false

## Fitted only to show what a single proportional-odds summary conceals; the
## threshold-specific fits in @tbl-ordinal are the result.
zo <- ZH; zo$sev <- factor(pmin(zo$Severity, 4L), levels = 0:4, ordered = TRUE)
polr_fit <- MASS::polr(sev ~ Elevation + WindResid + PinePct + Basal +
                             TPI600 + Slope + TWI + RIX,
                       data = zo, method = "logistic", Hess = FALSE)
```

```{r}
#| label: tbl-ordinal
#| echo: false
#| tbl-cap: "Cumulative logits at each severity threshold, cells containing lodgepole pine only. The orthogonalised wind coefficient strengthens monotonically as the threshold rises and its block interval excludes zero above trace, which the binary model does not show."

CUTS <- tibble(k = 1:4,
               label = c("trace or worse", "light or worse",
                         "moderate or worse", "severe or worse"))

cum <- lapply(seq_len(nrow(CUTS)), function(i) {
  z <- ZH; z$y <- as.integer(z$Severity >= CUTS$k[i])
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Threshold = CUTS$label[i], Cells = sum(z$y),
         `Wind residual beta` = coef(fit)[["WindResid"]],
         lo = q[1], hi = q[2],
         `sign stability` = mean(sign(bs[, "WindResid"]) ==
                                 sign(coef(fit)[["WindResid"]]), na.rm = TRUE),
         `Pine cover beta` = coef(fit)[["PinePct"]],
         `Elevation beta`  = coef(fit)[["Elevation"]])
}) |> bind_rows()

cum |> transmute(Threshold, Cells,
                 `Wind residual beta` = sprintf("%+.3f", `Wind residual beta`),
                 `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
                 `sign stability` = sprintf("%.2f", `sign stability`),
                 `Pine cover beta` = sprintf("%+.3f", `Pine cover beta`),
                 Supported = ifelse(lo * hi > 0, "yes", "no")) |>
  kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

Collapsing the response to attacked or not costs the study its main result. Fitted at successively harder severity thresholds, the orthogonalised wind coefficient strengthens monotonically from `r sprintf("%+.3f", cum$"Wind residual beta"[1])` at trace or worse to `r sprintf("%+.3f", cum$"Wind residual beta"[4])` at severe or worse, a `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])`-fold increase (@tbl-ordinal). At every threshold above trace the block interval excludes zero. The first row is the binary model of @sec-krawchuk restated, since a cell reaching trace or worse is exactly a cell recorded as attacked, so the table extends that fit rather than competing with it. It is also the weakest of the four, and the one the refugia literature would conventionally have reported.

This is a dose-response gradient of the shape the hypothesis predicts. Wherever the wind surface has a component independent of elevation, that component is associated with less attack, and the association is strongest where attack was heaviest. It also explains the binary result: a cell that reached trace severity is barely distinguishable from an unattacked cell, so a threshold drawn between them puts most of the noise exactly where the model is asked to fit.

The slopes differ by a factor of `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])` across the thresholds, so proportional odds does not hold and a single ordinal coefficient would be the wrong summary. Fitting one anyway returns `r sprintf("%+.3f", coef(polr_fit)[["WindResid"]])`, close to the trace-threshold estimate and concealing everything above it. Two other terms move with the threshold. Lodgepole cover, the only term that cleared the interval test in the binary model, weakens as the threshold rises and reverses at severe or worse (`r sprintf("%+.3f", cum$"Pine cover beta"[4])`). One reading is ecological: host abundance governs whether a stand is attacked at all, not how hard. A second is an artefact of the inventory's vintage, and we cannot separate them. Composition was interpreted from 2014 photography, late in the outbreak window, so stands that lost their pine early carry a lower recorded lodgepole percentage than when the beetle reached them. That predicts a negative coefficient at the heaviest severities, which is what is observed. We report the reversal and decline to interpret it. Elevation turns positive at moderate or worse (`r sprintf("%+.3f", cum$"Elevation beta"[3])`), consistent with severe attack having concentrated in the mid-elevation pine belt rather than at either end.

We cannot attribute the gradient to wind rather than cold, and the Discussion explains why the ordinal result does not help. More seriously, @sec-nonlinear shows the sweep does not survive entering elevation flexibly, so what looks here like improved support is largely the unimodal elevation response returning through a predictor orthogonalised only against the linear part of elevation. Read this table together with that one.

## Wind effects by elevation {#sec-nonlinear}

```{r}
#| label: tbl-nonlinear
#| echo: false
#| tbl-cap: "The ordinal sweep refitted with elevation entered flexibly instead of linearly, and wind residualised against that same basis. Elevation explains eleven points more of the wind surface once curvature is allowed, and no threshold clears the interval test under either flexible basis."

## Everything above enters elevation as one linear term, and orthogonalises wind
## against that same straight line. But @fig-gradient shows the response is
## unimodal in elevation, and a straight line through a hump leaves a structured,
## elevation-shaped residual that any covariate correlated with the curvature can
## absorb. The wind surface is such a covariate by construction, because the
## downscaling is a nonlinear function of terrain. Residualising against a line
## cannot remove curvature the line never described. This table tests that.

elev_basis <- function(E, basis) switch(basis,
  linear = cbind(ElevL = E),
  quad   = cbind(ElevL = E, ElevQ = E^2),
  spline = { b <- splines::ns(E, df = 4); colnames(b) <- paste0("ElevS", 1:4); as.matrix(b) })

nl_row <- function(basis) {
  z  <- ZH
  EB <- elev_basis(z$Elevation, basis)
  z  <- cbind(z[, setdiff(names(z), "Elevation")], as.data.frame(EB))
  ## wind residualised on THIS basis, not on the straight line
  z$WindResid <- as.numeric(scale(residuals(lm(z$Wind ~ EB))))
  vars <- c(colnames(EB), "WindResid", "PinePct", "Basal", "TPI600", "Slope", "TWI", "RIX")
  r2 <- summary(lm(z$Wind ~ EB))$r.squared
  lapply(seq_len(nrow(CUTS)), function(i) {
    z$y <- as.integer(z$Severity >= CUTS$k[i])
    fit <- fit_fast(design_matrix(z, vars), z$y)
    bs  <- block_boot(z, vars)
    q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
    tibble(`Elevation basis` = basis, `R² wind on basis` = round(r2, 3),
           Threshold = CUTS$label[i],
           `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
           `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
           Supported = ifelse(q[1]*q[2] > 0, "yes", "no"))
  }) |> bind_rows()
}
nl <- bind_rows(lapply(c("linear","quad","spline"), nl_row))
nl |> kable(booktabs = TRUE, align = "llllll")
```

@sandel2015 gives the diagnostic that motivates this test. For a genuinely linear relationship a coefficient is extent-independent and only the coefficient of determination moves as extent changes; a coefficient that changes sign under truncation is a symptom of an unmodelled nonlinearity rather than a general law of extent. Our response is unimodal in elevation and our wind surface monotone in it, exactly the configuration that produces such a symptom, so the specification has to be tested rather than assumed.

It does not pass. Entering elevation as a quadratic raises the share of the wind surface that elevation explains from `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[1]))` to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[5]))`, and a natural spline on four degrees of freedom raises it to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[9]))`. Roughly eleven percentage points of the wind surface's variance is therefore nonlinear in elevation, and the linear orthogonalisation used above leaves all of it inside what we had been calling the component of wind independent of elevation.

When that residue is removed the wind result goes with it. The coefficient at severe or worse falls from `r nl$"Wind residual beta"[4]` to `r nl$"Wind residual beta"[8]`, and at moderate or worse from `r nl$"Wind residual beta"[3]` to `r nl$"Wind residual beta"[7]`, losing between three fifths and seven tenths of its magnitude. Under both flexible bases every interval covers zero at every threshold. The monotone strengthening across thresholds survives in the point estimates, but nothing in the sweep clears the test this paper set itself.

We report this as the result rather than relegating it to a sensitivity appendix, because it changes what the preceding two sections mean. The dose-response in @tbl-ordinal is not evidence about wind. It is very largely the unimodal elevation response reappearing through a predictor that was never orthogonal to the curvature of elevation, only to its slope. This study does not have a wind effect to report, and the appearance of one followed from how elevation was specified.

## Thermal load by aspect {#sec-aspect}

```{r}
#| label: tbl-aspect
#| echo: false
#| tbl-cap: "The host-present model with the two aspect-derived terms added, at each threshold it can carry. Heat load is positive and strengthens sharply with severity: warmer aspects burn harder at the same elevation. The orthogonalised wind term is not displaced by it. The shelter index points the other way, which the text takes seriously rather than explaining away."

FULLV <- c(MODEL_VARS, EXTRA)
ZHx   <- ZH[complete.cases(ZH[, FULLV]), ]

## Ten terms and 1,522 events cannot be fitted together: the severe threshold
## drives fitted probabilities to numerical zero. We detect that rather than
## report the resulting coefficients, and the severe row is dropped below.
sep_check <- function(z, vars, k) {
  z$y <- as.integer(z$Severity >= k)
  p <- fit_fast(design_matrix(z, vars), z$y)$fitted.values
  sum(p < 1e-6 | p > 1 - 1e-6)
}
sep <- sapply(1:4, function(k) sep_check(ZHx, FULLV, k))
names(sep) <- c("trace+","light+","moderate+","severe+")
OK_CUTS <- which(sep == 0)

asp_tab <- lapply(OK_CUTS, function(k) {
  z <- ZHx; z$y <- as.integer(z$Severity >= k)
  fit <- fit_fast(design_matrix(z, FULLV), z$y); bs <- block_boot(z, FULLV)
  ci <- function(n) quantile(bs[, n], c(.025, .975), na.rm = TRUE)
  one <- function(n) sprintf("%+.3f [%+.3f, %+.3f]", coef(fit)[[n]], ci(n)[1], ci(n)[2])
  tibble(Threshold = CUTS$label[k], Cells = sum(z$y),
         `Wind residual` = one("WindResid"), `Heat load` = one("HLI"),
         `Shelter` = one("SxMean"))
}) |> bind_rows()

asp_tab |> kable(booktabs = TRUE, align = "lllll",
                 format.args = list(big.mark = ","))
```

The inherited predictor set had no aspect, so until now the model could not tell a warm slope from a cold one at the same elevation. Adding two aspect-derived terms does three things.

First, it supplies the one piece of thermal information available without a temperature surface. Heat load is positive at every threshold and strengthens steeply with severity (@tbl-aspect). Since heat load is essentially uncorrelated with elevation (@tbl-orthogonality), this is not the elevation gradient under another name: at a given elevation, southwest-facing ground carries heavier mortality than northeast-facing ground. That is the sign cold limitation predicts, and the first evidence here that the thermal axis works independently of elevation.

Second, it leaves the wind coefficient standing. The orthogonalised wind term does not weaken when heat load is controlled for, so whatever it tracks is not the warm-aspect effect in disguise. It is, however, largely the curvature of elevation: this model still enters elevation linearly, and @sec-nonlinear shows the wind term does not survive a flexible basis. Controlling for aspect does not rescue it.

Third, and least comfortably, the two exposure measures disagree. The shelter index is negative, meaning sheltered ground carries *less* attack and exposed ground more, the opposite of what the orthogonalised wind term says and of what @krawchuk2020 predict. The two are only weakly correlated (`r sprintf("%.2f", cor(ZHx$WindResid, ZHx$SxMean))`), so they are not two measurements of one quantity that happen to disagree; they measure different things, and only one is a direct geometric description of exposure. We take this as evidence against treating any single surface as "wind exposure" and reporting its coefficient as though the phrase had one referent. It is also a caution against our own result: the wind residual's negative sign is not corroborated by the one predictor here that describes exposure geometrically rather than by downscaling.

The severe threshold is absent from @tbl-aspect because ten terms and `r sum(ZHx$Severity >= 4)` events drive the fit into quasi-separation, with `r format(sep[["severe+"]], big.mark = ",")` cells at numerically zero fitted probability. The eight-term model of @tbl-ordinal is free of this, which is why it retains its severe row and this one does not.

## Winter temperature thresholds {#sec-cold}

@tbl-cold summarises the record and @fig-station maps the network. No station lies inside the analysis grid, and the stations reaching the coldest minima are not the highest ones, which is the point the next paragraph develops.

```{r}
#| label: tbl-cold
#| echo: false
#| tbl-cap: "Observed winter minimum air temperatures from stations within 100 km, by elevation band, after excluding station-winters censored at the -20.0 °C sensor floor. Nothing approaches the temperatures conventionally cited as lethal for this insect, and elevation orders the minima only weakly."

st <- read.csv(here("02.inputs/climate/station-winter-minima.csv"), stringsAsFactors = FALSE)

## The censoring rule is applied here rather than taken on trust from the
## retrieval script. Many older British Columbia Wildfire Service sensors floor
## at exactly -20.0 C. A station-winter is unusable when its minimum sits
## exactly on that floor, because the true minimum is then unknown. A minimum
## colder than the floor shows the sensor did read past it, so readings at
## -20.0 in those winters are real measurements and the winter is kept. The
## floor is a sensor artefact of that network, so the rule does not touch the
## Environment and Climate Change Canada records.
FLOOR <- -20
st$censored <- st$network == "BCWS" &
               abs(st$min_temp_c - FLOOR) < 1e-6 &
               st$n_at_minus20 > 0
stopifnot(identical(st$censored, as.logical(st$censored_flag)))

## Isolated single-observation spikes are rejected at the observation level by
## the retrieval script, which is the only stage holding the sub-daily series a
## per-winter summary cannot carry; `min_temp_c_usable` is that de-spiked
## minimum. Everything else about the screen is visible above.
use <- st[!st$censored & !is.na(st$min_temp_c_usable), ]

use$band <- cut(use$elevation_m, c(0, 800, 1500, 2500),
                labels = c("below 800 m", "800 to 1,500 m", "above 1,500 m"))
use |> group_by(band) |>
  summarise(Stations = n_distinct(station_name), `Station-winters` = n(),
            `Coldest (°C)` = min(min_temp_c_usable),
            `Mean minimum (°C)` = round(mean(min_temp_c_usable), 1),
            .groups = "drop") |>
  kable(booktabs = TRUE, align = "lllll")
```

```{r}
#| label: fig-station
#| echo: false
#| fig-width: 5.9
#| fig-height: 5.1
#| fig-cap: "The weather station network behind the cold argument. All stations within 100 km reporting a usable winter minimum between 1999 and 2015, shaded by station elevation, with the analysis grid outlined. Only the two stations named in the text are labelled. The alpine station and the valley stations near the international boundary reach comparable minima, which is the cold-air pooling the text describes."

stn <- aggregate(cbind(latitude, longitude, elevation_m) ~ station_name,
                 data = st, FUN = function(x) x[1])
coldest_by <- aggregate(min_temp_c_usable ~ station_name, data = use, FUN = min)
stn <- merge(stn, coldest_by, by = "station_name", all.x = TRUE)
stn_sf <- st_transform(st_as_sf(stn, coords = c("longitude","latitude"), crs = 4326), 3005)
xy <- st_coordinates(stn_sf); stn_sf$X <- xy[, 1]; stn_sf$Y <- xy[, 2]
site_bc <- st_transform(st_as_sf(as.polygons(ext(G), crs = crs(G))), 3005)
prov_bc <- st_transform(ne_states(country = c("Canada","United States of America"),
                                  returnclass = "sf"), 3005)
bbs <- st_bbox(st_buffer(st_union(c(st_geometry(stn_sf), st_geometry(site_bc))), 25000))
sw  <- unname(bbs["xmax"] - bbs["xmin"])
key <- stn_sf[order(stn_sf$min_temp_c_usable)[1:2], ]

fig_station <- ggplot() +
  geom_sf(data = prov_bc, fill = "grey95", colour = "grey70", linewidth = .25) +
  geom_sf(data = site_bc, fill = NA, colour = "#C0392B", linewidth = .6) +
  geom_sf(data = stn_sf, aes(fill = elevation_m), shape = 21, size = 1.9,
          colour = "grey20", stroke = .25) +
  scale_fill_viridis_c(option = "viridis", name = "Station\nelevation (m)") +
  geom_label_repel(data = as.data.frame(key), aes(X, Y, label = tolower(station_name)),
                   size = 2, colour = "grey10", fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm"), box.padding = .5,
                   min.segment.length = 0, segment.colour = "grey30",
                   segment.size = .3, seed = 1) +
  annotate("label", x = bbs["xmin"] + .02*sw,
           y = bbs["ymin"] + .04*(bbs["ymax"] - bbs["ymin"]),
           label = "red outline:\nanalysis grid", size = 1.9, hjust = 0, vjust = 0,
           colour = "#C0392B", fill = "white", alpha = .85,
           label.padding = unit(.7, "mm")) +
  coord_sf(crs = 3005, xlim = bbs[c("xmin","xmax")], ylim = bbs[c("ymin","ymax")],
           datum = st_crs(4326), expand = FALSE) +
  north + map_theme +
  labs(title = "Weather stations screened for winter minima",
       subtitle = "EPSG:3005 NAD83 / BC Albers")
fig_station
ggsave("../03.outputs/PNG/fig-station-maps.png", fig_station,
       width = 150, height = 130, units = "mm", dpi = 500, bg = "white")
```

```{r}
#| label: cold-numbers
#| echo: false
#| output: false
coldest  <- use[which.min(use$min_temp_c_usable), ]
hi       <- use[use$elevation_m >= 1500, ]
lowcold  <- use[use$elevation_m < 800, ]
lowcold  <- lowcold[which.min(lowcold$min_temp_c_usable), ]
```

The cold explanation for the high-elevation suppression assumes it gets cold enough up there to kill beetles. It does not hold. Across `r nrow(use)` station-winters from `r n_distinct(use$station_name)` stations spanning `r min(use$elevation_m)` to `r max(use$elevation_m)` m, retrieved and screened by the script in `02.inputs/climate/`, the coldest defensible winter minimum anywhere in the region is `r sprintf("%.1f", coldest$min_temp_c_usable)` °C, at `r tolower(coldest$station_name)` at `r format(round(coldest$elevation_m), big.mark=",")` m in winter `r coldest$winter_year`. The number of station-winters reaching -35 °C or colder is `r sum(use$min_temp_c_usable <= -35)`, and the number reaching -40 °C or colder is `r sum(use$min_temp_c_usable <= -40)`. Above 1,500 m the mean winter minimum is `r sprintf("%.1f", mean(hi$min_temp_c_usable))` °C. The temperatures conventionally cited as lethal for larvae sit near -40 °C under bark, the figure at which @safranyik2006 report complete mortality, while @cooke2009mortality measure a median lethal supercooling point near -37 °C and caution against any fixed threshold. Both are roughly ten degrees colder than anything this record contains. Neither is an air temperature, but phloem tracks air within about two degrees, so the comparison holds; and cold mortality is graded rather than a threshold, so this is a margin against substantial mortality rather than a guarantee of none.

Two features of the record matter. Elevation orders these minima only weakly: `r tolower(lowcold$station_name)`, at `r format(round(lowcold$elevation_m), big.mark=",")` m, reaches `r sprintf("%.1f", lowcold$min_temp_c_usable)` °C, within `r sprintf("%.1f", abs(lowcold$min_temp_c_usable - coldest$min_temp_c_usable))` °C of the alpine station `r format(round(coldest$elevation_m - lowcold$elevation_m), big.mark=",")` m higher. That is cold-air pooling, and the reason temperature cannot be treated as a linear function of elevation here. And `r sum(st$censored)` station-winters were set aside because the sensors floor at exactly -20.0 °C, an artefact concentrated before 2008; a naive minimum over those would have manufactured a threshold that is not in the data.

This is a plausibility check, not a temperature surface. No station inside the study area reported during the outbreak window, and those that do are spread across roughly 100 km of separate massifs, so these values bound the regional cold climate rather than map it. Within that limit the conclusion is firm enough: direct freeze mortality is unlikely to be what suppresses attack at the top of this gradient.

## Area extents of coefficients {#sec-extent}

```{r}
#| label: tbl-extent
#| echo: false
#| tbl-cap: "The same model fitted at three extents. As the analysis area shrinks, the share of wind variance explained by elevation rises and the orthogonalised wind coefficient reverses sign. The 2 by 4 km clip is the extent at which the companion study was run."

extent_row <- function(z, lab) {
  r2  <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  ## The block count is reported because it, not the cell count, is the
  ## resampling unit. The clip holds few enough blocks that its interval is
  ## far weaker than its cell count suggests, and the reader must see that.
  tibble(Extent = lab, Cells = nrow(z), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         ## This is the variance inflation the *raw* wind term would carry
         ## against elevation. It is not the VIF of the fitted model, in which
         ## the residualised term is orthogonal to elevation by construction and
         ## its VIF is 1. It is reported to show the collinearity that
         ## orthogonalisation reparameterises away rather than removes.
         `VIF of raw wind` = round(1/(1-r2), 1),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]))
}
bind_rows(
  extent_row(Z,  sprintf("full landscape (%.0f km²)", nrow(S)*prod(res(G))/1e6)),
  extent_row(ZH, "landscape, host present"),
  extent_row(ZC, sprintf("2 by 4 km clip (%.1f km²)",
                         sum(S$InClip == 1) * prod(res(G)) / 1e6))
) |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

The confound is a function of extent, and so is the answer. Across the landscape, elevation explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = Z))$r.squared)` per cent of the variance in the wind surface; inside the 2 by 4 km clip it explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` per cent (@tbl-extent). The orthogonalised wind coefficient is negative at landscape extent and positive inside the clip.

Two claims live in that table and carry very different weight.

The collinearity result is firm. The share of wind variance elevation explains rises from `r sprintf("%.3f", summary(lm(Wind ~ Elevation, data = Z))$r.squared)` across the landscape to `r sprintf("%.3f", summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` inside the clip, and that quantity is a property of the two surfaces rather than an estimate with sampling error. It is the mechanism this paper is about, and does not depend on resolving the Krawchuk hypothesis.

The sign reversal is a comparison of two point estimates and is not, on this evidence, a demonstrated reversal. The clip contains only `r length(unique(ZC$block))` spatial blocks, so its block interval runs `r {q <- quantile(block_boot(ZC)[, "WindResid"], c(.025,.975), na.rm = TRUE); sprintf("[%+.3f, %+.3f]", q[1], q[2])}`: it spans zero, and it spans both landscape estimates. By the rule this paper set itself the clip coefficient is not distinguishable from zero, from the landscape estimate, or from its own opposite. What the clip demonstrates is that shrinking the extent destroys the ability to estimate the coefficient at all, a weaker claim than reversal and still damaging for the practice we criticise: a study at this extent would report a positive point estimate with an interval too wide to notice it had learned nothing. Establishing reversal proper would need an extent-by-wind interaction fitted on the pooled data and bootstrapped on common blocks. We have not done that and do not claim it.

## Grain effects on coefficients {#sec-grain}

```{r}
#| label: tbl-grain
#| echo: false
#| tbl-cap: "The host-present model refitted at three grains. Predictors are aggregated by mean, severity class by maximum, and the bootstrap blocks are rescaled to stay near 1 km. Unlike extent, grain changes nothing: collinearity, sign and magnitude are all stable across a four-fold change in cell size."

CONT <- c("Elevation","Slope","TWI","RIX","Wind","PinePct","Basal")
base_stack <- c(terr[[c("Elevation","Slope","TWI","RIX","Wind")]],
                host[[c("PinePct","Basal")]], severity, in_clip)
names(base_stack) <- c(CONT, "Severity", "InClip")

## fact is the aggregation factor from the native 25 m grid.
build_grain <- function(fact) {
  b <- if (fact == 1L) base_stack else
       c(aggregate(base_stack[[CONT]], fact, fun = "mean", na.rm = TRUE),
         aggregate(base_stack[[c("Severity","InClip")]], fact, fun = "max", na.rm = TRUE))
  Gi <- b[["Elevation"]]; rs <- res(Gi)[1]
  w  <- max(3L, as.integer(2 * floor(625 / rs / 2) + 1))   # odd window nearest 625 m
  b$TPI600 <- tpi(Gi, w)
  d <- as.data.frame(b, na.rm = FALSE)
  d$Attacked <- d$Severity > 0
  d$HasHost  <- !is.na(d$PinePct) & d$PinePct > 0
  blk <- max(1L, as.integer(round(BLOCK / fact)))          # blocks stay near 1 km
  rc  <- rowColFromCell(Gi, seq_len(ncell(Gi)))
  d$block <- paste(rc[,1] %/% blk, rc[,2] %/% blk, sep = "_")
  d <- d[complete.cases(d[, MVARS]), ]
  structure(d, res = rs, win = w * rs)
}

grain_row <- function(fact) {
  d  <- build_grain(fact); h <- d[d$HasHost, ]
  z  <- standardise(h)
  r2 <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y); bs <- block_boot(z)
  q  <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Grain = sprintf("%.0f m", attr(d, "res")),
         `TPI window` = sprintf("%.0f m", attr(d, "win")),
         `Host cells` = nrow(h), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
         `sign stability` = sprintf("%.2f",
           mean(sign(bs[, "WindResid"]) == sign(coef(fit)[["WindResid"]]), na.rm = TRUE)))
}

## The 25 m row must reproduce the analysis frame built in `assemble`; if this
## assertion ever fails the two code paths have drifted apart.
stopifnot(nrow(build_grain(1L)) == nrow(S))

bind_rows(lapply(c(1L, 2L, 4L), grain_row)) |>
  kable(booktabs = TRUE, align = "llllllll", format.args = list(big.mark = ","))
```

@work2011, working on carabid beetles, found the apparent importance of terrain derivatives changed with the resolution at which they were computed, because elevation error propagates into derived curvature differently at every grain. The concern transfers directly and is worth separating from extent, which is routinely confused with it.

Here they behave quite differently (@tbl-grain). Aggregating from 25 m to 100 m, a sixteen-fold reduction in cells, leaves the confound essentially untouched: elevation explains close to the same share of wind variance at every grain, and the orthogonalised wind coefficient neither reverses nor decays. The suppression at the top of the gradient is a landscape-scale feature and survives a coarser lens.

The contrast with @sec-extent is the point, and one we should not overstate. Enlarging the cell fourfold changes almost nothing; shrinking the study area destroys the estimate. Grain governs how finely a pattern is resolved, extent governs which part of the joint distribution of elevation and wind the model ever sees, and here only the second mattered.

That is a result about this landscape and this pair of surfaces, not a general ordering of the two axes. @wiens1989 set out the distinction and warned that correlations evident within one domain may disappear or change sign when the extent is enlarged past it, and @dungan2002 recommended that studies report unit size, shape, spacing and extent together for the same reason. The literature since has not settled which axis matters more, and where it has looked it has often found the opposite of what we find: @mcgarigal2016, reviewing several hundred multi-scale habitat studies, concluded that the emphasis on extent may have left grain under-examined. Nor is reporting the weak point we assumed: @estes2018 found resolution and extent reported at closely comparable rates, near three fifths in both cases.

What our result adds is narrower and more useful. @sandel2015 shows that for a genuinely linear relationship the coefficient is extent-independent and only the coefficient of determination moves with extent. A coefficient that changes sign under truncation is therefore a diagnostic for an unmodelled nonlinearity rather than an instance of a general law. That is very likely what is happening here: @fig-gradient shows attack unimodal in elevation while wind rises monotonically, so no monotone wind term can hold across the whole gradient, and the clip lands on one limb of the hump. Read that way the finding is not "extent reverses inferences" but the sharper "truncating a gradient onto one limb of a unimodal response will reverse the apparent sign of anything monotonically correlated with the gradient." Testing it properly means fitting the nonlinearity rather than assuming it away, which is the first thing we would do next.

# Discussion

The high-elevation suppression is real and it is not host limitation. Among cells carrying lodgepole pine, attack above `r format(ALPINE, big.mark=",")` m runs at `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation>=ALPINE]))` per cent against `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation<ALPINE]))` per cent below, while mean pine cover there is `r sprintf("%.0f", 100*(1 - PINE_HI/PINE_ADJ))` per cent lower than in the band immediately below and basal area roughly a fifth lower, so the fall in attack is several times larger than the fall in host. Under the @krawchuk2020 definition these are refugia: disturbed less severely than the surrounding landscape, with host present, and predictably located by an enduring topographic feature, matching @cartwright2018 in associating refugia with thinner stands.

What we cannot do is attribute them to wind. The ordinal result in @sec-ordinal changes one half of that difficulty and leaves the other where it was.

The statistical half looked, at first, much improved. Once severity is modelled as the ordinal variable it is, the orthogonalised wind term clears the interval test at every threshold above trace and strengthens monotonically with severity. @sec-nonlinear withdraws that reading. Entering elevation flexibly, and residualising wind against the same basis rather than against a straight line, removes between three fifths and seven tenths of the coefficient and leaves no threshold clearing the test. The ordinal coding does recover structure the binary discards, and @krawchuk2020's warning against binary refugia coding stands. But that structure is not attributable to wind: it is very largely the curvature of the elevation response, which the linear specification pushed into the wind term. We therefore do not report a wind effect, in either coding.

The identification half is untouched. The upper gradient is also where it is coldest, and cold limits this beetle through overwinter mortality and through the degree-day accumulation univoltine development requires [@carroll2004bionomics; @cooke2009mortality; @safranyik2006]. Cold predicts a dose-response in severity just as wind disruption does: fewer beetles surviving winter means both fewer stands attacked and lighter attack in those that are. The monotone strengthening across thresholds is therefore fully consistent with cold limitation and does nothing to separate the two. Orthogonalising against elevation does not orthogonalise against temperature, which is not a linear function of elevation where cold air pools.

We looked for a way out and do not think one exists at this extent. Separating exposure from cold needs a temperature surface whose fine-scale structure is not itself a function of the elevation model, and any surface of the form intercept plus lapse rate times elevation is an affine rescaling of the elevation layer, whatever the lapse rate's source. It is collinear by construction, and only a residual departure from that form would identify anything, which would itself require terrain as the interpolator. The gridded products with genuine sub-kilometre temperature structure all obtain it by regression onto a digital elevation model, so they reproduce the problem rather than solve it.

Two results bear on this. The first is that cold limitation here is almost certainly not freeze mortality. @sec-cold shows observed winter minima across the region fall roughly `r sprintf("%.0f", abs(-40 - coldest$min_temp_c_usable))` degrees short of the temperatures conventionally cited as lethal for this insect, at every elevation for which there is a record. If larvae are not killed outright, the mechanism that remains is the degree-day accumulation needed to complete a generation in one year.

The second is that aspect gives purchase on exactly that. Because heat load is a trigonometric function of the elevation model rather than a field draped over it, it carries thermal information independent of elevation, and @sec-aspect finds it predicts attack severity in the direction warmth-limited development implies. The thermal constraint that plausibly operates here is the one for which we have an elevation-independent proxy, and that proxy behaves as the constraint predicts.

This gain is bounded in three ways. The station record is a regional plausibility check, not a temperature surface, since no station inside the study area reported during the outbreak window. Heat load is a modelled radiation load, not a measured temperature. And neither touches the possibility that cold-air pooling produces local freeze events the station network never sampled, which is precisely the structure @sec-cold shows when a valley station at `r format(round(lowcold$elevation_m), big.mark=",")` m matches an alpine station `r format(round(coldest$elevation_m - lowcold$elevation_m), big.mark=",")` m above it.

The deeper obstacle is that the quantity governing overwinter survival is not air temperature but temperature under the bark, which is not observed at landscape scale anywhere. @goodsman2024, validating the standard Canadian Forest Service winter mortality model against observed mortality in Banff National Park, found its spatial predictions poor and attributed the failure to precisely this. The confound we report unresolved is unresolved in this terrain generally, and not an artefact of our data being thin.

There is also a class of mechanism that would produce our result without pheromone plumes at all. If exposed ground receives fewer arriving beetles, rather than the same number arriving to a disrupted signal, the observable consequence in a static model is identical: less attack on exposed convexities. We advance that as an alternative of our own, because the obvious precedent does not support it. @smithmckenna2013, working on white pine blister rust in whitebark pine treeline communities, established that topographic position governs delivery of a wind-borne propagule; but the direction they report is the opposite of the one this alternative needs, with exposed positions favouring *interception* rather than escape. Their agent is a fungus dispersed passively, so neither direction transfers cleanly to a flying insect that selects where it lands. What their result does establish is that the dispersal pathway is a live confound in terrain of this kind. Nothing in our design distinguishes it from pheromone disruption; doing so would need the dispersal process observed rather than inferred, through attack years, distance from the previous year's attack, or trapping.

The honest summary is that we have located candidate refugia and confirmed the host-density half of the hypothesis, but cannot support the wind half once elevation is fitted flexibly, and the mechanism remains confounded with at least two alternatives on the same gradient.

## Lower elevation effects {#sec-lowtrap}

The result that should travel furthest is at the bottom of the gradient. Attack there is near zero, and any refugia mapping exercise lacking a host layer would have identified the valley as a large, spatially coherent, topographically predictable refugium. It is not a refugium. It is a place with almost no pine. Conditioning on host raises attack in the lowest band from `r sprintf("%.1f", grad$"Attacked, all cells (%)"[1])` per cent to `r sprintf("%.1f", grad$"Attacked, host present (%)"[1])` per cent, which is within the range observed at mid elevations.

@krawchuk2020 anticipate this in framing, asking of any candidate refugium both what it is a refugium from and what it is a refugium for, but the framework does not force the check and the data required to perform it are not always at hand. On this landscape the difference between performing it and not is the difference between reporting a roughly `r sprintf("%.0f", sum(S$Elevation < 900) * prod(res(G)) / 1e6)` km2 valley refugium, the modelled ground below 900 m that @tbl-gradient shows carries almost no pine, and reporting none. We would put the requirement more strongly than the framework currently does: a refugium claim without an independent host or habitat layer is not interpretable.

## Limits to terrain predictors

The collinearity result in @sec-extent generalises beyond wind. A fine-resolution climate surface is produced by imposing terrain on a coarse field, so wherever the coarse field is nearly constant the product approaches a function of the terrain. That covers most gridded climate products used in landscape ecology. Three properties govern the severity, and all are checkable before any model is fitted: the ratio of analysis extent to the correlation length of the driving field, the relief of the study area, and which terrain variable does the work in the downscaling, which for wind is elevation and roughness but for radiation would be slope and aspect.

The consequence for synthesis is uncomfortable. If a confound scales with study extent, effect sizes for derived climate predictors are not comparable across studies of different extents, and small studies will systematically report the strongest apparent effects, a pattern easily mistaken for publication bias.

We would ask three things of any study entering such a surface into a disturbance model: report its regression on elevation, report the same diagnostic at more than one extent, and interpret only the orthogonalised coefficient. None requires new data, and here each changed the answer.

The plot record calibrates severity but not location, because coordinates for the 28 ground plots were not recoverable. The survey response is observer sketch-mapping with known positional and severity error, and rasterising polygons imposes a grain coarser than the 25 m cell; severity classes are assigned by eye from an aircraft, so the ordinal gradient in @sec-ordinal rests on judgement of proportion killed rather than measurement. The inventory is a modelled operational product whose species percentages are polygon-level attributes applied uniformly. The grain test in @sec-grain spans 25 m to 100 m, which brackets the resolutions in common use and the 900 m2 Landsat grain at which @krawchuk2020 note refugia have typically been described. Finally the analysis is correlational and the outbreak is one epidemic passage; frequency refugia in the sense of @krawchuk2020 cannot be addressed with a single window, and the static design cannot separate a wind effect on pheromone communication from one on dispersal.

The identification gained in @sec-aspect is partial. Heat load bears on the degree-day half of cold limitation, the half governing whether a generation completes in one year. It says almost nothing about the overwinter minimum half, the lethal events that kill larvae outright, which are driven by synoptic air mass intrusions and cold-air pooling rather than by radiation on a slope. Separating warm aspects from cool ones at the same elevation therefore addresses one thermal mechanism and leaves the other untouched, and rests on an assumed functional form, the @mccune2002 equation, rather than an independent measurement.

The shelter index is also imperfect as a test of the mechanism. It is omnidirectional because the Global Wind Atlas supplies no direction, and an omnidirectional index cannot represent a process that depends on where the wind comes from. A directional index on a known prevailing bearing would be sharper, and its disagreement with the wind residual might resolve either way under one.

# Conclusions

Beetle-caused mortality at Darkwoods is unimodal across the `r format(round(diff(range(S$Elevation))), big.mark = ",")` m of relief the modelled area spans. Below 1,900 m that pattern is the distribution of lodgepole pine, and conditioning on host removes almost all of it. Above 1,900 m it is not: attack more than halves while host cover thins only slightly, so the decline is far larger than a proportionate response to the decline in host. That is where the Krawchuk hypothesis predicts refugia should be.

We do not find a wind effect. The component of the Global Wind Atlas surface orthogonal to a linear elevation term is the strongest negative predictor of attack in host-bearing stands, and modelled as the ordinal severity class it clears the interval test at every threshold above trace. But the response is unimodal, and orthogonalising against a straight line leaves the curvature of elevation inside the wind term. Entering elevation as a quadratic or a spline, and residualising wind against that same basis, collapses the coefficient at every threshold. The apparent wind signal is a specification artefact. Even had it survived, attribution would have remained open, because cold limitation and dispersal interception both predict the same gradient.

Two further results narrow that gap without closing it. Observed winter minima from every station within 100 km fall about ten degrees short of the temperatures conventionally cited as lethal, so the cold mechanism that plausibly operates here is the accumulated warmth a generation needs, not freeze mortality. And a heat load index, a trigonometric function of the elevation model and therefore almost unrelated to elevation, predicts attack at the moderate severity threshold in the direction cold limitation implies, though not at the two lighter thresholds, so the thermal axis is doing work that is not the elevation gradient under another name. But a geometric shelter index computed from the same elevation model carries the opposite sign to the wind residual while being nearly uncorrelated with it, a warning that "wind exposure" is not one measurable thing and that our wind result lacks corroboration from the one predictor describing exposure directly.

The methodological findings are firmer, and all three are failures of design rather than data. Fitted inside a 2 by 4 km clip of the same landscape, elevation explains four fifths of the variance in the wind surface rather than two thirds, the wind coefficient's point estimate turns positive, and its interval widens until it excludes nothing; a four-fold change of grain moves none of this. Extent is the axis that governs whether the coefficient can be estimated here. We do not generalise that ordering: following @sandel2015, the likeliest reading is that truncating a gradient onto one limb of a unimodal response reverses the apparent sign of anything monotonically correlated with it, which is a statement about functional form rather than extent as such. Collapsing severity to attacked or not converts the study's strongest apparent relationship into a null. And without a host layer, the pine-poor valley floor would have been mapped as the largest refugium on the mountain.

# CRediT authorship declaration

Seamus Murphy: Conceptualization, Methodology, Software, Formal analysis, Investigation, Data curation, Writing - original draft, Writing - review & editing, Visualization.

# Competing interest declaration

The author declares no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Funding declaration

This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.

# Declaration of Generative AI

All prose, writing, results in the analysis and reporting contained within this document was derived and verified by the author. No generative tool was used to create, alter or manipulate research data or results. The author is responsible for the accuracy of all content.

# References

::: {#refs}
:::
