---
title: "Wind does not explain mountain pine beetle refugia: a specification artefact across 1,776 m of relief"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
    affiliation: "Independent Researcher, Nelson, British Columbia, Canada"
abstract: >
  Mass attack by mountain pine beetle (Dendroctonus ponderosae Hopkins) on lodgepole pine
  (Pinus contorta Douglas ex Loudon) is coordinated by an airborne pheromone plume, so refugia
  are hypothesised to form where stands are thin enough for wind to disrupt that signal, a
  mechanism proposed repeatedly and never entered into a model. We test it across 168 km2 of
  inventoried forest and 1,776 m of relief in southeastern British Columbia, combining an
  aerial overview survey of beetle-caused mortality from 1999 to 2015 with inventory host
  attributes and Global Wind Atlas speed at 10 m. Attack is unimodal, peaking
  near 1,430 m. Below 1,900 m that peak is the distribution of the host; above it, attack
  halves while cover holds, so coding host as cover alone overstates the result. Attack rises
  monotonically with stand quadratic mean diameter, from 15 to 41 per cent, stepping most
  steeply at the 25 cm source-sink threshold the species' bionomics predicts, and the share
  of stands above that threshold falls four-fold at the top of the gradient.
  Conditioning on the full host set removes about a quarter of the suppression. We find no
  wind effect.
  Orthogonalised against a linear elevation term, wind is the strongest negative predictor in
  host-bearing stands and clears the interval test at every severity threshold above trace,
  but a unimodal response leaves elevation's curvature inside that term, and refitting
  elevation as a quadratic or a spline collapses the coefficient everywhere. The signal is a
  specification artefact, and the atlas does not report the quantity the insect responds to,
  flight being arrested by instantaneous below-canopy speeds near 2 m per second. Thermal
  alternatives survive. A Landsat red-stage classification built to replace the survey fails
  against it. Elevation explains 0.63 of the wind surface's variance landscape-wide and 0.80
  within a 2 by 4 km clip.
keywords:
  - bark beetle
  - disturbance refugia
  - pheromone disruption
  - Landsat classification
  - spatial block bootstrap
  - collinearity
date: today
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: false
  comment: NA
always_allow_html: true
engine: knitr
citeproc: true
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/apa-6th-edition.csl
---

<!--
TARGET VENUE: Journal of Applied Entomology (Wiley), Original Article.
Built against 04.references/Submission Guidelines - Journal of Applied Entomology.pdf,
text layer read 2026-08-20. Rules quoted verbatim from that file:

  Word limit   "6000 limit (without references)"
  Abstract     "an abstract of no more than 300 words containing the major keywords"
  Keywords     "Please provide six keywords. These keywords should not include words
               in the title."
  Structure    "Introduction, Materials and Methods, Results, Discussion (each
               separately). A Conclusions section is optional."
  Running head "A short running title of less than 40 characters"
  References   "prepared according to the Publication Manual of the American
               Psychological Association (6th edition)"
  Data         "A link to your data must be included in your Data Availability
               Statement." Data on request or as supplementary files is NOT accepted.
               "Please insert data link with a DOI in the manuscript above references."
  COI          "Authors are required to provide a Conflict of Interest statement."
  Footnotes    "Footnotes to the text are not allowed."
  Species      "Upon its first use in the title, abstract, and text, the common name of
               a species should be followed by the scientific name (genus, species, and
               authority with correct use of parentheses)."
  Tables       Each complete with title and footnotes; "SD or SEM should be identified
               in the headings."
  Fig legends  "Legends should be supplied as a complete list in the text. Figures
               should be uploaded as separate files."
  Files        "Manuscripts can be uploaded either as a single document (containing the
               main text, tables and figures), or with figures and tables provided as
               separate files. Should your manuscript reach revision stage, figures and
               tables must be provided as separate files."
  Charges      "There are no charges / fees for publishing pages except for printed
               colour figures." Colour is free online; print colour is 150 GBP for the
               first figure and 50 GBP each thereafter.
  AI           Use of an AI tool "must be described, transparently and in detail, in the
               Methods or Acknowledgements section."

STRUCTURAL DEVIATION, flagged deliberately: the standing project rule is that a
manuscript carries a numbered literature review as its own section before the
Introduction. JAE mandates the four-section body above and has no slot for one. The
review therefore moves to Supporting Information as Appendix S1, keeps its numbered
findings R1 to R5, and is cited by number from the Introduction, Results and
Discussion. Substance kept, placement changed.

PROVENANCE: this file was assembled once by 05.tasks/scripts/assemble-short.py, which
copied every R chunk verbatim by label from beetle-topography-wind-study.qmd so that
the two documents cannot compute different numbers from the same code. From that point
this .qmd is the source of truth and is edited directly.

Running title, for the separate title page file:
  "Wind and beetle refugia in BC"  -- 29 characters, under the 40 allowed.
-->

```{r}
#| label: setup
#| echo: true
#| output: false

## Every quantity in this manuscript is computed here at render time.
## patchwork is loaded for its `+` operator on ggplots (@fig-gradient), which is
## why no patchwork function appears by name anywhere below.
pkgs <- c("terra","knitr","kableExtra","ggplot2","dplyr","patchwork","tibble","kernlab",
          "sf","ggspatial","tidyterra","ggrepel","rnaturalearth","scales","parallel")
invisible(lapply(pkgs, function(p) library(p, character.only = TRUE)))
## dpi 300 is the Elsevier minimum for a combination line/halftone figure.
## error = FALSE on purpose: in a manuscript whose every number is computed at
## render time, a chunk that fails quietly while downstream inline expressions
## carry stale values is the one failure this design exists to prevent.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      comment = NA, dpi = 500, fig.align = "center",
                      cache.path = "_cache/", autodep = TRUE, cache.lazy = FALSE)
## Caching is opt-in per chunk and is applied only to the model-fitting chunks, which
## between them refit the block bootstrap roughly four thousand times per render.
## `autodep = TRUE` makes knitr read each chunk's globals and invalidate a cached
## chunk when any chunk that created those globals changes, so a cached result
## cannot outlive the data it was computed from. That guarantee is the only reason
## caching is acceptable in a manuscript whose whole design is that no number can go
## stale. It is not absolute: knitr tracks R objects, not files on disk. If an input
## raster or CSV changes without its reading chunk changing, delete `_cache/` and
## render again. `make clean-cache` does that, and the submission render should
## always be run from an empty cache.
knitr::dep_auto()
options(scipen = 999)
theme_set(theme_minimal(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                plot.title = element_text(face = "bold", size = 10)))

B_BOOT   <- 150L    # spatial block bootstrap replicates
BLOCK    <- 40L     # block side in cells (40 x 25.1 m ~= 1 km)
OUTBREAK <- 1999:2015
## The elevation above which conditioning on host stops removing the decline.
## It is a band edge of the 200 m banding in @tbl-gradient and is quoted in the
## prose, so it is defined once here rather than typed into each comparison.
ALPINE   <- 1900
```

```{r}
#| label: paths
#| echo: true

DATA_ROOT <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
dpath <- function(...) file.path(DATA_ROOT, ...)
here  <- function(...) file.path("..", ...)     # repo-relative, from 01.manuscript/
stopifnot(dir.exists(dpath("terrain_environment")))

## The elevation surface is the analysis grid: everything else is resampled onto
## it. Loaded here so the Study area section can quote the relief from the raster
## rather than from memory.
G <- rast(dpath("terrain_environment", "Elevation.utm.tif"))
GRID_RELIEF <- diff(as.vector(minmax(G)))
```

```{r}
#| label: map-furniture
#| echo: true

## Cartography shared by the four map figures, defined once here so each figure can
## sit in the section it documents instead of being grouped at the end of Materials
## for want of its furniture. Graticule, EPSG, north arrow, bar scale and a
## representative fraction quoted against a stated printed width are set once and
## reused, so the panels are directly comparable.

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
clip_sf  <- st_as_sf(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)))
grid_sf  <- st_as_sf(as.polygons(ext(G), crs = crs(G)))
ev <- as.vector(ext(G)); gw <- ev["xmax"] - ev["xmin"]; gh <- ev["ymax"] - ev["ymin"]
## Graticule and axis ticks are drawn in the same coordinate reference system the
## panel declares, so the CRS named beneath each title is the CRS the reader is
## reading off the frame. Ticks are metres divided by 1,000; the subtitle says so.
km_lab  <- function(x) sprintf("%.0f", x / 1000)
sub_utm <- "EPSG:32611 WGS 84 / UTM 11N; ticks in km"
sub_alb <- "EPSG:3005 NAD83 / BC Albers; ticks in km"

map_theme <- theme_bw(base_size = 8) +
  theme(panel.grid = element_line(colour = "grey70", linewidth = .25, linetype = "dotted"),
        axis.title = element_blank(), axis.text = element_text(size = 6, colour = "grey25"),
        plot.title = element_text(face = "bold", size = 8.5, margin = margin(b = 1)),
        plot.subtitle = element_text(size = 6, colour = "grey35", margin = margin(b = 3)),
        legend.key.width = unit(3.2, "mm"), legend.key.height = unit(7, "mm"),
        legend.title = element_text(size = 6.5), legend.text = element_text(size = 6))

north <- annotation_north_arrow(location = "tr", which_north = "true",
  height = unit(6, "mm"), width = unit(5, "mm"),
  pad_x = unit(1.2, "mm"), pad_y = unit(1.2, "mm"),
  style = north_arrow_orienteering(text_size = 5, line_width = .6,
    fill = c("white", "grey20"), text_col = "grey20"))

## White plates behind the north arrow and the clip label. Without them the arrow is
## unreadable over the dark end of the wind ramp and the label collides with the
## severity polygons.
plate <- annotate("rect", xmin = ev["xmin"] + .885*gw, xmax = ev["xmax"],
                  ymin = ev["ymax"] - .115*gh, ymax = ev["ymax"],
                  fill = "white", alpha = .72, colour = NA)
clip_layer <- geom_sf(data = clip_sf, fill = NA, colour = "#C0392B", linewidth = .45)
study <- list(coord_sf(crs = 32611, datum = st_crs(32611), expand = FALSE,
                       xlim = ev[c("xmin","xmax")], ylim = ev[c("ymin","ymax")]),
              scale_x_continuous(breaks = seq(520000, 530000, 5000), labels = km_lab),
              scale_y_continuous(breaks = seq(5440000, 5450000, 5000), labels = km_lab),
              plate, north, map_theme)

## Colour limits from robust quantiles. The raw topographic position surface runs
## past 1,000 m at the grid edge, where the 625 m window is truncated, and on a
## full-range ramp that artefact flattens the whole interior to one colour.
panel <- function(r, title, legend, pal = "viridis", dir = 1, q = c(.02, .98)) {
  lim <- unname(quantile(values(r, mat = FALSE), q, na.rm = TRUE))
  ggplot() + geom_spatraster(data = r) +
    scale_fill_viridis_c(option = pal, direction = dir, na.value = "grey88",
                         name = legend, limits = lim, oob = scales::squish) +
    clip_layer + study + labs(title = title, subtitle = sub_utm)
}
```

```{r}
#| label: review-screen
#| echo: true
#| output: false

## The screening table is built by 02.inputs/literature/build-review-screening.py against the
## central store, and read here. Counts in the prose below are inline expressions over this
## object, so a change to the screening cannot leave the text behind.
rev  <- read.csv(here("02.inputs/literature/review-screening.csv"), stringsAsFactors = FALSE)
inc  <- rev[rev$stage2_include == "yes", ]

N_CAND    <- nrow(rev)
N_INC     <- nrow(inc)
N_FULL    <- sum(inc$full_text_checked == "yes")
N_WIND    <- sum(inc$wind_predictor == "yes")
N_NOWIND  <- N_INC - N_WIND
WIND_WHO  <- inc$key[inc$wind_predictor == "yes"]
YR_RANGE  <- range(suppressWarnings(as.integer(inc$year)), na.rm = TRUE)

## Every included study was read in full text, because the review's central claim is an
## absence and an absence established from titles is not evidence.
stopifnot(N_FULL == N_INC)
```

```{r}
#| label: word-budget
#| echo: true

## The 6,000-word limit is the constraint this file exists to satisfy, so it is
## measured here from the file's own source, at render time, and the render stops if
## the body is over budget. That is deliberately a different measurement from the one
## the post-render stamping script makes: the script counts the rendered document,
## which includes every line of printed code, while this counts body prose only, which
## is what the journal's limit is about. Both numbers are real and they answer
## different questions, so neither replaces the other.
##
## Note for whoever edits this comment: do not write the stamping script's placeholder
## token in any chunk that echoes. The script substitutes on the joined text of any
## paragraph containing it and rewrites that paragraph as a single run, which destroys
## the line breaks of the code listing it appears in.

jae_wordcount <- function(path = "manuscript-short.qmd",
                          from = "# Introduction",
                          to   = "# Conflict of interest") {
  src <- paste(readLines(path, warn = FALSE), collapse = "\n")
  src <- sub("^---.*?\n---\n", "", src)                    # YAML
  src <- gsub("<!--.*?-->", "", src)                       # HTML comments
  src <- gsub("```\\{r[^\n]*\n.*?\n```", "", src)          # code chunks
  ## perl = TRUE on the two line-anchored patterns below is load-bearing, not
  ## decoration. R's default engine lets `.` match a newline and anchors ^ and $
  ## to the whole string, so `gsub("^#+ .*$", "", body)` matches the first
  ## heading and then eats every remaining character of the manuscript. That is
  ## how this guard reported 0 words against a 6,000 limit and passed: a guard
  ## that cannot fail is not a guard. PCRE with (?m) anchors per line and stops
  ## `.` at newlines, which is the behaviour the pattern was always assuming.
  src <- gsub("(?m)^:::.*$", "", src, perl = TRUE)         # div fences
  src <- gsub("`r [^`]*`", "NUM", src)                     # inline code -> one word
  i <- regexpr(from, src, fixed = TRUE)
  j <- regexpr(to,   src, fixed = TRUE)
  if (i < 0) stop("word-budget: start marker not found: ", from)
  body <- if (j > i) substr(src, i, j - 1) else substr(src, i, nchar(src))
  body <- gsub("(?m)^#+ .*$", "", body, perl = TRUE)       # headings are not body prose
  n <- length(strsplit(trimws(body), "\\s+")[[1]])
  ## The failure this guard exists to catch is silence, so refuse a count that
  ## is implausible on its face rather than reporting it.
  if (n < 1000) stop("word-budget: counted ", n, " words, which cannot be right; ",
                     "the stripping regexes have eaten the body again")
  n
}

WC_BODY  <- jae_wordcount()
WC_LIMIT <- 6000L
cat(sprintf("Main text (Introduction to Conclusions, references excluded): %d / %d words\n",
            WC_BODY, WC_LIMIT))
if (WC_BODY > WC_LIMIT)
  stop(sprintf("Over the JAE Original Article limit by %d words.", WC_BODY - WC_LIMIT))
```

# Introduction

Disturbance refugia, defined by @krawchuk2020 as locations "disturbed less severely or less frequently than other areas within the surrounding landscape", organise much of the current thinking about how forests carry biological legacies through disturbance. The empirical base is predominantly a wildfire literature [@krawchuk2020; @littlefield2019]; for insect outbreaks it is thinner, and @krawchuk2020 observe that few studies have explicitly identified refugia from insect outbreaks.

For mountain pine beetle (*Dendroctonus ponderosae* Hopkins), @krawchuk2020 advance three hypotheses: refugia in areas with cooler temperatures that protect trees from water stress, in areas of lower host density "allowing for greater wind disruption of beetle pheromone communication and more vigorous tree growth and chemical defenses", or in areas with fewer large-diameter host trees. All three predict reduced attack on high, cool, sparsely stocked ground, and they are separable only by the mechanism each invokes. We test the second and treat the other two as the competing explanations against which it must be judged.

A structured review of the studies that model bark beetle attack against terrain, reported in full as Appendix S1, makes the evidence countable. Of the studies retained, one enters a wind term, and it is a companion study on this same landscape whose response is conifer regeneration rather than beetle attack (R1). The closest existing test, @cartwright2018, associated refugia with shaded slopes, convergent topographic positions and thinner stands; thin stands are where a wind effect would be expected and also where host is scarce, so separating the two requires an independent host layer (R2). No retained study codes host as the attribute the life history says governs brood production (R3), and none refits one model across extent, grain and response coding to report whether its conclusion survives (R4).

Interpreting any spatial pattern of attack requires the population biology that generates it. Colonisation proceeds by mass attack, a density phenomenon rather than a behaviour. A lodgepole pine (*Pinus contorta* Douglas ex Loudon) resists a single beetle through constitutive resin flow and an induced response that saturates the phloem with terpenes, so overwhelming that defence requires the near-simultaneous arrival of many individuals [@safranyik2006chap1; @clark2010]. Arrival is coordinated chemically: pioneer females oxidise the host monoterpene alpha-pinene to *trans*-verbenol, arriving males release *exo*-brevicomin, and host kairomones synergise both, so colonisation is normally complete within one to two days, until at approximately 60 attacks per square metre of bark *exo*-brevicomin, frontalin and the anti-aggregation pheromone verbenone reverse the signal and redirect arrivals to neighbouring trees [@safranyik2006chap1]. Because that signal is airborne it is in principle susceptible to wind, and that susceptibility is the premise under test.

Host quality sets a second constraint, acting on stem size rather than host abundance. Bark thickness, bark surface area, phloem thickness and sapwood moisture retention all increase with tree diameter and together determine how many beetles a colonised tree produces. @carroll2004bionomics summarise the consequence as a threshold: lodgepole pine of 25 cm diameter or less are on average beetle sinks, in that more beetles attack than emerge, whereas trees above 25 cm are sources. The threshold is stated for individual trees; we apply it to a stand attribute and treat it as an approximate ordering of stand suitability.

Temperature imposes two further constraints, acting on different parts of the life cycle. Development is bounded by heat accumulation, and where accumulation is insufficient, as at high elevation, populations become semivoltine, overwintering twice in cold-susceptible stages and losing the emergence synchrony on which mass attack depends [@safranyik2006chap1; @powell2014]. Overwinter survival is bounded by cold tolerance, which is acquired rather than fixed: glycerol accumulates through autumn, so hardiness peaks from December to February and is incomplete on either side, and tolerance is stage-specific, with lethal ranges of -23 to -29 °C for first instars, -23 to -34 °C for second instars and -29 to -40 °C for third and fourth [@safranyik2006chap1]. The conventionally quoted -40 °C is the cold end of the range for the late instars that normally overwinter, not a threshold for the population, and the same source is explicit that cold arriving before hardening completes does far more damage than the same temperature in midwinter.

Dispersal is bounded thermally before it is bounded by wind. @carroll2004bionomics give a flight envelope of 19 to 41 °C after @mccambridge1971, most beetles flying between 22 and 32 °C, so a cold slope suppresses arrival without any appeal to a pheromone plume. Dispersal is also density-dependent, switching between short-range movement below the canopy and long-range transport above it as population density rises [@powell2014; @johnson2026density], so the spatial signature of attack is itself a function of outbreak phase.

The wind ceiling itself is an inference across species rather than a measurement on this one. @carroll2004bionomics put the maximum wind speed for flight in a large-bodied bark beetle at approximately 2 m per second, citing work on *Dendroctonus pseudotsugae* Hopkins, and the comparative evidence assembled by @jones2019 is consistent but no firmer: in the one wind-tunnel trial on *D. ponderosae* the beetles flew longer at 0, 0.5 and 1 m per second than at 2, which is reduced flight duration rather than arrest.

Testing the hypothesis requires a spatial wind surface, and the available surfaces bound what can be inferred (R5). Gridded products, including the Global Wind Atlas, are produced by driving a mesoscale model from reanalysis, generalising the output to remove local terrain effects, then downscaling it over a digital elevation model and a roughness map [@badger2014; @davis2023]. Elevation is an input to the product rather than a covariate measured independently of it, and the dependence is most severe at small extents, where the driving field is nearly flat. Nor does the atlas report the quantity at issue, an instantaneous below-canopy speed during the July and August flight period, rather than an annual mean at 10 m over open ground. A third confound is specific to refugia work: a cell recorded as unattacked may simply contain no host, and @hadley1994 predicted heavier mortality on hot, dry, south-facing slopes and observed the reverse, because host distribution rather than moisture stress governed where attack succeeded.

We therefore address four questions on a single landscape spanning `r format(round(GRID_RELIEF), big.mark = ",")` m of relief. From R2, does terrain condition beetle-caused mortality once host abundance is accounted for, and where does host distribution cease to explain the pattern? From R3, does that result survive coding host as stem diameter rather than as cover? From R1 and R5, does the component of the wind surface independent of elevation predict reduced attack where host is present? And from R4, how far does that answer depend on decisions the data do not fix: the extent of the study area, the grain of the cells, and whether damage is coded as binary or as the ordinal severity class the survey records?

# Materials and methods

## Study area

The Darkwoods Conservation Area lies in the South Selkirk Mountains of southeastern British Columbia. The terrain grid spans 16 by 17.5 km and `r format(round(GRID_RELIEF), big.mark = ",")` m of relief, from valley bottom to alpine; the modelled area is a subset, because cells carrying no inventory basal area cannot enter a model conditioning on stand structure, and @tbl-relief reports its extent. The range crosses the full biogeoclimatic sequence of the southern interior wet belt, from the dry warm Interior Cedar-Hemlock subzones to the Engelmann Spruce-Subalpine Fir zone and its parkland variant [@ketcheson1991; @braumandl2005], and lodgepole pine occupies the middle of that sequence. Terrain surfaces and the site description are inherited from a companion analysis on the same ground [@murphy2026], in which disturbance was a predictor; here it is the response.

## Beetle-caused mortality

```{r}
#| label: read-response
#| echo: true

aos_box <- ext(project(as.polygons(ext(G), crs = crs(G)), "EPSG:3005"))
aos <- vect(here("02.inputs/aos/pest_infestation_poly.gdb"),
            layer = "pest_infestation_poly", extent = aos_box)
mpb <- project(aos[aos$PEST_SPECIES_CODE == "IBM", ], crs(G))   # IBM = mountain pine beetle
m   <- mpb[mpb$CAPTURE_YEAR %in% OUTBREAK, ]

## BC severity classes are ordinal: trace, light, moderate, severe, very severe
sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)
m$sev <- unname(sev_rank[m$PEST_SEVERITY_CODE]); m$sev[is.na(m$sev)] <- 0

severity  <- rasterize(m, G, field = "sev", fun = "max", background = 0)

c(polygons = nrow(m), years = length(unique(m$CAPTURE_YEAR)),
  hectares = round(sum(m$AREA_HA)))
```

The response is the British Columbia Aerial Overview Survey, the operational provincial record of forest health, in which observers sketch-map damage annually from aircraft and label each polygon by year, causal agent, severity class and host species. We take polygons coded `IBM`, mountain pine beetle, for the outbreak window 1999 to 2015, and rasterise the maximum severity class reached by each cell.

The survey was preferred to a spectral classification for a specific reason. Red attack has been mapped from Landsat with improving accuracy [@white2005; @meddens2013; @ye2021], but every such product is a model whose errors are spatially structured and not independent of terrain, because illumination, shadow and view angle vary with slope and aspect. For a question about terrain, a response drawn from a terrain-sensitive classifier risks recovering the classifier's own terrain sensitivity. Sketch-mapping has cruder errors, but they are observer errors rather than errors generated by the predictors under test. Twenty-eight field plots established in 2020 recorded the basal area of pine killed, spanning 0.6 to 47.4 m2/ha, and anchor the severity scale [@murphy2026]; their coordinates were not recoverable, so no validation of polygon placement is claimed.

```{r}
#| label: fig-site
#| echo: true
#| fig-width: 7.1
#| fig-height: 6.8
#| fig-cap: "The study area and the three spatial inputs. (a) Location within British Columbia; the marker is a 22 km symbol, not the footprint. (b) The elevation surface, which is the analysis grid. (c) Global Wind Atlas mean speed at 10 m; the shaded southern band lies outside the wind product's extent. (d) Beetle-caused mortality from the Aerial Overview Survey, 1999 to 2015, as the maximum severity class in any year. The outlined rectangle in (b) to (d) is the 2 by 4 km clip of @sec-extent. Each panel's graticule, axis ticks and stated coordinate reference system are the same; the system is named beneath the panel title and ticks are in kilometres."

## The wind raster begins 715 m north of the grid's southern edge, so resampling onto G
## turns that strip into NA cells and na.value draws it grey. Left transparent it reads
## as a rendering fault rather than as missing data.
Wind_r <- resample(rast(dpath("terrain_environment", "Wind.utm.tif")), G)

prov <- st_transform(ne_states(country = c("Canada","United States of America"),
                               returnclass = "sf"), 3005)
site <- st_transform(st_centroid(grid_sf), 3005)
lab  <- transform(as.data.frame(st_coordinates(site)), lab = "Darkwoods\nstudy area")
## BC is framed to the analysis grid's aspect ratio so the two rows of the figure are
## the same height and no band of white opens between them.
b0 <- st_bbox(prov[prov$name == "British Columbia", ]); tgt <- unname(gw/gh)
w0 <- unname(b0["xmax"]-b0["xmin"]); h0 <- unname(b0["ymax"]-b0["ymin"])
cx <- unname(b0["xmax"]+b0["xmin"])/2; cy <- unname(b0["ymax"]+b0["ymin"])/2
if (w0/h0 > tgt) h0 <- w0/tgt else w0 <- h0*tgt
pd <- .04*w0
bb <- c(xmin = cx-w0/2-pd, xmax = cx+w0/2+pd, ymin = cy-h0/2-pd, ymax = cy+h0/2+pd)

p_a <- ggplot() +
  geom_sf(data = prov[prov$name != "British Columbia", ], fill = "grey93",
          colour = "grey65", linewidth = .2) +
  geom_sf(data = prov[prov$name == "British Columbia", ], fill = "grey82",
          colour = "grey30", linewidth = .4) +
  geom_sf(data = st_buffer(site, 22000), fill = "#C0392B", colour = "grey15",
          linewidth = .3) +
  geom_label_repel(data = lab, aes(X, Y, label = lab), size = 2.1, lineheight = .95,
                   colour = "grey10", nudge_x = -330000, nudge_y = 300000,
                   segment.colour = "grey25", segment.size = .3, min.segment.length = 0,
                   box.padding = .4, seed = 1, fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm")) +
  annotate("text", x = bb["xmin"] + .34*(bb["xmax"]-bb["xmin"]),
           y = bb["ymin"] + .70*(bb["ymax"]-bb["ymin"]), label = "British\nColumbia",
           size = 2.3, colour = "grey30", fontface = "italic", lineheight = .95) +
  coord_sf(crs = 3005, xlim = bb[c("xmin","xmax")], ylim = bb[c("ymin","ymax")],
           datum = st_crs(3005), expand = FALSE) +
  scale_x_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  scale_y_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  north + map_theme +
  labs(title = "(a) Location", subtitle = sub_alb)

p_b <- ggplot() + geom_spatraster(data = G) +
  scale_fill_hypso_c(palette = "dem_screen", na.value = "transparent",
                     name = "Elevation\n(m)") +
  clip_layer + study + labs(title = "(b) Elevation, the analysis grid", subtitle = sub_utm)

p_c <- ggplot() + geom_spatraster(data = Wind_r) +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "grey88",
                       name = "Wind speed\n(m s⁻¹)") +
  clip_layer + study +
  labs(title = "(c) Global Wind Atlas mean speed at 10 m", subtitle = sub_utm)

sev_f <- as.factor(severity)
levels(sev_f) <- data.frame(ID = 0:5, class = c("none","trace","light","moderate",
                                                "severe","very severe"))
p_d <- ggplot() + geom_spatraster(data = sev_f) +
  scale_fill_manual(values = c(none = "grey92", trace = "#FEE0B6", light = "#FDB863",
                    moderate = "#E08214", severe = "#B35806", `very severe` = "#7F3B08"),
                    na.value = "transparent", name = "Severity\nclass",
                    na.translate = FALSE) +
  clip_layer + study +
  labs(title = "(d) Beetle-caused mortality, 1999 to 2015", subtitle = sub_utm)

fig_site <- (p_a | p_b) / (p_c | p_d)
fig_site
ggsave("../03.outputs/PNG/fig-site-maps.png", fig_site,
       width = 180, height = 172, units = "mm", dpi = 500, bg = "white")
```

## Red-stage classification

The survey resolves damage at the scale an observer can sketch from an aircraft, and the refugia question is about wind and terrain acting at tens of metres, so we classified red-stage attack annually from Landsat across the whole grid to see whether a finer response layer would support the test the polygons cannot. The result is reported here rather than in Results because it decides which response layer the paper uses. Every stage runs in Google Earth Engine through `rgee`, gated on the exported rasters.

```{r}
#| label: ee-pipeline-setup
#| echo: true

RED_DIR   <- here("02.inputs/beetle-classification/red-stage")
BASE_YEAR <- 2003          # differencing baseline, see below
CLASS_YRS <- 2005:2014
dec_files <- file.path(RED_DIR, sprintf("dec_%d.tif", CLASS_YRS))

## Sensor supplying each year, taken from Landsat scene metadata rather than
## from the parent study's Table 2, which lists Landsat 5 for a 2013 acquisition.
## Landsat 5 stopped acquiring in November 2011 and Landsat 8 did not begin until
## 2013, so 2012 is the one year only Landsat 7 covers. The parent study omitted
## 2012 for that reason and not for cloud: six Landsat 7 scenes met its own
## selection rule of under 20 per cent cloud between 1 June and 30 August, the
## cleanest at 1 per cent. Landsat 7 has flown scan-line-corrector-off since May
## 2003, so that year alone carries striping. Inventory in
## 02.inputs/beetle-classification/landsat-scene-inventory.csv.
SENSOR <- c(`2003`="LT05",`2005`="LT05",`2006`="LT05",`2007`="LT05",`2008`="LT05",
            `2009`="LT05",`2010`="LT05",`2011`="LT05",`2012`="LE07",
            `2013`="LC08",`2014`="LC08")
COLL   <- c(LT05="LANDSAT/LT05/C02/T1_L2", LE07="LANDSAT/LE07/C02/T1_L2",
            LC08="LANDSAT/LC08/C02/T1_L2")
SB     <- c("BLUE","GREEN","RED","NIR","SWIR1","SWIR2")
BANDS  <- c(SB, "NDMI", "dNDMI")
```

```{r}
#| label: ee-pipeline-functions
#| echo: true

## Stage 1, imagery. Collection 2 Level-2 surface reflectance, scaled by the
## published factors, with cloud, cloud shadow, cirrus and snow masked from
## QA_PIXEL bits 1 to 5. Thematic Mapper and Enhanced Thematic Mapper carry the
## reflective bands as SR_B1-5 and SR_B7; the Operational Land Imager shifts them
## to SR_B2-7 because it adds a coastal aerosol band. Both are renamed to common
## names so nothing downstream has to know which sensor it is reading.
ee_prep <- function(sensor) {
  bn <- if (sensor == "LC08") paste0("SR_B", 2:7) else paste0("SR_B", c(1:5, 7))
  rgee::ee_utils_pyfunc(function(im) {
    qa   <- im$select("QA_PIXEL")
    keep <- qa$bitwiseAnd(2L)$eq(0)$And(qa$bitwiseAnd(4L)$eq(0))$
      And(qa$bitwiseAnd(8L)$eq(0))$And(qa$bitwiseAnd(16L)$eq(0))$
      And(qa$bitwiseAnd(32L)$eq(0))
    im$select(bn, SB)$multiply(0.0000275)$add(-0.2)$updateMask(keep)$
      copyProperties(im, list("SUN_ELEVATION","SUN_AZIMUTH","system:time_start"))
  })
}

ee_year <- function(y, AOI) {
  s <- SENSOR[[as.character(y)]]
  ee$ImageCollection(COLL[[s]])$filterBounds(AOI)$
    filterDate(sprintf("%d-06-01", y), sprintf("%d-08-31", y))$
    filter(ee$Filter$lt("CLOUD_COVER", 60))$map(ee_prep(s))
}

## Stage 2, terrain correction. Mandatory here and not at the 7.78 km2 extent the
## parent study worked at. Across 1,776 m of relief, illumination varies enough
## with slope and aspect to put a terrain signal into reflectance, and this study
## tests terrain effects, so an uncorrected artefact would be indistinguishable
## from the result. SCS+C after Soenen et al. 2005: the C term is the ratio of
## intercept to slope in a regression of reflectance on the illumination term,
## fitted per band per scene. Where a scene is wholly masked over the grid the
## regression returns null; C is then set very large, which drives the correction
## factor to one, so a scene with nothing to fit is passed through unchanged
## rather than destroyed.
ee_illum <- function(im, slope, asp) {
  sz <- ee$Image$constant(90)$subtract(ee$Number(im$get("SUN_ELEVATION")))$multiply(pi/180)
  sa <- ee$Image$constant(ee$Number(im$get("SUN_AZIMUTH")))$multiply(pi/180)
  slope$cos()$multiply(sz$cos())$
    add(slope$sin()$multiply(sz$sin())$multiply(sa$subtract(asp)$cos()))$rename("cos_i")
}

ee_scs_c <- function(im, slope, asp, AOI) {
  ci <- ee_illum(im, slope, asp)
  sz <- ee$Image$constant(90)$subtract(ee$Number(im$get("SUN_ELEVATION")))$multiply(pi/180)
  n0 <- slope$cos()$multiply(sz$cos())
  one <- function(b) {
    f  <- ci$addBands(im$select(b))$reduceRegion(ee$Reducer$linearFit(), AOI, 90,
                                                 maxPixels = 1e9, bestEffort = TRUE)
    sc <- ee$Number(ee$Algorithms$If(f$get("scale"),  f$get("scale"),  0))
    of <- ee$Number(ee$Algorithms$If(f$get("offset"), f$get("offset"), 0))
    C  <- ee$Number(ee$Algorithms$If(sc$abs()$gt(1e-8), of$divide(sc), 1e6))$max(0)$min(1e6)
    im$select(b)$multiply(n0$add(C))$divide(ci$add(C))$rename(b)
  }
  bs <- one(SB[1]); for (b in SB[-1]) bs <- bs$addBands(one(b))
  bs$copyProperties(im, list("SUN_ELEVATION","SUN_AZIMUTH","system:time_start"))
}

## Stage 3, index and differencing. NDMI is (NIR - SWIR1) / (NIR + SWIR1) and each
## annual composite is differenced against the pre-outbreak baseline. The baseline
## is 2003, not the 2005 the parent study's text names twice. Its own Table 2
## labels 2003 pre-outbreak and 2005 early-outbreak, and its 2005 raster carries
## 534 cells coded red attack, 9.73 per cent of the cells it maps. A scene
## differenced against itself is zero everywhere, so 2005 cannot be its own
## baseline. See 02.inputs/beetle-classification/02-parent-raster-audit.R.
ee_stack <- function(y, BASE, slope, asp, AOI) {
  m  <- ee_year(y, AOI)$map(rgee::ee_utils_pyfunc(
          function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()
  nd <- m$normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")
  m$addBands(nd)$addBands(nd$subtract(BASE)$rename("dNDMI"))
}
```

The classifier is trained on the aerial overview survey itself, because an earlier 310-row spectral sample derives from a single uncorrected 2020 Landsat 8 scene and would be mapping attack from 2005 to 2014, eight of those years from a different satellite. Training on the survey keeps labels and imagery in the same year and the same radiometry.

```{r}
#| label: ee-classify
#| echo: true

SAMP_FILE <- file.path(RED_DIR, "training-sample.csv")

## Stage 4, labels. Positives are cells the survey mapped at moderate severity or
## worse in that year, negatives cells it mapped as undamaged, 250 of each per
## year with a quarter held out. 2014 takes the pooled model because the survey
## records no cell above trace over this grid that year.
##
## The sample is written to disk and the classifier is then trained from the
## table rather than from the sampling graph. That is not tidiness. An Earth
## Engine classifier carries its training computation into every image it is
## later applied to, so classifying an annual composite from a classifier trained
## directly on nine years of sampled composites re-runs all of it and the
## download fails with a 400. Trained from stored numbers, the same download
## succeeds.
if (!file.exists(SAMP_FILE)) {
  library(rgee); ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  dir.create(RED_DIR, recursive = TRUE, showWarnings = FALSE)
  AOI   <- ee$Geometry$Rectangle(c(-116.7861, 49.0884, -116.5649, 49.2460))
  DEM   <- ee$Image("USGS/SRTMGL1_003")
  slope <- ee$Terrain$slope(DEM)$multiply(pi/180)
  asp   <- ee$Terrain$aspect(DEM)$multiply(pi/180)
  BASE  <- ee_year(BASE_YEAR, AOI)$map(rgee::ee_utils_pyfunc(
             function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()$
             normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")

  set.seed(42); parts <- list()
  for (y in 2005:2013) {
    ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev",
                    fun = "max", background = 0)
    if (global(ry >= 3, "sum", na.rm = TRUE)[[1]] < 30) next
    pick <- function(mk, lab) {
      p <- spatSample(mk, 250, "random", na.rm = TRUE, xy = TRUE, values = FALSE)
      data.frame(x = p[,1], y = p[,2], lab = lab,
                 elev = extract(G, p[,1:2])[,1], yr = y)
    }
    d <- rbind(pick(ifel(ry >= 3, 1, NA), 1L), pick(ifel(ry == 0, 1, NA), 0L))
    d <- d[!is.na(d$elev), ]
    d$split <- as.integer(runif(nrow(d)) < 0.25)
    pts <- sf::st_transform(sf::st_as_sf(d, coords = c("x","y"), crs = crs(G)), 4326)
    parts[[length(parts) + 1]] <- ee_stack(y, BASE, slope, asp, AOI)$sampleRegions(
      collection = sf_as_ee(pts), properties = list("lab","elev","yr","split"),
      scale = 30, tileScale = 4, geometries = TRUE)
  }
  got <- ee_as_sf(ee$FeatureCollection(parts)$flatten(), quiet = TRUE)
  xy  <- sf::st_coordinates(got)
  write.csv(cbind(sf::st_drop_geometry(got), lon = xy[,1], lat = xy[,2]),
            SAMP_FILE, row.names = FALSE)
}
samp <- read.csv(SAMP_FILE)
stopifnot(all(c(BANDS, "lab", "elev", "yr", "split", "lon", "lat") %in% names(samp)))

## Stage 4b, the classifier. Fitted here rather than by ee$Classifier$libsvm,
## for a mechanical reason. An Earth Engine classifier carries its training
## features into every request that uses it, so each annual download embedded the
## whole labelled table and returned HTTP 400; the same eight-band composite
## downloads without complaint on its own. A linear kernel is a plane, so the
## fitted model transfers exactly as a band expression and Earth Engine is left
## doing arithmetic. The check below is that transfer, and it holds to machine
## precision. Predictors are standardised on the training rows only.
SVM_TR <- samp[samp$split == 0, ]; SVM_TE <- samp[samp$split == 1, ]
SVM_MU <- sapply(SVM_TR[BANDS], mean); SVM_SD <- sapply(SVM_TR[BANDS], sd)
zscore <- function(d) as.matrix(sweep(sweep(d[BANDS], 2, SVM_MU, "-"), 2, SVM_SD, "/"))
set.seed(1)
SVM_FIT <- kernlab::ksvm(zscore(SVM_TR), factor(SVM_TR$lab), type = "C-svc",
                         kernel = "vanilladot", C = 1, scaled = FALSE)
SVM_W <- colSums(kernlab::coef(SVM_FIT)[[1]] * kernlab::xmatrix(SVM_FIT)[[1]])
SVM_B <- -SVM_FIT@b
stopifnot(identical(
  as.integer(as.vector(zscore(SVM_TE) %*% SVM_W + SVM_B) > 0),
  as.integer(as.character(predict(SVM_FIT, zscore(SVM_TE))))))

## Collapsed into raw reflectance units, which is the form Earth Engine applies.
SVM_A <- SVM_W / SVM_SD
SVM_C <- SVM_B - sum(SVM_W * SVM_MU / SVM_SD)
stopifnot(max(abs((as.matrix(SVM_TE[BANDS]) %*% SVM_A + SVM_C) -
                  (zscore(SVM_TE) %*% SVM_W + SVM_B))) < 1e-8)

## Stage 5, validation. Computed here from the held-out quarter, so the agreement
## table below and the exported rasters come from one fitted object.
SVM_TE$pred <- as.integer(as.vector(as.matrix(SVM_TE[BANDS]) %*% SVM_A + SVM_C) > 0)

if (!all(file.exists(dec_files))) {
  library(rgee); ee_Initialize(project = "murphys-deforisk", quiet = TRUE)
  AOI   <- ee$Geometry$Rectangle(c(-116.7861, 49.0884, -116.5649, 49.2460))
  DEM   <- ee$Image("USGS/SRTMGL1_003")
  slope <- ee$Terrain$slope(DEM)$multiply(pi/180)
  asp   <- ee$Terrain$aspect(DEM)$multiply(pi/180)
  BASE  <- ee_year(BASE_YEAR, AOI)$map(rgee::ee_utils_pyfunc(
             function(im) ee$Image(ee_scs_c(im, slope, asp, AOI))))$median()$
             normalizedDifference(c("NIR","SWIR1"))$rename("NDMI")

  ## Stage 6, export. The decision plane is applied band by band, thresholded at
  ## zero, classified at the 30 m native grid, then resampled onto the analysis
  ## grid by nearest neighbour, which is the correct rule for a class raster and
  ## makes the alignment exact rather than nearly so.
  gbox <- ee$Geometry$Rectangle(c(xmin(G), ymin(G), xmax(G), ymax(G)),
                                proj = "EPSG:32611", geodesic = FALSE)
  for (y in CLASS_YRS) {
    st <- ee_stack(y, BASE, slope, asp, AOI)
    d  <- ee$Image$constant(SVM_C)
    for (bn in BANDS) d <- d$add(st$select(bn)$multiply(unname(SVM_A[bn])))
    ## The continuous decision value is exported, not a thresholded map, so the
    ## threshold stays a visible choice made below rather than one baked into a
    ## download. Scaled by 1,000 and offset by 10,000 because a GeoTIFF written
    ## from an integer image takes zero as its nodata value: exporting a signed
    ## surface directly would silently delete every cell sitting on the decision
    ## boundary, and exporting a 0/1 map deleted every unattacked cell outright
    ## on the first attempt, returning 100 per cent attacked in all ten years.
    u <- d$multiply(1000)$add(10000)$round()$toInt16()$rename("dec")$
      getDownloadURL(list(name = sprintf("dec_%d", y), crs = "EPSG:32611",
                          scale = 30, region = gbox, format = "GEO_TIFF"))
    tmp <- tempfile(fileext = ".tif"); download.file(u, tmp, mode = "wb", quiet = TRUE)
    ## Kept in the scaled integer form Earth Engine sent, deflated, because the
    ## same surface as 32-bit float is 2.2 MB a year and this is 0.2 MB. The
    ## offset is removed at read time below, not here.
    writeRaster(resample(rast(tmp), G, method = "bilinear"),
                file.path(RED_DIR, sprintf("dec_%d.tif", y)),
                overwrite = TRUE, datatype = "INT2S",
                gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2"))
  }
}

dec <- (rast(dec_files) - 10000) / 1000     # undo the export scaling and offset
names(dec) <- CLASS_YRS

## The model's own threshold is zero, which is what a balanced training sample
## implies. It is the wrong prior for this landscape. Training drew 250 attacked
## and 250 unattacked cells per year, so the implied prevalence is one half,
## while the survey maps attack on a few per cent of the grid. Thresholded at
## zero the classification calls a third to two thirds of the landscape attacked.
## Both maps are therefore carried: the model's own, and one whose threshold is
## calibrated per year so the classified area matches the area the survey mapped.
SURV_AREA <- sapply(CLASS_YRS, function(y) {
  ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev", fun = "max", background = 0)
  global(ry >= 3, "mean", na.rm = TRUE)[[1]]
})
names(SURV_AREA) <- CLASS_YRS
RAW_AREA <- sapply(CLASS_YRS, function(y)
  global(dec[[as.character(y)]] > 0, "mean", na.rm = TRUE)[[1]])
names(RAW_AREA) <- CLASS_YRS

red <- rast(lapply(CLASS_YRS, function(y) {
  d <- dec[[as.character(y)]]
  thr <- if (SURV_AREA[[as.character(y)]] > 0)
    global(d, function(v, ...) quantile(v, 1 - SURV_AREA[[as.character(y)]], na.rm = TRUE))[[1]]
    else Inf
  d > thr
}))
names(red) <- CLASS_YRS
## A classification that is entirely one class, or entirely masked, is a
## pipeline failure rather than a result, and must not reach the prose.
stopifnot(nlyr(red) == length(CLASS_YRS),
          all(dim(red)[1:2] == dim(G)[1:2]),
          isTRUE(all.equal(as.vector(ext(red)), as.vector(ext(G)), tolerance = 1e-6)),
          all(RAW_AREA > 0.001), all(RAW_AREA < 0.999),
          all(global(!is.na(red), "mean")[[1]] > 0.5))
```

```{r}
#| label: tbl-redstage
#| echo: true
#| tbl-cap: "Agreement between the Landsat red-stage classification and the aerial overview survey, by 300 m elevation band. The held-out quarter of the labelled sample, pooled over 2005 to 2013. This is agreement with the survey, not accuracy against ground truth: the survey is also the training source, and no independent reference exists at this extent."

v <- SVM_TE
v$classification <- v$pred
v$band <- cut(v$elev, seq(500, 2400, by = 300), dig.lab = 5)

kap <- function(o, p) {
  cm <- table(factor(o, 0:1), factor(p, 0:1))
  oa <- sum(diag(cm)) / sum(cm)
  pe <- sum(rowSums(cm) * colSums(cm)) / sum(cm)^2
  c(OA = oa, kappa = (oa - pe) / (1 - pe))
}
POOLED <- kap(v$lab, v$classification)

RED_TBL <- do.call(rbind, lapply(levels(v$band), function(b) {
  s <- v[!is.na(v$band) & v$band == b, ]
  if (nrow(s) < 10)
    return(data.frame(Band = b, n = nrow(s), Attacked = sum(s$lab == 1),
                      Agreement = NA_real_, Kappa = NA_real_,
                      Note = "too few reference points, uncertain"))
  k <- kap(s$lab, s$classification)
  data.frame(Band = b, n = nrow(s), Attacked = sum(s$lab == 1),
             Agreement = round(k["OA"], 3), Kappa = round(k["kappa"], 3), Note = "")
}))
kable(RED_TBL, row.names = FALSE,
      col.names = c("Elevation band (m)", "n", "Attacked", "Agreement", "Kappa", ""))
```

```{r}
#| label: tbl-redstage-area
#| echo: true
#| tbl-cap: "Where the Landsat classification and the aerial overview survey disagree, by year. Survey area is the share of the grid the survey mapped at moderate severity or worse. Model area is what the classifier calls attacked at its own threshold of zero. Hit rate is the share of calibrated classified cells that fall inside a survey polygon, and lift is that rate divided by the survey's own prevalence, so 1.00 is chance."

RED_CMP <- do.call(rbind, lapply(CLASS_YRS, function(y) {
  yc <- as.character(y)
  ry <- rasterize(m[m$CAPTURE_YEAR == y, ], G, field = "sev", fun = "max", background = 0)
  S  <- ry >= 3
  p  <- SURV_AREA[[yc]]
  if (p == 0) return(data.frame(Year = y, Survey = 0, Model = round(100*RAW_AREA[[yc]], 1),
                                Hit = NA_real_, Lift = NA_real_))
  hits <- global(S & red[[yc]], "sum", na.rm = TRUE)[[1]]
  nC   <- global(red[[yc]], "sum", na.rm = TRUE)[[1]]
  data.frame(Year = y, Survey = round(100*p, 3), Model = round(100*RAW_AREA[[yc]], 1),
             Hit = round(100*hits/nC, 2), Lift = round((hits/nC)/p, 2))
}))
kable(RED_CMP, row.names = FALSE,
      col.names = c("Year", "Survey area (%)", "Model area (%)",
                    "Hit rate (%)", "Lift"))
```

The classification does not replace the survey, and the reason is not a marginal shortfall in accuracy. Pooled over the held-out quarter it agrees with the survey on `r sprintf("%.1f", 100 * POOLED["OA"])` per cent of cells, kappa `r sprintf("%.2f", POOLED["kappa"])`, but @tbl-redstage-area shows that figure is flattered by the sampling. Training took 250 attacked and 250 unattacked cells per year, an implied prevalence of one half against a true `r sprintf("%.2f", 100 * min(SURV_AREA[SURV_AREA > 0]))` to `r sprintf("%.2f", 100 * max(SURV_AREA))` per cent, so at its own threshold the classifier calls `r sprintf("%.0f", 100 * min(RAW_AREA))` to `r sprintf("%.0f", 100 * max(RAW_AREA))` per cent of the landscape attacked. Calibrating the threshold each year so that classified area equals surveyed area, it lands on `r sprintf("%.1f", min(RED_CMP$Hit, na.rm = TRUE))` to `r sprintf("%.1f", max(RED_CMP$Hit, na.rm = TRUE))` per cent of the cells the survey mapped, a lift of `r sprintf("%.2f", min(RED_CMP$Lift, na.rm = TRUE))` to `r sprintf("%.2f", max(RED_CMP$Lift, na.rm = TRUE))` against a chance value of 1.00, and `r sum(RED_CMP$Lift == 0, na.rm = TRUE)` of `r sum(!is.na(RED_CMP$Lift))` years share no cells with the survey at all. The clearest diagnostic is 2012, the only year covered by Landsat 7 alone, whose scan line corrector has been broken since May 2003: it is the year the classifier calls most attacked, at `r sprintf("%.0f", 100 * RAW_AREA[["2012"]])` per cent, and the year the survey maps least, at `r sprintf("%.2f", 100 * SURV_AREA[["2012"]])` per cent. The classifier is reading striping as change, and an instrument genuinely resolving attack the surveyor missed would still overlap the attack the surveyor found.

We therefore keep the survey as the response and report the classification as a comparison layer, which also means the timing analysis a per-cell year of first detection would allow is unavailable.

## Host attributes

```{r}
#| label: read-host
#| echo: true

vri <- project(vect(here("02.inputs/vri/vri_darkwoods.geojson")), crs(G))
vd  <- as.data.frame(vri)
num <- function(x) suppressWarnings(as.numeric(x))

## VRI records up to three species per polygon with percentage cover
accumulate <- function(pattern) {
  out <- rep(0, nrow(vd))
  for (i in 1:3) {
    cd <- vd[[paste0("SPECIES_CD_", i)]]
    pc <- num(vd[[paste0("SPECIES_PCT_", i)]]); pc[is.na(pc)] <- 0
    out <- out + ifelse(!is.na(cd) & grepl(pattern, cd), pc, 0)
  }
  out
}
vri$PinePct <- accumulate("^PL")               # lodgepole, the primary host
vri$Basal   <- num(vd$BASAL_AREA)

## Quadratic mean diameter of stems 12.5 cm and over. Carried because stem size,
## not cover, is what decides whether a stand produces more beetles than it
## consumes: below about 25 cm a lodgepole stand is a sink, above it a source
## [@carroll2004bionomics]. Cover and basal area cannot express that.
vri$QMD <- num(vd$QUAD_DIAM_125)

host <- rast(lapply(c("PinePct","Basal","QMD"), function(f) rasterize(vri, G, field = f)))
names(host) <- c("PinePct","Basal","QMD")
c(vri_polygons = nrow(vri), with_lodgepole = sum(vri$PinePct > 0),
  with_diameter = sum(!is.na(vri$QMD)))
```

Host abundance comes from the provincial Vegetation Resources Inventory, layer `VEG_COMP_LYR_R1_POLY`. We accumulate the percentage cover of lodgepole pine across the three species slots each polygon records, and carry stand basal area and quadratic mean diameter, the latter recorded for stems 12.5 cm and over. Cover and basal area describe how much host is present; neither describes whether that host can produce beetles, which is the gap R3 identifies. Diameter is recorded for `r format(sum(!is.na(vri$QMD)), big.mark=",")` of the `r format(nrow(vri), big.mark=",")` polygons, covering the analysis frame almost exactly.

The inventory's vintage matters. These polygons carry an interpretation date of August 2014, so composition comes from photography flown in the fourteenth year of a seventeen-year outbreak. A stand whose pine was killed early was interpreted after the fact, so host is under-recorded exactly where attack was heaviest and earliest, which is a candidate explanation for the reversal of the lodgepole coefficient at the severe threshold reported below. It does not touch the presence-absence distinction the layer is mainly used for.

```{r}
#| label: fig-host
#| echo: true
#| fig-width: 7.1
#| fig-height: 3.8
#| fig-cap: "The forest inventory host layers, rasterised onto the analysis grid. (a) Lodgepole pine cover, accumulated across the three species slots each polygon records. (b) Stand basal area, which defines the analysis frame: cells with no recorded basal area are dropped. Unshaded ground is unmapped or non-productive. Projection and scale as in @fig-site."

fig_host <-
  (ggplot() + geom_spatraster(data = host$PinePct) +
     scale_fill_viridis_c(option = "viridis", direction = -1, na.value = "grey88",
                          name = "Lodgepole\ncover (%)") +
     clip_layer + study + labs(title = "(a) Lodgepole pine cover", subtitle = sub_utm)) |
  (ggplot() + geom_spatraster(data = host$Basal) +
     scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "grey88",
                          name = "Basal area\n(m² ha⁻¹)") +
     clip_layer + study + labs(title = "(b) Stand basal area", subtitle = sub_utm))
fig_host
ggsave("../03.outputs/PNG/fig-host-maps.png", fig_host,
       width = 180, height = 96, units = "mm", dpi = 500, bg = "white")
```

## Terrain and wind

```{r}
#| label: read-terrain
#| echo: true

to_grid <- function(f) {
  r <- rast(dpath("terrain_environment", f))
  b <- project(as.polygons(ext(G), crs = crs(G)), crs(r))
  project(crop(r, b, snap = "out"), G, method = "bilinear")
}
terr <- rast(list(Elevation = G,               Slope = to_grid("Slope.utm.tif"),
                  TWI       = to_grid("TWI.utm.tif"),
                  RIX       = to_grid("Rutm.tif"),
                  Wind      = to_grid("Wind.utm.tif")))
names(terr) <- c("Elevation","Slope","TWI","RIX","Wind")

## Topographic position: mean deviation from a surrounding annulus.
tpi <- function(r, w) {
  k <- matrix(1, w, w); k[ceiling(w/2), ceiling(w/2)] <- 0
  r - focal(r, k / sum(k), fun = sum, na.policy = "omit", na.rm = TRUE)
}
terr$TPI600 <- tpi(G, 25)     # ~625 m neighbourhood
```

```{r}
#| label: aspect-terms
#| echo: true

## Two surfaces built from aspect, which the inherited set lacks entirely.
## Both are computed directly from the elevation model rather than draped over
## it, which is why @tbl-orthogonality can show them independent of elevation.
d2r <- pi / 180
asp  <- rast(dpath("terrain_environment", "Aspect.utm.tif"))
slp  <- rast(dpath("terrain_environment", "Slope.utm.tif"))
stopifnot(identical(dim(asp), dim(G)), identical(as.vector(ext(asp)), as.vector(ext(G))))

## Heat load after McCune and Keon, aspect folded about the SW-NE axis so that
## southwest is hottest. Flat cells carry a -1 sentinel; the sin(slope) factor
## nulls the aspect term there, so the substitution below is inert.
asp[asp < 0] <- 225
fold <- abs(180 - abs(asp - 225))
LAT  <- mean(crds(project(as.polygons(ext(G), crs = crs(G)), "EPSG:4326"))[, 2])
terr$HLI <- 0.339 + 0.808 * cos(LAT*d2r) * cos(slp*d2r) -
            0.196 * sin(LAT*d2r) * sin(slp*d2r) -
            0.482 * cos(fold*d2r) * sin(slp*d2r)

## Winstral shelter index: the maximum upwind slope angle within SEARCH metres
## along a bearing. Positive means terrain rises upwind, so the cell is
## sheltered; negative means it is exposed. The Global Wind Atlas gives speed
## but no direction, so rather than assume a prevailing bearing we compute all
## eight and average, giving an omnidirectional shelter index.
SEARCH <- 500
shift_mat <- function(m, dr, dc) {
  nr <- nrow(m); nc <- ncol(m); out <- matrix(NA_real_, nr, nc)
  ri <- seq_len(nr); ci <- seq_len(nc); sr <- ri + dr; sc <- ci + dc
  vr <- sr >= 1 & sr <= nr; vc <- sc >= 1 & sc <= nc
  out[ri[vr], ci[vc]] <- m[sr[vr], sc[vc]]
  out
}
zm <- as.matrix(G, wide = TRUE); rstep <- res(G)[1]
sx_dir <- function(az) {
  th <- az * d2r; best <- matrix(-Inf, nrow(zm), ncol(zm))
  for (k in 1:floor(SEARCH / rstep)) {
    dc <- round(k * sin(th)); dr <- round(-k * cos(th))
    if (dc == 0 && dr == 0) next
    best <- pmax(best, atan((shift_mat(zm, dr, dc) - zm) / (k * rstep)) / d2r, na.rm = TRUE)
  }
  best[!is.finite(best)] <- NA_real_
  best
}
sx <- rast(lapply(seq(0, 315, by = 45), function(a) {
  r <- G; values(r) <- as.vector(t(sx_dir(a))); r }))
terr$SxMean <- app(sx, mean, na.rm = TRUE)
```

Terrain surfaces are inherited unchanged from @murphy2026, derived from the 1 m Natural Resources Canada High Resolution Digital Elevation Model [@nrcan2017], smoothed, depression-filled by the algorithm of @wang2006 and aggregated to 25 m, with compound indices following @beers1973 and @parker1982. We add a topographic position index at a 625 m neighbourhood, which the inherited set lacked, and two aspect terms, because that set carries nothing distinguishing a warm slope from a cold one at the same elevation and the confound this paper cannot resolve is between wind exposure and cold. Heat load follows @mccune2002, aspect folded about the southwest to northeast axis so that southwest is hottest. The shelter index follows @winstral2002, the maximum upwind slope angle within 500 m along a bearing, averaged over all eight cardinal and intercardinal bearings because the atlas supplies no direction.

The decisive property of both is how they are built. They are derived from the elevation model by trigonometry rather than draped over it by downscaling, and @tbl-orthogonality shows the consequence: their variance is essentially unrelated to elevation, where the wind product's is largely elevation. A terrain-derived predictor is not automatically an elevation transform; the distinction is in the construction. Ambient wind is Global Wind Atlas mean speed at 10 m above ground level [@globalwindatlas], built by the downscaling chain described in the Introduction.

## Climate records

Because the interpretation turns on whether it is cold enough at the top of the gradient to kill beetles, we assembled the observational record. Hourly observations from the British Columbia Wildfire Service network and daily observations from Environment and Climate Change Canada were retrieved for all stations within 100 km for the winters of the outbreak window. The script in `02.inputs/climate/` streams roughly 2.7 GB and writes the per-station-winter summary the manuscript reads; the fetch cannot sit inside a document render, but every decision about which records to believe is made below, in code.

Two screens matter. Many older Wildfire Service sensors floor at exactly -20.0 °C, so a station-winter whose minimum sits on that floor has an unknown true minimum and is excluded. Isolated spikes, colder than both neighbouring observations by more than 15 °C, are rejected as instrument error during retrieval; one such value, a -40.0 °C daily minimum on a day whose maximum was -4.5 °C, would otherwise have produced exactly the conclusion the record does not support. A second script records the date each station-winter reached its minimum, needed because lethal temperature is conditional on the state of cold-hardening. The window is December to February, which bounds what the record can support.

## Analysis frame and inference

```{r}
#| echo: true
#| output: false
#| fig-width: 7.1
#| fig-height: 4.6

fig_terrain <-
  (panel(terr$Slope,  "(a) Slope", "Degrees", "cividis") |
   panel(terr$TWI,    "(b) Topographic wetness index", "TWI", "mako", -1) |
   panel(terr$RIX,    "(c) Ruggedness", "Index", "rocket", -1)) /
  (panel(terr$TPI600, "(d) Topographic position, 625 m", "TPI (m)", "viridis") |
   panel(terr$HLI,    "(e) Heat load index", "HLI", "inferno") |
   panel(terr$SxMean, "(f) Shelter index, omnidirectional", "Sx (deg)", "cividis", -1))
fig_terrain
ggsave("../03.outputs/PNG/fig-terrain-maps.png", fig_terrain,
       width = 180, height = 118, units = "mm", dpi = 500, bg = "white")
```

```{r}
#| label: assemble
#| echo: true

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
in_clip  <- rasterize(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)),
                      G, field = 1, background = 0)

X <- as.data.frame(c(terr, host, severity, in_clip), na.rm = FALSE)
names(X)[(ncol(X) - 1):ncol(X)] <- c("Severity","InClip")
X$Attacked <- X$Severity > 0
X$HasHost  <- !is.na(X$PinePct) & X$PinePct > 0

## spatial blocks for resampling-based inference
rc <- rowColFromCell(G, seq_len(ncell(G)))
X$block <- paste(rc[,1] %/% BLOCK, rc[,2] %/% BLOCK, sep = "_")

PRED  <- c("Elevation","Slope","TWI","RIX","Wind","TPI600")
EXTRA <- c("HLI","SxMean")          # aspect-derived; enter only the extended model
MVARS <- c(PRED, "PinePct", "Basal")   # defines the analysis frame, deliberately unchanged

## Cells lacking an inventory basal area are unmapped or non-productive and cannot
## enter a model that conditions on stand structure. Dropping them is reported here
## rather than left to silent na.action inside glm.
n_before <- sum(complete.cases(X[, PRED]) & !is.na(X$PinePct))
S <- X[complete.cases(X[, MVARS]), ]
H <- S[S$HasHost, ]
c(with_terrain_and_host = n_before, dropped_no_basal_area = n_before - nrow(S),
  analysis_cells = nrow(S), km2 = round(nrow(S) * prod(res(G)) / 1e6, 1),
  attacked = sum(S$Attacked), host_cells = nrow(H))
```

```{r}
#| label: tbl-relief
#| echo: true
#| tbl-cap: "The analysis landscape, and the 2 by 4 km clip drawn from it that the companion study used. The clip samples under two thirds of the relief and under 4 per cent of the cells."

tibble(
  Extent = c("full landscape", "2 by 4 km clip"),
  Cells  = c(nrow(S), sum(S$InClip == 1)),
  `Area (km²)` = round(c(nrow(S), sum(S$InClip == 1)) * prod(res(G)) / 1e6, 1),
  `Elevation (m)` = c(sprintf("%.0f to %.0f", min(S$Elevation), max(S$Elevation)),
                      sprintf("%.0f to %.0f", min(S$Elevation[S$InClip==1]),
                                              max(S$Elevation[S$InClip==1]))),
  `Relief (m)` = round(c(diff(range(S$Elevation)), diff(range(S$Elevation[S$InClip==1])))),
  `Attacked (%)` = round(100 * c(mean(S$Attacked), mean(S$Attacked[S$InClip==1])), 1)
) |> kable(booktabs = TRUE, align = "llllll", format.args = list(big.mark = ","))
```

The response is heavily clustered, so cell counts vastly overstate independent information. All coefficients carry percentile intervals from a spatial block bootstrap: the landscape is cut into square tiles of `r BLOCK` by `r BLOCK` cells, roughly 1 km on a side, and whole tiles are drawn at random with replacement `r B_BOOT` times, so neighbouring cells travel together and are never counted as separate evidence. We report the point estimate, the block interval and the proportion of resamples reproducing its sign, and treat a term as supported only when the interval excludes zero.

```{r}
#| label: fn-boot
#| echo: true

## The bootstrap refits several hundred times on >100,000 rows, so it works on a
## model matrix built once and calls glm.fit directly. This is arithmetically
## identical to glm() with the same formula, and roughly thirty times faster.
MODEL_VARS <- c("Elevation","WindResid","PinePct","Basal","TPI600","Slope","TWI","RIX")

design_matrix <- function(z, vars = MODEL_VARS)
  cbind(`(Intercept)` = 1, as.matrix(z[, vars, drop = FALSE]))

fit_fast <- function(Xm, y) glm.fit(Xm, y, family = binomial())

## The resamples are drawn first, in one sequence from one seed, and only then
## fitted. That ordering is the whole point: the random number stream is consumed
## in exactly the order the original serial loop consumed it, so every interval in
## this paper is bit-identical to the serial implementation, and the fitting, which
## is the expensive part, can be spread across cores. Seeding inside each worker
## instead would be simpler and would change every reported interval.
##
## Measured on the analysis frame: the matrix subsetting costs about 2 per cent of
## a replicate and glm.fit about 98, so there is nothing to gain from cleverer
## indexing, and two attempts at it were slower than the original. Four cores gives
## about 1.7 times; eight is slower again on this machine, which has four
## performance cores.
BOOT_CORES <- max(1L, min(4L, parallel::detectCores(logical = FALSE) - 1L))

block_boot <- function(z, vars = MODEL_VARS, B = B_BOOT, seed = 42) {
  Xm <- design_matrix(z, vars); y <- z$y
  idx_of <- split(seq_len(nrow(z)), z$block)
  blocks <- names(idx_of)
  set.seed(seed)
  draws <- lapply(seq_len(B), function(b) sample(blocks, length(blocks), replace = TRUE))
  one <- function(k) {
    idx <- unlist(idx_of[k], use.names = FALSE)
    f <- try(fit_fast(Xm[idx, , drop = FALSE], y[idx]), silent = TRUE)
    if (inherits(f, "try-error")) rep(NA_real_, ncol(Xm)) else coef(f)
  }
  ## mclapply forks, which is unavailable on Windows; fall back rather than fail.
  res <- if (BOOT_CORES > 1L && .Platform$OS.type == "unix")
           parallel::mclapply(draws, one, mc.cores = BOOT_CORES) else lapply(draws, one)
  matrix(unlist(res), B, ncol(Xm), byrow = TRUE,
         dimnames = list(NULL, colnames(Xm)))
}

model_table <- function(z, vars = MODEL_VARS, seed = 42) {
  Xm <- design_matrix(z, vars)
  fit <- fit_fast(Xm, z$y)
  cf  <- coef(fit)
  se  <- sqrt(diag(chol2inv(fit$qr$qr[1:fit$rank, 1:fit$rank, drop = FALSE])))
  bs  <- block_boot(z, vars, seed = seed)
  tibble(term = names(cf), beta = as.numeric(cf), naive_z = as.numeric(cf / se),
         lo = apply(bs, 2, quantile, .025, na.rm = TRUE),
         hi = apply(bs, 2, quantile, .975, na.rm = TRUE),
         sign_stab = sapply(names(cf), function(n)
                       mean(sign(bs[, n]) == sign(cf[n]), na.rm = TRUE)))
}

fmt_model <- function(mt) {
  mt |> filter(term != "(Intercept)") |>
    transmute(Term = term, beta = sprintf("%+.3f", beta),
              `naive z` = sprintf("%+.1f", naive_z),
              `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
              `sign stability` = sprintf("%.2f", sign_stab),
              Supported = ifelse(lo * hi > 0, "yes", "no"))
}
```

## Orthogonalisation and response specification

Finding R5 is the reason this step exists. Because the wind surface is built over the elevation surface, its coefficient in a model containing elevation is estimated from whatever small part of the joint distribution lets the two vary separately. We therefore replace wind by the residual of wind regressed on elevation, rescaled to unit variance, keeping only the part of the wind surface a straight line in elevation cannot predict.

The substitution is sometimes presented as a remedy for collinearity, which it is not. It is a linear reparameterisation of the same model, so by the standard partialling result of Frisch, Waugh and Lovell every supported-or-not verdict is numerically identical in both parameterisations and orthogonalisation buys no precision whatever. What changes is interpretation: the wind coefficient now refers to the part of the surface elevation does not explain, which is the quantity the hypothesis is about.

@krawchuk2020 warn against coding refugia as a binary, since thresholding discards the signal distinguishing a lightly touched stand from a destroyed one. The survey carries an ordinal severity class, so we fit the binary attacked model for comparability with the refugia literature and, alongside it, a cumulative logit at each severity threshold, which asks a yes-or-no question at each cut in turn. Fitting those thresholds separately rather than imposing one slope is deliberate: proportional odds means one slope for every step of the scale, and comparing separate fits is the standard diagnostic for whether it holds.

```{r}
#| label: design
#| echo: true

standardise <- function(dd) {
  vars <- intersect(c(PRED, EXTRA, "PinePct", "Basal"), names(dd))
  z <- as.data.frame(scale(dd[, vars]))
  z$y     <- as.integer(dd$Attacked)
  z$block <- dd$block
  z$Severity <- dd$Severity          # carried for the ordinal fits in @sec-ordinal
  z$WindResid <- as.numeric(scale(residuals(lm(Wind ~ Elevation, data = z))))
  z
}
Z  <- standardise(S)     # whole landscape
ZH <- standardise(H)     # cells containing lodgepole
ZC <- standardise(S[S$InClip == 1, ])   # the 2 by 4 km clip
round(cor(Z[, c("Elevation","Wind","WindResid")]), 3)
```

```{r}
#| label: tbl-orthogonality
#| echo: true
#| tbl-cap: "How much of each candidate predictor is elevation. The wind product is a mesoscale field draped over terrain and is largely an elevation transform. The heat load and shelter indices are computed from the same elevation model by trigonometry and are almost unrelated to it."

tibble(Surface = c("Global Wind Atlas wind speed", "Heat load index (McCune and Keon)",
                   "Shelter index (Winstral, omnidirectional)", "Topographic position, 625 m"),
       `How it is built` = c("mesoscale field downscaled over the DEM",
                             "trigonometric function of DEM aspect and slope",
                             "maximum upwind slope from the DEM within 500 m",
                             "deviation from surrounding DEM mean"),
       `R² on elevation` = sapply(c("Wind","HLI","SxMean","TPI600"), function(v)
         round(summary(lm(as.formula(paste(v, "~ Elevation")), data = Z))$r.squared, 3))) |>
  kable(booktabs = TRUE, align = "lll")
```

# Results

Three results below are built and then withdrawn, which is deliberate and is signposted as each arrives: the wind coefficient is fitted, strengthened across severity thresholds, then collapsed by a specification test; the cold check is run, then bounded by the structure of the record; the clip analysis produces an estimate, then shows it cannot be estimated. The sections carrying a result we defend are the elevational distribution, the stem diameter effects and the grain comparison.

## Elevational distribution {#sec-host}

```{r}
#| label: tbl-gradient
#| echo: true
#| tbl-cap: "Attack across the elevation gradient, before and after conditioning on lodgepole presence. Conditioning flattens the gradient below 1,900 m but not above it, where attack halves while lodgepole cover rises and stands thin."

S$band <- cut(S$Elevation, breaks = seq(500, 2400, 200), dig.lab = 5)
grad <- S |> group_by(band) |>
  summarise(Cells = n(),
            `Pine cover (%)` = round(mean(PinePct), 1),
            `Basal area` = round(mean(Basal, na.rm = TRUE), 1),
            `Wind (m/s)` = round(mean(Wind), 2),
            `Attacked, all cells (%)` = round(100 * mean(Attacked), 1),
            `Attacked, host present (%)` =
              ifelse(sum(HasHost) > 300, round(100 * mean(Attacked[HasHost]), 1), NA),
            .groups = "drop") |>
  filter(Cells > 500)
grad |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))

## Spread of host-conditioned attack below 1,900 m, quoted in the text below.
hp_low <- S |> filter(HasHost, Elevation < 1900) |>
  mutate(b = cut(Elevation, seq(500, 1900, 200))) |>
  group_by(b) |> summarise(n = n(), pct = 100 * mean(Attacked), .groups = "drop") |>
  filter(n > 300)
HP_RANGE <- range(hp_low$pct)

## Host cover above the threshold, compared two ways, because the two
## comparisons disagree and reporting only the flattering one would be the
## same denominator problem examined in @sec-lowtrap. PINE_LO is
## the whole gradient below the threshold, which is dominated by the pine-poor
## valley; PINE_ADJ is the 150 m band immediately below it.
PINE_HI  <- mean(H$PinePct[H$Elevation >= ALPINE])
PINE_LO  <- mean(H$PinePct[H$Elevation <  ALPINE])
PINE_ADJ <- mean(H$PinePct[H$Elevation >= ALPINE - 150 & H$Elevation < ALPINE])
```

Across the full gradient attack is strongly unimodal (@tbl-gradient, @fig-gradient), rising from under three per cent in the valley to a peak near 1,430 m and falling to zero above 2,100 m.

Conditioning on host presence separates that curve into two phenomena. Below roughly 1,900 m the hump is host distribution: among cells containing lodgepole, attack runs between `r sprintf("%.0f", HP_RANGE[1])` and `r sprintf("%.0f", HP_RANGE[2])` per cent with no strong trend upward, and the rise in the raw curve is the rise in pine cover. Attack is `r sprintf("%.1f", 100*mean(S$Attacked[S$HasHost]))` per cent where lodgepole is present against `r sprintf("%.1f", 100*mean(S$Attacked[!S$HasHost]))` per cent where it is absent.

Above `r format(ALPINE, big.mark = ",")` m conditioning does not remove the decline. Attack among host-bearing cells falls by about half, while stand basal area falls from `r sprintf("%.1f", mean(H$Basal[H$Elevation<ALPINE], na.rm=TRUE))` to `r sprintf("%.1f", mean(H$Basal[H$Elevation>=ALPINE], na.rm=TRUE))`. Whether that is the pattern the hypothesis requires depends on which comparison is made, so both are reported. Against the whole gradient below, mean lodgepole cover in host-bearing cells is `r sprintf("%.1f", PINE_HI)` per cent above the threshold against `r sprintf("%.1f", PINE_LO)` per cent below, essentially unchanged; against the `r format(ALPINE - 150, big.mark = ",")` to `r format(ALPINE, big.mark = ",")` m band immediately below, where cover is `r sprintf("%.1f", PINE_ADJ)` per cent, the top of the gradient carries about `r sprintf("%.0f", 100 * (1 - PINE_HI / PINE_ADJ))` per cent less pine. The first comparison is inflated by the pine-poor valley it includes. The honest reading is that host thins somewhat at the top while attack falls a great deal more, so the conjunction the hypothesis requires holds in the weaker form.

```{r}
#| label: fig-gradient
#| echo: true
#| fig-width: 10.5
#| fig-height: 4.2
#| fig-cap: "(a) Attack across the gradient, raw and conditioned on lodgepole presence, with mean lodgepole cover. (b) The same data against wind speed: wind rises monotonically with elevation while attack turns over, so no monotone wind relationship can hold across the gradient. Shading marks the elevation range of the 2 by 4 km clip."

fine <- S |> mutate(b = cut(Elevation, breaks = seq(500, 2350, 100))) |>
  group_by(b) |>
  summarise(el = mean(Elevation), raw = 100*mean(Attacked), pine = mean(PinePct),
            wind = mean(Wind), n = n(),
            cond = ifelse(sum(HasHost) > 150, 100*mean(Attacked[HasHost]), NA_real_),
            .groups = "drop") |> filter(n > 300)
crng <- range(S$Elevation[S$InClip == 1])

pa <- ggplot(fine) +
  annotate("rect", xmin = crng[1], xmax = crng[2], ymin = -Inf, ymax = Inf,
           fill = "#b2182b", alpha = .10) +
  geom_line(aes(el, raw,  colour = "all cells"), linewidth = .9) +
  geom_line(aes(el, cond, colour = "host present"), linewidth = .9) +
  geom_line(aes(el, pine, colour = "lodgepole cover (%)"), linetype = "22", linewidth = .8) +
  scale_colour_manual(values = c("all cells" = "#2b4162", "host present" = "#b2182b",
                                 "lodgepole cover (%)" = "#3f8f9b"), name = NULL) +
  labs(x = "Elevation (m)", y = "Per cent", title = "(a) Attack and lodgepole cover against elevation") +
  theme(legend.position = "bottom")

pb <- ggplot(fine, aes(wind, raw)) +
  geom_path(colour = "#3f8f9b", linewidth = .8) +
  geom_point(aes(colour = el), size = 1.8) +
  scale_colour_gradientn(colours = c("#2b4162","#3f8f9b","#c6d8af","#e8c07d","#8c4b3c"),
                         name = "Elev (m)") +
  labs(x = "Mean wind speed (m/s)", y = "Attacked (%)",
       title = "(b) Attack against mean wind speed")
pa + pb
```

## Stem diameter effects {#sec-size}

The result above rests on cover holding while attack halves, and cover is the wrong host variable to rest it on: a subalpine and a montane stand can carry identical cover and basal area while differing entirely in stem size, which is what decides whether a stand returns more beetles than it consumes [@carroll2004bionomics].

```{r}
#| label: tbl-size
#| echo: true
#| tbl-cap: "Attack by stand quadratic mean diameter, in cells carrying lodgepole pine. The published source-sink crossover for lodgepole is about 25 cm [@carroll2004bionomics]. The observed distribution reproduces that threshold without it being imposed."

HQ <- H[!is.na(H$QMD), ]
HQ$dclass <- cut(HQ$QMD, c(0, 15, 20, 25, 30, 35, Inf),
                 labels = c("≤ 15", "15 to 20", "20 to 25",
                            "25 to 30", "30 to 35", "> 35"))

HQ |> group_by(`Quadratic mean diameter (cm)` = dclass) |>
  summarise(Cells = n(), `Attacked (%)` = round(100*mean(Attacked), 1),
            `Lodgepole cover (%)` = round(mean(PinePct), 1),
            `Basal area (m² ha⁻¹)` = round(mean(Basal), 1), .groups = "drop") |>
  kable(booktabs = TRUE, align = "lllll", format.args = list(big.mark = ","))
```

Attack rises monotonically with stem size, from `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$QMD <= 15]))` per cent in the smallest class to `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$QMD > 35]))` per cent in the largest, stepping most steeply at the 25 cm boundary the entomology names. Cover does not order the same way, and in the largest diameter class it is at its lowest, so this is not cover under another name.

```{r}
#| label: tbl-size-elev
#| echo: true
#| tbl-cap: "The high-elevation decline, decomposed. Host cover is flat across the threshold; stem size is not. The final column restricts to stands above the 25 cm source-sink crossover, where the decline persists."

sz <- function(d) tibble(
  Cells = nrow(d),
  `Attacked (%)`         = round(100*mean(d$Attacked), 1),
  `Lodgepole cover (%)`  = round(mean(d$PinePct), 1),
  `Basal area (m² ha⁻¹)` = round(mean(d$Basal), 1),
  `Mean diameter (cm)`   = round(mean(d$QMD), 1),
  `Stands > 25 cm (%)`   = round(100*mean(d$QMD > 25), 1),
  `Attacked, > 25 cm only (%)` = round(100*mean(d$Attacked[d$QMD > 25]), 1))

bind_rows(sz(HQ[HQ$Elevation <  ALPINE, ]), sz(HQ[HQ$Elevation >= ALPINE, ])) |>
  mutate(Band = c(sprintf("below %s m", format(ALPINE, big.mark=",")),
                  sprintf("above %s m", format(ALPINE, big.mark=","))), .before = 1) |>
  kable(booktabs = TRUE, align = "llllllll", format.args = list(big.mark = ","))
```

```{r}
#| label: size-conditioning
#| echo: true
#| output: false

## How much of the high-elevation suppression is stem size? Fit the same
## indicator three ways and read the attenuation. Nested fits on one frame, so
## the coefficients are directly comparable.
HQ$Hi <- HQ$Elevation >= ALPINE
hi_alone <- coef(glm(Attacked ~ Hi, HQ, family = binomial))[["HiTRUE"]]
hi_qmd   <- coef(glm(Attacked ~ Hi + QMD, HQ, family = binomial))[["HiTRUE"]]
hi_full  <- coef(glm(Attacked ~ Hi + QMD + PinePct + Basal,
                     HQ, family = binomial))[["HiTRUE"]]
```

Mean quadratic diameter falls from `r sprintf("%.1f", mean(HQ$QMD[HQ$Elevation < ALPINE]))` cm below the threshold to `r sprintf("%.1f", mean(HQ$QMD[HQ$Elevation >= ALPINE]))` cm above it, and the share of stands above the source-sink crossover falls from `r sprintf("%.1f", 100*mean(HQ$QMD[HQ$Elevation < ALPINE] > 25))` per cent to `r sprintf("%.1f", 100*mean(HQ$QMD[HQ$Elevation >= ALPINE] > 25))` per cent, while cover over the same step moves only from `r sprintf("%.1f", mean(HQ$PinePct[HQ$Elevation < ALPINE]))` to `r sprintf("%.1f", mean(HQ$PinePct[HQ$Elevation >= ALPINE]))` per cent. Part of the suppression we had attributed to a non-host mechanism is therefore host quality, and reporting cover alone concealed it.

The claim survives the correction. Within source stands alone, attack still falls from `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$Elevation < ALPINE & HQ$QMD > 25]))` to `r sprintf("%.1f", 100*mean(HQ$Attacked[HQ$Elevation >= ALPINE & HQ$QMD > 25]))` per cent across the threshold. Fitted as a single indicator the log-odds of attack above `r format(ALPINE, big.mark=",")` m is `r sprintf("%+.3f", hi_alone)`; adding diameter leaves `r sprintf("%+.3f", hi_qmd)`, and adding cover and basal area as well leaves `r sprintf("%+.3f", hi_full)`. Stem size absorbs about `r sprintf("%.0f", 100*(1 - hi_qmd/hi_alone))` per cent of the effect and the full host set about `r sprintf("%.0f", 100*(1 - hi_full/hi_alone))` per cent; the remaining `r sprintf("%.0f", 100*hi_full/hi_alone)` per cent is not host in any coding.

## Wind disruption test {#sec-krawchuk}

This is the first fit of the term R1 identified as proposed but never entered.

```{r}
#| echo: true
#| output: false

mt_h <- model_table(ZH)
SUPP_KRAWCHUK <- fmt_model(mt_h) |> kable(booktabs = TRUE, align = "llllll")
```

```{r}
#| label: krawchuk-numbers
#| cache: true
#| echo: true
#| output: false
wr <- mt_h[mt_h$term == "WindResid", ]
pp <- mt_h[mt_h$term == "PinePct", ]
```

Restricted to cells containing host, the orthogonalised wind term is the largest negative coefficient in the model, `r sprintf("%+.3f", wr$beta)` per standard deviation, comparable to the positive effect of lodgepole cover at `r sprintf("%+.3f", pp$beta)` (Table S1). Its sign is reproduced in `r sprintf("%.0f", 100*wr$sign_stab)` per cent of block resamples, but its 95 per cent block interval runs `r sprintf("[%+.3f, %+.3f]", wr$lo, wr$hi)` and reaches zero. On this fit the hypothesis is supported in direction and unresolved in magnitude, and the specification test below withdraws even the direction, so read this as the conventional analysis we go on to test rather than as a result we stand behind. Host cover is the only term whose interval excludes zero.

## Severity thresholds {#sec-ordinal}

```{r}
#| label: polr-fit
#| cache: true
#| echo: true
#| output: false

## Fitted only to show what a single proportional-odds summary conceals; the
## threshold-specific fits in @tbl-ordinal are the result.
zo <- ZH; zo$sev <- factor(pmin(zo$Severity, 4L), levels = 0:4, ordered = TRUE)
polr_fit <- MASS::polr(sev ~ Elevation + WindResid + PinePct + Basal +
                             TPI600 + Slope + TWI + RIX,
                       data = zo, method = "logistic", Hess = FALSE)
```

```{r}
#| label: tbl-ordinal
#| cache: true
#| echo: true
#| tbl-cap: "Cumulative logits at each severity threshold, cells containing lodgepole pine only. The orthogonalised wind coefficient strengthens monotonically as the threshold rises and its block interval excludes zero above trace, which the binary model does not show."

CUTS <- tibble(k = 1:4,
               label = c("trace or worse", "light or worse",
                         "moderate or worse", "severe or worse"))

cum <- lapply(seq_len(nrow(CUTS)), function(i) {
  z <- ZH; z$y <- as.integer(z$Severity >= CUTS$k[i])
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Threshold = CUTS$label[i], Cells = sum(z$y),
         `Wind residual beta` = coef(fit)[["WindResid"]],
         lo = q[1], hi = q[2],
         `sign stability` = mean(sign(bs[, "WindResid"]) ==
                                 sign(coef(fit)[["WindResid"]]), na.rm = TRUE),
         `Pine cover beta` = coef(fit)[["PinePct"]],
         `Elevation beta`  = coef(fit)[["Elevation"]])
}) |> bind_rows()

cum |> transmute(Threshold, Cells,
                 `Wind residual beta` = sprintf("%+.3f", `Wind residual beta`),
                 `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
                 `sign stability` = sprintf("%.2f", `sign stability`),
                 `Pine cover beta` = sprintf("%+.3f", `Pine cover beta`),
                 Supported = ifelse(lo * hi > 0, "yes", "no")) |>
  kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

Collapsing the response to attacked or not costs the study its main apparent result. Fitted at successively harder severity thresholds, the orthogonalised wind coefficient strengthens monotonically from `r sprintf("%+.3f", cum$"Wind residual beta"[1])` at trace or worse to `r sprintf("%+.3f", cum$"Wind residual beta"[4])` at severe or worse, a `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])`-fold increase (@tbl-ordinal), and at every threshold above trace the block interval excludes zero. The first row is the binary model restated, since a cell reaching trace or worse is exactly a cell recorded as attacked, and it is the weakest of the four and the one the refugia literature would conventionally have reported.

The slopes differ by a factor of `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])` across the thresholds, so proportional odds does not hold and a single ordinal coefficient would be the wrong summary; fitting one anyway returns `r sprintf("%+.3f", coef(polr_fit)[["WindResid"]])`, close to the trace estimate and concealing everything above it. Lodgepole cover, the only term clearing the interval test in the binary model, weakens as the threshold rises and reverses at severe or worse (`r sprintf("%+.3f", cum$"Pine cover beta"[4])`), which is either ecological or an artefact of the 2014 inventory vintage; we report it and decline to interpret it.

## Nonlinear elevation specification {#sec-nonlinear}

```{r}
#| label: tbl-nonlinear
#| cache: true
#| echo: true
#| tbl-cap: "The ordinal sweep refitted with elevation entered flexibly instead of linearly, and wind residualised against that same basis. Elevation explains eleven points more of the wind surface once curvature is allowed, and no threshold clears the interval test under either flexible basis."

## Everything above enters elevation as one linear term, and orthogonalises wind
## against that same straight line. But @fig-gradient shows the response is
## unimodal in elevation, and a straight line through a hump leaves a structured,
## elevation-shaped residual that any covariate correlated with the curvature can
## absorb. The wind surface is such a covariate by construction, because the
## downscaling is a nonlinear function of terrain. Residualising against a line
## cannot remove curvature the line never described. This table tests that.

elev_basis <- function(E, basis) switch(basis,
  linear = cbind(ElevL = E),
  quad   = cbind(ElevL = E, ElevQ = E^2),
  spline = { b <- splines::ns(E, df = 4); colnames(b) <- paste0("ElevS", 1:4); as.matrix(b) })

nl_row <- function(basis) {
  z  <- ZH
  EB <- elev_basis(z$Elevation, basis)
  z  <- cbind(z[, setdiff(names(z), "Elevation")], as.data.frame(EB))
  ## wind residualised on THIS basis, not on the straight line
  z$WindResid <- as.numeric(scale(residuals(lm(z$Wind ~ EB))))
  vars <- c(colnames(EB), "WindResid", "PinePct", "Basal", "TPI600", "Slope", "TWI", "RIX")
  r2 <- summary(lm(z$Wind ~ EB))$r.squared
  lapply(seq_len(nrow(CUTS)), function(i) {
    z$y <- as.integer(z$Severity >= CUTS$k[i])
    fit <- fit_fast(design_matrix(z, vars), z$y)
    bs  <- block_boot(z, vars)
    q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
    tibble(`Elevation basis` = basis, `R² wind on basis` = round(r2, 3),
           Threshold = CUTS$label[i],
           `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
           `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
           Supported = ifelse(q[1]*q[2] > 0, "yes", "no"))
  }) |> bind_rows()
}
nl <- bind_rows(lapply(c("linear","quad","spline"), nl_row))
nl |> kable(booktabs = TRUE, align = "llllll")
```

@sandel2015 gives the diagnostic that motivates this test: for a genuinely linear relationship a coefficient is extent-independent, so a coefficient that changes sign under truncation is a symptom of an unmodelled nonlinearity. Our response is unimodal in elevation and our wind surface monotone in it, exactly the configuration that produces such a symptom, so the specification has to be tested rather than assumed.

It does not pass. Entering elevation as a quadratic raises the share of the wind surface that elevation explains from `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[1]))` to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[5]))`, and a natural spline on four degrees of freedom raises it to `r sprintf("%.3f", as.numeric(nl$"R² wind on basis"[9]))`. Roughly eleven percentage points of the wind surface's variance is therefore nonlinear in elevation, and the linear orthogonalisation used above leaves all of it inside what we had been calling the component of wind independent of elevation.

When that residue is removed the wind result goes with it. The coefficient at severe or worse falls from `r nl$"Wind residual beta"[4]` to `r nl$"Wind residual beta"[8]`, and at moderate or worse from `r nl$"Wind residual beta"[3]` to `r nl$"Wind residual beta"[7]`, and under both flexible bases every interval covers zero at every threshold. This changes what the preceding two sections mean. The dose-response is not evidence about wind; it is very largely the unimodal elevation response reappearing through a predictor that was never orthogonal to the curvature of elevation, only to its slope.

## Aspect-derived predictors {#sec-aspect}

```{r}
#| label: tbl-aspect
#| cache: true
#| echo: true
#| tbl-cap: "The host-present model with the two aspect-derived terms added, at each threshold it can carry. Heat load is positive and strengthens sharply with severity: warmer aspects burn harder at the same elevation. The orthogonalised wind term is not displaced by it. The shelter index carries the opposite sign, and this result is examined rather than set aside."

FULLV <- c(MODEL_VARS, EXTRA)
ZHx   <- ZH[complete.cases(ZH[, FULLV]), ]

## Ten terms and 1,522 events cannot be fitted together: the severe threshold
## drives fitted probabilities to numerical zero. We detect that rather than
## report the resulting coefficients, and the severe row is dropped below.
sep_check <- function(z, vars, k) {
  z$y <- as.integer(z$Severity >= k)
  p <- fit_fast(design_matrix(z, vars), z$y)$fitted.values
  sum(p < 1e-6 | p > 1 - 1e-6)
}
sep <- sapply(1:4, function(k) sep_check(ZHx, FULLV, k))
names(sep) <- c("trace+","light+","moderate+","severe+")
OK_CUTS <- which(sep == 0)

asp_tab <- lapply(OK_CUTS, function(k) {
  z <- ZHx; z$y <- as.integer(z$Severity >= k)
  fit <- fit_fast(design_matrix(z, FULLV), z$y); bs <- block_boot(z, FULLV)
  ci <- function(n) quantile(bs[, n], c(.025, .975), na.rm = TRUE)
  one <- function(n) sprintf("%+.3f [%+.3f, %+.3f]", coef(fit)[[n]], ci(n)[1], ci(n)[2])
  tibble(Threshold = CUTS$label[k], Cells = sum(z$y),
         `Wind residual` = one("WindResid"), `Heat load` = one("HLI"),
         `Shelter` = one("SxMean"))
}) |> bind_rows()

asp_tab |> kable(booktabs = TRUE, align = "lllll",
                 format.args = list(big.mark = ","))
```

Adding two aspect-derived terms does three things. First, it supplies the one piece of thermal information available without a temperature surface: heat load is positive at every threshold and strengthens steeply with severity (@tbl-aspect), and since heat load is essentially uncorrelated with elevation (@tbl-orthogonality) this is not the elevation gradient under another name. At a given elevation, southwest-facing ground carries heavier mortality than northeast-facing ground, which is the sign warmth-limited development predicts. Second, it leaves the wind coefficient standing, so whatever that term tracks is not the warm-aspect effect in disguise. Third, and least comfortably, the two exposure measures disagree: the shelter index is negative, meaning sheltered ground carries *less* attack, the opposite of the orthogonalised wind term and of the hypothesis. The two are only weakly correlated (`r sprintf("%.2f", cor(ZHx$WindResid, ZHx$SxMean))`), so they are not two measurements of one quantity that happen to disagree, and we take this as evidence against treating any single surface as "wind exposure" as though the phrase had one referent. The severe threshold is absent from @tbl-aspect because ten terms and `r sum(ZHx$Severity >= 4)` events drive the fit into quasi-separation, meaning some cells are fitted at effectively zero probability and the estimate runs away.

## Winter temperature {#sec-cold}

```{r}
#| echo: true
#| output: false

st <- read.csv(here("02.inputs/climate/station-winter-minima.csv"), stringsAsFactors = FALSE)

## The censoring rule is applied here rather than taken on trust from the
## retrieval script. Many older British Columbia Wildfire Service sensors floor
## at exactly -20.0 C. A station-winter is unusable when its minimum sits
## exactly on that floor, because the true minimum is then unknown. A minimum
## colder than the floor shows the sensor did read past it, so readings at
## -20.0 in those winters are real measurements and the winter is kept. The
## floor is a sensor artefact of that network, so the rule does not touch the
## Environment and Climate Change Canada records.
FLOOR <- -20
st$censored <- st$network == "BCWS" &
               abs(st$min_temp_c - FLOOR) < 1e-6 &
               st$n_at_minus20 > 0
stopifnot(identical(st$censored, as.logical(st$censored_flag)))

## Isolated single-observation spikes are rejected at the observation level by
## the retrieval script, which is the only stage holding the sub-daily series a
## per-winter summary cannot carry; `min_temp_c_usable` is that de-spiked
## minimum. Everything else about the screen is visible above.
use <- st[!st$censored & !is.na(st$min_temp_c_usable), ]

## Warm sensor failure. The retrieval screen inspects only the minimum, so a
## sensor that fails warm passes through it invisibly and contributes a
## spuriously mild minimum. NORNS, the highest station in the network at 2,423 m,
## did this through 2015: a February mean daily minimum of +30.0 C and a winter
## "minimum" of +4.6 C. The bias runs towards the conclusion that the landscape
## never gets cold enough, so it is screened here and its effect is reported.
## `build-monthly-climatology.R` derives the monthly means from the same cached
## extracts and carries the flag; the cut is read off that distribution.
clim <- read.csv(here("02.inputs/climate/station-monthly-climatology.csv"),
                 stringsAsFactors = FALSE)
bad_sw <- unique(clim[clim$warm_fail == "TRUE" | clim$warm_fail == TRUE,
                      c("station_name", "winter_year")])
WARM_DROPPED <- nrow(merge(use, bad_sw, by = c("station_name", "winter_year")))
MEAN_HI_RAW  <- mean(use$min_temp_c_usable[use$elevation_m >= 1500])
use <- use[!(paste(use$station_name, use$winter_year) %in%
             paste(bad_sw$station_name, bad_sw$winter_year)), ]
## No physically impossible winter minimum may survive the screen.
stopifnot(max(use$min_temp_c_usable) < 0)

use$band <- cut(use$elevation_m, c(0, 800, 1500, 2500),
                labels = c("below 800 m", "800 to 1,500 m", "above 1,500 m"))
## Held for the supplementary section rather than printed here, so the main
## text keeps to the ten floats the guidelines allow.
SUPP_COLD <- use |> group_by(band) |>
  summarise(Stations = n_distinct(station_name), `Station-winters` = n(),
            `Coldest (°C)` = min(min_temp_c_usable),
            `Mean minimum (°C)` = round(mean(min_temp_c_usable), 1),
            .groups = "drop") |>
  kable(booktabs = TRUE, align = "lllll")
```

```{r}
#| echo: true
#| output: false
#| fig-width: 5.9
#| fig-height: 5.1

stn <- aggregate(cbind(latitude, longitude, elevation_m) ~ station_name,
                 data = st, FUN = function(x) x[1])
coldest_by <- aggregate(min_temp_c_usable ~ station_name, data = use, FUN = min)
stn <- merge(stn, coldest_by, by = "station_name", all.x = TRUE)
stn_sf <- st_transform(st_as_sf(stn, coords = c("longitude","latitude"), crs = 4326), 3005)
xy <- st_coordinates(stn_sf); stn_sf$X <- xy[, 1]; stn_sf$Y <- xy[, 2]
site_bc <- st_transform(st_as_sf(as.polygons(ext(G), crs = crs(G))), 3005)
prov_bc <- st_transform(ne_states(country = c("Canada","United States of America"),
                                  returnclass = "sf"), 3005)
bbs <- st_bbox(st_buffer(st_union(c(st_geometry(stn_sf), st_geometry(site_bc))), 25000))
sw  <- unname(bbs["xmax"] - bbs["xmin"])
key <- stn_sf[order(stn_sf$min_temp_c_usable)[1:2], ]

fig_station <- ggplot() +
  geom_sf(data = prov_bc, fill = "grey95", colour = "grey70", linewidth = .25) +
  geom_sf(data = site_bc, fill = NA, colour = "#C0392B", linewidth = .6) +
  geom_sf(data = stn_sf, aes(fill = elevation_m), shape = 21, size = 1.9,
          colour = "grey20", stroke = .25) +
  scale_fill_viridis_c(option = "viridis", name = "Station\nelevation (m)") +
  geom_label_repel(data = as.data.frame(key), aes(X, Y, label = tolower(station_name)),
                   size = 2, colour = "grey10", fill = "white", alpha = .9,
                   label.size = 0, label.padding = unit(.7, "mm"), box.padding = .5,
                   min.segment.length = 0, segment.colour = "grey30",
                   segment.size = .3, seed = 1) +
  annotate("label", x = bbs["xmin"] + .02*sw,
           y = bbs["ymin"] + .04*(bbs["ymax"] - bbs["ymin"]),
           label = "red outline:\nanalysis grid", size = 1.9, hjust = 0, vjust = 0,
           colour = "#C0392B", fill = "white", alpha = .85,
           label.padding = unit(.7, "mm")) +
  coord_sf(crs = 3005, xlim = bbs[c("xmin","xmax")], ylim = bbs[c("ymin","ymax")],
           datum = st_crs(3005), expand = FALSE) +
  scale_x_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  scale_y_continuous(breaks = scales::breaks_pretty(3), labels = km_lab) +
  north + map_theme +
  labs(title = "Weather stations screened for winter minima",
       subtitle = sub_alb)
fig_station
ggsave("../03.outputs/PNG/fig-station-maps.png", fig_station,
       width = 150, height = 130, units = "mm", dpi = 500, bg = "white")
```

```{r}
#| label: cold-numbers
#| echo: true
#| output: false

coldest <- use[which.min(use$min_temp_c_usable), ]
hi      <- use[use$elevation_m >= 1500, ]
lowcold <- use[use$elevation_m < 800, ]
lowcold <- lowcold[which.min(lowcold$min_temp_c_usable), ]

## Date of each winter minimum. Lethal temperature is conditional on the state of
## cold-hardening, so a minimum without a date cannot be compared with a lethal
## range. The per-station-winter summary the fetch writes does not carry the
## timestamp; `add-minimum-dates.R` recovers it from the same cached hourly
## extracts and is documented alongside the fetch.
dates <- read.csv(here("02.inputs/climate/station-winter-minimum-dates.csv"),
                  stringsAsFactors = FALSE)
useD <- merge(use, dates[, c("station_id","winter_year","min_month","min_date")],
              by = c("station_id","winter_year"))
## The merge must not silently drop or duplicate station-winters.
stopifnot(nrow(useD) == sum(use$network == "BCWS"))
DEC <- 100 * mean(useD$min_month == 12)
```

The cold explanation assumes it gets cold enough at the top of the gradient to kill beetles. Assessed against the stage-specific ranges given in the Introduction rather than against a single benchmark, the record supports a narrower conclusion than we first drew.

```{r}
#| label: fig-climate
#| echo: true
#| fig-width: 7.2
#| fig-height: 8.4
#| fig-cap: "The observational climate record behind the cold check. (a) Winter climograph by elevation band: mean daily maximum, monthly mean and mean daily minimum air temperature, with monthly precipitation as bars rising from the floor of the temperature axis and read against the right-hand scale. Only December, January and February were retrieved, so the shoulder months in which a given temperature is most lethal are absent. The apparent decline of winter precipitation with elevation is a gauge artefact: the high-elevation stations are tipping-bucket instruments, which under-catch snowfall severely, and the bars are shown for the seasonal shape within a band rather than for comparison between bands. (b) Winter minima by year, one point per station-winter, against the lethal ranges reported by @safranyik2006chap1 for first-instar larvae and for the third and fourth instars that normally overwinter. No station-winter enters the late-instar range. (c) The same minima against station elevation, showing that elevation orders them only weakly because cold air pools in the valleys."

MONTH_ORDER <- c(12, 1, 2)
band_of <- function(e) cut(e, c(0, 800, 1500, 2500),
                           labels = c("below 800 m", "800 to 1,500 m", "above 1,500 m"))

cl <- clim[!(clim$warm_fail %in% c(TRUE, "TRUE")) & clim$n_days >= 20, ]
cl$band  <- band_of(cl$elevation_m)
cl$month <- factor(cl$month, levels = MONTH_ORDER,
                   labels = c("Dec", "Jan", "Feb"))

## Precipitation is plotted on a secondary axis, so it is rescaled onto the
## temperature axis once here rather than inside the geom.
PR_MAX <- max(tapply(cl$total_precip_mm, list(cl$band, cl$month), mean, na.rm = TRUE),
              na.rm = TRUE)
T_LO   <- -20
PR_SCALE <- abs(T_LO) / PR_MAX

clm <- cl |>
  group_by(band, month) |>
  summarise(tmax = mean(mean_daily_max_c, na.rm = TRUE),
            tavg = mean(mean_temp_c,      na.rm = TRUE),
            tmin = mean(mean_daily_min_c, na.rm = TRUE),
            prcp = mean(total_precip_mm,  na.rm = TRUE),
            .groups = "drop")

## Bars are drawn from the floor of the temperature axis upward. A geom_col here
## would draw from zero downward, so a dry month would render as the tallest bar.
clm$xi <- as.integer(clm$month)

p_clim <- ggplot(clm, aes(month, group = band)) +
  geom_rect(aes(xmin = xi - .3, xmax = xi + .3,
                ymin = T_LO, ymax = prcp * PR_SCALE + T_LO),
            fill = "grey82", colour = NA) +
  geom_ribbon(aes(ymin = tmin, ymax = tmax), fill = "#2C7FB8", alpha = .18) +
  geom_line(aes(y = tmax), colour = "#C0392B", linewidth = .5) +
  geom_line(aes(y = tmin), colour = "#2C7FB8", linewidth = .5) +
  geom_line(aes(y = tavg), colour = "grey20", linewidth = .7) +
  geom_point(aes(y = tavg), colour = "grey20", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = .3) +
  facet_wrap(~band, nrow = 1) +
  scale_y_continuous(
    name = "Air temperature (°C)", limits = c(T_LO, NA),
    sec.axis = sec_axis(~ (. - T_LO) / PR_SCALE, name = "Precipitation (mm)")) +
  labs(x = NULL, title = "(a) Winter climograph by elevation band",
       subtitle = "Red: mean daily maximum. Black: monthly mean. Blue: mean daily minimum. Bars: precipitation") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank(),
        plot.subtitle = element_text(colour = "grey35"))

## Lethal ranges are stage-specific and are quoted from the same source the text
## uses, so the figure and the prose cannot drift apart.
leth <- data.frame(
  stage = factor(c("first instar", "third and fourth instars"),
                 levels = c("first instar", "third and fourth instars")),
  lo = c(-29, -40), hi = c(-23, -29))

useP <- use
useP$band <- band_of(useP$elevation_m)

p_series <- ggplot() +
  geom_rect(data = leth, aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi,
                             fill = stage), alpha = .18) +
  scale_fill_manual(values = c("first instar" = "#F39C12",
                               "third and fourth instars" = "#2C7FB8"),
                    name = "Lethal range") +
  geom_point(data = useP, aes(winter_year, min_temp_c_usable, colour = elevation_m),
             size = 1.5, alpha = .9) +
  scale_colour_viridis_c(name = "Station\nelevation (m)") +
  scale_x_continuous(breaks = seq(2000, 2015, 5)) +
  labs(x = "Winter (year of January)", y = "Winter minimum (°C)",
       title = "(b) Observed winter minima against stage-specific lethal ranges") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank())

p_elev <- ggplot(useP, aes(elevation_m, min_temp_c_usable)) +
  geom_rect(data = leth, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi, fill = stage),
            alpha = .18, show.legend = FALSE) +
  scale_fill_manual(values = c("first instar" = "#F39C12",
                               "third and fourth instars" = "#2C7FB8")) +
  geom_point(aes(colour = winter_year), size = 1.5, alpha = .9) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "grey20", linewidth = .5) +
  scale_colour_viridis_c(option = "magma", name = "Winter") +
  labs(x = "Station elevation (m)", y = "Winter minimum (°C)",
       title = "(c) Winter minimum against station elevation") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank())

fig_climate <- p_clim / p_series / p_elev + plot_layout(heights = c(1, 1, 1))
fig_climate
ggsave("../03.outputs/PNG/fig-climate.png", fig_climate,
       width = 183, height = 213, units = "mm", dpi = 500, bg = "white")
```

```{r}
#| label: climate-lapse
#| echo: true
#| output: false

## The lapse rate the record supports, quoted in the text. Reported because the
## alternative to a station record is an assumed lapse rate, and a referee is
## entitled to see what this landscape's own stations give.
lap <- lm(min_temp_c_usable ~ elevation_m, data = use)
LAPSE   <- unname(coef(lap)[2]) * 1000
LAPSE_R2 <- summary(lap)$r.squared
```

Across the screened record the winter minimum falls by `r sprintf("%.1f", abs(LAPSE))` °C per 1,000 m of station elevation, and elevation explains `r sprintf("%.2f", LAPSE_R2)` of the variance in it, which is the quantitative form of cold-air pooling. Across `r nrow(use)` station-winters from `r n_distinct(use$station_name)` stations spanning `r min(use$elevation_m)` to `r max(use$elevation_m)` m, the coldest defensible minimum in the region is `r sprintf("%.1f", coldest$min_temp_c_usable)` °C, at `r tolower(coldest$station_name)` at `r format(round(coldest$elevation_m), big.mark=",")` m in winter `r coldest$winter_year`. Above 1,500 m the mean winter minimum is `r sprintf("%.1f", mean(hi$min_temp_c_usable))` °C, and no station-winter reaches -35 °C.

That coldest value sits inside the lethal range for first and second instars and at the warm edge of the range for third and fourth. It is not far short of any lethal range. Timing qualifies this further: `r sprintf("%.0f", DEC)` per cent of Wildfire Service station-winters reach their minimum in December, at the front of the hardening window, as do `r sum(useD$min_month[order(useD$min_temp_c_usable)][1:10] == 12)` of the ten coldest. Elevation orders these minima only weakly, since `r tolower(lowcold$station_name)` at `r format(round(lowcold$elevation_m), big.mark=",")` m reaches `r sprintf("%.1f", lowcold$min_temp_c_usable)` °C, within `r sprintf("%.1f", abs(lowcold$min_temp_c_usable - coldest$min_temp_c_usable))` °C of the alpine station `r format(round(coldest$elevation_m - lowcold$elevation_m), big.mark=",")` m higher. Two sensor faults were screened out: `r sum(st$censored)` station-winters flooring at exactly -20.0 °C, and `r WARM_DROPPED` failing warm, retaining which would have raised the mean minimum above 1,500 m to `r sprintf("%.1f", MEAN_HI_RAW)` °C, a bias running towards the conclusion this section reaches.

The retrieval window runs from 1 December to the end of February, so October, November, March and April are absent, and those are precisely the months in which a given temperature is most lethal; no station inside the study area reported during the outbreak window. The defensible conclusion is correspondingly narrow: deep-winter air temperature in this region does not approach the lethal range for late-instar larvae, which does not establish that freeze mortality is absent.

## Sensitivity to extent and grain {#sec-extent}

```{r}
#| label: tbl-extent
#| cache: true
#| echo: true
#| tbl-cap: "The same model fitted at three extents. As the analysis area shrinks, the share of wind variance explained by elevation rises and the orthogonalised wind coefficient reverses sign. The 2 by 4 km clip is the extent at which the companion study was run."

extent_row <- function(z, lab) {
  r2  <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  ## The block count is reported because it, not the cell count, is the
  ## resampling unit. The clip holds few enough blocks that its interval is
  ## far weaker than its cell count suggests, and the reader must see that.
  tibble(Extent = lab, Cells = nrow(z), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         ## This is the variance inflation the *raw* wind term would carry
         ## against elevation. It is not the VIF of the fitted model, in which
         ## the residualised term is orthogonal to elevation by construction and
         ## its VIF is 1. It is reported to show the collinearity that
         ## orthogonalisation reparameterises away rather than removes.
         `VIF of raw wind` = round(1/(1-r2), 1),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]))
}
bind_rows(
  extent_row(Z,  sprintf("full landscape (%.0f km²)", nrow(S)*prod(res(G))/1e6)),
  extent_row(ZH, "landscape, host present"),
  extent_row(ZC, sprintf("2 by 4 km clip (%.1f km²)",
                         sum(S$InClip == 1) * prod(res(G)) / 1e6))
) |> kable(booktabs = TRUE, align = "lllllll", format.args = list(big.mark = ","))
```

Across the landscape, elevation explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = Z))$r.squared)` per cent of the variance in the wind surface; inside the 2 by 4 km clip it explains `r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` per cent (@tbl-extent), and the orthogonalised wind coefficient is negative at landscape extent and positive inside the clip. The collinearity result is firm, being a property of the two surfaces rather than an estimate with sampling error.

```{r}
#| label: clip-interval
#| cache: true
#| echo: true
#| output: false

## This interval used to be computed inside an inline expression in the sentence
## below. That hid 150 model refits inside a line of prose, where no reader could
## see the cost and no cache could reach it.
CLIP_BLOCKS <- length(unique(ZC$block))
CLIP_CI     <- local({
  q <- quantile(block_boot(ZC)[, "WindResid"], c(.025, .975), na.rm = TRUE)
  sprintf("[%+.3f, %+.3f]", q[1], q[2])
})
```

The sign reversal is not, on this evidence, a demonstrated reversal. The clip holds only `r CLIP_BLOCKS` spatial blocks, so its block interval runs `r CLIP_CI`, spanning zero and both landscape estimates. What it demonstrates is that shrinking the extent destroys the ability to estimate the coefficient at all, and a study at this extent would report a positive point estimate with an interval too wide to notice it had learned nothing.

```{r}
#| label: grain-fit
#| cache: true
#| echo: true
#| output: false

## Not `tbl-grain-fit`: a `tbl-` prefix declares a float reference target, and this
## chunk sets `output: false`, so Quarto warned "FloatRefTarget with no content".
## The prefix is reserved; a chunk that emits nothing must not claim one.
## The four option lines above must stay above these comments: Quarto only reads
## `#|` options from the top of a cell, and until 2026-08-19 they sat below and
## were silently ignored, so this chunk was neither labelled, cached nor silenced.

CONT <- c("Elevation","Slope","TWI","RIX","Wind","PinePct","Basal")
base_stack <- c(terr[[c("Elevation","Slope","TWI","RIX","Wind")]],
                host[[c("PinePct","Basal")]], severity, in_clip)
names(base_stack) <- c(CONT, "Severity", "InClip")

## fact is the aggregation factor from the native 25 m grid.
build_grain <- function(fact) {
  b <- if (fact == 1L) base_stack else
       c(aggregate(base_stack[[CONT]], fact, fun = "mean", na.rm = TRUE),
         aggregate(base_stack[[c("Severity","InClip")]], fact, fun = "max", na.rm = TRUE))
  Gi <- b[["Elevation"]]; rs <- res(Gi)[1]
  w  <- max(3L, as.integer(2 * floor(625 / rs / 2) + 1))   # odd window nearest 625 m
  b$TPI600 <- tpi(Gi, w)
  d <- as.data.frame(b, na.rm = FALSE)
  d$Attacked <- d$Severity > 0
  d$HasHost  <- !is.na(d$PinePct) & d$PinePct > 0
  blk <- max(1L, as.integer(round(BLOCK / fact)))          # blocks stay near 1 km
  rc  <- rowColFromCell(Gi, seq_len(ncell(Gi)))
  d$block <- paste(rc[,1] %/% blk, rc[,2] %/% blk, sep = "_")
  d <- d[complete.cases(d[, MVARS]), ]
  structure(d, res = rs, win = w * rs)
}

grain_row <- function(fact) {
  d  <- build_grain(fact); h <- d[d$HasHost, ]
  z  <- standardise(h)
  r2 <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y); bs <- block_boot(z)
  q  <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Grain = sprintf("%.0f m", attr(d, "res")),
         `TPI window` = sprintf("%.0f m", attr(d, "win")),
         `Host cells` = nrow(h), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
         `sign stability` = sprintf("%.2f",
           mean(sign(bs[, "WindResid"]) == sign(coef(fit)[["WindResid"]]), na.rm = TRUE)))
}

## The 25 m row must reproduce the analysis frame built in `assemble`; if this
## assertion ever fails the two code paths have drifted apart.
stopifnot(nrow(build_grain(1L)) == nrow(S))

SUPP_GRAIN <- bind_rows(lapply(c(1L, 2L, 4L), grain_row)) |>
  kable(booktabs = TRUE, align = "llllllll", format.args = list(big.mark = ","))
```

Grain behaves quite differently (Table S3): aggregating from 25 m to 100 m leaves the confound essentially untouched, and the orthogonalised wind coefficient neither reverses nor decays. Enlarging the cell changes almost nothing; shrinking the study area destroys the estimate. That is a result about this landscape and this pair of surfaces, not a general ordering of the two axes [@mcgarigal2016].

# Discussion

Read against the review, the result is a negative answer to a question the literature had been answering in the affirmative without measurement. R1 recorded that the wind-disruption mechanism has been invoked for years while entering no model; fitted here, on a landscape chosen for its relief, it does not survive a specification that allows the response its observed curvature. Four findings stand: attack is unimodal in elevation; below 1,900 m the pattern is the distribution of lodgepole pine; attack rises monotonically with stem diameter and steps at the published source-sink threshold, so host must be coded as size and not only as cover; and above 1,900 m, `r sprintf("%.0f", 100*hi_full/hi_alone)` per cent of the suppression remains after every host term we have. Three do not: the orthogonalised wind coefficient does not survive fitting elevation flexibly, the conclusion that freeze mortality can be excluded does not survive the stage-specific and seasonal structure of lethal temperature, and no estimate at all survives truncating the landscape to a 2 by 4 km clip. Nothing here identifies the mechanism behind the high-elevation refugia.

The high-elevation suppression is real and only partly host limitation. Among cells carrying lodgepole pine, attack above `r format(ALPINE, big.mark=",")` m runs at `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation>=ALPINE]))` per cent against `r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation<ALPINE]))` per cent below; cover barely moves across that step and stem size does. Under the @krawchuk2020 definition what remains after conditioning are refugia: disturbed less severely than the surrounding landscape, with host present, and predictably located by an enduring topographic feature. That matches @cartwright2018 in associating refugia with thinner stands, and says something more specific, because thinner here means smaller-stemmed as well as sparser.

What we cannot do is attribute the remainder to wind. The ordinal coding does recover structure the binary discards, so @krawchuk2020's warning against binary refugia coding stands, but that structure is not attributable to wind: residualising against a flexible elevation basis rather than a straight line removes most of the coefficient and leaves no threshold clearing the test.

The identification problem is untouched, and the life history makes the competing explanation more specific than we first allowed. Of the two thermal mechanisms set out above, overwinter mortality graded by instar [@safranyik2006chap1; @cooke2009mortality] and the loss of univoltine development, the second predicts our result without any appeal to wind. Amman's relationship between life-cycle duration, elevation and latitude, reproduced by @safranyik2006chap1, places the ceiling of univoltine development within the elevation range this study spans at this latitude, and a transition to semivoltine ground produces exactly what we observe: a step down in attack, with host present, at an elevation fixed by terrain, and graded in severity because the mechanism acts on how many beetles emerge together. Cold in either form predicts a dose-response in severity just as wind disruption does, so the monotone strengthening across thresholds does nothing to separate them.

We looked for a way out and do not think one exists at this extent. Separating exposure from cold needs a temperature surface whose fine-scale structure is not itself a function of the elevation model, and the gridded products with genuine sub-kilometre structure obtain it by regression onto a digital elevation model, so they reproduce the problem rather than solve it. Aspect gives partial purchase on one mechanism, heat load being trigonometric in the elevation model rather than draped over it, but says nothing about the overwinter minimum. The deeper obstacle is that the quantity governing overwinter survival is temperature under the bark, observed at landscape scale nowhere: @goodsman2024, validating the standard Canadian Forest Service winter mortality model against observed mortality in Banff National Park, found its spatial predictions poor and attributed the failure to precisely this.

Two computations would settle more than prose can, and neither is possible from the record assembled here. The univoltinism criterion asks where a generation completes in one year and needs growing-season degree-days, emergence scaling with accumulation above 14.4 °C [@safranyik2006chap1]; our archive was streamed for December to February only. The overwinter mortality model asks what share of larvae survive each winter and needs a daily series with modelled phloem temperature rather than the per-station-winter minima we hold. Both are data-collection tasks rather than design problems, and until they are run we report the thermal alternatives as live and unquantified rather than as established.

A second class of mechanism would produce our result without pheromone plumes at all, because wind acts on the beetle before it acts on the plume. Bark beetles do not fly in winds exceeding their own flight speed, and dispersal within a stand is itself wind-structured, beetles drifting downwind below the crowns until they meet an attractive plume and then flying upwind to its source [@carroll2004bionomics; @jones2019]. If exposed ground receives fewer arriving beetles, rather than the same number arriving to a disrupted signal, the observable consequence in a static model is identical, and the Allee effect in colonisation means a modest reduction in arrivals produces a large reduction in successful attack [@powell2014]. We cannot compare our wind values against the flight threshold, and that mismatch is itself the point.

The failure of the Landsat classification is a result worth reporting in its own right. A finer response layer is the obvious remedy for a survey that sketch-maps at the wrong scale, and the one built here is close to uninformative once its area is matched to the survey's; terrain correction does not escape the circularity either, being itself a model of the terrain under test. Studies proposing to substitute a classified response for an operational survey should report hit rate and lift at matched prevalence, because a pooled kappa from a balanced sample conceals the failure entirely.

## Host-absence confounding {#sec-lowtrap}

The result that should travel furthest is at the bottom of the gradient. Attack there is near zero, and any refugia mapping exercise lacking a host layer would have identified the valley as a large, spatially coherent, topographically predictable refugium. It is not a refugium; it is a place with almost no pine. Conditioning on host raises attack in the lowest band from `r sprintf("%.1f", grad$"Attacked, all cells (%)"[1])` to `r sprintf("%.1f", grad$"Attacked, host present (%)"[1])` per cent, within the range observed at mid elevations, so the difference between performing that check and not is the difference between reporting a roughly `r sprintf("%.0f", sum(S$Elevation < 900) * prod(res(G)) / 1e6)` km2 valley refugium and reporting none. A refugium claim without an independent host or habitat layer is not interpretable.

## Limitations

The collinearity result generalises beyond wind. A fine-resolution climate surface is produced by imposing terrain on a coarse field, so wherever that field is nearly constant the product approaches a function of the terrain, which covers most gridded climate products used in landscape ecology. Effect sizes for derived climate predictors are therefore not comparable across studies of different extents. We would ask three things of any study entering such a surface into a disturbance model: report its regression on elevation, report the same diagnostic at more than one extent, and interpret only the orthogonalised coefficient.

The strongest limitation is about the insect rather than the data. Everything estimated here describes an epidemic population, and endemic populations occupy weakened, small-diameter stems and select hosts differently [@safranyik2006chap1; @howe2022], so the diameter relationship should not be read as a constant of the species; the 1999 to 2015 window is a single epidemic passage, and of @krawchuk2020's distinction between refugia by severity and by frequency it can address only the first. Quadratic mean diameter is a stand attribute applied uniformly across a polygon, and the 25 cm crossover is an average across studies rather than a constant of this landscape [@carroll2004bionomics]. The survey response is observer sketch-mapping with known positional and severity error.

# Conclusions

Beetle-caused mortality at Darkwoods is unimodal across the `r format(round(diff(range(S$Elevation))), big.mark = ",")` m of relief the modelled area spans. Below 1,900 m that pattern is the distribution of lodgepole pine; above it, the margin is narrower than cover alone implies. Stem size is the host attribute that matters and it had not been used: attack rises from `r sprintf("%.0f", 100*mean(HQ$Attacked[HQ$QMD <= 15]))` to `r sprintf("%.0f", 100*mean(HQ$Attacked[HQ$QMD > 35]))` per cent across the diameter range, stepping at the 25 cm crossover [@carroll2004bionomics], and conditioning on the full host set removes about `r sprintf("%.0f", 100*(1 - hi_full/hi_alone))` per cent of the high-elevation suppression. We do not find a wind effect: the component of the atlas surface orthogonal to a linear elevation term clears the interval test at every severity threshold above trace, but entering elevation as a quadratic or a spline collapses it everywhere, so the apparent signal is a specification artefact. Even had it survived it would not have measured the mechanism, because wind arrests bark beetle flight near 2 m per second below the canopy while the atlas supplies an annual mean at 10 m over open ground. Thermal alternatives are not eliminated, and separating them needs the degree-day and overwinter-mortality computations this record cannot support.

# Conflict of interest

The author declares no conflict of interest.

# Author contribution

**Seamus Murphy:** Conceptualization; Data curation; Formal analysis; Investigation; Methodology; Software; Validation; Visualization; Writing - original draft; Writing - review and editing.

# Acknowledgements

Terrain surfaces and the site description are inherited from a companion analysis on the same ground [@murphy2026]. No external funding supported this work. Generative artificial intelligence tools were used to assist with drafting, code review and copy-editing of this manuscript. The author directed all analysis, verified every reported quantity against the code that produces it, and is fully responsible for the content.

# Data availability statement

PLACEHOLDER, BLOCKING BEFORE SUBMISSION. The Journal of Applied Entomology requires a resolvable link with a DOI inserted above the references, and states that data on request or supplied as supplementary files is not accepted. A repository deposit must therefore exist before submission. Wiley pays Dryad archiving charges for papers published in this journal, and a "Private for Peer Review" link should be quoted here while the paper is under review. The analysis code is in this document; the derived rasters, the classified red-stage series and the climate summaries are in the project repository and require a deposit with a minted DOI.

# References

::: {#refs}
:::

# Figure legends

Figure 1. The study area and the three spatial inputs. (a) Location within British Columbia; the marker is a 22 km symbol, not the footprint. (b) The elevation surface, which is the analysis grid. (c) Global Wind Atlas mean speed at 10 m; the shaded southern band lies outside the wind product's extent. (d) Beetle-caused mortality from the Aerial Overview Survey, 1999 to 2015, as the maximum severity class in any year. The outlined rectangle in (b) to (d) is the 2 by 4 km clip. Each panel's graticule, axis ticks and stated coordinate reference system are the same; ticks are in kilometres.

Figure 2. The forest inventory host layers, rasterised onto the analysis grid. (a) Lodgepole pine cover, accumulated across the three species slots each polygon records. (b) Stand basal area, which defines the analysis frame: cells with no recorded basal area are dropped. Unshaded ground is unmapped or non-productive. Projection and scale as in Figure 1.

Figure 3. Attack against elevation. The unimodal response, shown for all cells and for host-bearing cells separately, against the monotone rise of the wind surface over the same gradient.

Figure 4. The winter temperature record. The shape of the winter by elevation band, the year-to-year sequence of minima against the stage-specific lethal ranges, and the ordering of minima by station elevation.

# Supporting information {#sec-review}

Appendix S1 carries the literature review below with its numbered findings and search protocol, together with Tables S1 to S3 and Figures S1 and S2. Supporting information is supplied as separate files and hosted without editing or typesetting.

The review is written against this study's four objectives rather than as a survey of the species, and its findings are numbered R1 to R5 and cited by number in the main text. The protocol, screening decisions and per-study attribute extraction are recorded in `02.inputs/literature/review-protocol.md`; every count reported here is computed at render time from `02.inputs/literature/review-screening.csv`.

The corpus is the central store described below, holding 579 distinct papers at the date of the search. Matching titles and filenames on *beetle*, *Dendroctonus*, *MPB*, *Ips typographus*, *Scolytinae* and *refugi* returned `r N_CAND` candidates. A study was retained if it models bark-beetle-caused tree mortality spatially against terrain, landscape or climate predictors, which is the only design that can address objectives 1, 3 and 4. `r N_INC` studies qualified, spanning `r YR_RANGE[1]` to `r YR_RANGE[2]`, and all `r N_FULL` were read in full text. Excluded studies are not discarded: papers on mechanism and physiology carry the life history the Introduction rests on, and papers on detection carry the error properties of the response variable.

```{r}
#| label: tbl-review
#| echo: true
#| tbl-cap: "The studies retained by the review, with the attributes the objectives turn on, extracted from full text. The wind column records whether a wind term entered the model, not whether the word appears: every occurrence of `wind` in all eleven texts was read in context, and the per-study result is recorded in the review protocol."

inc |>
  transform(Study = ifelse(nzchar(key), key, "Zabihi 2021 (unverified)")) |>
  (\(d) d[order(suppressWarnings(as.integer(d$year))), ])() |>
  transform(Year = year,
            `Wind in model` = ifelse(wind_predictor == "yes", "yes", "no"),
            `Occurrences of \"wind\"` = wind_mentions,
            `Terrain terms` = terrain_terms,
            Pages = n_pages) |>
  subset(select = c("Study", "Year", "Terrain terms", "Wind in model",
                    "Occurrences of \"wind\"", "Pages")) |>
  kable(booktabs = TRUE, row.names = FALSE, align = "llccrr")
```

**R1. The wind-disruption mechanism has been proposed but never fitted.** Of the `r N_INC` studies that model bark beetle attack against terrain, `r N_WIND` enters a wind term, and that one is the companion study on this same landscape whose response is conifer regeneration rather than beetle attack [@murphy2026]. In the remaining `r N_NOWIND`, wind is either absent from the text entirely or occurs only in a reference-list entry or a discussion sentence. @cartwright2018 is the clearest case: the predictor set is slope, topographic position, heat load, compound topographic index and soil bulk density, and wind appears three times, all in the discussion. @krawchuk2020 states the hypothesis and fits no model. The mechanism has therefore been carried in this literature for at least `r YR_RANGE[2] - 2018` years as an explanation, without once being entered as a predictor.

**R2. The host-density half of the hypothesis already has support, and it is confounded.** The one included study designed to identify insect refugia and model their controls found them associated with thinner stands, shaded slopes and convergent topographic positions [@cartwright2018]. Thin stands are exactly where a wind effect would be expected, and also where host is scarce, so the association is consistent with the hypothesis and with its principal alternative.

**R3. Host is never coded as the attribute the life history says matters.** None of the `r N_INC` included studies mentions the 25 cm source-sink threshold or quadratic mean diameter, though the threshold has been in the standard bionomics review since 2003 [@carroll2004bionomics]. Host enters as cover, as basal area, or as a susceptibility class.

**R4. Scale is recognised in this literature but conclusions are not tested against it.** Scale vocabulary appears in six of the `r N_INC` included studies, and the included extents run from a single landscape to the whole of British Columbia. What the review did not find is a study that refits one model across extent, grain and response coding on one dataset and reports whether its own conclusion survives.

**R5. No available wind product reports the quantity the hypothesis is about.** The hypothesis concerns an instantaneous speed, below the canopy, during the July and August flight period, at the height a beetle flies. Gridded products report a long-run mean at 10 m over open ground, and they are produced by downscaling a coarse field over a digital elevation model [@badger2014; @davis2023], which makes them partly a transform of the terrain they are offered alongside.

The corpus is one curated library rather than a systematic search of an abstracting database, so this is a review of what was read, made checkable, rather than a claim about the complete literature. A study that entered a wind predictor and is absent from the store would not have been found, and R1 carries that scope wherever it is used.
