---
# The manuscript's own title, unchanged. It is 57 characters, inside the 95-character
# cap Frontiers in Ecology and the Environment sets in its "Titles and Abstracts"
# section, so nothing in the guidelines requires it to be rewritten.
#
# The subtitle was dropped here until 2026-08-30, because Frontiers in Ecology and the
# Environment makes no provision for one and title plus subtitle is 233 characters against
# its 95-character cap. Seamus reinstated it on that date so every draft carries the same
# title and subtitle. The cap conflict is real and unresolved, and this file will not
# satisfy that guideline until the subtitle is cut or the venue changes.
title: "Testing the wind disruption hypothesis for beetle refugia"
subtitle: "Stand density, terrain and wind as competing controls on mountain pine beetle attack (Dendroctonus ponderosae) in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
keywords:
  - Dendroctonus ponderosae
  - forest disturbance
  - insect outbreaks
  - Landsat time series
  - refugia
  - British Columbia

date: today
keep-md: true

format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: false
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/frontiers-in-ecology-and-the-environment.csl
df-print: kable
---

```{r}
#| label: host-structure-switch
#| echo: false
#| output: false

## Fit on the annual VRI snapshots, one per study year, not the static 2025 composite.
## Set 2026-08-29. Without this the shared preamble falls back to the composite, which
## holds stand structure at its 2025 projection and so grows it forward through the
## outbreak it is meant to predict. That fallback is silent, and it had this file
## reporting a density-by-wind interaction of -0.094 and a diameter peak of 45.9 per
## cent where the Journal of Applied Entomology submission reports -0.049 and 39.7,
## from the same data. Two drafts of one study must not disagree on their own numbers.
USE_VRI_TIMESERIES <- TRUE
```

{{< include _sections/_preamble.qmd >}}

# Abstract {.unnumbered}

Refugia from mountain pine beetle (*Dendroctonus ponderosae*) are hypothesised to form
where tree defences hold: on shaded ground that spares trees water stress, or in thin stands
with few large hosts, where wind may disrupt the aggregation pheromone. None has been fitted
spatially. We tested all three over `r fmt(AREA_HA)` ha of the Selkirk Mountains, British
Columbia, at 30 m across `r EP_EPOCHS` sixteen-day Landsat epochs and `r EP_YEARS` outbreak
years, with a terrain-resolved wind field from hourly station records. Wind disruption held,
conditionally. Stand density interacted negatively with wind, `r be1(EINT[1])` for stem
density and `r be1(EINT[2])` for standing volume, both surviving a within-season spread term;
neither protected alone. A refugium here is a conjunction of stand and weather, not a fixed
place. Host held as a threshold, attack peaking at `r qb("attack_pc", "25-30")` per cent at
25 to 30 cm. Shading failed backwards: shaded ground carried more attack.

```{r}
#| label: word-budget
#| echo: false
#| output: false

## The 2,500-word limit is the constraint this file exists to satisfy, so it is measured
## from the file's own source at render time and the render stops if the body is over.
## This is a different measurement from the post-render stamping script's: that counts
## the rendered document, this counts body prose only, which is what the limit is about.
##
## Headings are counted here although the guidelines do not name them, because the
## guidelines exclude only the abstract, title, authors, references and figure legends.
## Counting them is the conservative reading and costs about forty words.
##
## Note for whoever edits this comment: do not write the stamping script's placeholder
## token in any chunk that echoes. The script rewrites any paragraph containing it as a
## single run, which destroys the line breaks of the listing it appears in.

fee_wordcount <- function(path = "beetle-topography-wind-study-short-short.qmd",
                          from = "# Introduction",
                          to   = "# Open research") {
  src <- paste(readLines(path, warn = FALSE), collapse = "\n")
  src <- sub("^---.*?\n---\n", "", src)                    # YAML
  src <- gsub("<!--.*?-->", "", src)                       # HTML comments
  src <- gsub("```\\{r[^\n]*\n.*?\n```", "", src)          # code chunks and their captions
  ## perl = TRUE on the line-anchored patterns is load-bearing. R's default engine lets
  ## `.` cross newlines and anchors ^ to the whole string, so a heading pattern without
  ## it eats the rest of the manuscript and the guard reports zero and passes. A guard
  ## that cannot fail is not a guard.
  src <- gsub("(?m)^:::.*$", "", src, perl = TRUE)         # div fences
  src <- gsub("`r [^`]*`", "NUM", src)                     # inline code counts as one word
  i <- regexpr(from, src, fixed = TRUE)
  j <- regexpr(to,   src, fixed = TRUE)
  if (i < 0) stop("word-budget: start marker not found: ", from)
  if (j < 0) stop("word-budget: end marker not found: ", to)
  body <- substr(src, i, j - 1)
  body <- gsub("(?m)^#+ \\{.*$", "", body, perl = TRUE)    # unnumbered heading attributes
  n <- length(strsplit(trimws(body), "\\s+")[[1]])
  if (n < 800) stop("word-budget: counted ", n, " words, which cannot be right; ",
                    "the stripping regexes have eaten the body again")
  n
}

fee_refcount <- function(path = "beetle-topography-wind-study-short-short.qmd") {
  src <- paste(readLines(path, warn = FALSE), collapse = "\n")
  keys <- regmatches(src, gregexpr("@[a-z]+[0-9]{4}[a-z0-9]*", src))[[1]]
  length(unique(keys))
}

## The caps a Research Communication is held to, re-read live from the guidelines page on
## 2026-08-29: "Not exceed 2500 words in the main text", "Include an abstract with a
## maximum of 150 words", "Contain a maximum of 25 references". They live here rather than
## in the shared preamble because the preamble also serves the Journal of Applied
## Entomology draft, whose caps are 6000 and 300, and it no longer defines these at all.
WC_LIMIT  <- 2500L
REF_LIMIT <- 25L

WC_BODY <- fee_wordcount()
N_REF   <- fee_refcount()

cat(sprintf("Main text (Introduction to Conclusions): %d / %d words; %d / %d references\n",
            WC_BODY, WC_LIMIT, N_REF, REF_LIMIT), file = stderr())
if (WC_BODY > WC_LIMIT)
  stop(sprintf("Over the Research Communication limit by %d words.", WC_BODY - WC_LIMIT))
if (N_REF > REF_LIMIT)
  stop(sprintf("Over the 25-reference limit by %d.", N_REF - REF_LIMIT))
```

# Introduction

Mountain pine beetle (*Dendroctonus ponderosae* Hopkins) has killed more lodgepole pine
(*Pinus contorta* Douglas ex Loudon) in British Columbia than any other agent on record
[@taylor2003; @sambaraju2021]. Where a landscape is not killed uniformly, the surviving
patches shape what the forest becomes, and the framework for them is disturbance refugia,
places buffered from disturbance over time [@krawchuk2020]. Explaining refugia rather than
locating them needs mechanistic links to terrain, soil and stand structure [@cartwright2018].

@krawchuk2020 proposed such a mechanism: refugia could occur "in areas with cooler
temperatures (eg from topographic shading) that protect trees from water stress; in areas
with lower host density, allowing for greater wind disruption of beetle pheromone
communication and more vigorous tree growth and chemical defenses; and in areas with fewer
large-diameter host trees". Three mechanisms, two of them terrain, and none has been fitted
spatially. Of the studies our screen retained, only a companion study on this landscape
enters a wind term, and its response is conifer regeneration rather than attack [@murphy2026].

Why a plume should matter at all is a population argument. A lodgepole pine resists a
single beetle through resin flow and an induced response that saturates the phloem with
terpenes, so overwhelming that defence takes the near-simultaneous arrival of many
individuals. Arrival is coordinated chemically, pioneer females oxidising a host
monoterpene to *trans*-verbenol while arriving males release *exo*-brevicomin, and
colonisation is normally complete within one to two days [@safranyik2006chap1]. Because
that signal is airborne it is in principle susceptible to wind, and that susceptibility is
the premise under test. Stem size sets a second constraint: bark thickness, phloem
thickness and moisture retention all rise with diameter, and @carroll2004bionomics put the
consequence as a threshold, pine at or below 25 cm being on average beetle sinks and larger
trees sources.

The wind half is a claim about canopy, not open air. @cartwright2018 found insect refugia in
stands of low basal area: "thinner stands also increase wind penetration, helping to disperse
beetle pheromones and disrupt chemical communications needed to coordinate attacks."
@powell2014 state the converse as the condition for outbreak, "higher local host density,
which minimizes pheromone plume dispersion, reduces wind, and promotes successful switching
to nearby hosts". A model that controls stand density out and then reads a terrain
coefficient as a wind effect has removed the pathway it set out to test.

Terrain also acts through heat. Flight is thermally gated at 19 to 41 degrees C, and "most
flights occur on bright sunny days, and peak flight is in the early to mid afternoon"
[@safranyik2006chap1; @mccambridge1971]. Radiation during the flight window is therefore a
different quantity from radiation across the season, and the season quantity is the one the
shading pathway concerns. Cool sites also push the beetle toward a two-year cycle
[@sambaraju2021], so shading and tree vigour predict the same sign by independent routes
[@bartos1989].

No product reports the wind the hypothesis concerns, an instantaneous below-canopy speed at
flight height during the flight period. Gridded climatologies report a long-run mean at 10 m
over open ground, downscaled over a digital elevation model, which makes them partly a
transform of the terrain offered alongside them [@badger2014; @davis2023]. Nor does the
literature state the interval over which wind should be summarised. We therefore take wind
from station observations of each 16-day period, modify it over the terrain, and test the
response at Landsat's own 16-day repeat rather than annually. Appendix S1 gives the direction
expected of every attribute entered and the reasoning behind it.

# Methods

@tbl-inventory sets out the datasets this study combines and the resolution of each.
The asymmetry it shows is the one the analysis turns on: the response varies every sixteen
days, the inventory once a year, the station wind every hour, and the terrain not at all.

```{r}
#| label: tbl-inventory
#| echo: false
#| tbl-cap: "The datasets this study combines, with the structure and resolution of each. Every count is read from the files at render time. Spatial resolution is the grid the variable is analysed on; temporal resolution is the interval at which it varies. Note that the response varies every 16 days, the inventory once a year and the terrain not at all, which is the asymmetry the wind analysis turns on."

data_inventory(BC, SA) |>
  save_tbl("table-1-inventory") |>
  kable(booktabs = TRUE, align = "llllllr", row.names = FALSE)
```


The study area is `r fmt(AREA_HA)` ha of the Selkirk Mountains in southeastern British
Columbia, `r fmt(N_CELL)` cells at 30 m in EPSG:`r EPSG`, spanning `r fmt(ELEV_R[1])` to
`r fmt(ELEV_R[2])` m (Figure 1). The grid and the perimeter are the parent study's
[@murphy2026], anchored on the 2015 Mt Midgeley fire and expanded to the elevation band that
site occupies, because 480 ha cannot carry the stand-density contrast the mechanism runs on.

The response is moderate-to-high beetle disturbance, classified from Landsat normalised
difference moisture index by the parent study's method. It covers `r length(YRS)` outbreak
years, `r paste(range(YRS), collapse = " to ")`, excluding 2012, the one year flown only by
Landsat 7 with its scan-line corrector off. Pooled prevalence is
`r sprintf("%.1f", 100*PREV)` per cent over `r fmt(N_ROWS)` cell-years (Appendix S3). Years
are never unioned: unioning destroys the year-to-year spread this study measures.

Stand structure comes from the provincial Vegetation Resources Inventory (Appendix S4). Six
attributes carry the mechanism and the stem-size threshold: basal area, crown closure, live
stems per hectare, quadratic mean diameter over stems 12.5 cm and up, stand age, and
susceptible pine basal area. We described terrain with `r length(GEO_V)` geomorphometric surfaces, computed
with SAGA GIS over the full elevation model and clipped afterwards, so a search radius near
the boundary sees real ground. They cover exposure, shape, landform and aspect, the last
entering as northward and eastward components and as the heat load index of @mccune2002. Radiation is computed twice, because two mechanisms need it:
flight-window insolation over 1 July to 15 August restricted to 12:00 to 17:00, the hours
@safranyik2006chap1 identify as the flight peak, and growing-season insolation as the
whole-day total from 1 May to 30 September. A single annual index cannot separate them, the
flight-window surface spanning a
`r sprintf("%.0f", max(d$solar_flight_direct)/min(d$solar_flight_direct))`-fold range here
against `r sprintf("%.1f", max(d$solar_season_total)/min(d$solar_season_total))`-fold for the
season total.

```{r}
#| label: micromet-load
#| echo: false
#| output: false

## The MicroMet summary the paragraph below reads. It is loaded here rather than in the
## shared preamble because the preamble does not carry it: the Journal of Applied
## Entomology draft loads it in its own @sec-micromet chunk, and this file inherited the
## sentence without the chunk. Only `spatial_range` is used here, so the direction rasters
## the sibling chunk also reads are not opened.
mmsum <- read.csv(file.path(BC, "covariates", "wind-micromet", "micromet_summary.csv"))
```

Station wind interpolated from four to seven valley stations is nearly flat within a year, so
wind is also computed as a field varying in space, using the MicroMet model of @liston2006.
Its wind component is seven equations rather than a program, implemented directly. Terrain
slope, azimuth and curvature give a weighting factor $W_w = 1 + \gamma_s\Omega_s +
\gamma_c\Omega_c$ with $\gamma_s = \gamma_c = 0.5$, a modified speed $W_t = W_wW$, and a
direction diversion $\theta_d = -0.5\,\Omega_s\sin[2(\xi-\theta)]$. $W_w$ depends on
direction and not speed, so it is computed once for each of 16 direction bins and each hourly
observation multiplied by the surface for its own bin; nothing is averaged before the terrain
acts on it. The field varies `r sprintf("%.1f", min(mmsum$spatial_range))` to
`r sprintf("%.1f", max(mmsum$spatial_range))` km/h across the grid within a year, and is not
a repackaged terrain index, correlating `r cc("mm_flight_mean", "wind_effect")` with the
windward-leeward index.

`r N_CAND` candidate variables in six groups, one per pathway, passed a four-stage selection
required to retain at least one variable from each group, so a collinearity filter could not
delete a hypothesis. Fitting used a class-balanced sample of `r fmt(N_BAL)` cell-years,
because an unbalanced fit at `r sprintf("%.0f", 100*PREV)` per cent prevalence reports the
intercept rather than the covariates. Four annual logistic models each add one claim to the
last: host size, shading and landform (M0), stand density (M1), terrain shape, exposure and
flight radiation (M2), and the interactions (M3). Two further annual models add previous-year
attack and 90 m neighbourhood pressure after the autologistic design used for this province
(M4), and replace station wind with the terrain-resolved field (M5). The 16-day models (E1,
E2) refit the same covariates over `r EP_EPOCHS` epochs, E2 adding the previous epoch of the
same season. Model comparison is in Appendix S2.

# Results

Every pathway earns its place. AIC falls `r fmt(aic$AIC[1] - aic$AIC[2])` when stand density
enters, `r fmt(aic$AIC[2] - aic$AIC[3])` more with terrain and flight radiation, and
`r fmt(aic$AIC[3] - aic$AIC[4])` more with the interactions.

Stand basal area carries the density pathway (Table 1), at `r b3(DENS[1])` log-odds per
standard deviation (p = `r p3(DENS[1])`). Live stems per hectare carries
`r b3("VRI_LIVE_STEMS_PER_HA")` in the same direction, while crown closure alone runs the
other way at `r b3("CROWN_CLOSURE")` (p = `r p3("CROWN_CLOSURE")`). Standing volume does not
survive selection once each year carries its own inventory. As main effects the density
measures therefore disagree, which is a weaker reading than it first appears, because the
mechanism's prediction is not about a main effect at all.

Radiation enters once. Flight-window direct radiation carries
`r b3("solar_flight_direct")`, which is the thermal gate on flight. Growing-season radiation
did not survive selection, so the shading pathway is carried by northness, with which it
correlates `r cc("solar_season_direct", "northness")`. Northness is `r b3("northness")`:
shaded, north-facing ground carries more attack, against the prediction. Terrain predicts
attack independently of both, the windward-leeward index at `r b3("wind_effect")` and
elevation at `r b3("elevation")`, while terrain ruggedness is `r b3("tri")` and not
distinguishable from zero (p = `r p3("tri")`).

A terrain index is partly an insolation measurement. Fitted without flight-window radiation,
stand density interacts with terrain exposure at `r bnr(I_GEO)`. With radiation in the model
that interaction falls to `r b3(I_GEO)` (p = `r p3(I_GEO)`) and density interacts with
radiation at `r b3(paste0(DENS[1], ":", FSUN[1]))`. On a range whose prevailing bearing is
`r sprintf("%.0f", WDIR)` degrees the windward slopes face west, which are also the slopes
taking afternoon sun, so about two fifths of the apparent exposure effect was the sun. What
remains is not distinguishable from zero, so the annual model neither supports the mechanism
nor rejects it. The next paragraph shows why.

Temporal resolution decided the outcome (Figure 2). At 16 days, stand density interacts with
terrain-resolved wind at `r be1(EINT[1])` for stem density (p = `r pe1(EINT[1])`) and
`r be1(EINT[2])` for standing volume (p = `r pe1(EINT[2])`), both negative, which is the
mechanism's signature: a dense stand is worth less where and when the wind blows harder. Both
survive entering the previous epoch of the same season, which carries
`r be2("ep_lag_nbr90")`, so the result is not the outbreak's own spread wearing a wind
coefficient, and they discriminate at `r sprintf("%.3f", AUCE1)` and
`r sprintf("%.3f", AUCE2)` respectively. With one map a year, the same wind field,
covariates and classification
threshold give `r b5(INT_MM)` (p = `r p5(INT_MM)`), the opposite sign. Nothing about the
landscape or the wind product changed between the two. What changed is that an annual
response can only ask whether windy summers carry less attack than calm ones, and a summer
contains both windy and calm weeks.

Attack also recurs where it was, by a margin that changes how every other coefficient reads.
Entering persistence and 90 m spread separately gives `r b4("lag_self")` and
`r b4("lag_nbr90")` and raises discrimination from `r sprintf("%.3f", AUC3l)` to
`r sprintf("%.3f", AUC4)`. Which environmental terms survive is the informative part, and
the landform terms do not: valley depth keeps `r ret("valley_depth")` of its value while
mid-slope position moves to `r ret("midslope_position")`, growing rather than shrinking. Part
of what the landform terms measured was where the outbreak had already been.

Host coded as diameter holds as a threshold rather than a slope. Attack peaks in the 25 to
30 cm class at `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent,
stepping up from `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "20-25"])` per cent
below the 25 cm source-sink boundary of @carroll2004bionomics and falling to
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == ">40"])` per cent above 40 cm
(Appendix S5). The linear coefficient is negative, `r b3("QUAD_DIAM_125")`, because a
straight line through a humped response returns the slope of its falling limb.

# Conclusions

Of the three mechanisms @krawchuk2020 name, two hold at the resolution the sensor and the
insect share and one does not. Host density holds in the conditional form the mechanism
specifies rather than as a bare main effect, which is the important distinction: as main
effects the three density measures disagree in sign, but what a dense canopy buys shrinks as
the wind rises, and that is the claim. Large-diameter host holds as a threshold at 25 to
30 cm. Topographic shading fails, and it fails on a surrogate: growing-season radiation did
not survive selection, so the pathway is carried by northness, which correlates
`r cc("solar_season_direct", "northness")` with it. Northness tests aspect, and aspect
carries more than shade, so this is the least secure of the three verdicts.

The methodological finding is the more transferable one, and it has three parts. A terrain
index is not an environmental measurement: entering flight-window radiation cut the density
by terrain-exposure interaction from `r bnr(I_GEO)` to `r b3(I_GEO)`. A landform variable in
a spreading outbreak records the outbreak's own history: entering previous-year pressure left
valley depth at `r ret("valley_depth")` of its value while mid-slope position moved to
`r ret("midslope_position")`. And an annual response cannot test a mechanism that operates over weeks: the
same wind field, covariates and threshold gave `r b5(INT_MM)` annually against
`r be1(EINT[1])` at sixteen days. A test that aggregates past the flight period is not a
weaker version of the right test. It is a different question, and this study answers it in
the opposite direction.

Since no previous study has fitted a wind term at all, nothing establishes the resolution at
which one should be fitted. Our answer is that it must be no coarser than the flight period.
For managers reading refugia maps built from annual disturbance products, that matters
directly: a map that shows no wind signal may be showing the calendar rather than the
landscape.

Four limits stay attached. Flight-window radiation and terrain exposure are not separable
here, correlating `r cc("solar_flight_direct", "wind_effect")`, so the positive flight
radiation coefficient is consistent with the thermal gate and with exposure alike, and the
annual density by exposure interaction is not distinguishable from zero once radiation
enters. The wind field is modelled from station data rather than measured on the ridge, so
what is established is that the modelled field behaves as the mechanism requires, not that
the air did. Matching the inventory to its own year is what lets basal area carry the density
pathway at all, at a univariate
`r sprintf("%.3f", uni$auc[uni$term == "BASAL_AREA"])`; held at the 2025 composite it is
indistinguishable from chance. And every conclusion rests on one
mountain range over `r EP_YEARS` years and `r EP_EPOCHS` epochs.

# Open research

This study complies with the Ecological Society of America's Open Research Policy. Every
dataset is public and none was collected by the authors. Beetle disturbance is classified
from Landsat Collection 2 Level-2 surface reflectance. Stand structure is the provincial
Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved from the British
Columbia Data Catalogue web feature service. Terrain derives from the Natural Resources
Canada High Resolution Digital Elevation Model [@nrcan2017], with geomorphometry computed by
SAGA GIS. Wind is Environment and Climate Change Canada hourly station data. All processing
scripts and the derived model tables that reproduce every number reported here will be
deposited in Dryad on acceptance, and are available to reviewers on request.

```{r}
#| label: figure-1
#| echo: false
#| fig-cap: "Landscape, terrain and stand surfaces across the study area, all EPSG:3153 at 30 m over Esri World Shaded Relief. (a) elevation; (b) terrain ruggedness index; (c) windward-leeward index at the prevailing bearing; (d) flight-window direct radiation, the thermal gate on flight; (e) growing-season direct radiation, the shading pathway; (f) the MicroMet wind weighting factor at the prevailing bearing; (g) stand basal area; (h) quadratic mean diameter, whose 25 cm source-sink threshold falls near the midpoint of the scale; (i) live stems per hectare. The white outline is the study perimeter and the red outline the 2015 Mt Midgeley burn. Contours are at 200 m. Every panel carries its own north arrow and the same numerical scale, 1:150,000, which holds at a printed panel width of 66 mm."
#| fig-height: 8.4
#| fig-width: 7.5

geo   <- rast(file.path(SA, "geomorphometry_context.tif"))
vri   <- rast(file.path(SA, "vri_context.tif"))
elevc <- rast(file.path(SA, "dem_context.tif"))
ctx   <- map_context(per, burn, SA)
## The unmasked weighting from 49-context-covariates.R, not the stored per-bin stack,
## which is masked to the perimeter and would leave this the one panel with a hole in it.
mmb   <- rast(file.path(SA, "micromet_weight_context.tif"))

m <- function(r, title, pal = "viridis")
  academic_map(r, title, ctx, pal, base_size = 7)

(m(elevc, "(a) Elevation (m)") |
 m(geo[["tri"]], "(b) Terrain ruggedness index", "magma") |
 m(geo[["wind_effect"]], "(c) Windward-leeward index", "cividis")) /
(m(geo[["solar_flight_direct"]], "(d) Flight-window radiation (kWh/m2)", "inferno") |
 m(geo[["solar_season_direct"]], "(e) Growing-season radiation (kWh/m2)", "inferno") |
 m(mmb, "(f) MicroMet wind weighting", "mako")) /
(m(vri[["BASAL_AREA"]], "(g) Stand basal area (m2/ha)", "mako") |
 m(vri[["QUAD_DIAM_125"]], "(h) Quadratic mean diameter (cm)", "viridis") |
 m(vri[["VRI_LIVE_STEMS_PER_HA"]], "(i) Live stems per hectare", "viridis"))
```

```{r}
#| label: figure-2
#| echo: false
#| fig-cap: "**Figure 2.** The refugia mechanism as fitted at 16 days. (a) Predicted probability of moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem density, all other terms at their means; the lines cross, so wind raises attack in thin stands and lowers it in dense ones, which is the interaction. (b) The same for standing volume. (c) Coefficients of the 16-day model, with and without the within-season spread term. (d) Epoch prevalence against epoch mean wind, one point per epoch."
#| fig-height: 6

nd_base <- be[1, , drop = FALSE]
for (v in EV) nd_base[[v]] <- 0
pred_panel <- function(dv, lab) {
  qs <- quantile(be[[dv]], c(0.1, 0.9))
  gr <- expand.grid(w = seq(min(be$ep_wind_mean), max(be$ep_wind_mean), length.out = 60),
                    d = qs)
  nd <- nd_base[rep(1, nrow(gr)), ]
  nd$ep_wind_mean <- gr$w; nd[[dv]] <- gr$d
  nd$geomorphon <- factor(levels(be$geomorphon)[1], levels = levels(be$geomorphon))
  gr$p <- predict(E1, newdata = nd, type = "response")
  gr$Density <- factor(gr$d, labels = c("low (10th pct)", "high (90th pct)"))
  ggplot(gr, aes(w, p, colour = Density)) + geom_line(linewidth = 0.8) +
    scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
    labs(x = "Epoch wind (SD units)", y = "P(moderate-to-high)", title = lab) +
    theme_minimal(base_size = 9)
}
pa <- pred_panel("VRI_LIVE_STEMS_PER_HA", "(a) Stem density")
pb <- pred_panel("LIVE_STAND_VOLUME_125", "(b) Standing volume")

cf <- bind_rows(
  ce1 |> mutate(model = "E1 all epochs"),
  ce2 |> mutate(model = "E2 with lag")) |>
  filter(!grepl("^geomorphon|Intercept", term)) |>
  group_by(term) |> filter(any(abs(beta) > 0.05)) |> ungroup()
pc <- ggplot(cf, aes(reorder(term, beta), beta, colour = model)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_pointrange(aes(ymin = beta - 1.96*se, ymax = beta + 1.96*se),
                  position = position_dodge(width = 0.5), size = 0.25) +
  coord_flip() + scale_colour_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = NULL, y = "Log-odds per SD", title = "(c) 16-day model coefficients") +
  theme_minimal(base_size = 8) + theme(legend.position = "bottom")

pd <- epsum |>
  inner_join(epwind, by = c("year","epoch")) |>
  ggplot(aes(mean_kmh, prevalence)) +
  geom_point(aes(colour = factor(year)), size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.5) +
  labs(x = "Epoch mean wind (km/h)", y = "Epoch prevalence",
       colour = "Year", title = "(d) Epochs") +
  theme_minimal(base_size = 8)

(pa | pb) / (pc | pd)
```

```{r}
#| label: table-1
#| echo: false
#| tbl-cap: "**Table 1.** The terms the three mechanisms turn on, and the specification contrasts the paper rests on. Coefficients are log-odds per standard deviation on a class-balanced sample. The annual models are M3, M3 without flight-window radiation, and M5; the 16-day models are E1 and E2. The three mechanisms are those of Krawchuk et al. (2020). Growing-season direct radiation did not survive variable selection, so the shading pathway is represented by northness."

## Built from the fitted objects through the same accessors the prose uses, so no
## coefficient in this table is transcribed by hand. A term absent from a model returns
## an empty string rather than an error, which is the visible failure mode.
row <- function(block, label, beta, z, p)
  data.frame(Block = block, Term = label, Beta = beta, z = z, p = p)

## A term absent from the model it is asked for returns an empty cell rather than an error,
## which is visible in the rendered table. term-guard above is what makes the absence itself
## fail the render, so an empty cell here can only mean a term the guard does not police.

g <- function(f, v) { x <- f(v); if (length(x)) x else "" }

tbl1 <- bind_rows(
  ## Growing-season direct radiation did not survive selection, so the shading pathway
  ## is represented by northness, its strongest correlate among the surfaces that did.
  row("Mechanism 1: shading", "Northness (shading surrogate)",
      g(b3, "northness"), g(z3, "northness"), g(p3, "northness")),
  row("Mechanism 2: host density", "Standing volume",
      g(b3, "LIVE_STAND_VOLUME_125"), g(z3, "LIVE_STAND_VOLUME_125"), g(p3, "LIVE_STAND_VOLUME_125")),
  row("Mechanism 2: host density", "Live stems per hectare",
      g(b3, "VRI_LIVE_STEMS_PER_HA"), g(z3, "VRI_LIVE_STEMS_PER_HA"), g(p3, "VRI_LIVE_STEMS_PER_HA")),
  row("Mechanism 2: host density", "Crown closure",
      g(b3, "CROWN_CLOSURE"), g(z3, "CROWN_CLOSURE"), g(p3, "CROWN_CLOSURE")),
  row("Mechanism 3: host size", "Quadratic mean diameter",
      g(b3, "QUAD_DIAM_125"), g(z3, "QUAD_DIAM_125"), g(p3, "QUAD_DIAM_125")),
  row("Mechanism 3: host size", "Susceptible pine basal area",
      g(b3, "PINE_BA"), g(z3, "PINE_BA"), g(p3, "PINE_BA")),
  row("Terrain", "Flight-window direct radiation",
      g(b3, "solar_flight_direct"), g(z3, "solar_flight_direct"), g(p3, "solar_flight_direct")),
  row("Terrain", "Terrain ruggedness index",
      g(b3, "tri"), g(z3, "tri"), g(p3, "tri")),
  row("Terrain", "Windward-leeward index",
      g(b3, "wind_effect"), g(z3, "wind_effect"), g(p3, "wind_effect")),
  row("Terrain", "Elevation",
      g(b3, "elevation"), g(z3, "elevation"), g(p3, "elevation")),
  row("Radiation and exposure", "Density x terrain exposure, no radiation",
      g(bnr, I_GEO), g(znr, I_GEO), g(pnr, I_GEO)),
  row("Radiation and exposure", "Density x terrain exposure, with radiation",
      g(b3, I_GEO), g(z3, I_GEO), g(p3, I_GEO)),
  row("Resolution decides the sign", "Density x wind, annual (M5)",
      g(b5, INT_MM), g(z5, INT_MM), g(p5, INT_MM)),
  row("Resolution decides the sign", "Stem density x wind, 16-day (E1)",
      g(be1, EINT[1]), g(ze1, EINT[1]), g(pe1, EINT[1])),
  row("Resolution decides the sign", "Standing volume x wind, 16-day (E1)",
      g(be1, EINT[2]), g(ze1, EINT[2]), g(pe1, EINT[2])),
  row("Resolution decides the sign", "Standing volume x wind, 16-day with lag (E2)",
      g(be2, EINT[2]), g(ze2, EINT[2]), g(pe2, EINT[2])))

tbl1$Block[duplicated(tbl1$Block)] <- ""
kable(tbl1, booktabs = TRUE, align = "llrrr",
      col.names = c("", "Term", "Beta", "z", "p"))
```

# Supporting information {.unnumbered}

Five components accompany this paper and are supplied as stand-alone documents at
submission. They are reproduced below for review.

## Appendix S1

```{r}
#| label: appendix-s1
#| echo: false
#| tbl-cap: "**Appendix S1.** Landscape attributes entered in this study, the direction expected of each, and the reasoning behind that expectation."

tibble::tribble(
  ~Attribute, ~Expected, ~Rationale,
  "Stand basal area, volume, crown closure, stems", "positive",
  "Denser canopy holds the pheromone plume together; thinner stands admit wind that disperses it [@krawchuk2020; @cartwright2018; @powell2014].",
  "Quadratic mean diameter", "positive above 25 cm",
  "Trees under 25 cm are beetle sinks, over 25 cm are sources [@carroll2004bionomics].",
  "Lodgepole pine cover and basal area", "positive",
  "Attack cannot occur where the host is absent; a cell without pine is not a refugium [@cartwright2018].",
  "Flight-period direct radiation", "positive",
  "Flight is gated between 19 and 41 degrees C and peaks on bright afternoons, so sunlit slopes are reachable [@safranyik2006chap1; @mccambridge1971].",
  "Growing-season direct radiation", "negative",
  "Shading cools, relieving water stress and funding defence, and cool sites also push the beetle toward a two-year cycle [@krawchuk2020; @sambaraju2021].",
  "Terrain exposure to wind", "negative",
  "Wind disrupts the pheromone plume, so exposed ground should carry less attack [@krawchuk2020].",
  "Terrain ruggedness", "uncertain",
  "The strongest terrain term in the companion study on this landscape, but fitted there against regeneration rather than attack [@murphy2026].",
  "Convergence, wetness, valley depth, slope position", "positive in convergent terrain",
  "Infested groups gather in draws and gullies, and deep snow insulates overwintering brood [@safranyik2006chap1].",
  "Flight-period wind speed", "negative",
  "The mechanism acts during dispersal, and no study specifies the interval, so it is taken at the flight period [@krawchuk2020; @jones2019].",
  "Elevation", "uncertain",
  "A composite of temperature, snowpack, season length and host distribution that this design cannot separate [@sambaraju2021]."
) |>
  kable(booktabs = TRUE, align = "lll", row.names = FALSE)
```

## Appendix S2

```{r}
#| label: appendix-s2
#| echo: false
#| tbl-cap: "Model comparison. Each model adds one pathway to the previous. AIC ranks on likelihood and a parameter penalty; the remaining columns are predictive error on the fitted probabilities. RMSE is the root mean squared error and is the square root of the Brier score; RMSE (%) expresses it against the prevalence of the response. Brier skill is the improvement over predicting the prevalence for every cell, where 0 is no better than the base rate. MAPE and Theil's U are not reported: both divide by the observed value, which is zero for the majority class of a binary response, so both are undefined without discarding that class."

fit_metrics_table(
  fit_metrics(M0, "M0 host size + shading + landform"),
  fit_metrics(M1, "M1 + stand density"),
  fit_metrics(M2, "M2 + terrain and flight radiation"),
  fit_metrics(M3, "M3 + interactions")) |>
  kable(booktabs = TRUE, align = "lrrrrrrrr", row.names = FALSE)
```

## Appendix S3

```{r}
#| label: appendix-s3
#| echo: false
#| tbl-cap: "**Appendix S3.** Moderate-to-high beetle disturbance by year inside the study perimeter."

prev_yr |>
  transmute(Year = year, Cells = fmt(n),
            `Moderate-to-high (%)` = sprintf("%.1f", 100*prev)) |>
  kable(booktabs = TRUE, align = "lrr", row.names = FALSE)
```

## Appendix S4

```{r}
#| label: appendix-s4
#| echo: false
#| tbl-cap: "Stand structure across the study perimeter, from the Vegetation Resources Inventory. n is cell-years. SD is the standard deviation of the landscape; SE is the standard error of the mean and is small by construction at this n, so it is not precision about any one cell. Skewness and kurtosis are the bias-corrected third and fourth standardised moments; kurtosis is excess, so 0 is Gaussian."

describe_vars(d, VRI_V, VRI_LABELS) |>
  describe_table() |>
  kable(booktabs = TRUE, align = "lrrrrrrrrr", row.names = FALSE)
```

## Appendix S5

```{r}
#| label: appendix-s5
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced sample. The 25 cm boundary is the source-sink threshold of the species' bionomics. Intervals are Wilson score intervals on the class proportion. Note that 30 m cells in a spreading outbreak are not independent, so the accompanying tests are anti-conservative."

qmd_bin |>
  diameter_table() |>
  kable(booktabs = TRUE, align = "lrrrr", row.names = FALSE)
```

# References
