---
# THE SUBMISSION. Journal of Applied Entomology, Original Article.
#
# Chosen on 2026-08-28 over beetle-topography-wind-study-short.qmd, which is archived. That
# draft held stand structure at the 2025 WFS composite, so every year of a cell carried the
# same basal area, volume, stems and diameter and those values had been grown forward
# through the outbreak they were meant to predict. It reported a density by wind interaction
# roughly twice the size of the one here.
#
# This draft fits every model on the annual VRI snapshots, one per study year, and reports
# the smaller effect. It also discriminates better: AUC 0.6730 against 0.6685.
#
# Body 5,797 words against the 6,000 limit; abstract 295 against 300 after the 2026-08-29 redraft.
title: "Testing the wind disruption hypothesis for beetle refugia"
subtitle: "Stand density, terrain and wind as competing controls on mountain pine beetle attack (Dendroctonus ponderosae) in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
# Journal of Applied Entomology requires exactly six keywords and states that they
# "should not include words in the title". The title carries testing, wind, disruption,
# hypothesis, beetle and refugia, so none of those may appear here, which is why the
# insect is named by its binomial.
#
# ONE DELIBERATE BREACH. "refugia" is a keyword and it is also in the title. Seamus
# instructed this on 2026-08-30, having been told twice that the rule bars it, after
# rejecting "holdouts" as meaningless. It is legal the moment "refugia" leaves the title.
#
# Set on 2026-08-30 from the keyword lists of the target literature, read from the PDFs.
# "forest disturbance" is Meddens et al. 2012 and 2013 and Meddens and Hicke 2014;
# "Landsat time series" is Meddens and Hicke 2014 and Liang et al. 2016; "British Columbia"
# is Meddens et al. 2012; "Dendroctonus ponderosae" is Johnson et al. 2026. "insect outbreaks" is from the title and body of Krawchuk et al.
# 2020, which prints no keyword list at all.
#
# "mountain pine beetle", "bark beetle" and "disturbance refugia" are the three keywords
# this literature would most expect and all three are barred by the title. Using them
# requires dropping "beetle" and "refugia" from the title.
keywords:
  - Dendroctonus ponderosae
  - forest disturbance
  - insect outbreaks
  - Landsat time series
  - refugia
  - British Columbia

date: today
keep-md: true

format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: false
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
# Journal of Applied Entomology requires APA 6th: "References should be prepared
# according to the Publication Manual of the American Psychological Association (6th
# edition)". Both drafts carried the Springer style until 2026-08-28, and apa-6th-edition.csl
# had been sitting unused in 04.references since 20 August. Note that apa.csl in the same
# folder is APA FIFTH edition despite its name.
csl: ../04.references/apa-6th-edition.csl
df-print: kable
---

```{r}
#| label: host-structure-switch
#| echo: false
#| output: false

## This draft is the year-matched refit. The shared preamble reads this switch and fits
## every model on the annual VRI snapshots instead of the 2025 composite, so the abstract,
## the Results all report the same models. Remove it and this file silently
## becomes a copy of the Journal of Applied Entomology draft with two extra sections that
## disagree with the rest of it.
USE_VRI_TIMESERIES <- TRUE
```

{{< include _sections/_pipeline.qmd >}}

{{< include _sections/_preamble.qmd >}}

```{r}
#| label: vri-refit
#| echo: false
#| output: false

## The two comparisons this draft exists to report, read from the files the scripts wrote
## rather than retyped. 55-build-vri-timeseries-table.R builds the year-matched tables and
## 56-refit-vri-timeseries.R fits both models; 52- and 53- do the same for the wind window.
VRIS <- read.csv(file.path(BC, "model-data", "vri_refit_comparison.csv"))
VSRC <- read.csv(file.path(BC, "model-data", "vri_year_source.csv"))
FWS  <- read.csv(file.path(BC, "model-data", "flight_window_sensitivity.csv"))

vr <- function(m, col) VRIS[[col]][VRIS$model == m]
b  <- function(x) sprintf("%+.4f", x)
pp <- function(x) if (x < 0.001) "< 0.001" else sprintf("= %.3f", x)

VRI_YEARS <- nrow(VSRC)
VRI_SUB   <- VSRC$year[VSRC$substituted]
VRI_SUBN  <- length(VRI_SUB)
```

# Abstract {.unnumbered}

```{r}
#| label: abstract-wordcount
#| include: false

## The abstract is the last thing measured, because Journal of Applied Entomology
## caps it at 300 words and an abstract over the cap is a desk rejection. The count
## is taken from the rendered text, not from the source, so inline expressions are
## counted as the words they become.
ABS_LIMIT <- 300
```

Refugia from mountain pine beetle (*Dendroctonus ponderosae* Hopkins) are hypothesised to
form where tree defences hold, on shaded ground that spares trees water stress, or in thin
stands with few large hosts, where wind may disrupt the aggregation pheromone or vigour
sustain chemical defence. None has been fitted spatially. All three were tested across
`r fmt(AREA_HA)` ha of the Selkirk Mountains, British Columbia, using `r EP_EPOCHS`
sixteen-day Landsat epochs over `r spellnum(EP_YEARS)` outbreak years, year-matched forest
inventory, `r length(GEO_V)` geomorphometric variables, and a terrain-resolved MicroMet wind
field from hourly station records. The wind disruption hypothesis held, conditionally.
Attack fell where a thin stand and strong flight-period wind coincided, the density-by-wind
interaction `r be1(EINT[1])` for stem density (`r pe1T(EINT[1])`) and `r be1(EINT[2])` for
standing volume (`r pe1T(EINT[2])`), both surviving a within-season contagion term. Neither
protected alone. Wind alone carried marginally more attack (`r be1("ep_wind_mean")`), and
thin stands in still air did not. A refugium here is a conjunction of stand and weather, not
a fixed place. The model cannot separate plume disruption from the vigour that also
accompanies low density. Host size held as a threshold, not a gradient. Attack peaked at
`r qb("attack_pc", "25-30")` per cent in the 25 to 30 cm quadratic mean diameter class, the
source-sink boundary of @carroll2004bionomics, and fell away on both sides. Above that class
pine basal area falls from `r qb("pine_ba", "25-30")` to `r qb("pine_ba", ">40")` square
metres per hectare while total basal area rises, as expected of a pioneer, because the
larger stands are not pine. The rise across `r QMD_THRESH` cm is the size threshold, the
fall above it host scarcity. Temperature and hydrology failed in reverse. North-facing
slopes, which the water-stress mechanism predicts should escape, carried more attack,
`r b3("northness")`. Refugia here were warm, not cool, so terrain wind indices at these
elevations partly carry insolation.

# Keywords {.unnumbered}

Dendroctonus ponderosae, forest disturbance, insect outbreaks, Landsat time series,
refugia, British Columbia

<!-- The journal requires six keywords in the main document file and states that none may
     appear in the title. Quarto's docx writer puts the YAML `keywords` field into the
     file's core properties only, where no editor sees it, so they are printed here as
     well. The two lists must be kept identical. Audited 2026-08-28. -->

# Introduction

Mountain pine beetle (*Dendroctonus ponderosae* Hopkins) has killed more lodgepole pine
(*Pinus contorta* Douglas ex Loudon) in British Columbia than any other agent on record
[@taylor2003; @sambaraju2021]. An outbreak never kills a landscape evenly. The patches that
survive are the source of structure and seed for the next forest, and they are called
disturbance refugia, places buffered from disturbance over time [@krawchuk2020]. Finding a
refugium is not the same as explaining one. Explanation needs a mechanism that ties survival
to something measurable on the ground, such as terrain, soil or the structure of the stand
itself [@cartwright2018].

@krawchuk2020 proposed such a mechanism, that refugia could occur "in areas with cooler
temperatures (eg from topographic shading) that protect trees from water stress; in areas
with lower host density, allowing for greater wind disruption of beetle pheromone
communication and more vigorous tree growth and chemical defenses; and in areas with fewer
large-diameter host trees" (p. 239). These are treated here as three testable hypotheses.
Under H1, topographic shading reduces attack by relieving water stress on cool ground. Under
H2, low host density reduces attack by admitting the wind that disperses the aggregation
pheromone. Under H3, a scarcity of large-diameter hosts reduces attack by limiting brood
production. Two of the three act through terrain, and none has been fitted to a map. Only
one of the studies the screen retained enters a wind term, a companion study on this same
landscape whose response is conifer regeneration rather than beetle attack [@murphy2026].

The wind half is a claim about canopy, not open air. @cartwright2018 is the one retained study designed
to model what controls an insect refugium, and it located refugia in stands of low basal
area, reasoning that "thinner stands also increase wind penetration, helping to disperse
beetle pheromones and disrupt chemical communications needed to coordinate attacks." @powell2014 state the converse as the
condition for outbreak, that "higher local host density, which minimizes pheromone plume
dispersion, reduces wind, and promotes successful switching to nearby hosts, positively
influences outbreak propensity." A model that controls stand density out and then reads a
terrain coefficient as a wind effect has removed the pathway it set out to test.

Terrain acts through more than airflow. Flight has a temperature window. The beetle flies
between 19 and 41 degrees C, mostly between 22 and 32, and @safranyik2006chap1 and
@mccambridge1971 report that "most flights occur on bright sunny days, and peak flight is in
the early to mid afternoon". Radiation during the
flight window is therefore a different quantity from radiation across the season, and the
season quantity is the one the shading pathway concerns. Light, temperature and wind are not
competing explanations but one bundle. Stand density governs "tree vigour and within-stand
microclimate, which in turn influence success of bark beetle dispersal, host selection, attack
or brood development" [@safranyik2006chap1]; @bartos1989 set the two halves of that bundle
against each other in their own title, microclimate against tree vigour. A third route leads
to the same place. Cool sites slow development enough to push the beetle onto a two-year
cycle [@sambaraju2021]. So shade and vigour both predict less attack on cool ground, by
routes this design cannot separate.

Two further constraints decide what a wind term can be measured over, and both are
temporal rather than spatial. The first is the day. @gray1972 recorded emergence in
*D. ponderosae* against ambient temperature and found that on days whose maximum stayed
within optimal limits, emergence "began when temperatures rose above 20' C, then increased
in the morning hours with increasing temperatures, and ceased in the afternoon at about the
same threshold temperature", with "Peak emergence occurred between 11 a. m. and 2 p. m.
when 61 percent (232 out of 379) of the beetles were collected". They also record "the
sharp decline in activity which occurred when ambient temperatures exceeded 30' C". Flight
is therefore gated at both ends and confined to a few hours, and a daily or monthly mean
wind speed averages across hours in which no beetle is flying.

The second is the season, and it is a population constraint rather than an individual one.
Mass attack is a threshold phenomenon. @howe2022 conclude that "a combination of
stand-level spatial aggregation, behavioral shifts, and higher quality of attainable hosts
defines a critical threshold beyond which continual population growth becomes
self-driving", and @cooke2025 define the irruption threshold as "the population density at
which endemic populations may transition towards the epidemic state". @carroll2004bionomics
place the same argument on the thermal side, requiring "thermal environments conducive to
overwintering survival and with sufficient heat accumulation to maintain a synchronous
univoltine life cycle". Two things follow for this design. Attack in one year is not
independent of attack in the year before, which is why previous-year and neighbourhood
pressure enter the models. And an environmental variable that acts on this system acts by
regulating that threshold rather than by adding to attack. @cooke2025 report that "in every
study where an Allee effect was demonstrated, investigators also identified at least one
extrinsic environmental factor (e.g., winter weather, summer drought, microclimatic effect)
that was regulating its strength". A wind effect on mass attack is therefore expected as an
interaction with host density, which is what this study fits, and not as a main effect.

No product reports the wind the hypothesis concerns, an instantaneous below-canopy speed at
flight height during the flight period. Gridded climatologies report a long-run mean at 10 m
over open ground and are downscaled over a digital elevation model, making them partly a
transform of the terrain offered alongside them [@badger2014; @davis2023]. Nor does the
literature state the interval over which wind should be summarised. Wind is therefore taken
from station observations of each 16-day period, modified over the terrain with the MicroMet
model of @liston2006, and the response is measured at Landsat's own 16-day repeat, which is
the finest interval the sensor and the flight period share.

@tbl-hypotheses gives the direction expected of each attribute and the reasoning behind it.
This study asks five questions. Does stand density predict moderate-to-high disturbance, as
the pheromone mechanism requires? Does topographic shading predict it, against the competing
microclimate and adaptive-seasonality arguments? Does terrain shape predict it once density
and radiation are already in the model? Is the density effect conditional on wind, which is
the form the mechanism actually takes?

```{r}
#| label: tbl-hypotheses
#| echo: false
#| tbl-cap: "Landscape attributes entered in this study, the direction expected of each, and the reasoning behind that expectation."

tibble::tribble(
  ~Attribute, ~Expected, ~Rationale,
  "Stand basal area, volume, crown closure, stems", "positive",
  "Denser canopy holds the pheromone plume together; thinner stands admit wind that disperses it [@krawchuk2020; @cartwright2018; @powell2014].",
  "Quadratic mean diameter", "positive above 25 cm",
  "Trees under 25 cm are beetle sinks, over 25 cm are sources [@carroll2004bionomics].",
  "Lodgepole pine cover and basal area", "positive",
  "Attack cannot occur where the host is absent; a cell without pine is not a refugium [@cartwright2018].",
  "Flight-period direct radiation", "positive",
  "Flight is gated between 19 and 41 degrees C and peaks on bright afternoons, so sunlit slopes are reachable [@safranyik2006chap1; @mccambridge1971].",
  "Growing-season direct radiation", "negative",
  "Shading cools, relieving water stress and funding defence, and cool sites also push the beetle toward a two-year cycle [@krawchuk2020; @sambaraju2021].",
  "Terrain exposure to wind", "negative",
  "Wind disrupts the pheromone plume, so exposed ground should carry less attack [@krawchuk2020].",
  "Terrain ruggedness", "uncertain",
  "The strongest terrain term in the companion study on this landscape, but fitted there against regeneration rather than attack [@murphy2026].",
  "Convergence, wetness, valley depth, slope position", "positive in convergent terrain",
  "Infested groups gather in draws and gullies, and deep snow insulates overwintering brood [@safranyik2006chap1].",
  "Flight-period wind speed", "negative",
  "The mechanism acts during dispersal, and no study specifies the interval, so it is taken at the flight period [@krawchuk2020; @jones2019].",
  "Elevation", "uncertain",
  "A composite of temperature, snowpack, season length and host distribution that this design cannot separate [@sambaraju2021]."
) |>
  save_tbl("table-2-hypotheses") |>
  kable(booktabs = TRUE, align = "lll", row.names = FALSE)
```

# Materials

## Study area

@tbl-inventory sets out the datasets this study combines and the resolution of each. The
asymmetry it shows is the one the analysis depends on. The response varies every sixteen
days, the inventory once a year, the station wind every hour, and the terrain not at all.

```{r}
#| label: tbl-inventory
#| echo: false
#| tbl-cap: "The datasets this study combines, with the structure and resolution of each. Every count is read from the files at render time. Spatial resolution is the grid the variable is analysed on; temporal resolution is the interval at which it varies. Note that the response varies every 16 days, the inventory once a year and the terrain not at all, which is the asymmetry the wind analysis depends on."

data_inventory(BC, SA) |>
  save_tbl("table-1-inventory") |>
  kable(booktabs = TRUE, align = "llllllr", row.names = FALSE)
```



The study area is `r fmt(AREA_HA)` ha of the Selkirk Mountains in southeastern British
Columbia, `r fmt(N_CELL)` cells at 30 m, spanning `r fmt(ELEV_R[1])` to
`r fmt(ELEV_R[2])` m and `r fmt(diff(ELEV_R))` m of relief.

The coordinate reference system is EPSG:`r EPSG`, `r CRSNAME`. This is the parent study's
grid, adopted so that results are directly comparable to it. @murphy2026 report that "All
disturbance and covariate rasters were aligned to a common 30 m grid in EPSG:3153, the
native resolution of the Landsat-derived dNBR and dNDMI products", and that "The plot anchor
was georeferenced with a handheld GPS receiver in EPSG:3153".

The perimeter is anchored on the 2015 Mt Midgeley fire, `r fmt(BURN_HA)` ha, which is the
parent study's site. That study reports its site as spanning "elevations from 830 to 1744 m
a.s.l."; the digital elevation model used here returns `r fmt(burn_e[1])` to
`r fmt(burn_e[2])` m inside the same perimeter, which reproduces the statement. The present
study extends beyond that hard boundary, because 480 ha does not contain the stand-density
contrast the pheromone mechanism requires. The extension is not arbitrary. The perimeter is
the burn buffered 5 km and then cut to the elevation band the parent's site occupies, so the
added ground is ecologically the same kind of ground. Without that cut a 1 km buffer already
reaches the Kootenay Lake surface at 534 m, which is not a site elevation.

```{r}
#| label: fig-study-area
#| echo: false
#| fig-cap: "Landscape, terrain and stand surfaces across the study area, all EPSG:3153 at 30 m over Esri World Shaded Relief. (a) elevation; (b) terrain ruggedness index; (c) windward-leeward index at the prevailing bearing; (d) flight-window direct radiation, the thermal gate on flight; (e) growing-season direct radiation, the shading pathway; (f) the MicroMet wind weighting factor at the prevailing bearing; (g) stand basal area; (h) quadratic mean diameter, whose 25 cm source-sink threshold falls near the midpoint of the scale; (i) live stems per hectare. The white outline is the study perimeter and the red outline the 2015 Mt Midgeley burn. Contours are at 200 m. Every panel has its own north arrow and the same numerical scale, 1:150,000, which is correct at a printed panel width of 66 mm."
#| fig-height: 8.4
#| fig-width: 7.5

geo   <- rast(file.path(SA, "geomorphometry_context.tif"))
vri   <- rast(file.path(SA, "vri_context.tif"))
elevc <- rast(file.path(SA, "dem_context.tif"))
ctx   <- map_context(per, burn, SA)
## The unmasked weighting from 49-context-covariates.R, not the stored per-bin stack,
## which is masked to the perimeter and would leave this the one panel with a hole in it.
mmb   <- rast(file.path(SA, "micromet_weight_context.tif"))

m <- function(r, title, pal = "viridis")
  academic_map(r, title, ctx, pal, base_size = 7)

(m(elevc, "(a) Elevation (m)") |
 m(geo[["tri"]], "(b) Terrain ruggedness index", "magma") |
 m(geo[["wind_effect"]], "(c) Windward-leeward index", "cividis")) /
(m(geo[["solar_flight_direct"]], "(d) Flight-window radiation (kWh/m2)", "inferno") |
 m(geo[["solar_season_direct"]], "(e) Growing-season radiation (kWh/m2)", "inferno") |
 m(mmb, "(f) MicroMet wind weighting", "mako")) /
(m(vri[["BASAL_AREA"]], "(g) Stand basal area (m2/ha)", "mako") |
 m(vri[["QUAD_DIAM_125"]], "(h) Quadratic mean diameter (cm)", "viridis") |
 m(vri[["VRI_LIVE_STEMS_PER_HA"]], "(i) Live stems per hectare", "viridis"))
```



## Beetle disturbance


The response is annual moderate-to-high beetle disturbance, classified from Landsat
normalised difference moisture index by the method of the parent study and rebuilt for this
project. It covers `r spellnum(length(YRS))` outbreak years, `r paste(range(YRS), collapse = " to ")`,
excluding 2012, the one year covered only by Landsat 7, which has flown with its scan-line
corrector off since May 2003. Annual prevalence inside the perimeter runs
`r sprintf("%.1f", 100*min(prev_yr$prev))` to `r sprintf("%.1f", 100*max(prev_yr$prev))` per
cent, pooled `r sprintf("%.1f", 100*PREV)` per cent over `r fmt(N_ROWS)` cell-years.

Years are never unioned. Unioning destroys the year-to-year spread this study measures, and
the union layer over the wider grid reached 73 per cent of the landscape, which is not
credible for a beetle outbreak.

The provincial aerial overview survey is not used as a response and supplies no training
label. It is retained only as a visual check.

```{r}
#| label: tbl-prevalence
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by year inside the study perimeter."

prev_yr |>
  transmute(Year = year, Cells = fmt(n),
            `Moderate-to-high (%)` = sprintf("%.1f", 100*prev)) |>
  save_tbl("table-3-prevalence") |>
  kable(booktabs = TRUE, align = "lrr", row.names = FALSE)
```

## Stand structure


Stand structure comes from the provincial Vegetation Resources Inventory layer
`VEG_COMP_LYR_R1_POLY`, retrieved by web feature service over the perimeter. Six attributes
cover the mechanism and the stem-size threshold. These are total basal area, crown closure,
live stems per hectare, quadratic mean diameter over stems 12.5 cm and up, stand age, and
susceptible pine basal area formed as total basal area times the pine share of cover.

Every model is fitted on year-matched stand structure. Each cell-year takes the inventory
snapshot the province published for that year, depleted for harvest and projected for growth
to it, so the host terms vary in time as well as in space.

`r if (VRI_SUBN) sprintf("One year is substituted. The %d delivery lists basal area and live stems as columns and fills neither, 0 and 3 per cent of polygons against 70 to 97 per cent in every other year, so the preceding year is carried forward rather than averaged. The inventory is itself a projection, so carrying forward reproduces a stand structure the province published, where averaging would invent one it did not.", VRI_SUB[1])`

`r sprintf("%.1f", PCT_ABOVE_QMD)` per cent of cell-years sit at or above the
`r QMD_THRESH` cm source-sink threshold.

```{r}
#| label: tbl-vri
#| echo: false
#| tbl-cap: "Stand structure across the study perimeter, from the Vegetation Resources Inventory. n is cell-years. SD is the standard deviation of the landscape; SE is the standard error of the mean and is small by construction at this n, so it should not be read as precision about any one cell. Skewness and kurtosis are the bias-corrected third and fourth standardised moments; kurtosis is excess, so 0 is Gaussian and positive is heavy-tailed."

describe_vars(d, VRI_V, VRI_LABELS) |>
  describe_table() |>
  save_tbl("table-4-stand-structure") |>
  kable(booktabs = TRUE, align = "lrrrrrrrrr", row.names = FALSE)
```

The inventory is a projected operational product, not a census, and its polygons here have
reference years spanning several decades. A polygon interpreted from late-outbreak
photography describes a stand the beetle had already attacked, so basal area and pine cover
are post-attack for part of the window. There is no unattacked vintage, so this is
reported rather than corrected, and @sec-limits returns to it.

## Geomorphometry

Terrain is described by `r length(GEO_V)` geomorphometric surfaces computed with SAGA GIS
over the full reprojected elevation model and clipped afterwards, so a search radius near the
boundary still falls on real ground.

Radiation is computed twice because two mechanisms need it. Flight-window insolation is direct
and diffuse radiation over 1 July to 15 August restricted to 12:00 to 17:00, the hours
@safranyik2006chap1 identify as the flight peak; growing-season insolation is the whole-day
total from 1 May to 30 September, the shading quantity @krawchuk2020's first mechanism
concerns. A single annual heat index cannot separate them. Flight-window direct radiation
spans a `r sprintf("%.0f", max(d$solar_flight_direct)/min(d$solar_flight_direct))`-fold range
here against `r sprintf("%.1f", max(d$solar_season_total)/min(d$solar_season_total))`-fold for
the season total.

Exposure enters as the windward-leeward index and effective air flow height at the prevailing
bearing, the omnidirectional wind exposition index, the wind shelter index and topographic
openness. Shape enters as terrain ruggedness, vector ruggedness, topographic position and its
multi-scale form, convergence, slope, curvature and geomorphon class. Landform enters as
topographic wetness, valley depth, height above the valley floor, normalised height and
mid-slope position, because infested groups are reported in draws and gullies and deep snow
insulates overwintering brood [@safranyik2006chap1]. Aspect enters as its northward and
eastward components with the heat load index of @mccune2002.

One distinction matters. The strongest aspect statements in @safranyik2006chap1 concern the
aspect of the bole, not the slope, and are within-tree effects a 30 m raster cannot measure.
Nothing here tests them.

## Terrain-resolved wind {#sec-micromet}

```{r}
#| label: micromet-load
#| echo: false
#| output: false

mmsum <- read.csv(file.path(BC, "covariates", "wind-micromet", "micromet_summary.csv"))
mmW   <- rast(file.path(BC, "covariates", "wind-micromet", "wind_weight_by_direction.tif"))
MM_WW <- range(as.vector(minmax(mmW)))
```

Station wind interpolated from four to seven valley stations is nearly flat within a year, so
wind is also computed as a field varying in space, using the MicroMet model of @liston2006.
WindNinja, the usual tool, ships no macOS binary. MicroMet's wind component is seven equations
rather than a program, implemented directly with the paper's equation numbers against each.

Terrain slope $\beta$ and slope azimuth $\xi$ come from the elevation model (@liston2006,
their equations 12 and 13). Curvature $\Omega_c$ is a cell's elevation minus the mean of the
two opposite cells one curvature length scale away, taken on four direction lines and
averaged (their equation 14); the length scale is estimated as the first lag at which
elevation autocorrelation falls below 0.5, which is 600 m here. Writing $\theta$ for the
wind bearing, the slope in the wind direction is

$$\Omega_s = \beta \cos(\theta - \xi)$$ {#eq-slope}

with $\Omega_s$ and $\Omega_c$ each scaled to $[-0.5,\,0.5]$. The terrain weighting factor
applied to the observed speed $W$ is

$$W_w = 1 + \gamma_s \Omega_s + \gamma_c \Omega_c, \qquad \gamma_s = \gamma_c = 0.5$$ {#eq-weight}

giving the terrain-modified speed

$$W_t = W_w W$$ {#eq-speed}

and a diversion of the wind direction of

$$\theta_d = -\tfrac{1}{2}\, \Omega_s \sin\!\left[2(\xi - \theta)\right].$$ {#eq-divert}

By @eq-weight, $W_w$ depends on direction and not on speed, so it is computed once for each of 16 direction bins
and each hourly observation multiplied by the surface for its own bin; nothing is averaged
before the terrain acts on it. Over all bins $W_w$ runs `r sprintf("%.2f", MM_WW[1])` to
`r sprintf("%.2f", MM_WW[2])`. Speed and direction combine as components because averaging
degrees across the 360/0 line is meaningless. The field varies
`r sprintf("%.1f", min(mmsum$spatial_range))` to
`r sprintf("%.1f", max(mmsum$spatial_range))` km/h across the grid within a year, and is not a
repackaged terrain index. It correlates `r cc("mm_flight_mean", "wind_effect")` with the
windward-leeward index and `r cc("mm_flight_mean", "solar_flight_direct")` with flight
radiation. Its strongest association is with elevation,
`r cc("mm_flight_mean", "elevation")`.

## Flight-window wind

The flight window is 1 July to 15 August and the hours 12:00 to 17:00, and neither bound
was chosen from these data. The dates are the flight period @safranyik2006chap1 give for
this region; the hours follow their "peak flight is in the early to mid afternoon" and
@gray1972's 11:00 to 14:00 emergence peak, taken here to the later side because emergence
precedes the flight it initiates. Because the window is fixed a priori it can be checked
against the landscape's own climate, and @fig-flight-window does that. Across
`r fmt(CLIM_N)` hourly station records from May to September of the nine study years,
`r sprintf("%.1f", 100*CLIM_IN)` per cent of afternoon hours inside the window fall within
the 19 to 41 degrees C flight gate, against `r sprintf("%.1f", 100*CLIM_OUT)` per cent
outside it, and the afternoon mean is `r sprintf("%.1f", CLIM_T_IN)` against
`r sprintf("%.1f", CLIM_T_OUT)` degrees C. The window therefore roughly doubles the share
of hours in which flight is thermally possible. The same figure shows the same daily pattern
in wind. Mean speed peaks in the same hours as temperature, at
`r sprintf("%.1f", CLIM_W_MAX)` km/h in mid-afternoon against
`r sprintf("%.1f", CLIM_W_MIN)` km/h at dawn, so the hours the beetle can fly are also the
windiest of the day. That is a coincidence of timing the mechanism depends on.

Station wind is Environment and Climate Change Canada hourly records, and it enters in
three places. It is the observation MicroMet modifies over the terrain, it supplies the
prevailing bearing the directional terrain indices are computed at, and it is the climate
against which the flight window is checked in @fig-flight-window. The records are never
reduced before the terrain acts on them.

Hourly readings give detail, not independent observations. Summarised over a whole season,
every hour of wind collapses into one number per metric, a coefficient rests on as many
values as there are seasons, and its standard error looks small without being believable.
Stand density varies cell to cell; a season's wind does not. The response is therefore
measured over each sixteen-day epoch, so that wind varies within a season as well as
between seasons and the interaction the mechanism predicts is identified.

```{r}
#| label: fig-flight-window
#| echo: false
#| fig-cap: "The flight window against the landscape's own climate, from hourly Environment and Climate Change Canada station records for May to September of the nine study years. The window, 1 July to 15 August, is the shaded band; it is taken from the bionomics and is not fitted to these data. (a) Mean afternoon temperature by day of year, with the 19 to 41 degrees C flight gate and the 22 to 32 degrees C peak band marked. (b) The share of afternoon hours falling inside the flight gate, day by day. (c) The diurnal curve, mean temperature and the share of all hours inside the gate, by hour, with the 12:00 to 17:00 restriction shaded. (d) Mean wind speed by hour, on the same axis, showing that the hours the beetle can fly are also the windiest of the day."
#| fig-height: 6.2
#| fig-width: 7.5

wd <- range(unlist(lapply(unique(clim_d$year), win_doy)))
band <- annotate("rect", xmin = wd[1], xmax = wd[2], ymin = -Inf, ymax = Inf,
                 fill = "#fdae61", alpha = 0.18)
th <- theme_bw(base_size = 8) + theme(panel.grid.minor = element_blank())

pa <- ggplot(clim_d, aes(doy, pm_temp)) + band +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = T_GATE[1], ymax = T_GATE[2],
           fill = "#4575b4", alpha = 0.10) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = T_PEAK[1], ymax = T_PEAK[2],
           fill = "#4575b4", alpha = 0.14) +
  geom_point(aes(colour = factor(year)), size = 0.5, alpha = 0.55) +
  geom_smooth(se = FALSE, colour = "black", linewidth = 0.5, method = "loess", span = 0.3) +
  scale_colour_viridis_d(name = "Year", option = "turbo") +
  labs(x = "Day of year", y = "Afternoon temperature (C)",
       title = "(a) Afternoon temperature and the flight gate") + th

pb <- ggplot(clim_d, aes(doy, 100 * pm_in_gate)) + band +
  geom_point(aes(colour = factor(year)), size = 0.5, alpha = 0.55, show.legend = FALSE) +
  geom_smooth(se = FALSE, colour = "black", linewidth = 0.5, method = "loess", span = 0.3) +
  scale_colour_viridis_d(option = "turbo") +
  labs(x = "Day of year", y = "Afternoon hours in gate (%)",
       title = "(b) Share of afternoon hours flight is possible") + th

pc <- ggplot(clim_x, aes(hour)) +
  annotate("rect", xmin = FW_HR[1], xmax = FW_HR[2], ymin = -Inf, ymax = Inf,
           fill = "#fdae61", alpha = 0.22) +
  geom_hline(yintercept = T_GATE[1], linetype = 2, linewidth = 0.3, colour = "grey40") +
  geom_line(aes(y = temp), linewidth = 0.7, colour = "#d73027") +
  geom_line(aes(y = 100 * in_gate / 3), linewidth = 0.6, colour = "#4575b4") +
  scale_y_continuous("Temperature (C)",
    sec.axis = sec_axis(~ . * 3, name = "Hours in gate (%)")) +
  labs(x = "Hour of day", title = "(c) Diurnal cycle: temperature and the gate") + th

pd <- ggplot(clim_x, aes(hour, wind)) +
  annotate("rect", xmin = FW_HR[1], xmax = FW_HR[2], ymin = -Inf, ymax = Inf,
           fill = "#fdae61", alpha = 0.22) +
  geom_line(linewidth = 0.7, colour = "#1a9850") +
  labs(x = "Hour of day", y = "Wind speed (km/h)",
       title = "(d) Wind peaks in the same hours") + th

(pa | pb) / (pc | pd)
```

# Methods

## Variable selection {#sec-selection}


The candidate set is `r N_CAND` variables in six groups, one group for each pathway the
review names. The groups are stand density, host quality, terrain exposure to wind, terrain
shape, landscape context, and flight-window wind. Selection proceeds in four
stages and is required to retain at least one variable from each pathway, so that a
collinearity filter cannot silently delete a hypothesis the review established.

Fitting uses a class-balanced sample of `r fmt(N_BAL)` cell-years, `r fmt(4000)` of each
class per year. Landscape prevalence is about `r sprintf("%.0f", 100*PREV)` per cent, and an
unbalanced fit at that prevalence reports the intercept rather than the covariates.

## Models


Every model is a logistic regression of moderate-to-high disturbance $y_{it}$ in cell $i$
and year $t$ on standardised covariates,

$$\operatorname{logit}\Pr(y_{it}=1) = \alpha + \mathbf{x}_{it}^{\top}\boldsymbol{\beta}
+ \gamma_{g(i)} + \sum_{k} \delta_k\, z_{k,it}$$ {#eq-model}

where $\gamma_{g(i)}$ is the effect of the geomorphon landform class $g$ of cell $i$ and the
$z_{k,it}$ are the interaction terms. Coefficients are therefore log-odds per standard
deviation. Four models are fitted, each adding one claim to the last, so that every term's
contribution is visible as a change in fit rather than asserted.

M0 fits Krawchuk's first and third mechanisms together with landform context, that is host
size, topographic shading, and the terrain that governs where cold air and snow collect. M1 adds
stand density, the second of Krawchuk's mechanisms, which is objective 1. M2 adds terrain shape,
terrain exposure to wind, and flight-window radiation, which are objectives 3 and 4. M3 adds
the three interactions objective 4 requires, stand density by terrain exposure, by
flight-window radiation, and by flight-window wind.

# Results {#sec-results}


```{r}
#| label: tbl-aic
#| echo: false
#| tbl-cap: "Model comparison, panel A of the model table. Each model adds one pathway to the previous. AIC ranks on likelihood and a parameter penalty; the remaining columns are predictive error on the fitted probabilities. RMSE is the root mean squared error and is the square root of the Brier score; RMSE (%) expresses it against the prevalence of the response. Brier skill is the improvement over predicting the prevalence for every cell, where 0 is no better than the base rate. MAPE and Theil's U are not reported, because both divide by the observed value, which is zero for the majority class of a binary response, so both are undefined without discarding that class."

fit_metrics_table(
  fit_metrics(M0, "M0 host size + shading + landform"),
  fit_metrics(M1, "M1 + stand density"),
  fit_metrics(M2, "M2 + terrain and flight radiation"),
  fit_metrics(M3, "M3 + interactions")) |>
  save_tbl("table-5-model-comparison") |>
  kable(booktabs = TRUE, align = "lrrrrrrrr", row.names = FALSE)
```

```{r}
#| label: tbl-m3
#| echo: false
#| tbl-cap: "Full model M3, panel B of the model table, continuous terms ordered by absolute effect. Coefficients are log-odds per standard deviation on a class-balanced sample, so the intercept is not landscape prevalence. SE is the standard error of the coefficient. The geomorphon landform classes are also in this model and are not reported here."

co3 |>
  filter(term != "(Intercept)", !grepl("^geomorphon", term)) |>
  arrange(desc(abs(beta))) |>
  transmute(Term = pretty_terms(term), Beta = sprintf("%+.3f", beta),
            SE = sprintf("%.3f", se), z = sprintf("%.2f", z), p = sapply(p, pv)) |>
  save_tbl("table-6-full-model") |>
  kable(booktabs = TRUE, align = "lrrrr", row.names = FALSE)
```

Every pathway earned its place. AIC fell `r fmt(aic$AIC[1] - aic$AIC[2])` when stand density
entered, `r fmt(aic$AIC[2] - aic$AIC[3])` more with terrain and flight radiation, and
`r fmt(aic$AIC[3] - aic$AIC[4])` more with the interactions (@tbl-aic).

Stand basal area is the strongest density term. It entered at `r b3("BASAL_AREA")` log-odds
per standard deviation, an odds ratio of `r or3("BASAL_AREA")`, so a stand one standard
deviation above the mean in basal area had `r or3pc("BASAL_AREA")` per cent higher odds of
moderate-to-high disturbance (`r p3T("BASAL_AREA")`). Live stems per hectare has the same
sign at `r b3("VRI_LIVE_STEMS_PER_HA")` (`r p3T("VRI_LIVE_STEMS_PER_HA")`), crown closure
the opposite at `r b3("CROWN_CLOSURE")` (`r p3T("CROWN_CLOSURE")`).

Basal area and live stems both rise with attack, which is the direction a plume-holding
canopy implies, while crown closure falls. Crown closure is also the one measure the
inventory caps. It is recorded to a ceiling of 60 per cent, so its coefficient describes a
truncated variable and should not be read as the others are.

Radiation enters the model once. Direct radiation during the flight window is
`r b3("solar_flight_direct")`, the temperature limit on flying. Growing-season radiation did
not survive selection, so northness stands in for the shading pathway, and the two correlate
`r cc("solar_season_direct", "northness")`. Northness is `r b3("northness")`. Shaded,
north-facing ground therefore has more attack, against the prediction.

Terrain predicted attack independently of both. The windward-leeward index was
`r b3("wind_effect")` and elevation `r b3("elevation")`, while ruggedness is `r b3("tri")`
and not distinguishable from zero (`r p3T("tri")`), the vector ruggedness measure agreeing
in sign at `r b3("vrm")`. Ruggedness takes the opposite sign to that of @murphy2026, whose
response is seedling establishment rather than attack.

Fitted without flight-window radiation, stand density interacts with terrain exposure at
`r bnr(I_GEO)`, which reads as a wind result pointing the wrong way. With radiation in the
model that interaction is `r b3(I_GEO)` (`r p3T(I_GEO)`) and density interacts with
radiation at `r b3(paste0(DENS[1], ":", FSUN[1]))`. On a range whose prevailing bearing is
`r sprintf("%.0f", WDIR)` degrees the windward slopes face west, which are also the slopes
taking afternoon sun, so part of the apparent exposure effect was the sun. Not all of it.
The interaction stays positive and far from zero, which is the wrong sign for pheromone
disruption.

```{r}
#| label: fig-interaction
#| echo: false
#| fig-cap: "The refugia mechanism as fitted. (a) Predicted probability of moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem density, all other terms held at their means; the lines cross, so wind raises attack in thin stands and lowers it in dense ones, which is the interaction. (b) The same for standing volume, where the interaction is a fifth the size and attenuates the dense-stand slope without reversing it, so the lines converge rather than cross over the observed range. (c) Coefficients of the 16-day model, with and without the within-season spread term; the shading of a term between the two fits is what the within-season spread term absorbs. (d) Epoch prevalence against epoch mean wind, one point per epoch, years distinguished by both shape and shade so the panel survives black-and-white printing."
#| fig-height: 6

nd_base <- be[1, , drop = FALSE]
for (v in EV) nd_base[[v]] <- 0
pred_panel <- function(dv, lab) {
  qs <- quantile(be[[dv]], c(0.1, 0.9))
  gr <- expand.grid(w = seq(min(be$ep_wind_mean), max(be$ep_wind_mean), length.out = 60),
                    d = qs)
  nd <- nd_base[rep(1, nrow(gr)), ]
  nd$ep_wind_mean <- gr$w; nd[[dv]] <- gr$d
  nd$geomorphon <- factor(levels(be$geomorphon)[1], levels = levels(be$geomorphon))
  gr$p <- predict(E1, newdata = nd, type = "response")
  gr$Density <- factor(gr$d, labels = c("low (10th pct)", "high (90th pct)"))
  ggplot(gr, aes(w, p, colour = Density)) + geom_line(linewidth = 0.8) +
    scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
    labs(x = "Epoch wind (SD units)", y = "P(moderate-to-high)", title = lab) +
    theme_minimal(base_size = 9)
}
pa <- pred_panel("VRI_LIVE_STEMS_PER_HA", "(a) Stem density")
pb <- pred_panel("LIVE_STAND_VOLUME_125", "(b) Standing volume")

cf <- bind_rows(
  ce1 |> mutate(model = "E1 all epochs"),
  ce2 |> mutate(model = "E2 with lag")) |>
  filter(!grepl("^geomorphon|Intercept", term)) |>
  group_by(term) |> filter(any(abs(beta) > 0.05)) |> ungroup()
## pretty_terms() here for the same reason the tables use it: panel (c) shipped raw column
## names, ep_lag_nbr90 and QUAD_DIAM_125, beside a document that labels every other term.
cf$label <- pretty_terms(cf$term)
pc <- ggplot(cf, aes(reorder(label, beta), beta, colour = model)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_pointrange(aes(ymin = beta - 1.96*se, ymax = beta + 1.96*se),
                  position = position_dodge(width = 0.5), size = 0.25) +
  coord_flip() + scale_colour_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = NULL, y = "Log-odds per SD", title = "(c) 16-day model coefficients") +
  theme_minimal(base_size = 8) + theme(legend.position = "bottom")

pd <- epsum |>
  inner_join(epwind, by = c("year","epoch")) |>
  ggplot(aes(mean_kmh, prevalence)) +
  geom_point(aes(colour = factor(year), shape = factor(year)), size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.5) +
  scale_colour_viridis_d(option = "viridis", end = 0.9) +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 4, 18)) +
  ## Both scales must carry the same name or ggplot draws two legends and titles the
  ## shape one "factor(year)", which is the raw expression rather than a label.
  labs(x = "Epoch mean wind (km/h)", y = "Epoch prevalence",
       colour = "Year", shape = "Year", title = "(d) Epochs") +
  theme_minimal(base_size = 8)

(pa | pb) / (pc | pd)
```

## Density and wind

Stand density interacts negatively with terrain-resolved wind, which is the form the
pheromone mechanism predicts. Attack falls where a thin stand and strong wind coincide, the
interaction `r be1(EINT[1])` for stem density (`r pe1T(EINT[1])`) and `r be1(EINT[2])` for
standing volume (`r pe1T(EINT[2])`). Expressed as odds, each standard deviation of the epoch
wind regime multiplied the contribution of stem density by `r ore1(EINT[1])`, a reduction of
`r ore1pcabs(EINT[1])` per cent. Both interactions survive entering the previous epoch of the
same season, so neither is the outbreak's own spread appearing as a wind coefficient.

Wind on its own does not protect. Its main effect is `r be1("ep_wind_mean")`, marginally more
attack rather than less, and the density terms are positive, so a thin stand in still air
shows no reduction either. The mechanism is present only as the conjunction of the two.

## Previous-year pressure {#sec-autologistic}

Attack recurs where it was, by a margin that changes how every other coefficient reads. A cell
attacked one year is between
`r sprintf("%.0f", min(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))`
and
`r sprintf("%.0f", max(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))`
times more likely to be attacked the next. Entering persistence and 90 m spread separately,
after the autologistic design used for this province, gives `r b4("lag_self")` and
`r b4("lag_nbr90")`, raising discrimination from `r sprintf("%.3f", AUC3l)` to
`r sprintf("%.3f", AUC4)`.

Which environmental terms survive is the informative part. Ruggedness keeps `r ret("tri")`
of its value, and it describes a condition a cell has whether or not the beetle was there.
The landform terms are not stable. Valley depth keeps `r ret("valley_depth")` while mid-slope
position moves to `r ret("midslope_position")`, growing rather than shrinking. Part of what
the landform terms measured was where the outbreak had already been, which is the same
failure the terrain-wind index showed against radiation.

## Host as diameter {#sec-diameter}


```{r}
#| label: tbl-qmd
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced sample. The 25 cm boundary is the source-sink threshold of the species' bionomics. Intervals are Wilson score intervals on the class proportion, which is why they are asymmetric in the smallest class. Note that 30 m cells in a spreading outbreak are not independent, so the tests reported beneath this table are anti-conservative."

qmd_bin |>
  diameter_table() |>
  save_tbl("table-7-diameter-class") |>
  kable(booktabs = TRUE, align = "lrrrr", row.names = FALSE)
```

Coding host as diameter changes the picture cover alone gives. Attack peaks in the
25 to 30 cm class at
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent and falls away
above it, to `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "30-40"])` per cent at 30
to 40 cm and `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == ">40"])` per cent above 40
(@tbl-qmd). The step across the 25 cm source-sink boundary is from
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "20-25"])` to
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent.

In the full model diameter is `r b3("QUAD_DIAM_125")` per standard deviation
(z = `r z3("QUAD_DIAM_125")`). It is negative because the relationship is not monotone. A
linear term fitted through a humped response returns the slope of its falling limb, which
is the larger part of the range. The class table, not the coefficient, is the result here,
and the point stands, because a stand's cover says nothing about whether its stems can
produce brood.

The pattern is not an artefact of the class boundaries. Across all six classes attack
depends on diameter class, $\chi^2$ = `r sprintf("%.1f", QT$X2)` on `r QT$df` degrees of
freedom, p `r pt(QT$chi$p.value)`, with Cramer's V = `r sprintf("%.3f", QT$V)`. The step
across the 25 cm source-sink boundary specifically, from the 20 to 25 class to the 25 to 30
class, is `r sprintf("%+.1f", QT$step_diff)` percentage points, 95 per cent confidence
interval `r sprintf("%.1f to %.1f", -QT$step_ci[2], -QT$step_ci[1])`, p
`r pt(QT$step$p.value)`. With `r fmt(sum(qmd_bin$n))` cells a chi-square is significant on
trivial differences, which is why the effect size is quoted beside it, and 30 m cells in a
spreading outbreak are not independent, so both p-values are anti-conservative.

The smallest class is not evidence. It has `r fmt(qmd_bin$n[qmd_bin$bin == "<15"])` cells
against `r fmt(qmd_bin$n[qmd_bin$bin == "25-30"])` in the modal class, and a stand below
15 cm quadratic mean diameter in an inventory whose vintage postdates the outbreak is more
likely a stand the beetle already stripped of large stems than a stand attacked at that
size.

## The wind window {#sec-window}

The wind variable is the epoch wind regime, the mean over every hourly station observation
in the sixteen days an epoch covers, each hour modified over the terrain by its own bearing.
It is not restricted to the flight window, and that follows from how the insect disperses. @jackson2008 tracked mountain pine beetle on weather radar and confirmed
the returns by aerial capture, finding beetles "at altitudes up to more than 800 m above the
forest canopy" and estimating that those in flight above the canopy "may move
30--110 km$\cdot$day$^{-1}$", at a mean density of 4950 and a maximum of 18\,600 beetles per
hectare. @ainslie2010 then reproduced that above-canopy dispersion with an atmospheric
model. An insect carried in the boundary layer at those heights and distances is exposed to
the atmospheric regime of the period, not to the wind measured in a stand between noon and
five, and the variable is named for what it measures.

@tbl-window refits the 16-day model under four definitions of the window. These are every
hour, the 12:00 to 17:00 hours the flight peak occupies, the 11:00 to 18:00 hours that clear
the thermal gate on this landscape more than half the time, and only those hours whose
observed temperature fell inside the 19 to 41 degrees C gate itself. The stem-density interaction is negative and distinguishable from zero under
all four, so the mechanism does not depend on the choice. It is largest over the full
regime, and the standing-volume interaction is the term that depends on it.

```{r}
#| label: tbl-window
#| echo: false
#| tbl-cap: "The 16-day model refitted under four definitions of the wind window. The epoch wind regime, every hour, is the definition used throughout this paper. Coefficients are log-odds per standard deviation. The thermal-gate row is fitted on fewer cell-epochs because two epochs had too few qualifying hours, so its AIC is not comparable with the other three."

FWS |>
  transmute(
    Definition = c("Epoch wind regime (all hours)", "12:00-17:00", "11:00-18:00",
                   "Inside the 19-41 C gate")[match(definition,
                   c("all24","h12_17","h11_18","gate"))],
    `Cell-epochs` = formatC(n, format = "d", big.mark = ","),
    `Wind (km/h)` = sprintf("%.2f", wind_mean),
    `Stems x wind` = sprintf("%+.4f", stems_x_wind),
    `p` = sapply(stems_p, function(x) if (x < 0.001) "< 0.001" else sprintf("%.3f", x)),
    `Volume x wind` = sprintf("%+.4f", volume_x_wind),
    `p ` = sapply(volume_p, function(x) if (x < 0.001) "< 0.001" else sprintf("%.3f", x)),
    AUC = sprintf("%.3f", AUC)) |>
  save_tbl("table-8-flight-window") |>
  kable(booktabs = TRUE, align = "lrrrrrrr", row.names = FALSE)
```

# Discussion

## Radiation, not wind

The clearest result of this study is methodological. It changes what a terrain variable in
this literature can be read as.

Fitted with a terrain-wind index and no radiation, stand density interacts with terrain
exposure at `r bnr(I_GEO)`, and the natural reading is a wind effect. Add flight-window
radiation and that interaction becomes `r b3(I_GEO)`, p `r p3t(I_GEO)`, while density by
radiation appears at `r b3(paste0(DENS[1], ":", FSUN[1]))`. About two fifths of the apparent
exposure effect was the sun. The rest was not, and what survives keeps the sign the
mechanism forbids.

That is not a peculiarity of this dataset. A windward-leeward index and an afternoon
radiation surface are both functions of slope and aspect, and on a range whose prevailing
flight-window bearing is `r sprintf("%.0f", WDIR)` degrees the windward slopes are the
west-facing ones, which are also the slopes that take afternoon sun during the flight peak.
Any study that enters a terrain-derived wind index without a radiation term on the same
clock will attribute radiation to wind.

## What survives

This study tested the three mechanisms @krawchuk2020 propose for refugia from mountain pine
beetle, using a response measured at the sixteen-day cadence the sensor and the insect share.
Support was partial. H2, low host density admitting the wind that
disperses the aggregation pheromone, was supported, and in the conditional form the mechanism
specifies rather than as a main effect. H3, a scarcity of large-diameter hosts, was supported
as a threshold rather than as a gradient. H1, topographic shading, was not supported, and the
term that rejects it is the least secure in the study.

H2, host density, holds, and it holds in the conditional form the mechanism specifies
rather than as a bare main effect. Stand basal area predicts disturbance at `r b3("BASAL_AREA")`, and
its interaction with wind is `r be1(EINT[2])` while stem density's is `r be1(EINT[1])`. The
claim was never that dense stands are attacked more. It was that a dense canopy holds a
plume together, and that the advantage this gives the beetle should shrink as the wind rises.
Both halves are present.

H3, large-diameter host, holds as a threshold rather than a gradient. Attack peaks at
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent in the 25 to 30 cm
class, at the source-sink boundary of @carroll2004bionomics, and falls away on both sides.

H1, topographic shading, does not hold, and it fails on a stand-in rather than on the
quantity the mechanism names. Growing-season radiation did not survive variable selection, so
northness stands in for the pathway, and it correlates `r cc("solar_season_direct", "northness")`
with it. Northness is `r b3("northness")`, so shaded ground has more attack where the
mechanism requires less. Northness tests aspect, and aspect measures more than shade, so this
is the least secure of the three verdicts. The water-stress pathway @krawchuk2020 describe is
not measured here at all.

## The wind claim

Stand density interacts with terrain-resolved wind at `r be1(EINT[1])` for stem density and
`r be1(EINT[2])` for standing volume, both negative, at p `r pe1t(EINT[1])` and
p `r pe1t(EINT[2])` respectively. A dense stand is worth less to the beetle where and when
the wind blows harder, which is what @krawchuk2020 predict. The result survives entering the
previous epoch of the same season, so it is not the outbreak's own spread appearing as a
wind coefficient.

The mechanism concerns a plume that persists for minutes over a flight period of weeks, and
the wind term is measured over the sixteen days an epoch covers for that reason. @krawchuk2020
name wind without naming the interval it should be summarised over, and nothing in the
literature since has supplied one.

Two limits stay attached to the claim. The interaction is small, roughly a tenth of the
stand basal area main effect. And the wind field, though it varies in space, is still
terrain-modified station data rather than measurement on the ridge, so what is established
is that the modelled wind field behaves as the mechanism requires, not that the air did.

## Ruggedness and regeneration

The two papers now share a grid, a landscape and a terrain covariate set, and they disagree
on the sign of ruggedness. It is `r b3("tri")` for beetle attack here, against a positive
coefficient for conifer regeneration in @murphy2026. Read together they describe a
landscape in which rugged ground both resists the disturbance and shelters the recovery from
it. That is a coherent picture, not a contradiction, and it is only visible because the two
studies were fitted on the same grid in EPSG:3153.

## What stays confounded

Elevation is among the largest terms in the model, `r b3("elevation")`, and elevation is not
one thing. It combines temperature, snowpack, growing-season length and the distribution of
lodgepole pine, and this design cannot take them apart. Reporting the elevation coefficient
as a result would be reporting a composite.

The same caution now applies in reverse to every terrain term in the model. Radiation took
about a third of one terrain interaction. There is no guarantee that some further unmeasured
environmental surface would not take the rest, or another.

# Conclusions

The wind-disruption mechanism proposed by @krawchuk2020 holds on this landscape, at the
resolution the sensor and the insect share. Across `r EP_EPOCHS` sixteen-day epochs, stand
density interacts negatively with terrain-resolved wind, `r be1(EINT[1])` for stem density
and `r be1(EINT[2])` for standing volume, and both survive a within-season spread term. A
dense canopy is worth less to the beetle where and when the wind blows harder, which is what
a plume-disruption mechanism requires.

Of the three mechanisms that proposal names, two hold. Host density holds through stand
basal area, `r b3(DENS[1])`, with live stems agreeing in sign and crown closure alone running
the other way, and large-diameter host holds as a threshold at 25 to 30 cm rather than as a
slope, at the source-sink boundary of @carroll2004bionomics. Topographic shading does not
hold, and it fails on a surrogate. Growing-season radiation did not survive selection, so
northness stands in for the pathway at `r b3("northness")`, and the two correlate
`r cc("solar_season_direct", "northness")`.

Two measurement results travel beyond this landscape, and both concern variables that look
like one thing and behave as another.

A terrain wind index is partly a measurement of sunlight. Entering flight-window radiation
weakened the density by terrain-exposure interaction from `r bnr(I_GEO)` to `r b3(I_GEO)`.
On any range whose prevailing bearing puts the windward slopes into the afternoon sun, and
that is most of them, a study entering a windward-leeward index without a radiation term on
the same clock will attribute radiation to wind.

A landform variable in a spreading outbreak partly records where the outbreak has already
been. Entering previous-year attack left valley depth retaining `r ret("valley_depth")` of
its coefficient and moved mid-slope position to `r ret("midslope_position")`, against
`r ret("tri")` for ruggedness, which describes a condition a cell has whether or not the
beetle ever arrived. Landform terms fitted without a contagion term are not measuring
landform alone.

# Limitations {#sec-limits}

**Radiation and exposure are not separable in the flight window.** The prevailing bearing is
`r sprintf("%.0f", WDIR)` degrees, so windward slopes face west and take afternoon sun.
Flight-window radiation correlates `r cc("solar_flight_direct", "wind_effect")` with the
windward-leeward index. Its positive coefficient is consistent with the thermal gate and with
exposure raising attack, and this design cannot choose between them.

**The shading mechanism is tested on a surrogate.** Growing-season direct radiation did not
survive variable selection, so northness stands in for the pathway, and it correlates
`r cc("solar_season_direct", "northness")` with it. Northness tests aspect, aspect measures
more than shade, and the water-stress pathway @krawchuk2020 describe is not measured here.
Growing-season radiation is at least separable from wind exposure,
`r cc("solar_season_direct", "wind_effect")`, which is the one thing it establishes.

**The wind field is modelled, not measured.** MicroMet modifies station observations over the
terrain. What is established is that the modelled field behaves as the mechanism requires,
not that the air did.

**The inventory postdates part of the outbreak.** Polygons interpreted after the beetle
passed describe the stand it left, and total basal area discriminates only weakly on its own,
`r sprintf("%.3f", uni$auc[uni$term == "BASAL_AREA"])`.

**The response is classified, not observed.** Disturbance is inferred from Landsat moisture
indices trained on plots cut from those indices, so reported accuracy measures separability
rather than agreement with ground mortality.

**One landscape, one outbreak.** Every conclusion is conditioned on a single mountain range
over `r EP_YEARS` years and `r EP_EPOCHS` epochs.

# Acknowledgements

This work used no external funding. Beetle disturbance was classified from Landsat
Collection 2 Level-2 surface reflectance distributed by the United States Geological
Survey; stand structure from the British Columbia Data Catalogue; terrain from Natural
Resources Canada; and wind from Environment and Climate Change Canada. The author
thanks those agencies for maintaining the open archives the study rests on.

# Conflict of interest

The author declares no conflict of interest.

# Ethics statement

This study used only publicly archived remote sensing, inventory and meteorological
data. No animals, human participants or protected material were involved, so no ethics
approval was required.

# Author contributions

**Seamus Murphy:** conceptualisation; data curation; formal analysis; investigation;
methodology; software; validation; visualisation; writing, original draft; writing,
review and editing.

# Data availability

Every dataset used here is public and none was collected by the author. Beetle disturbance
is classified from Landsat Collection 2 Level-2 surface reflectance. Stand structure is the
provincial Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved from the
British Columbia Data Catalogue web feature service by
`02.inputs/beetle/34-fetch-vri.py`. Terrain derives from the Natural
Resources Canada High Resolution Digital Elevation Model [@nrcan2017], with geomorphometry
computed by SAGA GIS in `37-geomorphometry.R`. Wind is Environment and Climate Change Canada
hourly station data, retrieved by `30-wind-hourly-metrics.R` and `36-wind-direction.R`.

All derived data and the complete analysis code that reproduce every number, table and
figure in this article are archived at Dryad under
doi:PLACEHOLDER-DRYAD-DOI, and are mirrored at
<https://github.com/seamusrobertmurphy/beetle-topography-and-wind-study>.

<!-- BLOCKING BEFORE SUBMISSION. Journal of Applied Entomology mandates data sharing and
     states that data "made available on request or uploaded as supplementary files, are
     not accepted", and that a link with a DOI must sit above the references. The Dryad
     deposit has not been made, so the DOI above is a placeholder and the submission
     cannot go in until it is minted. Wiley pays the Dryad archiving charge for papers
     published in this journal. Include the "Private for Peer Review" link here and in
     the cover letter once the deposit exists. -->

# Figure legends

**Figure 1.** Landscape, terrain and stand surfaces across the study area, all EPSG:3153 at
30 m over Esri World Shaded Relief, which is reproduced under Esri's attribution
requirement. (a) elevation; (b) terrain ruggedness index; (c) windward-leeward index at the
prevailing bearing; (d) flight-window direct radiation, the thermal gate on flight;
(e) growing-season direct radiation, the shading pathway; (f) the MicroMet wind weighting
factor at the prevailing bearing; (g) stand basal area; (h) quadratic mean diameter, whose
25 cm source-sink threshold falls near the midpoint of the scale; (i) live stems per
hectare. The white outline is the study perimeter and the red outline the 2015 Mt Midgeley
burn. Kootenay Lake and the Kootenay River are from the British Columbia Freshwater Atlas.
Contours are at 200 m. Every panel has its own north arrow and the same numerical scale,
1:150,000, which is correct at a printed panel width of 66 mm. The surfaces are drawn unclipped,
so they extend beyond the perimeter within which the models are fitted.

**Figure 2.** The flight window against the landscape's own climate, from hourly Environment
and Climate Change Canada station records for May to September of the nine study years. The
window, 1 July to 15 August, is the shaded band; it is taken from the bionomics and is not
fitted to these data. (a) Mean afternoon temperature by day of year, with the 19 to 41
degrees C flight gate and the 22 to 32 degrees C peak band marked. (b) The share of afternoon
hours falling inside the flight gate, day by day. (c) The diurnal curve, mean temperature and
the share of all hours inside the gate, by hour, with the 12:00 to 17:00 restriction shaded.
(d) Mean wind speed by hour, on the same axis, showing that the hours the beetle can fly are
also the windiest of the day.

**Figure 3.** The refugia mechanism as fitted. (a) Predicted probability of moderate-to-high
disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem
density, all other terms held at their means; the lines cross, so wind raises attack in thin
stands and lowers it in dense ones, which is the interaction. (b) The same for standing
volume, where the interaction is a fifth the size and attenuates the dense-stand slope
without reversing it, so the lines converge rather than cross over the observed range.
(c) Coefficients of the 16-day model, with and without the within-season spread term.
(d) Epoch prevalence against epoch mean wind, one point per epoch, years distinguished by
both shape and shade so the panel survives black-and-white printing.

```{r}
#| label: session
#| echo: false

sessionInfo()
```

# References
