---
title: "Testing the wind disruption hypothesis for beetle refugia"
subtitle: "Stand density, terrain shape and flight-window wind as competing controls on mountain pine beetle attack (Dendroctonus ponderosae) in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
# Journal of Applied Entomology requires exactly six keywords and states that they
# "should not include words in the title". The title carries testing, wind, disruption,
# hypothesis, beetle and refugia, so none of those may appear here. The earlier set
# broke that rule three times over.
keywords:
  - Dendroctonus ponderosae
  - Landsat time series
  - stand density
  - geomorphometry
  - temporal resolution
  - pheromone communication

date: today
keep-md: true

format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: false
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/springer-basic-author-date.csl
df-print: kable
---

```{r}
#| label: setup
#| echo: false
#| output: false

## Every quantity in this manuscript is computed here at render time. No number in the
## prose is typed: each is an inline expression over an object built in a chunk below.
## error = FALSE on purpose. In a document whose numbers are all computed, a chunk that
## fails quietly while downstream inline expressions carry stale values is the single
## failure this design exists to prevent.

suppressPackageStartupMessages({
  library(terra); library(sf); library(dplyr); library(tidyr); library(ggplot2)
  library(knitr); library(mgcv); library(car); library(ranger); library(patchwork)
  library(tidyterra)
})
set.seed(42)
knitr::opts_chunk$set(dpi = 300, fig.width = 9, fig.height = 6)

here <- function(...) file.path("..", ...)          # repo-relative, from 01.manuscript/
BC   <- here("02.inputs", "beetle-classification")
SA   <- file.path(BC, "study-area")
LIT  <- here("02.inputs", "literature")

fmt <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = ",")

## Tables and figures have to be uploaded to Journal of Applied Entomology as separate
## files. Rather than rebuild them in a second document, which would let the two drift,
## every table is written to 03.outputs/TBL as it is printed, and submission-jae/tables.qmd
## reads those files back. save_tbl returns its argument so it can sit inside a pipe.
TBLDIR <- here("03.outputs", "TBL")
dir.create(TBLDIR, showWarnings = FALSE, recursive = TRUE)
save_tbl <- function(x, name) {
  write.csv(x, file.path(TBLDIR, paste0(name, ".csv")), row.names = FALSE)
  x
}

pv  <- function(p) if (p < 1e-16) "< 1e-16" else sprintf("%.3g", p)
```


<!-- Computation runs here, in one block, because the Abstract below is computed
     from the results and knitr evaluates inline expressions in document order.
     Every chunk keeps its own label, code and comments; only its position moved. -->

```{r}
#| label: review-load
#| echo: false
#| output: false

## Two screening products are read, both built by scripts committed with the manuscript
## so that every count below is checkable rather than remembered.
##
##   review-screening.csv    the systematic screen of the terrain-and-attack corpus,
##                           02.inputs/literature/build-review-screening.py
##   wind-temporal-summary   the temporal-resolution extraction, built by
##                           02.inputs/literature/extract-wind-temporal.py
##   mechanism-summary.csv   the mechanism extraction, built by
##                           02.inputs/literature/extract-mechanism.py, which reads the
##                           full text of every mountain pine beetle and refugia paper
##                           in the central store and writes out every sentence that
##                           mentions wind together with a mechanism term.

rev <- read.csv(file.path(LIT, "review-screening.csv"), stringsAsFactors = FALSE)
wt  <- read.csv(file.path(LIT, "wind-temporal-summary.csv"), stringsAsFactors = FALSE)
mech_sent <- read.csv(file.path(LIT, "mechanism-evidence.csv"), stringsAsFactors = FALSE)
mech <- read.csv(file.path(LIT, "mechanism-summary.csv"), stringsAsFactors = FALSE)

N_INC    <- sum(rev$stage2_include == "yes", na.rm = TRUE)
N_WIND   <- sum(rev$stage2_include == "yes" & rev$wind_predictor == "yes", na.rm = TRUE)
N_NOWIND <- N_INC - N_WIND
YR_RANGE <- range(rev$year[rev$stage2_include == "yes"], na.rm = TRUE)

MECH_PAPERS <- nrow(mech)
MECH_SENT   <- nrow(mech_sent)
N_PHER  <- sum(mech$pheromone > 0)
N_DENS  <- sum(mech$density  > 0)
N_TERR  <- sum(mech$terrain  > 0)
N_FLIGHT<- sum(mech$flight   > 0)
```

```{r}
#| label: review-wind-temporal
#| echo: false
#| output: false

WT_TOTAL  <- nrow(wt)
WT_NOWIND <- sum(wt$n_wind == 0, na.rm = TRUE)
WT_WIND   <- WT_TOTAL - WT_NOWIND
## "scales" is the temporal vocabulary the extraction found; empty means the study
## mentions wind and says nothing about over what interval.
WT_SILENT <- sum(wt$n_wind > 0 & (is.na(wt$scales) | wt$scales == ""), na.rm = TRUE)
```

```{r}
#| label: study-area
#| echo: false
#| output: false

msk  <- rast(file.path(SA, "perimeter_mask.tif"))
elev <- rast(file.path(SA, "elevation.tif"))
per  <- st_read(file.path(SA, "study_perimeter.gpkg"), quiet = TRUE)
DATA <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
burn <- st_transform(st_union(st_read(file.path(DATA, "fire_perimiter", "Fire.Perimiter.shp"),
                                      quiet = TRUE)), 3153)

N_CELL  <- sum(!is.na(values(msk)))
AREA_HA <- N_CELL * 900 / 1e4
BURN_HA <- as.numeric(st_area(burn)) / 1e4
ELEV_R  <- as.vector(minmax(elev))
EPSG    <- crs(msk, describe = TRUE)$code
CRSNAME <- crs(msk, describe = TRUE)$name
burn_e  <- range(values(mask(crop(elev, vect(burn)), vect(burn))), na.rm = TRUE)
```

```{r}
#| label: response
#| echo: false
#| output: false

d <- read.csv(file.path(BC, "model-data", "model_table.csv"), stringsAsFactors = FALSE)
YRS      <- sort(unique(d$year))
N_ROWS   <- nrow(d)
PREV     <- mean(d$modhigh)
prev_yr  <- d |> group_by(year) |> summarise(prev = mean(modhigh), n = n())
```

```{r}
#| label: vri-summary
#| echo: false
#| output: false

VRI_V <- c("BASAL_AREA","CROWN_CLOSURE","VRI_LIVE_STEMS_PER_HA","QUAD_DIAM_125",
           "PROJ_AGE_1","PROJ_HEIGHT_1","LIVE_STAND_VOLUME_125","PinePct","PINE_BA")
QMD_THRESH <- 25
PCT_ABOVE_QMD <- 100 * mean(d$QUAD_DIAM_125 >= QMD_THRESH)
```

```{r}
#| label: geo-summary
#| echo: false
#| output: false

## Read from the raster, never typed. A hardcoded list here reported 20 surfaces after
## the stack had grown to 29, because the radiation and slope-position layers were added
## to the pipeline and not to the list.
GEO_V <- names(rast(file.path(BC, "geomorphometry", "geomorphometry.tif")))
wdir <- read.csv(file.path(BC, "covariates", "wind_direction_flight_window.csv"))
WDIR <- as.numeric(readLines(file.path(BC, "covariates", "wind_direction_pooled.txt")))
```

```{r}
#| label: wind-summary
#| echo: false
#| output: false

WIND_V <- c("jun","jul","aug","flight_mean","flight_p95","flight_calm","flight_windy")
wind_yr <- d |> group_by(year) |>
  summarise(across(all_of(WIND_V), mean))
```

```{r}
#| label: selection-setup
#| echo: false
#| output: false

## The candidate set, grouped by the pathway each variable belongs to. Grouping is not
## cosmetic: the selection below is required to retain at least one variable from each
## pathway, so a collinearity filter cannot silently delete a hypothesis.

## The candidate set, grouped by the mechanism each variable belongs to. Grouping is not
## cosmetic: selection is required to retain at least one variable from each group, so a
## collinearity filter cannot silently delete a hypothesis.
##
## The first three groups are Krawchuk et al.'s (2020) own three mechanisms, in their
## order: topographic shading, host density, and large-diameter host. The remaining
## groups carry terrain shape and the thermal gate on
## flight, plus the landscape context that has to be controlled.

PATH <- list(
  ## Topographic shading. Season radiation is the tree's water-stress load; sky view
  ## and openness are the geometry that produces it.
  shading  = c("solar_season_direct","solar_season_total","sky_view","heat_load",
               "northness","eastness"),
  ## Host density: the pathway the pheromone argument says wind acts through.
  density  = c("BASAL_AREA","CROWN_CLOSURE","VRI_LIVE_STEMS_PER_HA","LIVE_STAND_VOLUME_125"),
  ## Large-diameter host, against the 25 cm source-sink threshold.
  hostsize = c("QUAD_DIAM_125","PROJ_HEIGHT_1","PROJ_AGE_1","PinePct","PINE_BA"),
  ## The thermal gate on flight: radiation during the flight window, in the hours
  ## the flight peak occupies, rather than an annual index.
  flightsun = c("solar_flight_direct","solar_flight_diffuse"),
  ## Terrain exposure to wind.
  wind_geo = c("wind_effect","wind_afh","wind_exposition","wind_shelter",
               "openness_pos","openness_neg"),
  ## Terrain shape, including the parent study's ruggedness term.
  shape    = c("tri","vrm","tpi","mstpi","convergence","slope","curv_prof","curv_plan"),
  ## The landforms the literature names, plus elevation.
  landform = c("elevation","twi","valley_depth","height_valley_floor",
               "normalised_height","midslope_position"),
  ## Flight-window wind at hourly source resolution, interpolated from stations. Nearly
  ## flat within any year, so it carries the between-year signal only.
  wind_t   = c("flight_mean","flight_p95","flight_calm","flight_windy","jun","jul","aug"),
  ## Flight-window wind after terrain modification by MicroMet, which does vary within
  ## the year across the grid. Only the mean is carried: the calm fraction is its mirror
  ## image at r = -0.98 and entering both puts two collinear terms in with the same sign.
  wind_mm  = c("mm_flight_mean")
)
CAND <- unlist(PATH, use.names = FALSE)
stopifnot(all(CAND %in% names(d)))
N_CAND <- length(CAND)

## Balanced sample per year. The landscape prevalence is about 11 per cent, and an
## unbalanced fit at that prevalence reports the intercept, not the covariates.
bal <- d |> group_by(year, modhigh) |>
  slice_sample(n = 4000, replace = FALSE) |> ungroup()
N_BAL <- nrow(bal)

z <- function(x) (x - mean(x)) / sd(x)
bz <- bal; for (v in CAND) bz[[v]] <- z(bz[[v]])

## Geomorphons are a landform CLASS, not a magnitude, so they cannot be z-scored,
## correlated or ranked by variance inflation alongside the continuous candidates. They
## enter every model as a factor instead, bypassing the numeric selection stages, with
## the most common class as the reference level. The ten-class scheme is Jasiewicz and
## Stepien's; classes 1 and 10, flat and pit, do not occur on this perimeter.
GEO_LAB <- c("flat","peak","ridge","shoulder","spur","slope","hollow","footslope",
             "valley","pit")
bz$geomorphon <- factor(GEO_LAB[bal$geomorphons], levels = GEO_LAB)
bz$geomorphon <- droplevels(bz$geomorphon)
bz$geomorphon <- relevel(bz$geomorphon,
                         ref = names(sort(table(bz$geomorphon), decreasing = TRUE))[1])
GEOM_REF <- levels(bz$geomorphon)[1]
GEOM_N   <- nlevels(bz$geomorphon)
```

```{r}
#| label: sel-univariate
#| echo: false
#| output: false

uni <- lapply(CAND, function(v) {
  m <- glm(reformulate(v, "modhigh"), data = bz, family = binomial)
  s <- coef(summary(m))
  data.frame(term = v,
             group = names(PATH)[sapply(PATH, function(g) v %in% g)][1],
             beta = s[v, "Estimate"], p = s[v, "Pr(>|z|)"],
             auc = as.numeric(pROC::auc(pROC::roc(bz$modhigh, fitted(m), quiet = TRUE))))
}) |> bind_rows() |> arrange(desc(auc))
UNI_KEEP <- uni$term[uni$p < 0.01]
```

```{r}
#| label: sel-collinear
#| echo: false
#| output: false

cm <- cor(bz[, CAND], use = "complete.obs")
hc <- hclust(as.dist(1 - abs(cm)), method = "average")
CUT <- 0.25                                  # clusters at |r| >= 0.75
cl  <- cutree(hc, h = CUT)

## Within each cluster keep the variable with the highest univariate AUC, so the survivor
## is chosen on its relationship to the response rather than on alphabetical order.
keep_cluster <- sapply(split(names(cl), cl), function(vs)
  vs[which.max(uni$auc[match(vs, uni$term)])])
CLUST_KEEP <- intersect(unname(keep_cluster), UNI_KEEP)
N_CLUST <- length(unique(cl))
```

```{r}
#| label: sel-vif
#| echo: false
#| output: false

vif_drop <- function(vars, thresh = 5) {
  dropped <- character(0)
  repeat {
    if (length(vars) < 2) break
    m <- glm(reformulate(vars, "modhigh"), data = bz, family = binomial)
    v <- car::vif(m)
    if (max(v) < thresh) break
    worst <- names(v)[which.max(v)]
    ## A pathway is never emptied: if dropping this variable would remove the last
    ## member of its group, the next worst is taken instead.
    grp <- names(PATH)[sapply(PATH, function(g) worst %in% g)][1]
    if (sum(vars %in% PATH[[grp]]) <= 1) {
      v <- v[setdiff(names(v), worst)]
      if (!length(v) || max(v) < thresh) break
      worst <- names(v)[which.max(v)]
    }
    vars <- setdiff(vars, worst); dropped <- c(dropped, worst)
  }
  list(keep = vars, dropped = dropped)
}
vf <- vif_drop(CLUST_KEEP)
VIF_KEEP <- vf$keep
vif_final <- car::vif(glm(reformulate(VIF_KEEP, "modhigh"), data = bz, family = binomial))
```

```{r}
#| label: sel-importance
#| echo: false
#| cache: true
#| output: false

rf <- ranger::ranger(as.factor(modhigh) ~ ., data = bz[, c("modhigh", VIF_KEEP)],
                     num.trees = 500, importance = "permutation",
                     probability = TRUE, seed = 42)
imp <- data.frame(term = names(rf$variable.importance),
                  importance = as.numeric(rf$variable.importance)) |>
  left_join(uni |> select(term, group), by = "term") |>
  arrange(desc(importance))
RF_AUC <- as.numeric(pROC::auc(pROC::roc(bz$modhigh, rf$predictions[, 2], quiet = TRUE)))
FINAL <- imp$term
```

```{r}
#| label: model-spec
#| echo: false
#| output: false

DENS <- intersect(FINAL, PATH$density)
WGEO <- intersect(FINAL, PATH$wind_geo)
WTMP <- intersect(FINAL, PATH$wind_t)
WMM  <- intersect(FINAL, PATH$wind_mm)
SHAD <- intersect(FINAL, PATH$shading)
FSUN <- intersect(FINAL, PATH$flightsun)
HSIZ <- intersect(FINAL, PATH$hostsize)
SHAP <- intersect(FINAL, PATH$shape)
LFRM <- intersect(FINAL, PATH$landform)
```

```{r}
#| label: fit-models
#| echo: false
#| cache: true
#| output: false

## Geomorphon class is in every model, including M0: it is a description of landform,
## not a hypothesis being added, and leaving it out of the baseline would let it be
## credited to whichever pathway entered next.
f <- function(vars, extra = character(0))
  reformulate(c(vars, "geomorphon", extra), "modhigh")

## M0 is host size and topographic shading together with the
## landform context that must be controlled before any of the rest can be read.
V_CTX  <- c(HSIZ, SHAD, LFRM)
M0 <- glm(f(V_CTX), data = bz, family = binomial)
M1 <- glm(f(c(V_CTX, DENS)), data = bz, family = binomial)
M2 <- glm(f(c(V_CTX, DENS, WGEO, SHAP, FSUN)), data = bz, family = binomial)

INT <- character(0)
if (length(DENS) && length(WGEO)) INT <- c(INT, paste0(DENS[1], ":", WGEO[1]))
if (length(DENS) && length(WTMP)) INT <- c(INT, paste0(DENS[1], ":", WTMP[1]))
## The interaction objective 4 requires, plus the one the thermal gate implies: whether the flight-
## window radiation a slope receives changes what stand density buys, since a stand the
## beetle cannot reach at flight temperature is not a stand density can protect.
if (length(DENS) && length(FSUN)) INT <- c(INT, paste0(DENS[1], ":", FSUN[1]))
M3 <- glm(f(c(V_CTX, DENS, WGEO, SHAP, FSUN, WTMP), INT),
          data = bz, family = binomial)

## The counterfactual the Results section turns on: the same model with flight-window
## radiation removed. Fitted here so the coefficient quoted for it is computed, not
## remembered from an earlier draft.
M3_norad <- glm(f(c(setdiff(c(V_CTX, DENS, WGEO, SHAP), FSUN), WTMP),
                  setdiff(INT, paste0(DENS[1], ":", FSUN[1]))),
                data = bz, family = binomial)
cnr <- as.data.frame(coef(summary(M3_norad)))
names(cnr) <- c("beta", "se", "z", "p"); cnr$term <- rownames(cnr)
bnr <- function(v) sprintf("%+.3f", cnr$beta[cnr$term == v])
znr <- function(v) sprintf("%.2f", cnr$z[cnr$term == v])
pnr <- function(v) pv(cnr$p[cnr$term == v])

## M4 adds previous-year beetle pressure, following the autologistic design of
## Aukema et al. (2007), who fit environmental covariates "in addition to 1st order
## spatial and lag-1 and lag-2 temporal terms". Persistence and spread enter separately:
## lag_self is whether the cell itself was attacked last year, lag_nbr90 the fraction of
## its 90 m neighbourhood that was, excluding itself. 2006 has no predecessor and drops,
## so M4 is fitted on the seven year-pairs and is not AIC-comparable with M0 to M3.
bl <- bz[!is.na(bz$lag_self), ]
LAGV <- c("lag_self","lag_nbr90","lag_nbr510")
for (v in LAGV) bl[[v]] <- (bl[[v]] - mean(bl[[v]])) / sd(bl[[v]])
M3l <- glm(formula(M3), data = bl, family = binomial)
M4  <- glm(f(c(V_CTX, DENS, WGEO, SHAP, FSUN, WTMP, LAGV), INT),
           data = bl, family = binomial)

## M5 replaces the station wind terms with the terrain-resolved MicroMet field, which is
## the only wind variable in this study that varies within a year. Its main effect is
## therefore identified in space, unlike the station terms, and its interaction with stand
## density is the spatial form of the test the station terms could only make in time.
INT_MM <- if (length(DENS) && length(WMM)) paste0(DENS[1], ":", WMM[1]) else character(0)
INT_NOW <- INT[!grepl(paste0(":", paste(PATH$wind_t, collapse = "|:")), INT)]
M5 <- glm(f(c(V_CTX, DENS, WGEO, SHAP, FSUN, LAGV, WMM), c(INT_NOW, INT_MM)),
          data = bl, family = binomial)
co5 <- as.data.frame(coef(summary(M5)))
names(co5) <- c("beta","se","z","p"); co5$term <- rownames(co5)
b5 <- function(v) sprintf("%+.3f", co5$beta[co5$term == v])
z5 <- function(v) sprintf("%.2f", co5$z[co5$term == v])
p5 <- function(v) pv(co5$p[co5$term == v])
p5t <- function(v) pt(co5$p[co5$term == v])
AUC5 <- as.numeric(pROC::auc(pROC::roc(bl$modhigh, fitted(M5), quiet = TRUE)))
co4 <- as.data.frame(coef(summary(M4)))
names(co4) <- c("beta", "se", "z", "p"); co4$term <- rownames(co4)
b4 <- function(v) sprintf("%+.3f", co4$beta[co4$term == v])
z4 <- function(v) sprintf("%.2f", co4$z[co4$term == v])
p4 <- function(v) pv(co4$p[co4$term == v])
p4t <- function(v) pt(co4$p[co4$term == v])
co3l <- as.data.frame(coef(summary(M3l)))
names(co3l) <- c("beta","se","z","p"); co3l$term <- rownames(co3l)
b3l <- function(v) sprintf("%+.3f", co3l$beta[co3l$term == v])

## What autocorrelation does to every environmental coefficient, on the same rows.
shift <- merge(co3l[, c("term","beta")], co4[, c("term","beta")], by = "term",
               suffixes = c("_env", "_auto"))
shift <- shift[!grepl("^\\(Intercept|^geomorphon|^lag_", shift$term), ]
shift$retained <- shift$beta_auto / shift$beta_env
shift <- shift[order(-abs(shift$beta_env)), ]
AUC3l <- as.numeric(pROC::auc(pROC::roc(bl$modhigh, fitted(M3l), quiet = TRUE)))
AUC4  <- as.numeric(pROC::auc(pROC::roc(bl$modhigh, fitted(M4),  quiet = TRUE)))

## ---- the 16-day series -------------------------------------------------------------
## The annual response gives eight observations in time, so a wind effect can only be
## tested as a comparison between summers. Landsat repeats every 16 days, which is the
## finest cadence the sensor supports, and 42- to 45- rebuild the whole chain at that
## cadence: imagery, classification, terrain-resolved wind and the modelling table. Wind
## then varies both across the grid within an epoch and between the 59 epochs, so its
## effect is identified in space and in time rather than across eight annual values.
de <- read.csv(file.path(BC, "model-data", "epoch_model_table.csv"))
EP_N     <- nrow(de); EP_EPOCHS <- length(unique(de$t)); EP_YEARS <- length(unique(de$year))
epsum    <- read.csv(file.path(BC, "epoch-response", "epoch_summary.csv"))
epwind   <- read.csv(file.path(BC, "covariates", "wind-epoch", "epoch_wind_summary.csv"))

EV <- c("elevation","PINE_BA","LIVE_STAND_VOLUME_125","VRI_LIVE_STEMS_PER_HA",
        "QUAD_DIAM_125","CROWN_CLOSURE","tri","vrm","tpi","midslope_position",
        "valley_depth","curv_prof","solar_season_direct","solar_flight_direct",
        "wind_effect","ep_wind_mean")
be <- de; for (v in EV) be[[v]] <- (be[[v]] - mean(be[[v]])) / sd(be[[v]])
be$geomorphon <- droplevels(factor(GEO_LAB[round(de$geomorphons)], levels = GEO_LAB))
be$geomorphon <- relevel(be$geomorphon,
                         ref = names(sort(table(be$geomorphon), decreasing = TRUE))[1])
EINT <- c("VRI_LIVE_STEMS_PER_HA:ep_wind_mean", "LIVE_STAND_VOLUME_125:ep_wind_mean")

E1 <- glm(reformulate(c(EV, "geomorphon", EINT), "modhigh"), data = be, family = binomial)

## The same model with the previous epoch in the same season entered, so the interaction
## is read after within-season spread rather than alongside it.
esel <- !is.na(be$ep_lag_self) & !is.na(be$ep_lag_nbr90)
bel <- be[esel, ]
bel$geomorphon <- relevel(droplevels(bel$geomorphon),
                          ref = names(sort(table(droplevels(bel$geomorphon)), decreasing = TRUE))[1])
for (v in c("ep_lag_self","ep_lag_nbr90")) bel[[v]] <- (bel[[v]] - mean(bel[[v]])) / sd(bel[[v]])
E2 <- glm(reformulate(c(EV, "geomorphon", "ep_lag_self", "ep_lag_nbr90", EINT), "modhigh"),
          data = bel, family = binomial)

ce1 <- as.data.frame(coef(summary(E1))); names(ce1) <- c("beta","se","z","p"); ce1$term <- rownames(ce1)
ce2 <- as.data.frame(coef(summary(E2))); names(ce2) <- c("beta","se","z","p"); ce2$term <- rownames(ce2)
be1 <- function(v) sprintf("%+.3f", ce1$beta[ce1$term == v])
ze1 <- function(v) sprintf("%.2f", ce1$z[ce1$term == v])
pe1 <- function(v) pv(ce1$p[ce1$term == v])
pe1t <- function(v) pt(ce1$p[ce1$term == v])
be2 <- function(v) sprintf("%+.3f", ce2$beta[ce2$term == v])
ze2 <- function(v) sprintf("%.2f", ce2$z[ce2$term == v])
pe2 <- function(v) pv(ce2$p[ce2$term == v])
pe2t <- function(v) pt(ce2$p[ce2$term == v])
AUCE1 <- as.numeric(pROC::auc(pROC::roc(be$modhigh,  fitted(E1), quiet = TRUE)))
AUCE2 <- as.numeric(pROC::auc(pROC::roc(bel$modhigh, fitted(E2), quiet = TRUE)))
EP_LAG_N <- nrow(bel)

aic <- data.frame(Model = c("M0 host size + shading + landform", "M1 + stand density",
                            "M2 + terrain and flight radiation", "M3 + interactions"),
                  df = c(length(coef(M0)), length(coef(M1)), length(coef(M2)), length(coef(M3))),
                  AIC = c(AIC(M0), AIC(M1), AIC(M2), AIC(M3)))
aic$dAIC <- aic$AIC - min(aic$AIC)

co3 <- as.data.frame(coef(summary(M3)))
names(co3) <- c("beta", "se", "z", "p")
co3$term <- rownames(co3)

## Accessors, so no coefficient is ever transcribed into the prose by hand.
b3 <- function(v) sprintf("%+.3f", co3$beta[co3$term == v])
z3 <- function(v) sprintf("%.2f", co3$z[co3$term == v])
p3 <- function(v) pv(co3$p[co3$term == v])
## The "t" variants carry the relational operator for use inside a sentence.
p3t <- function(v) pt(co3$p[co3$term == v])
DENS_V <- if (length(DENS)) DENS[1] else NA_character_
WGEO_V <- if (length(WGEO)) WGEO[1] else NA_character_
WTMP_V <- if (length(WTMP)) WTMP[1] else NA_character_
I_GEO  <- paste0(DENS_V, ":", WGEO_V)
I_TMP  <- paste0(DENS_V, ":", WTMP_V)

## Retention of an environmental coefficient once previous-year pressure is in the model.
## Defined here rather than beside the fit because it needs the interaction names above.
## What each radiation surface is actually correlated with on this landscape. Computed
## rather than asserted, because the interpretation of the shading result turns entirely
## on whether radiation is separable from wind exposure here, and on this site it is
## separable for one of the two surfaces and not for the other.
cc <- function(a, b) sprintf("%+.3f", cor(d[[a]], d[[b]]))
ccn <- function(a, b) cor(d[[a]], d[[b]])

ret <- function(v) sprintf("%.2f", shift$retained[shift$term == v])
I_TMP_R <- ret(I_TMP)
```

```{r}
#| label: diameter
#| echo: false
#| cache: true
#| output: false

bz$above_qmd <- as.integer(bal$QUAD_DIAM_125 >= QMD_THRESH)
Mc <- glm(f(setdiff(c(V_CTX, DENS, WGEO), "QUAD_DIAM_125")), data = bz, family = binomial)
Md <- glm(f(c(setdiff(c(V_CTX, DENS, WGEO), "QUAD_DIAM_125"), "above_qmd")),
          data = bz, family = binomial)
qmd_bin <- bal |>
  mutate(bin = cut(QUAD_DIAM_125, breaks = c(0, 15, 20, 25, 30, 40, Inf),
                   labels = c("<15","15-20","20-25","25-30","30-40",">40"))) |>
  group_by(bin) |> summarise(n = n(), attack = mean(modhigh))
```

```{r}
#| label: sensitivity
#| echo: false
#| cache: true
#| output: false

## Objective 6. The two decisions not fixed by the data are which covariates enter and
## how wind is summarised in time. Both are varied and the density coefficient reported.

dens_beta <- function(mod, v) if (v %in% rownames(coef(summary(mod))))
  coef(summary(mod))[v, "Estimate"] else NA_real_

sens <- bind_rows(
  data.frame(Specification = "Full model M3",
             beta = dens_beta(M3, DENS[1])),
  data.frame(Specification = "No terrain terms",
             beta = dens_beta(M1, DENS[1])),
  data.frame(Specification = "Elevation as a spline",
             beta = {
               m <- mgcv::gam(as.formula(paste("modhigh ~ s(elevation) + geomorphon +",
                    paste(setdiff(c(V_CTX, DENS, WGEO), "elevation"), collapse = "+"))),
                    data = bz, family = binomial)
               s <- summary(m)$p.table
               if (DENS[1] %in% rownames(s)) s[DENS[1], "Estimate"] else NA_real_ })
)

for (w in WIND_V) {
  m <- glm(f(c(V_CTX, DENS, WGEO, w)), data = bz, family = binomial)
  sens <- bind_rows(sens, data.frame(
    Specification = paste("Wind summarised as", w), beta = dens_beta(m, DENS[1])))
}
```

```{r}
#| label: term-guard
#| echo: false
#| output: false

## Every term this manuscript names is checked against the model it is quoted from,
## because the failure mode is silent. pv() errors on a zero-length subset but b3() does
## not: sprintf over numeric(0) returns character(0), which knitr renders as nothing at
## all. Before 2026-08-27 this file quoted a growing-season radiation coefficient that is
## not in its own model and printed an empty gap mid-sentence in the docx. A guard that
## only catches the erroring half is not a guard, so the presence of every named term is
## asserted here instead.

## Statistical reporting follows the journal's own recent practice, measured from 11
## open-access Journal of Applied Entomology papers of 2025 and 2026 by
## 02.inputs/literature/audit-jae-stats-style.py. Across 76 p-value expressions the
## form is a decimal to three places with "p < 0.001" as the floor; scientific notation
## appears once in 76. R's own %.3g gives "8.29e-08" and "< 1e-16", which is the R
## default rather than the journal's convention, so it is replaced here.
##
## Two functions because a p-value is written differently in a sentence and in a table.
## In prose the relational operator has to travel with the number, or "p = " followed by
## "< 0.001" prints "p = < 0.001".
pv <- function(p) if (!length(p) || is.na(p)) NA_character_ else
  if (p < 0.001) "< 0.001" else sprintf("%.3f", p)
pt <- function(p) if (!length(p) || is.na(p)) NA_character_ else
  if (p < 0.001) "< 0.001" else sprintf("= %.3f", p)

named <- function(tbl, terms, what) {
  missing <- setdiff(terms, tbl$term)
  if (length(missing))
    stop("term-guard: ", what, " does not contain ", paste(missing, collapse = ", "),
         ". The prose quotes it, so either the selection changed or the prose is wrong.")
  invisible(TRUE)
}

named(co3, c("solar_flight_direct", "northness", "LIVE_STAND_VOLUME_125",
             "VRI_LIVE_STEMS_PER_HA", "CROWN_CLOSURE", "QUAD_DIAM_125", "PINE_BA",
             "tri", "vrm", "wind_effect", "elevation", I_GEO,
             paste0(DENS[1], ":", FSUN[1])), "the full annual model M3")
named(cnr, I_GEO, "the annual model fitted without flight-window radiation")
named(co5, c(WMM, INT_MM), "the terrain-resolved annual model M5")
named(co4, c("lag_self", "lag_nbr90"), "the autologistic annual model M4")
named(ce1, EINT, "the 16-day model E1")
named(ce2, c(EINT[2], "ep_lag_nbr90"), "the 16-day model E2 with the within-season lag")
stopifnot(all(c("tri", "valley_depth", "midslope_position") %in% shift$term))

## Growing-season direct radiation did not survive selection into M3, so the shading
## pathway is carried by northness. That is asserted rather than assumed, because the
## whole shading result is read off northness below.
stopifnot(!("solar_season_direct" %in% co3$term))
```

# Abstract {.unnumbered}

```{r}
#| label: abstract-wordcount
#| echo: false
#| output: false

## The abstract is the last thing measured, because Journal of Applied Entomology
## caps it at 300 words and an abstract over the cap is a desk rejection. The count
## is taken from the rendered text, not from the source, so inline expressions are
## counted as the words they become.
ABS_LIMIT <- 300
```

Refugia from mountain pine beetle (*Dendroctonus ponderosae* Hopkins) are hypothesised to
form where stands are thin enough for wind to disrupt the pheromone plume coordinating mass
attack. The hypothesis names topographic shading, low host density and few large-diameter
hosts, and none has been fitted. We tested all three over `r fmt(AREA_HA)` ha of the Selkirk
Mountains, British Columbia, at 30 m across `r fmt(diff(ELEV_R))` m of relief, using Landsat
at its 16-day repeat: `r EP_EPOCHS` epochs across `r EP_YEARS` outbreak years,
`r fmt(EP_N)` cell-epochs. Predictors were inventory stand structure,
`r length(GEO_V)` geomorphometric surfaces, solar radiation computed separately for the
flight window and the growing season, and a terrain-resolved wind field from the MicroMet
model driven by hourly station data for each epoch's own sixteen days.

The wind-disruption mechanism was supported. Stand density interacted negatively with wind,
`r be1(EINT[1])` for stem density (p `r pe1t(EINT[1])`) and `r be1(EINT[2])` for standing
volume (p `r pe1t(EINT[2])`), and both survived a within-season spread term carrying
`r be2("ep_lag_nbr90")`. Host held through standing volume, `r b3("LIVE_STAND_VOLUME_125")`,
and attack peaked at
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent at 25 to 30 cm
diameter, the source-sink threshold. Topographic shading failed on its surrogate: growing-season
radiation did not survive selection, and northness, which correlates
`r cc("solar_season_direct", "northness")` with it, is `r b3("northness")`, so shaded ground
carries more attack.

Temporal resolution decided the outcome. With one map a year, the same wind field, covariates
and threshold gave `r b5(INT_MM)` (p `r p5t(INT_MM)`), the opposite sign. Two further
specification tests showed why proxies mislead here: flight-window radiation cut the density
by terrain-exposure interaction from `r bnr(I_GEO)` to `r b3(I_GEO)` without changing its
sign, and previous-year pressure left valley depth at `r ret("valley_depth")` of its value
while mid-slope position moved to `r ret("midslope_position")`. A wind test aggregated to the year answers a different question from the one
the mechanism poses, and answers it wrongly.

# Introduction

Mountain pine beetle (*Dendroctonus ponderosae* Hopkins) has killed more lodgepole pine
(*Pinus contorta* Douglas ex Loudon) in British Columbia than any other agent on record
[@taylor2003; @sambaraju2021]. Where a landscape is not killed uniformly the surviving
patches shape what the forest becomes, and one framework for them is disturbance refugia,
places buffered from disturbance over time [@krawchuk2020]. Explaining refugia rather than
merely locating them requires mechanistic links to landscape drivers such as terrain, soil
and stand structure [@cartwright2018].

@krawchuk2020 proposed such a mechanism: refugia could occur "in areas with cooler
temperatures (eg from topographic shading) that protect trees from water stress; in areas
with lower host density, allowing for greater wind disruption of beetle pheromone
communication and more vigorous tree growth and chemical defenses; and in areas with fewer
large-diameter host trees". Three mechanisms, two of them terrain, and none fitted spatially:
of the studies our screen retained, only a companion study on this landscape enters a wind
term, and its response is conifer regeneration rather than attack [@murphy2026].

The wind half is a claim about canopy, not open air. @cartwright2018, the one retained study
designed to model insect-refugium controls, found them in stands of low basal area: "thinner
stands also increase wind penetration, helping to disperse beetle pheromones and disrupt
chemical communications needed to coordinate attacks." @powell2014 state the converse as the
condition for outbreak: "higher local host density, which minimizes pheromone plume
dispersion, reduces wind, and promotes successful switching to nearby hosts, positively
influences outbreak propensity." A model that controls stand density out and then reads a
terrain coefficient as a wind effect has removed the pathway it set out to test.

Terrain acts through more than airflow. Flight is thermally gated at 19 to 41 degrees C, most
of it between 22 and 32, and "most flights occur on bright sunny days, and peak flight is in
the early to mid afternoon" [@safranyik2006chap1; @mccambridge1971]. Radiation during the
flight window is therefore a different quantity from radiation across the season, and the
season quantity is the one the shading pathway concerns. Light, temperature and wind are named
as one triad rather than as alternatives, stand density affecting "tree vigour and within-stand
microclimate, which in turn influence success of bark beetle dispersal, host selection, attack
or brood development" [@safranyik2006chap1], an argument @bartos1989 put in their title as
microclimate against tree vigour. Cool sites also push the beetle toward a two-year cycle
[@sambaraju2021], so shading and vigour predict the same sign by independent routes.

No product reports the wind the hypothesis concerns, an instantaneous below-canopy speed at
flight height during the flight period. Gridded climatologies report a long-run mean at 10 m
over open ground and are downscaled over a digital elevation model, making them partly a
transform of the terrain offered alongside them [@badger2014; @davis2023]. Nor does the
literature state the interval over which wind should be summarised. We therefore take wind
from station observations of each 16-day period, modify it over the terrain with the MicroMet
model [@liston2006], and test the response at Landsat's own 16-day repeat rather than annually.

@tbl-hypotheses gives the direction expected of each attribute and the reasoning behind it.
The objectives were to test whether stand density predicts moderate-to-high disturbance as the
pheromone mechanism requires, whether topographic shading predicts it against the microclimate
and adaptive-seasonality arguments, whether terrain shape predicts it independently of density
and radiation, whether the density effect is conditional on wind, and how much the answer
depends on temporal resolution.

```{r}
#| label: tbl-hypotheses
#| echo: false
#| tbl-cap: "Landscape attributes entered in this study, the direction expected of each, and the reasoning behind that expectation."

tibble::tribble(
  ~Attribute, ~Expected, ~Rationale,
  "Stand basal area, volume, crown closure, stems", "positive",
  "Denser canopy holds the pheromone plume together; thinner stands admit wind that disperses it [@krawchuk2020; @cartwright2018; @powell2014].",
  "Quadratic mean diameter", "positive above 25 cm",
  "Trees under 25 cm are beetle sinks, over 25 cm are sources [@carroll2004bionomics].",
  "Lodgepole pine cover and basal area", "positive",
  "Attack cannot occur where the host is absent; a cell without pine is not a refugium [@cartwright2018].",
  "Flight-period direct radiation", "positive",
  "Flight is gated between 19 and 41 degrees C and peaks on bright afternoons, so sunlit slopes are reachable [@safranyik2006chap1; @mccambridge1971].",
  "Growing-season direct radiation", "negative",
  "Shading cools, relieving water stress and funding defence, and cool sites also push the beetle toward a two-year cycle [@krawchuk2020; @sambaraju2021].",
  "Terrain exposure to wind", "negative",
  "Wind disrupts the pheromone plume, so exposed ground should carry less attack [@krawchuk2020].",
  "Terrain ruggedness", "uncertain",
  "The strongest terrain term in the companion study on this landscape, but fitted there against regeneration rather than attack [@murphy2026].",
  "Convergence, wetness, valley depth, slope position", "positive in convergent terrain",
  "Infested groups gather in draws and gullies, and deep snow insulates overwintering brood [@safranyik2006chap1].",
  "Flight-period wind speed", "negative",
  "The mechanism acts during dispersal, and no study specifies the interval, so it is taken at the flight period [@krawchuk2020; @jones2019].",
  "Elevation", "uncertain",
  "A composite of temperature, snowpack, season length and host distribution that this design cannot separate [@sambaraju2021]."
) |>
  save_tbl("table-1-hypotheses") |>
  kable(booktabs = TRUE, align = "lll")
```

# Materials

## Study area


The study area is `r fmt(AREA_HA)` ha of the Selkirk Mountains in southeastern British
Columbia, `r fmt(N_CELL)` cells at 30 m, spanning `r fmt(ELEV_R[1])` to
`r fmt(ELEV_R[2])` m and `r fmt(diff(ELEV_R))` m of relief.

The coordinate reference system is EPSG:`r EPSG`, `r CRSNAME`. This is the parent study's
grid, adopted so that results are directly comparable to it: "All disturbance and covariate
rasters were aligned to a common 30 m grid in EPSG:3153, the native resolution of the
Landsat-derived dNBR and dNDMI products", and "The plot anchor was georeferenced with a
handheld GPS receiver in EPSG:3153" [@murphy2026].

The perimeter is anchored on the 2015 Mt Midgeley fire, `r fmt(BURN_HA)` ha, which is the
parent study's site. That study reports its site as spanning "elevations from 830 to 1744 m
a.s.l."; the digital elevation model used here returns `r fmt(burn_e[1])` to
`r fmt(burn_e[2])` m inside the same perimeter, which reproduces the statement. The present
study expands beyond that hard boundary, because 480 ha cannot carry the stand-density
contrast the pheromone mechanism runs on. It does not expand arbitrarily: the perimeter is the
burn buffered 5 km and then cut to the elevation band the parent's site occupies, so the
added ground is ecologically the same kind of ground. Without that cut a 1 km buffer already
reaches the Kootenay Lake surface at 534 m, which is not a site elevation.

```{r}
#| label: fig-study-area
#| echo: false
#| fig-cap: "The study area. (a) The perimeter, the 2015 Mt Midgeley burn that anchors it, and elevation. (b) Terrain ruggedness index, the strongest terrain predictor in the parent study. (c) Stand basal area from the Vegetation Resources Inventory, the density term the pheromone-disruption mechanism runs through. All panels are EPSG:3153 at 30 m; ticks are kilometres."
#| fig-height: 5

geo <- rast(file.path(BC, "geomorphometry", "geomorphometry.tif"))
vri <- rast(file.path(SA, "vri_covariates.tif"))

base_map <- function(r, title, pal) {
  ggplot() +
    geom_spatraster(data = r) +
    geom_sf(data = burn, fill = NA, colour = "red", linewidth = 0.4) +
    scale_fill_viridis_c(option = pal, na.value = "transparent", name = NULL) +
    labs(title = title, subtitle = sprintf("EPSG:%s; ticks in km", EPSG)) +
    scale_x_continuous(labels = function(x) round(x/1000)) +
    scale_y_continuous(labels = function(x) round(x/1000)) +
    theme_minimal(base_size = 9) +
    theme(legend.key.width = unit(0.3, "cm"), axis.title = element_blank())
}

base_map(elev, "(a) Elevation (m)", "viridis") +
  base_map(geo[["tri"]], "(b) Terrain ruggedness index", "magma") +
  base_map(vri[["BASAL_AREA"]], "(c) Stand basal area (m2/ha)", "mako")
```

```{r}
#| label: fig-layers
#| echo: false
#| fig-cap: "The predictor surfaces, all EPSG:3153 at 30 m inside the study perimeter, with the 2015 burn outlined in red. (a) flight-window direct radiation, the thermal gate on flight; (b) growing-season direct radiation, the shading pathway; (c) the MicroMet wind weighting factor at the prevailing bearing; (d) terrain ruggedness; (e) stand basal area; (f) quadratic mean diameter, with the 25 cm source-sink threshold at the midpoint of the scale."
#| fig-height: 6.5

lay <- function(r, title, pal = "viridis") {
  ggplot() + geom_spatraster(data = r) +
    geom_sf(data = burn, fill = NA, colour = "red", linewidth = 0.3) +
    scale_fill_viridis_c(option = pal, na.value = "transparent", name = NULL) +
    labs(title = title) + theme_minimal(base_size = 8) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          panel.grid = element_blank(), legend.key.width = unit(0.25, "cm"))
}
mmw <- rast(file.path(BC, "covariates", "wind-micromet", "wind_weight_by_direction.tif"))
mmb <- mmw[[which.min(abs(as.numeric(sub("bin_", "", names(mmw))) - WDIR))]]

(lay(geo[["solar_flight_direct"]], "(a) Flight-window radiation (kWh/m2)", "inferno") |
 lay(geo[["solar_season_direct"]], "(b) Growing-season radiation (kWh/m2)", "inferno") |
 lay(mmb, "(c) MicroMet wind weighting", "mako")) /
(lay(geo[["tri"]], "(d) Terrain ruggedness", "magma") |
 lay(vri[["BASAL_AREA"]], "(e) Basal area (m2/ha)") |
 lay(vri[["QUAD_DIAM_125"]], "(f) Quadratic mean diameter (cm)"))
```

## Beetle disturbance


The response is annual moderate-to-high beetle disturbance, classified from Landsat
normalised difference moisture index by the method of the parent study and rebuilt for this
project. It covers `r length(YRS)` outbreak years, `r paste(range(YRS), collapse = " to ")`,
excluding 2012, the one year covered only by Landsat 7, which has flown with its scan-line
corrector off since May 2003. Annual prevalence inside the perimeter runs
`r sprintf("%.1f", 100*min(prev_yr$prev))` to `r sprintf("%.1f", 100*max(prev_yr$prev))` per
cent, pooled `r sprintf("%.1f", 100*PREV)` per cent over `r fmt(N_ROWS)` cell-years.

Years are never unioned. Unioning destroys the year-to-year spread this study measures, and
the union layer over the wider grid reached 73 per cent of the landscape, which is not
credible for a beetle outbreak.

The provincial aerial overview survey is not used as a response and supplies no training
label. It is retained only as a visual check.

```{r}
#| label: tbl-prevalence
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by year inside the study perimeter."

prev_yr |>
  transmute(Year = year, Cells = fmt(n),
            `Moderate-to-high (%)` = sprintf("%.1f", 100*prev)) |>
  save_tbl("table-2-prevalence") |>
  kable(booktabs = TRUE, align = "lrr")
```

## Stand structure


Stand structure comes from the provincial Vegetation Resources Inventory layer
`VEG_COMP_LYR_R1_POLY`, retrieved by web feature service over the perimeter. Six attributes
carry the mechanism and the stem-size threshold: total basal area, crown closure, live stems per
hectare, quadratic mean diameter over stems 12.5 cm and up, stand age, and susceptible pine
basal area formed as total basal area times the pine share of cover.

`r sprintf("%.1f", PCT_ABOVE_QMD)` per cent of cell-years sit at or above the
`r QMD_THRESH` cm source-sink threshold.

```{r}
#| label: tbl-vri
#| echo: false
#| tbl-cap: "Stand structure across the study perimeter, from the Vegetation Resources Inventory."

d |>
  select(all_of(VRI_V)) |>
  pivot_longer(everything()) |>
  group_by(Attribute = name) |>
  summarise(Min = sprintf("%.1f", min(value)), Median = sprintf("%.1f", median(value)),
            Mean = sprintf("%.1f", mean(value)), Max = sprintf("%.1f", max(value))) |>
  save_tbl("table-3-stand-structure") |>
  kable(booktabs = TRUE, align = "lrrrr")
```

The inventory is a projected operational product, not a census, and its polygons here carry
reference years spanning several decades. A polygon interpreted from late-outbreak
photography describes a stand the beetle had already worked through, so basal area and pine
cover are post-attack for part of the window. There is no unattacked vintage, so this is
reported rather than corrected, and @sec-limits returns to it.

## Geomorphometry

Terrain is described by `r length(GEO_V)` geomorphometric surfaces computed with SAGA GIS
over the full reprojected elevation model and clipped afterwards, so a search radius near the
boundary sees real ground.

Radiation is computed twice because two mechanisms need it. Flight-window insolation is direct
and diffuse radiation over 1 July to 15 August restricted to 12:00 to 17:00, the hours
@safranyik2006chap1 identify as the flight peak; growing-season insolation is the whole-day
total from 1 May to 30 September, the shading quantity @krawchuk2020's first mechanism
concerns. A single annual heat index cannot separate them: flight-window direct radiation
spans a `r sprintf("%.0f", max(d$solar_flight_direct)/min(d$solar_flight_direct))`-fold range
here against `r sprintf("%.1f", max(d$solar_season_total)/min(d$solar_season_total))`-fold for
the season total.

Exposure enters as the windward-leeward index and effective air flow height at the prevailing
bearing, the omnidirectional wind exposition index, the wind shelter index and topographic
openness. Shape enters as terrain ruggedness, vector ruggedness, topographic position and its
multi-scale form, convergence, slope, curvature and geomorphon class. Landform enters as
topographic wetness, valley depth, height above the valley floor, normalised height and
mid-slope position, because infested groups are reported in draws and gullies and deep snow
insulates overwintering brood [@safranyik2006chap1]. Aspect enters as its northward and
eastward components with the heat load index of @mccune2002.

One distinction matters. The strongest aspect statements in @safranyik2006chap1 concern the
aspect of the bole, not the slope, and are within-tree effects a 30 m raster cannot measure.
Nothing here tests them.

## Terrain-resolved wind {#sec-micromet}

```{r}
#| label: micromet-load
#| echo: false
#| output: false

mmsum <- read.csv(file.path(BC, "covariates", "wind-micromet", "micromet_summary.csv"))
mmW   <- rast(file.path(BC, "covariates", "wind-micromet", "wind_weight_by_direction.tif"))
MM_WW <- range(as.vector(minmax(mmW)))
```

Station wind interpolated from four to seven valley stations is nearly flat within a year, so
wind is also computed as a field varying in space, using the MicroMet model of @liston2006.
WindNinja, the usual tool, ships no macOS binary. MicroMet's wind component is seven equations
rather than a program, implemented directly with the paper's equation numbers against each.

Terrain slope and slope azimuth come from the elevation model (equations 12, 13). Curvature is
a cell's elevation minus the mean of the two opposite cells one curvature length scale away,
on four direction lines and averaged (equation 14); the length scale is estimated as the first
lag at which elevation autocorrelation falls below 0.5, 600 m here. Slope in the wind direction
is $\Omega_s = \beta\cos(\theta-\xi)$ (equation 15), both scaled to $[-0.5, 0.5]$. The
weighting factor is $W_w = 1 + \gamma_s\Omega_s + \gamma_c\Omega_c$ with
$\gamma_s=\gamma_c=0.5$ (equation 16), the modified speed $W_t = W_wW$ (equation 17), and
direction is diverted by $\theta_d = -0.5\,\Omega_s\sin[2(\xi-\theta)]$ (equation 18).

$W_w$ depends on direction and not speed, so it is computed once for each of 16 direction bins
and each hourly observation multiplied by the surface for its own bin; nothing is averaged
before the terrain acts on it. Over all bins $W_w$ runs `r sprintf("%.2f", MM_WW[1])` to
`r sprintf("%.2f", MM_WW[2])`. Speed and direction combine as components because averaging
degrees across the 360/0 line is meaningless. The field varies
`r sprintf("%.1f", min(mmsum$spatial_range))` to
`r sprintf("%.1f", max(mmsum$spatial_range))` km/h across the grid within a year, and is not a
repackaged terrain index: it correlates `r cc("mm_flight_mean", "wind_effect")` with the
windward-leeward index and `r cc("mm_flight_mean", "solar_flight_direct")` with flight
radiation. Its strongest association is with elevation,
`r cc("mm_flight_mean", "elevation")`.

## Flight-window wind

Station wind is Environment and Climate Change Canada hourly records, reduced only at the
last step into monthly means for June, July and August and four flight-window metrics for
1 July to 15 August: mean speed, the 95th percentile, and the fractions of hours below
5 km h and above 15 km h.

Hourly resolution buys metrics, not observations. With an annual response every hourly
reading collapses into one of `r length(YRS)` values per metric, so a wind main effect is
identified only across years and its standard error should not be believed alone. Stand
density varies cell to cell; a season's wind does not. That asymmetry is why the annual
models report an interaction rather than a main effect, and why the response was rebuilt at
16 days.

# Methods

## Variable selection {#sec-selection}


The candidate set is `r N_CAND` variables in six groups, one group per pathway the review
names: stand density, host quality, terrain exposure to wind, terrain shape, landscape
context, and flight-window wind. Selection proceeds in four
stages and is required to retain at least one variable from each pathway, so that a
collinearity filter cannot silently delete a hypothesis the review established.

Fitting uses a class-balanced sample of `r fmt(N_BAL)` cell-years, `r fmt(4000)` of each
class per year. Landscape prevalence is about `r sprintf("%.0f", 100*PREV)` per cent, and an
unbalanced fit at that prevalence reports the intercept rather than the covariates.

## Models


Four models are fitted, each adding one claim to the last, so that every term's
contribution is visible as a change in fit rather than asserted.

M0 carries Krawchuk's first and third mechanisms together with landform context: host size,
topographic shading, and the terrain that governs where cold air and snow collect. M1 adds
stand density, the second of Krawchuk's mechanisms, which is objective 1. M2 adds terrain shape,
terrain exposure to wind, and flight-window radiation, which are objectives 3 and 4. M3 adds
the interactions objective 4 requires: stand density by terrain exposure, by flight-window
radiation, and by flight-window wind.

# Results {#sec-results}


```{r}
#| label: tbl-aic
#| echo: false
#| tbl-cap: "Model comparison. Each model adds one pathway to the previous."

aic |>
  transmute(Model, `Parameters` = df, AIC = fmt(AIC), `dAIC` = fmt(dAIC)) |>
  save_tbl("table-4-model-comparison") |>
  kable(booktabs = TRUE, align = "lrrr")
```

```{r}
#| label: tbl-m3
#| echo: false
#| tbl-cap: "Full model, continuous terms. Coefficients are per standard deviation on a class-balanced sample, so the intercept is not landscape prevalence. The geomorphon landform classes are also in this model and are not reported here."

co3 |>
  filter(term != "(Intercept)", !grepl("^geomorphon", term)) |>
  arrange(desc(abs(beta))) |>
  transmute(Term = term, Beta = sprintf("%+.3f", beta), SE = sprintf("%.3f", se),
            z = sprintf("%.2f", z), p = sapply(p, pv)) |>
  save_tbl("table-5-full-model") |>
  kable(booktabs = TRUE, align = "lrrrr")
```

Every pathway earns its place: AIC falls `r fmt(aic$AIC[1] - aic$AIC[2])` when stand density
enters, `r fmt(aic$AIC[2] - aic$AIC[3])` more with terrain and flight radiation, and
`r fmt(aic$AIC[3] - aic$AIC[4])` more with the interactions (@tbl-aic).

Stand density is supported in one of its three measures. Standing volume carries
`r b3("LIVE_STAND_VOLUME_125")` log-odds per standard deviation, stem count
`r b3("VRI_LIVE_STEMS_PER_HA")`, and crown closure `r b3("CROWN_CLOSURE")`
(p `r p3t("CROWN_CLOSURE")`). The last two are negative, so as main effects the three
density measures disagree in sign and only standing volume points the way the mechanism
predicts.
Total basal area did not survive selection, its univariate discrimination being
`r sprintf("%.3f", uni$auc[uni$term == "BASAL_AREA"])`. What predicts attack is standing wood
in large stems, not how many stems there are.

Radiation enters once. Flight-window direct radiation carries `r b3("solar_flight_direct")`,
which is the thermal gate on flight. Growing-season radiation did not survive selection, so
the shading pathway is carried by northness, with which it correlates
`r cc("solar_season_direct", "northness")`. Northness is `r b3("northness")`: shaded,
north-facing ground carries more attack, against the prediction.

Terrain predicts attack independently of both. The windward-leeward index is
`r b3("wind_effect")` and elevation `r b3("elevation")`, while ruggedness is `r b3("tri")`
and not distinguishable from zero (p `r p3t("tri")`), the vector ruggedness measure agreeing
in sign at `r b3("vrm")`.
Ruggedness takes the opposite sign to the parent study's, whose response is seedling
establishment rather than attack [@murphy2026].

Fitted without flight-window radiation, stand density interacts with terrain exposure at
`r bnr(I_GEO)`, which reads as a wind result pointing the wrong way. With radiation in the
model that interaction is `r b3(I_GEO)` (p `r p3t(I_GEO)`) and density interacts with
radiation at `r b3(paste0(DENS[1], ":", FSUN[1]))`. On a range whose prevailing bearing is
`r sprintf("%.0f", WDIR)` degrees the windward slopes face west, which are also the slopes
taking afternoon sun, so part of the apparent exposure effect was the sun. Not all of it:
the interaction stays positive and far from zero, which is the wrong sign for pheromone
disruption.

```{r}
#| label: fig-interaction
#| echo: false
#| fig-cap: "The refugia mechanism as fitted. (a) Predicted probability of moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem density, all other terms held at their means; the lines cross, so wind raises attack in thin stands and lowers it in dense ones, which is the interaction. (b) The same for standing volume. (c) Coefficients of the 16-day model, with and without the within-season spread term. (d) Epoch prevalence against epoch mean wind, one point per epoch."
#| fig-height: 6

nd_base <- be[1, , drop = FALSE]
for (v in EV) nd_base[[v]] <- 0
pred_panel <- function(dv, lab) {
  qs <- quantile(be[[dv]], c(0.1, 0.9))
  gr <- expand.grid(w = seq(min(be$ep_wind_mean), max(be$ep_wind_mean), length.out = 60),
                    d = qs)
  nd <- nd_base[rep(1, nrow(gr)), ]
  nd$ep_wind_mean <- gr$w; nd[[dv]] <- gr$d
  nd$geomorphon <- factor(levels(be$geomorphon)[1], levels = levels(be$geomorphon))
  gr$p <- predict(E1, newdata = nd, type = "response")
  gr$Density <- factor(gr$d, labels = c("low (10th pct)", "high (90th pct)"))
  ggplot(gr, aes(w, p, colour = Density)) + geom_line(linewidth = 0.8) +
    scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
    labs(x = "Epoch wind (SD units)", y = "P(moderate-to-high)", title = lab) +
    theme_minimal(base_size = 9)
}
pa <- pred_panel("VRI_LIVE_STEMS_PER_HA", "(a) Stem density")
pb <- pred_panel("LIVE_STAND_VOLUME_125", "(b) Standing volume")

cf <- bind_rows(
  ce1 |> mutate(model = "E1 all epochs"),
  ce2 |> mutate(model = "E2 with lag")) |>
  filter(!grepl("^geomorphon|Intercept", term)) |>
  group_by(term) |> filter(any(abs(beta) > 0.05)) |> ungroup()
pc <- ggplot(cf, aes(reorder(term, beta), beta, colour = model)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_pointrange(aes(ymin = beta - 1.96*se, ymax = beta + 1.96*se),
                  position = position_dodge(width = 0.5), size = 0.25) +
  coord_flip() + scale_colour_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = NULL, y = "Log-odds per SD", title = "(c) 16-day model coefficients") +
  theme_minimal(base_size = 8) + theme(legend.position = "bottom")

pd <- epsum |>
  inner_join(epwind, by = c("year","epoch")) |>
  ggplot(aes(mean_kmh, prevalence)) +
  geom_point(aes(colour = factor(year)), size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.5) +
  labs(x = "Epoch mean wind (km/h)", y = "Epoch prevalence",
       colour = "Year", title = "(d) Epochs") +
  theme_minimal(base_size = 8)

(pa | pb) / (pc | pd)
```

## Wind as a measurement {#sec-micromet-result}

```{r}
#| label: tbl-micromet-model
#| echo: false
#| tbl-cap: "The wind terms under two specifications, both with previous-year pressure in the model. M4 carries station wind, which is flat within a year and identified only across years. M5 carries the terrain-resolved MicroMet field, which varies within the year and is identified in space."

data.frame(
  Model = c("M4 station wind", "M4 station wind",
            "M5 terrain-resolved wind", "M5 terrain-resolved wind"),
  Term = c(WTMP_V, I_TMP, WMM, INT_MM),
  Beta = c(b4(WTMP_V), b4(I_TMP), b5(WMM), b5(INT_MM)),
  z = c(z4(WTMP_V), z4(I_TMP), z5(WMM), z5(INT_MM)),
  p = c(p4(WTMP_V), p4(I_TMP), p5(WMM), p5(INT_MM))) |>
  save_tbl("table-6-wind-terms") |>
  kable(booktabs = TRUE, align = "lllrr")
```

This section reports the same wind field at the annual resolution the study first used, and
it is retained because the contrast with the 16-day models is the paper's methodological
result.

The terrain-resolved field carries `r b5(WMM)` log-odds per standard deviation
(z = `r z5(WMM)`, p `r p5t(WMM)`) when the response is one map per year. Windier ground
carries more moderate-to-high disturbance. The interaction the mechanism turns on is
`r b5(INT_MM)` (z = `r z5(INT_MM)`, p `r p5t(INT_MM)`), positive where the
pheromone-disruption mechanism requires negative.

At 16 days the same wind field, the same covariates and the same classification threshold
give `r be1(EINT[1])` and `r be1(EINT[2])`. Nothing about the landscape or the
wind product changed between the two. What changed is that an annual response can only ask
whether windy summers carry less attack than calm ones, and a summer contains both windy and
calm weeks. Averaging across them destroys exactly the contrast the mechanism operates on.

The station terms behave the same way and for the same reason. In this specification they
collapse to `r WTMP_V` at `r b4(WTMP_V)`, which carries no flight-window information at all.

The lesson generalises past this dataset. The mechanism concerns a plume that persists for
minutes, acting on a flight period of weeks. A test that aggregates the response to a year is
not a weaker version of the right test; it is a different question, and this study answers it
in the opposite direction.

## Previous-year pressure {#sec-autologistic}

Attack recurs where it was, by a margin that changes how every other coefficient reads. A cell
attacked one year is between
`r sprintf("%.0f", min(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))`
and
`r sprintf("%.0f", max(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))`
times more likely to be attacked the next. Entering persistence and 90 m spread separately,
after the autologistic design used for this province, gives `r b4("lag_self")` and
`r b4("lag_nbr90")`, raising discrimination from `r sprintf("%.3f", AUC3l)` to
`r sprintf("%.3f", AUC4)`.

Which environmental terms survive is the informative part. Ruggedness keeps `r ret("tri")`
of its value, a term describing a condition a cell has whether or not the beetle was there.
The landform terms are not stable: valley depth keeps `r ret("valley_depth")` while mid-slope
position moves to `r ret("midslope_position")`, growing rather than shrinking. Part of what
the landform terms measured was where the outbreak had already been, which is the same
failure the terrain-wind index showed against radiation, one level up.

## Host as diameter {#sec-diameter}


```{r}
#| label: tbl-qmd
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced sample. The 25 cm boundary is the source-sink threshold of the species' bionomics."

qmd_bin |>
  transmute(`QMD class (cm)` = bin, Cells = fmt(n),
            `Attacked (%)` = sprintf("%.1f", 100*attack)) |>
  save_tbl("table-7-diameter-class") |>
  kable(booktabs = TRUE, align = "lrr")
```

Coding host as diameter changes the picture cover alone gives. Attack peaks in the
25 to 30 cm class at
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent and falls away
above it, to `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "30-40"])` per cent at 30
to 40 cm and `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == ">40"])` per cent above 40
(@tbl-qmd). The step across the 25 cm source-sink boundary is from
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "20-25"])` to
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent.

In the full model diameter carries `r b3("QUAD_DIAM_125")` per standard deviation
(z = `r z3("QUAD_DIAM_125")`). It is negative because the relationship is not monotone: a
linear term fitted through a humped response returns the slope of its falling limb, which
is the larger part of the range. The class table, not the coefficient, is the result here,
and the point stands, because a stand's cover says nothing about whether its stems can
produce brood.

The smallest class is not evidence. It holds `r fmt(qmd_bin$n[qmd_bin$bin == "<15"])` cells
against `r fmt(qmd_bin$n[qmd_bin$bin == "25-30"])` in the modal class, and a stand below
15 cm quadratic mean diameter in an inventory whose vintage postdates the outbreak is more
likely a stand the beetle already stripped of large stems than a stand attacked at that
size.

# Discussion

## Radiation, not wind

The clearest result of this study is methodological, and it changes what a terrain variable
in this literature is allowed to be taken for.

Fitted with a terrain-wind index and no radiation, stand density interacts with terrain
exposure at `r bnr(I_GEO)`, and the natural reading is a wind effect. Add flight-window
radiation and that interaction becomes `r b3(I_GEO)`, p `r p3t(I_GEO)`, while density by
radiation appears at `r b3(paste0(DENS[1], ":", FSUN[1]))`. About a third of the apparent
exposure effect was the sun. The rest was not, and what survives keeps the sign the
mechanism forbids.

That is not a peculiarity of this dataset. A windward-leeward index and an afternoon
radiation surface are both functions of slope and aspect, and on a range whose prevailing
flight-window bearing is `r sprintf("%.0f", WDIR)` degrees the windward slopes are the
west-facing ones, which are also the slopes that take afternoon sun during the flight peak.
Any study that enters a terrain-derived wind index without a radiation term on the same
clock will attribute radiation to wind.

## What survives

@krawchuk2020 name three mechanisms. At the resolution the sensor and the insect share, two
of the three hold and the third does not.

Host density holds, and it holds in the way the mechanism specifies rather than as a bare
main effect. Standing volume predicts disturbance at `r b3("LIVE_STAND_VOLUME_125")`, and
its interaction with wind is `r be1(EINT[2])` while stem density's is `r be1(EINT[1])`. The
claim was never that dense stands are attacked more; it was that a dense canopy holds a
plume together, and that what a dense canopy buys should shrink as the wind rises. Both
halves are present.

Large-diameter host holds as a threshold rather than a slope. Attack peaks at
`r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent in the 25 to 30 cm
class, at the source-sink boundary of @carroll2004bionomics, and falls away on both sides.

Topographic shading does not hold, and it fails on a surrogate rather than on the surface
the mechanism names. Growing-season radiation did not survive variable selection, so the
pathway is carried by northness, which correlates `r cc("solar_season_direct", "northness")`
with it. Northness is `r b3("northness")`, so shaded ground carries more attack where the
mechanism requires less. Northness tests aspect, and aspect carries more than shade, so this
is the least secure of the three verdicts; the water-stress pathway @krawchuk2020 describe is
not measured here at all.

## The wind claim

The mechanism holds, and the reason earlier specifications rejected it is temporal
resolution rather than anything about the landscape.

At Landsat's own 16-day cadence, stand density interacts with terrain-resolved wind at
`r be1(EINT[1])` for stem density and `r be1(EINT[2])` for standing volume, both negative,
at p `r pe1t(EINT[1])` and p `r pe1t(EINT[2])` respectively. A dense stand is worth
less where and when the wind blows harder, which is what @krawchuk2020 predict. The result
survives entering the previous epoch of the same season, so it is not the outbreak's own
spread wearing a wind coefficient.

Every coarser specification in this study rejected the same mechanism. With one map a year
and station wind, the interaction was `r b4(I_TMP)` on eight annual values, right sign and
weak identification. With one map a year and the terrain-resolved field it was
`r b5(INT_MM)`, wrong sign. The difference is not the wind product and not the covariate
set, both of which are shared: it is that a single annual map forces the comparison to be
made between summers, and a summer is not the unit the insect flies in.

That has a consequence beyond this paper. The mechanism concerns a plume that persists for
minutes over a flight period of weeks. Any test that aggregates the response to a year is
testing whether windy years have less attack than calm years, which is a different and much
weaker question, and it will answer no. The literature has never fitted a wind term at all,
so nothing in it establishes the resolution at which one should be fitted; this study's
answer is that it must be at least as fine as the flight period itself.

Two limits stay attached to the claim. The interaction is small, roughly a tenth of the
standing volume main effect. And the wind field, though it varies in space, is still
terrain-modified station data rather than measurement on the ridge, so what is established
is that the modelled wind field behaves as the mechanism requires, not that the air did.

## Ruggedness and regeneration

The two papers now share a grid, a landscape and a terrain covariate set, and they disagree
on the sign of ruggedness: `r b3("tri")` for beetle attack here against a positive
coefficient for conifer regeneration there [@murphy2026]. Read together they describe a
landscape in which rugged ground both resists the disturbance and shelters the recovery from
it. That is a coherent picture, not a contradiction, and it is only visible because the two
studies were fitted on the same grid in EPSG:3153.

## What stays confounded

Elevation is among the largest terms in the model, `r b3("elevation")`, and elevation is not
one thing. It carries temperature, snowpack, growing-season length and the distribution of
lodgepole pine, and this design cannot take them apart. Reporting the elevation coefficient
as a result would be reporting a composite.

The same caution now applies in reverse to every terrain term in the model. Radiation took
about a third of one terrain interaction. There is no guarantee that some further unmeasured
environmental surface would not take the rest, or another.

# Conclusions

The wind-disruption mechanism proposed by @krawchuk2020 holds on this landscape, at the
resolution the sensor and the insect share. Across `r EP_EPOCHS` sixteen-day epochs, stand
density interacts negatively with terrain-resolved wind, `r be1(EINT[1])` for stem density
and `r be1(EINT[2])` for standing volume, and both survive a within-season spread term. A
dense canopy is worth less to the beetle where and when the wind blows harder, which is what
a plume-disruption mechanism requires.

Of the three mechanisms that proposal names, two hold. Host density holds through standing
volume rather than crown closure or stem count, which run the other way as main effects, and
large-diameter host holds as a threshold at 25 to 30 cm rather than as a slope, at the
source-sink boundary of @carroll2004bionomics. Topographic shading does not hold, and it
fails on a surrogate: growing-season radiation did not survive selection, so the pathway is
carried by northness, `r b3("northness")`, which correlates
`r cc("solar_season_direct", "northness")` with it.

The methodological finding is the more transferable one, and it has three parts. A terrain
index is partly an insolation measurement: entering flight-window radiation cut the density
by terrain-exposure interaction from `r bnr(I_GEO)` to `r b3(I_GEO)`, leaving the sign
unchanged. A landform variable in a spreading outbreak records the outbreak's own history:
entering previous-year pressure left valley depth at `r ret("valley_depth")` of its value
while mid-slope position moved to `r ret("midslope_position")`, against `r ret("tri")` for
ruggedness. And an annual response cannot test a mechanism that operates over weeks: the
same wind field, covariates and threshold gave `r b5(INT_MM)` at annual resolution against
`r be1(EINT[1])` at sixteen days.

Since no previous study has fitted a wind term at all [@krawchuk2020; @cartwright2018], there
is no precedent establishing the resolution at which one should be fitted. This study's answer
is that it must be no coarser than the flight period, and that a test which aggregates past it
will reject a mechanism that is there.

# Limitations {#sec-limits}

**Radiation and exposure are not separable in the flight window.** The prevailing bearing is
`r sprintf("%.0f", WDIR)` degrees, so windward slopes face west and take afternoon sun.
Flight-window radiation correlates `r cc("solar_flight_direct", "wind_effect")` with the
windward-leeward index. Its positive coefficient is consistent with the thermal gate and with
exposure raising attack, and this design cannot choose between them.

**The shading mechanism is tested on a surrogate.** Growing-season direct radiation did not
survive variable selection, so the pathway is carried by northness, which correlates
`r cc("solar_season_direct", "northness")` with it. Northness tests aspect, aspect carries
more than shade, and the water-stress pathway @krawchuk2020 describe is not measured here.
Growing-season radiation is at least separable from wind exposure,
`r cc("solar_season_direct", "wind_effect")`, which is the one thing it establishes.

**The wind field is modelled, not measured.** MicroMet modifies station observations over the
terrain. What is established is that the modelled field behaves as the mechanism requires,
not that the air did.

**The inventory postdates part of the outbreak.** Polygons interpreted after the beetle
passed describe the stand it left. Total basal area discriminates below chance,
`r sprintf("%.3f", uni$auc[uni$term == "BASAL_AREA"])`, most likely for this reason.

**The response is classified, not observed.** Disturbance is inferred from Landsat moisture
indices trained on plots cut from those indices, so reported accuracy measures separability
rather than agreement with ground mortality.

**One landscape, one outbreak.** Every conclusion is conditioned on a single mountain range
over `r EP_YEARS` years and `r EP_EPOCHS` epochs.

# Acknowledgements

This work used no external funding. Beetle disturbance was classified from Landsat
Collection 2 Level-2 surface reflectance distributed by the United States Geological
Survey; stand structure from the British Columbia Data Catalogue; terrain from Natural
Resources Canada; and wind from Environment and Climate Change Canada. The author
thanks those agencies for maintaining the open archives the study rests on.

# Conflict of interest

The author declares no conflict of interest.

# Ethics statement

This study used only publicly archived remote sensing, inventory and meteorological
data. No animals, human participants or protected material were involved, so no ethics
approval was required.

# Author contributions

**Seamus Murphy:** conceptualisation; data curation; formal analysis; investigation;
methodology; software; validation; visualisation; writing, original draft; writing,
review and editing.

# Data availability

Every dataset used here is public and none was collected by the author. Beetle disturbance
is classified from Landsat Collection 2 Level-2 surface reflectance. Stand structure is the
provincial Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved from the
British Columbia Data Catalogue web feature service by
`02.inputs/beetle-classification/34-fetch-vri.py`. Terrain derives from the Natural
Resources Canada High Resolution Digital Elevation Model [@nrcan2017], with geomorphometry
computed by SAGA GIS in `37-geomorphometry.R`. Wind is Environment and Climate Change Canada
hourly station data, retrieved by `30-wind-hourly-metrics.R` and `36-wind-direction.R`.

All derived data and the complete analysis code that reproduce every number, table and
figure in this article are archived at Dryad under
doi:PLACEHOLDER-DRYAD-DOI, and are mirrored at
<https://github.com/seamusrobertmurphy/beetle-topography-and-wind-study>.

<!-- BLOCKING BEFORE SUBMISSION. Journal of Applied Entomology mandates data sharing and
     states that data "made available on request or uploaded as supplementary files, are
     not accepted", and that a link with a DOI must sit above the references. The Dryad
     deposit has not been made, so the DOI above is a placeholder and the submission
     cannot go in until it is minted. Wiley pays the Dryad archiving charge for papers
     published in this journal. Include the "Private for Peer Review" link here and in
     the cover letter once the deposit exists. -->

# Figure legends

**Figure 1.** The study area, all panels EPSG:3153 at 30 m, with the 2015 Mt Midgeley
burn outlined in red and Kootenay Lake and the Kootenay River from the British Columbia
Freshwater Atlas. (a) elevation; (b) terrain ruggedness index; (c) stand basal area from
the Vegetation Resources Inventory.

**Figure 2.** The predictor surfaces, all EPSG:3153 at 30 m. (a) flight-window direct
radiation; (b) growing-season direct radiation; (c) the MicroMet wind weighting factor at
the prevailing bearing; (d) terrain ruggedness; (e) stand basal area; (f) quadratic mean
diameter.

**Figure 3.** The refugia mechanism as fitted. (a) Predicted probability of
moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th
percentiles of stem density, all other terms held at their means. (b) The same for
standing volume. (c) Coefficients of the 16-day model, with and without the within-season
spread term. (d) Epoch prevalence against epoch mean wind, one point per epoch.

```{r}
#| label: session
#| echo: false

sessionInfo()
```

# References
