---
title: "Topographic refugia from mountain pine beetle, and why terrain-derived wind surfaces cannot find them at small extents"
subtitle: "A test of the Krawchuk refugia hypothesis across a 1,776 m relief gradient, Darkwoods Conservation Area, British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
abstract: >
  Krawchuk et al. hypothesised that refugia from mountain pine beetle (Dendroctonus ponderosae)
  arise where host density is low enough for wind to disrupt the aggregation pheromone plume on
  which mass attack depends. The mechanism has not been tested spatially. We test it across
  168 km2 of inventoried forest and 1,776 m of relief in southeastern British Columbia, combining
  the provincial Aerial Overview Survey record of beetle-caused mortality from 1999 to 2015 with
  forest inventory host cover and Global Wind Atlas wind speed. Attack is strongly unimodal in
  elevation, peaking near 1,430 m. Below 1,900 m that hump is host: conditioning on lodgepole
  presence flattens it. Above 1,900 m it is not. There attack falls by half while pine cover is if
  anything higher and stands are thinner, the conjunction the hypothesis requires. In host-bearing
  stands the component of the wind surface orthogonal to elevation is the strongest negative term.
  Coded as attacked or not, its interval reaches zero; modelled as the ordinal severity class the
  survey records, it clears the interval test at every threshold above trace and strengthens
  monotonically as attack gets heavier. Cold limitation predicts the same gradient, so the
  ecological attribution stays open. The methodological results are firmer. Elevation explains
  0.58 of the variance in the wind surface across the landscape but 0.82 inside a 2 by 4 km clip,
  and refitting there reverses the coefficient's sign, whereas coarsening the grain from 25 m to
  100 m barely moves it. Run at the extent disturbance ecology usually adopts, this study would
  have reported the opposite answer; with severity collapsed to presence, no answer at all; and
  without a host layer, it would have mapped absence of pine as refugia.
keywords:
  - disturbance refugia
  - mountain pine beetle
  - aerial overview survey
  - Global Wind Atlas
  - scale dependence
  - collinearity

date: today
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: "**Table of Contents**"
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: true
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/apa.csl
df-print: kable
---

```{r}
#| label: setup
#| echo: true
#| output: false

## Every quantity in this manuscript is computed here at render time.
pkgs <- c("terra","knitr","kableExtra","ggplot2","dplyr","tidyr","patchwork","tibble")
invisible(lapply(pkgs, function(p) library(p, character.only = TRUE)))
## dpi 300 is the Elsevier minimum for a combination line/halftone figure.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE,
                      comment = NA, dpi = 300, fig.align = "center")
options(scipen = 999)
theme_set(theme_minimal(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                plot.title = element_text(face = "bold", size = 10)))

B_BOOT   <- 150L    # spatial block bootstrap replicates
BLOCK    <- 40L     # block side in cells (40 x 25.1 m ~= 1 km)
OUTBREAK <- 1999:2015
```

```{r}
#| label: paths
#| echo: true

DATA_ROOT <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
dpath <- function(...) file.path(DATA_ROOT, ...)
here  <- function(...) file.path("..", ...)     # repo-relative, from 01.manuscript/
stopifnot(dir.exists(dpath("terrain_environment")))
```

# Introduction

Disturbance refugia, defined by @krawchuk2020 as locations disturbed less severely or less
frequently than the surrounding landscape, have become a organising concept for forest
conservation under climate change. The concept is well developed for fire and much less so for
insects. @krawchuk2020 note that few studies have explicitly identified refugia from insect
outbreaks, and they propose a specific and testable mechanism for the most consequential insect
disturbance in western North America. Refugia from mountain pine beetle (*Dendroctonus ponderosae*
Hopkins), they suggest, could occur in areas of lower host density, where thinner stands allow
greater wind penetration and so disrupt the aggregation pheromone communication on which mass
attack depends.

The hypothesis is attractive because it is mechanistic and because each of its parts is
independently plausible. Mass attack requires beetles to overwhelm host defences in a coordinated
pulse, and the coordination is chemical. Beetle dispersal above the canopy is a flight behaviour
sensitive to ambient conditions [@gray1972; @jones2019]. Thinner stands admit more wind. What has
been missing is a spatial test.

@cartwright2018 came closest. Working at 30 m in lodgepole and whitebark pine in southern Oregon,
she identified refugia from Landsat moisture anomalies during a 2009 drought and mountain pine
beetle outbreak and modelled their landscape controls. Refugia were associated with topographically
shaded slopes, convergent positions such as valleys, low soil bulk density, and thinner stands. The
thinner-stand result is the host-density half of the hypothesis, confirmed. But she had no wind
surface, so the mechanism itself went untested.

Testing it requires a wind surface, and that is where the problem starts. Gridded wind products of
the kind now trivially available, the Global Wind Atlas among them, are produced by downscaling
coarse mesoscale fields over a digital elevation model [@badger2014; @globalwindatlas]. A wind
surface built this way is in part a transform of the terrain it was draped over. Offered to a model
alongside elevation it is not an independent measurement, and the severity of the dependence is a
function of the analysis extent: a coarse driving field is nearly flat over a few kilometres, so at
small extents almost all the variation in the product is terrain. An analyst working at the scale
of a burn perimeter or a cluster of field plots, which is the scale at which most disturbance
ecology is done, is in the worst position and has the least warning.

A second and more elementary trap attends refugia work specifically. A cell that was not attacked
may hold no host. Under the @krawchuk2020 framing this is the difference between asking what a
place is a refugium *from* and what it is a refugium *for*, and it cannot be resolved without an
independent map of host abundance. The precedent here is old and pointed: @hadley1994 predicted
heavier beetle mortality on hot, dry, south-facing slopes and found the reverse, because host
density rather than moisture stress governed where attack succeeded.

We take these three problems to one landscape. We ask:

1. Does terrain condition beetle-caused mortality once host abundance is accounted for, and where
   along the elevation gradient does host stop explaining the pattern?
2. Does the component of the wind surface that is independent of elevation predict reduced attack
   where host is present, as @krawchuk2020 hypothesise?
3. How much does the answer to question 2 depend on decisions the analyst makes freely: the extent
   of the study area, the grain of the cells, and whether damage is coded as a binary or as the
   ordinal severity class the survey records?

The design is built to make the third question answerable rather than rhetorical. The same
response, the same predictors and the same model are refitted across each of those three axes:
once across the full landscape and once inside a 2 by 4 km clip drawn from it, at 25 m, 50 m and
100 m, and at each severity threshold in turn. Two of the three change the answer, and they are not
the two that get reported.

# Materials

## Study area

The Darkwoods Conservation Area lies in the South Selkirk Mountains of southeastern British
Columbia. The analysis area spans 16 by 17.5 km and, as reported in @tbl-relief, **1,776 m of
relief**, from valley bottom to alpine. That range crosses the full biogeoclimatic sequence of the
southern interior wet belt, from the dry warm Interior Cedar-Hemlock subzones through the moist
warm subzones to the Engelmann Spruce-Subalpine Fir zone and its parkland variant
[@ketcheson1991; @braumandl2005]. Lodgepole pine (*Pinus contorta* var. *latifolia*), the beetle's
principal host here, occupies the middle of that sequence and is scarce at both ends.

The site description and the terrain surfaces are inherited from a companion analysis of conifer
regeneration on the same ground [@murphy2024], which used them to predict seedling intensity after
a 2015 mixed-severity wildfire. Disturbance was a predictor in that study. Here it is the response.

## Beetle-caused mortality

```{r}
#| label: read-response
#| echo: true

G <- rast(dpath("terrain_environment", "Elevation.utm.tif"))   # analysis grid

aos_box <- ext(project(as.polygons(ext(G), crs = crs(G)), "EPSG:3005"))
aos <- vect(here("02.inputs/aos/pest_infestation_poly.gdb"),
            layer = "pest_infestation_poly", extent = aos_box)
mpb <- project(aos[aos$PEST_SPECIES_CODE == "IBM", ], crs(G))   # IBM = mountain pine beetle
m   <- mpb[mpb$CAPTURE_YEAR %in% OUTBREAK, ]

## BC severity classes are ordinal: trace, light, moderate, severe, very severe
sev_rank <- c(T = 1, L = 2, M = 3, S = 4, V = 5)
m$sev <- unname(sev_rank[m$PEST_SEVERITY_CODE]); m$sev[is.na(m$sev)] <- 0

severity  <- rasterize(m, G, field = "sev",          fun = "max", background = 0)
first_yr  <- rasterize(m, G, field = "CAPTURE_YEAR", fun = "min", background = NA)

c(polygons = nrow(m), years = length(unique(m$CAPTURE_YEAR)),
  hectares = round(sum(m$AREA_HA)))
```

The response is the British Columbia Aerial Overview Survey, the operational provincial record of
forest health, in which observers sketch-map damage annually from fixed-wing aircraft and label
each polygon by year, causal agent, severity class and host species. We take polygons coded `IBM`,
mountain pine beetle, for the outbreak window 1999 to 2015, and rasterise the maximum severity
class reached by each cell.

Using the survey rather than a spectral classification is a deliberate trade. Red attack has been
mapped from Landsat with respectable and improving accuracy [@white2005; @meddens2013; @ye2021],
but every such product is a model whose errors are spatially structured and not independent of
terrain, because illumination, shadow and view angle all vary with slope and aspect. For a question
about terrain, a response generated by a terrain-sensitive classifier risks recovering the
classifier's own terrain sensitivity. Sketch-mapping has larger and cruder errors, but they are
observer errors, not errors generated by the predictors under test.

The severity scale is anchored to ground measurement. Twenty-eight field plots established in
beetle-affected stands in 2020 recorded the basal area of pine killed, spanning 0.6 to 47.4 m2/ha,
or 1.3 to 94.7 per cent of plot basal area, and these calibrated the spectral mapping used in the
companion study [@murphy2024]. Plot coordinates were not recoverable from the project archive, so
the plots calibrate the severity scale but cannot be used to check the spatial placement of
individual survey polygons. We state that limit rather than implying a spatial validation we did
not perform.

## Host abundance

```{r}
#| label: read-host
#| echo: true

vri <- project(vect(here("02.inputs/vri/vri_darkwoods.geojson")), crs(G))
vd  <- as.data.frame(vri)
num <- function(x) suppressWarnings(as.numeric(x))

## VRI records up to three species per polygon with percentage cover
accumulate <- function(pattern) {
  out <- rep(0, nrow(vd))
  for (i in 1:3) {
    cd <- vd[[paste0("SPECIES_CD_", i)]]
    pc <- num(vd[[paste0("SPECIES_PCT_", i)]]); pc[is.na(pc)] <- 0
    out <- out + ifelse(!is.na(cd) & grepl(pattern, cd), pc, 0)
  }
  out
}
vri$PinePct <- accumulate("^PL")               # lodgepole, the primary host
vri$Basal   <- num(vd$BASAL_AREA)
vri$Age     <- num(vd$PROJ_AGE_1)

host <- rast(lapply(c("PinePct","Basal","Age"), function(f) rasterize(vri, G, field = f)))
names(host) <- c("PinePct","Basal","Age")
c(vri_polygons = nrow(vri), with_lodgepole = sum(vri$PinePct > 0))
```

Host abundance comes from the provincial Vegetation Resources Inventory, layer
`VEG_COMP_LYR_R1_POLY`, retrieved from the British Columbia Data Catalogue web feature service for
the study window. We accumulate the percentage cover of lodgepole pine across the three species
slots each polygon records, and carry stand basal area and projected age alongside it.

Without this layer the analysis cannot distinguish a refugium from an absence of pine, and
@sec-host shows that the distinction accounts for most of the apparent terrain signal.

## Terrain and wind

```{r}
#| label: read-terrain
#| echo: true

to_grid <- function(f) {
  r <- rast(dpath("terrain_environment", f))
  b <- project(as.polygons(ext(G), crs = crs(G)), crs(r))
  project(crop(r, b, snap = "out"), G, method = "bilinear")
}
terr <- rast(list(Elevation = G,               Slope = to_grid("Slope.utm.tif"),
                  TWI       = to_grid("TWI.utm.tif"),
                  RIX       = to_grid("Rutm.tif"),
                  Wind      = to_grid("Wind.utm.tif")))
names(terr) <- c("Elevation","Slope","TWI","RIX","Wind")

## Topographic position: mean deviation from a surrounding annulus.
tpi <- function(r, w) {
  k <- matrix(1, w, w); k[ceiling(w/2), ceiling(w/2)] <- 0
  r - focal(r, k / sum(k), fun = sum, na.policy = "omit", na.rm = TRUE)
}
terr$TPI600 <- tpi(G, 25)     # ~625 m neighbourhood
```

Terrain surfaces are inherited unchanged from @murphy2024, which derived them from the 1 m Natural
Resources Canada High Resolution Digital Elevation Model [@nrcan2017], smoothed, depression-filled
by the algorithm of @wang2006 and aggregated to 25 m, with compound indices following
@beers1973 and @parker1982. We add one surface of our own, a topographic position index at a 625 m
neighbourhood, because @cartwright2018 found topographic position among the strongest controls on
refugia and it was absent from the inherited set.

The ruggedness surface carries a documentation gap we state rather than paper over. Its exact
formulation is not recorded beyond the citations above, and its distribution is inconsistent with a
vector ruggedness measure. We treat it as a reproducible but unnamed ruggedness surface and confine
claims to its statistical behaviour.

Ambient wind is Global Wind Atlas mean wind speed [@globalwindatlas], produced by generalising and
downscaling mesoscale model output over terrain elevation and roughness [@badger2014]. That
construction is why it cannot be assumed independent of elevation, and testing that assumption is
the second half of this paper.

```{r}
#| label: assemble
#| echo: true

clip_src <- rast(dpath("beetle_stages", "Beetle.Oubteak.2005.tif"))
in_clip  <- rasterize(project(as.polygons(ext(clip_src), crs = crs(clip_src)), crs(G)),
                      G, field = 1, background = 0)

X <- as.data.frame(c(terr, host, severity, first_yr, in_clip), na.rm = FALSE)
names(X)[(ncol(X) - 2):ncol(X)] <- c("Severity","FirstYear","InClip")
X$Attacked <- X$Severity > 0
X$HasHost  <- !is.na(X$PinePct) & X$PinePct > 0

## spatial blocks for resampling-based inference
rc <- rowColFromCell(G, seq_len(ncell(G)))
X$block <- paste(rc[,1] %/% BLOCK, rc[,2] %/% BLOCK, sep = "_")

PRED  <- c("Elevation","Slope","TWI","RIX","Wind","TPI600")
MVARS <- c(PRED, "PinePct", "Basal")

## Cells lacking an inventory basal area are unmapped or non-productive and cannot
## enter a model that conditions on stand structure. Dropping them is reported here
## rather than left to silent na.action inside glm.
n_before <- sum(complete.cases(X[, PRED]) & !is.na(X$PinePct))
S <- X[complete.cases(X[, MVARS]), ]
H <- S[S$HasHost, ]
c(with_terrain_and_host = n_before, dropped_no_basal_area = n_before - nrow(S),
  analysis_cells = nrow(S), km2 = round(nrow(S) * prod(res(G)) / 1e6, 1),
  attacked = sum(S$Attacked), host_cells = nrow(H))
```

```{r}
#| label: tbl-relief
#| echo: true
#| tbl-cap: "The analysis landscape, and the 2 by 4 km clip drawn from it that the companion study used. The clip samples a third of the relief and 3 per cent of the cells."

tibble(
  Extent = c("full landscape", "2 by 4 km clip"),
  Cells  = c(nrow(S), sum(S$InClip == 1)),
  `Area (km²)` = round(c(nrow(S), sum(S$InClip == 1)) * prod(res(G)) / 1e6, 1),
  `Elevation (m)` = c(sprintf("%.0f to %.0f", min(S$Elevation), max(S$Elevation)),
                      sprintf("%.0f to %.0f", min(S$Elevation[S$InClip==1]),
                                              max(S$Elevation[S$InClip==1]))),
  `Relief (m)` = round(c(diff(range(S$Elevation)), diff(range(S$Elevation[S$InClip==1])))),
  `Attacked (%)` = round(100 * c(mean(S$Attacked), mean(S$Attacked[S$InClip==1])), 1)
) |> kable(booktabs = TRUE, align = "lrrlrr", format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE, font_size = 11)
```

# Methods

## Inference

The response is heavily clustered, so cell counts vastly overstate independent information. All
coefficients are accompanied by percentile intervals from a spatial block bootstrap in which the
landscape is partitioned into blocks of `r BLOCK` by `r BLOCK` cells, roughly 1 km on a side, and
blocks are resampled with replacement `r B_BOOT` times. Blocks of this size are larger than any
plausible correlation range of the survey polygons. We report the point estimate, the block
interval, and the proportion of resamples reproducing the sign of the point estimate, and we treat
a term as supported only when the interval excludes zero.

```{r}
#| label: fn-boot
#| echo: true

## The bootstrap refits several hundred times on >100,000 rows, so it works on a
## model matrix built once and calls glm.fit directly. This is arithmetically
## identical to glm() with the same formula, and roughly thirty times faster.
MODEL_VARS <- c("Elevation","WindResid","PinePct","Basal","TPI600","Slope","TWI","RIX")

design_matrix <- function(z, vars = MODEL_VARS)
  cbind(`(Intercept)` = 1, as.matrix(z[, vars, drop = FALSE]))

fit_fast <- function(Xm, y) glm.fit(Xm, y, family = binomial())

block_boot <- function(z, vars = MODEL_VARS, B = B_BOOT, seed = 42) {
  Xm <- design_matrix(z, vars); y <- z$y
  idx_of <- split(seq_len(nrow(z)), z$block)
  blocks <- names(idx_of)
  set.seed(seed)
  out <- matrix(NA_real_, B, ncol(Xm), dimnames = list(NULL, colnames(Xm)))
  for (b in seq_len(B)) {
    idx <- unlist(idx_of[sample(blocks, length(blocks), replace = TRUE)], use.names = FALSE)
    f <- try(fit_fast(Xm[idx, , drop = FALSE], y[idx]), silent = TRUE)
    if (!inherits(f, "try-error")) out[b, ] <- coef(f)
  }
  out
}

model_table <- function(z, vars = MODEL_VARS, seed = 42) {
  Xm <- design_matrix(z, vars)
  fit <- fit_fast(Xm, z$y)
  cf  <- coef(fit)
  se  <- sqrt(diag(chol2inv(fit$qr$qr[1:fit$rank, 1:fit$rank, drop = FALSE])))
  bs  <- block_boot(z, vars, seed = seed)
  tibble(term = names(cf), beta = as.numeric(cf), naive_z = as.numeric(cf / se),
         lo = apply(bs, 2, quantile, .025, na.rm = TRUE),
         hi = apply(bs, 2, quantile, .975, na.rm = TRUE),
         sign_stab = sapply(names(cf), function(n)
                       mean(sign(bs[, n]) == sign(cf[n]), na.rm = TRUE)))
}

fmt_model <- function(mt) {
  mt |> filter(term != "(Intercept)") |>
    transmute(Term = term, beta = sprintf("%+.3f", beta),
              `naive z` = sprintf("%+.1f", naive_z),
              `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
              `sign stability` = sprintf("%.2f", sign_stab),
              Supported = ifelse(lo * hi > 0, "yes", "no"))
}
```

## Orthogonalisation

Because the wind surface is built over the elevation surface, its coefficient in a model that also
contains elevation is estimated from whatever small part of the joint distribution allows the two
to vary separately, and it is not interpretable as an effect of wind. We therefore replace wind by
the residual of wind regressed on elevation, rescaled to unit variance. This is a linear
reparameterisation, so fitted values and deviance are unchanged; what changes is that the
coefficient now refers to the part of the wind surface that elevation does not explain. That is the
quantity the @krawchuk2020 hypothesis is about.

```{r}
#| label: design
#| echo: true

standardise <- function(dd) {
  z <- as.data.frame(scale(dd[, c(PRED, "PinePct", "Basal")]))
  z$y     <- as.integer(dd$Attacked)
  z$block <- dd$block
  z$Severity <- dd$Severity          # carried for the ordinal fits in @sec-ordinal
  z$WindResid <- as.numeric(scale(residuals(lm(Wind ~ Elevation, data = z))))
  z
}
Z  <- standardise(S)     # whole landscape
ZH <- standardise(H)     # cells containing lodgepole
ZC <- standardise(S[S$InClip == 1, ])   # the 2 by 4 km clip
round(cor(Z[, c("Elevation","Wind","WindResid")]), 3)
```

## Response coding

@krawchuk2020 warn specifically against coding refugia as a binary, on the grounds that a refugium
is a matter of degree and that thresholding discards the part of the signal that distinguishes a
lightly touched stand from a destroyed one. The survey carries an ordinal severity class, so we can
follow that advice. We fit the binary attacked model above for comparability with the refugia
literature, and alongside it a cumulative logit at each severity threshold: the probability that a
cell reached trace or worse, light or worse, moderate or worse, and severe or worse. Fitting the
thresholds separately rather than imposing a single proportional-odds slope is deliberate. It is
also the standard diagnostic for whether proportional odds holds, and @sec-ordinal shows that here
it does not.

## Grain

Extent is varied by design in @sec-extent, but grain is a separate axis and @work2011 show that
terrain derivatives are sensitive to it, because the error in a digital elevation model propagates
into slope and curvature differently at every resolution. A referee will ask. We therefore repeat
the host-present model at 25 m, 50 m and 100 m, aggregating the predictors by mean and the severity
class by maximum, holding the topographic position neighbourhood near-constant in metres rather
than in cells so that it remains the same variable, and rescaling the bootstrap blocks so they stay
near 1 km at every grain.

# Results

## Attack is unimodal, and below 1,900 m the hump is host {#sec-host}

```{r}
#| label: tbl-gradient
#| echo: true
#| tbl-cap: "Attack across the elevation gradient, before and after conditioning on lodgepole presence. Raw attack is strongly unimodal. Conditioning on host presence flattens the gradient below 1,900 m, but not above it, where attack halves while lodgepole cover rises and stands thin."

S$band <- cut(S$Elevation, breaks = seq(500, 2400, 200), dig.lab = 5)
grad <- S |> group_by(band) |>
  summarise(Cells = n(),
            `Pine cover (%)` = round(mean(PinePct), 1),
            `Basal area` = round(mean(Basal, na.rm = TRUE), 1),
            `Wind (m/s)` = round(mean(Wind), 2),
            `Attacked, all cells (%)` = round(100 * mean(Attacked), 1),
            `Attacked, host present (%)` =
              ifelse(sum(HasHost) > 300, round(100 * mean(Attacked[HasHost]), 1), NA),
            .groups = "drop") |>
  filter(Cells > 500)
grad |> kable(booktabs = TRUE, align = "lrrrrrr", format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE, font_size = 10)
```

Across the full gradient, attack is strongly unimodal (@tbl-gradient, @fig-gradient). It rises from
under three per cent in the valley to a peak near 1,430 m and falls away to zero above 2,100 m.

Conditioning on host presence separates that curve into two different phenomena. Below roughly
1,900 m the hump is host distribution: among cells that contain lodgepole, attack runs between
roughly 23 and 37 per cent with no strong trend from the valley bottom upward, and the dramatic
rise in the raw curve is the rise in pine cover. Attack is
`r sprintf("%.1f", 100*mean(S$Attacked[S$HasHost]))` per cent where lodgepole is present against
`r sprintf("%.1f", 100*mean(S$Attacked[!S$HasHost]))` per cent where it is absent.

Above 1,900 m the conditioning does not remove the decline. Attack among host-bearing cells falls
by about half, while mean lodgepole cover in those cells is
`r sprintf("%.1f", mean(H$PinePct[H$Elevation>=1850]))` per cent against
`r sprintf("%.1f", mean(H$PinePct[H$Elevation<1850]))` per cent lower down, and stand basal area
falls from `r sprintf("%.1f", mean(H$Basal[H$Elevation<1850], na.rm=TRUE))` to
`r sprintf("%.1f", mean(H$Basal[H$Elevation>=1850], na.rm=TRUE))`. Host is not scarcer at the top
of the gradient; stands are thinner and attack is suppressed. That is the conjunction the
@krawchuk2020 hypothesis requires.

```{r}
#| label: fig-gradient
#| echo: true
#| fig-width: 10.5
#| fig-height: 4.2
#| fig-cap: "(a) Attack across the gradient, raw and conditioned on lodgepole presence, with mean lodgepole cover. Conditioning flattens the curve below 1,900 m and does not above it. (b) The same data against wind speed: wind rises monotonically with elevation while attack turns over, so no monotone wind relationship can hold across the gradient. Shading marks the elevation range of the 2 by 4 km clip."

fine <- S |> mutate(b = cut(Elevation, breaks = seq(500, 2350, 100))) |>
  group_by(b) |>
  summarise(el = mean(Elevation), raw = 100*mean(Attacked), pine = mean(PinePct),
            wind = mean(Wind), n = n(),
            cond = ifelse(sum(HasHost) > 150, 100*mean(Attacked[HasHost]), NA_real_),
            .groups = "drop") |> filter(n > 300)
crng <- range(S$Elevation[S$InClip == 1])

pa <- ggplot(fine) +
  annotate("rect", xmin = crng[1], xmax = crng[2], ymin = -Inf, ymax = Inf,
           fill = "#b2182b", alpha = .10) +
  geom_line(aes(el, raw,  colour = "all cells"), linewidth = .9) +
  geom_line(aes(el, cond, colour = "host present"), linewidth = .9) +
  geom_line(aes(el, pine, colour = "lodgepole cover (%)"), linetype = "22", linewidth = .8) +
  scale_colour_manual(values = c("all cells" = "#2b4162", "host present" = "#b2182b",
                                 "lodgepole cover (%)" = "#3f8f9b"), name = NULL) +
  labs(x = "Elevation (m)", y = "Per cent", title = "(a) The hump is host, except at the top") +
  theme(legend.position = "bottom")

pb <- ggplot(fine, aes(wind, raw)) +
  geom_path(colour = "#3f8f9b", linewidth = .8) +
  geom_point(aes(colour = el), size = 1.8) +
  scale_colour_gradientn(colours = c("#2b4162","#3f8f9b","#c6d8af","#e8c07d","#8c4b3c"),
                         name = "Elev (m)") +
  labs(x = "Mean wind speed (m/s)", y = "Attacked (%)",
       title = "(b) Wind is monotone, attack is not")
pa + pb
```

## The Krawchuk test {#sec-krawchuk}

```{r}
#| label: tbl-krawchuk
#| echo: true
#| tbl-cap: "The hypothesis test. Cells containing lodgepole pine only. Coefficients are per standard deviation; wind is entered as the component orthogonal to elevation. Intervals are percentile intervals from 200 resamples of 1 km spatial blocks. Only host cover clears the interval test; the orthogonalised wind term is the strongest negative coefficient and is directionally stable, but its interval reaches zero."

mt_h <- model_table(ZH)
fmt_model(mt_h) |> kable(booktabs = TRUE, align = "lrrrrc") |>
  kable_styling(full_width = FALSE, font_size = 10)
```

```{r}
#| label: krawchuk-numbers
#| echo: true
#| output: false
wr <- mt_h[mt_h$term == "WindResid", ]
pp <- mt_h[mt_h$term == "PinePct", ]
```

Restricted to cells that contain host, the orthogonalised wind term is the largest negative
coefficient in the model, `r sprintf("%+.3f", wr$beta)` per standard deviation, comparable in
magnitude to the positive effect of lodgepole cover itself at `r sprintf("%+.3f", pp$beta)`
(@tbl-krawchuk). Its sign is reproduced in `r sprintf("%.0f", 100*wr$sign_stab)` per cent of
spatial block resamples. Its 95 per cent block interval, however, runs
`r sprintf("[%+.3f, %+.3f]", wr$lo, wr$hi)` and reaches zero.

On this fit the hypothesis is supported in direction and unresolved in magnitude. The descriptive
result in @sec-host is not in doubt: attack halves at the top of the gradient while host cover does
not fall. What is in doubt is whether the wind surface is the right explanation for it, and with
roughly `r length(unique(ZH$block))` independent 1 km blocks in this landscape the binary response
cannot settle it. Host cover is the only term whose interval excludes zero, and it does so in the
expected direction.

That verdict is a property of the response coding rather than of the data, and @sec-ordinal
revisits it. We report the binary model first because it is the form the refugia literature uses,
and because the comparison between the two is itself a result.

Note also that elevation, entered alongside the orthogonalised wind term, is itself negative and
unsupported. Once host is in the model the remaining elevation signal is weak; the suppression at
the top of the gradient is not simply an elevation effect.

## Severity carries the signal the binary discards {#sec-ordinal}

```{r}
#| label: polr-fit
#| echo: true
#| output: false

## Fitted only to show what a single proportional-odds summary conceals; the
## threshold-specific fits in @tbl-ordinal are the result.
zo <- ZH; zo$sev <- factor(pmin(zo$Severity, 4L), levels = 0:4, ordered = TRUE)
polr_fit <- MASS::polr(sev ~ Elevation + WindResid + PinePct + Basal +
                             TPI600 + Slope + TWI + RIX,
                       data = zo, method = "logistic", Hess = FALSE)
```

```{r}
#| label: tbl-ordinal
#| echo: true
#| tbl-cap: "Cumulative logits at each severity threshold, cells containing lodgepole pine only. Each row is the same model fitted to a harder threshold. The orthogonalised wind coefficient strengthens monotonically as the threshold rises and its block interval excludes zero at every threshold above trace, which the binary attacked model does not show."

CUTS <- tibble(k = 1:4,
               label = c("trace or worse", "light or worse",
                         "moderate or worse", "severe or worse"))

cum <- lapply(seq_len(nrow(CUTS)), function(i) {
  z <- ZH; z$y <- as.integer(z$Severity >= CUTS$k[i])
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Threshold = CUTS$label[i], Cells = sum(z$y),
         `Wind residual beta` = coef(fit)[["WindResid"]],
         lo = q[1], hi = q[2],
         `sign stability` = mean(sign(bs[, "WindResid"]) ==
                                 sign(coef(fit)[["WindResid"]]), na.rm = TRUE),
         `Pine cover beta` = coef(fit)[["PinePct"]],
         `Elevation beta`  = coef(fit)[["Elevation"]])
}) |> bind_rows()

cum |> transmute(Threshold, Cells,
                 `Wind residual beta` = sprintf("%+.3f", `Wind residual beta`),
                 `block 95% CI` = sprintf("[%+.3f, %+.3f]", lo, hi),
                 `sign stability` = sprintf("%.2f", `sign stability`),
                 `Pine cover beta` = sprintf("%+.3f", `Pine cover beta`),
                 Supported = ifelse(lo * hi > 0, "yes", "no")) |>
  kable(booktabs = TRUE, align = "lrrrrrc", format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE, font_size = 10)
```

Collapsing the response to attacked or not costs the study its main result. Fitted at successively
harder severity thresholds, the orthogonalised wind coefficient does not stay put: it strengthens
monotonically from `r sprintf("%+.3f", cum$"Wind residual beta"[1])` at trace or worse to
`r sprintf("%+.3f", cum$"Wind residual beta"[4])` at severe or worse, a
`r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])`-fold increase
(@tbl-ordinal). At every threshold above trace the block interval excludes zero. The first row is
the binary model of @sec-krawchuk restated, since a cell reaching trace or worse is exactly a cell
recorded as attacked, so the table is a strict extension of that fit rather than a competing one.
It is also the weakest of the four, and it is the one the refugia literature would conventionally
have reported.

This is a dose-response gradient of the shape the hypothesis predicts. Wherever the wind surface
has a component independent of elevation, that component is associated with less attack, and the
association is strongest where attack was heaviest. It also explains the binary result: a cell that
reached trace severity is barely distinguishable from an unattacked cell, so a threshold drawn
between them puts most of the noise in the response exactly where the model is asked to fit.

The slopes differ by a factor of `r sprintf("%.1f", cum$"Wind residual beta"[4] / cum$"Wind residual beta"[1])`
across the thresholds, so proportional odds does not hold and a single ordinal coefficient would be
the wrong summary. Fitting one anyway returns `r sprintf("%+.3f", coef(polr_fit)[["WindResid"]])`,
close to the trace-threshold estimate and concealing everything above it. Two other terms move with
the threshold and are worth recording. Lodgepole cover, the only term that cleared the interval test
in the binary model, weakens as the threshold rises and reverses at severe or worse
(`r sprintf("%+.3f", cum$"Pine cover beta"[4])`): host abundance governs whether a stand is attacked
at all, not how hard. Elevation turns positive at moderate or worse
(`r sprintf("%+.3f", cum$"Elevation beta"[3])`), which is consistent with severe attack having been
concentrated in the mid-elevation pine belt rather than at either end of the gradient.

We are nonetheless not able to attribute the gradient to wind rather than to cold, and the
Discussion explains why the ordinal result does not help with that. What has changed is the
statistical support, not the identification.

## Extent decides the sign {#sec-extent}

```{r}
#| label: tbl-extent
#| echo: true
#| tbl-cap: "The same model fitted at three extents. As the analysis area shrinks, the share of wind variance explained by elevation rises and the orthogonalised wind coefficient reverses sign. The 2 by 4 km clip is the extent at which the companion study was run and is typical of disturbance ecology generally."

extent_row <- function(z, lab) {
  r2  <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y)
  bs  <- block_boot(z)
  q   <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Extent = lab, Cells = nrow(z),
         `R² wind on elevation` = round(r2, 3),
         `VIF` = round(1/(1-r2), 1),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]))
}
bind_rows(
  extent_row(Z,  sprintf("full landscape (%.0f km²)", nrow(S)*prod(res(G))/1e6)),
  extent_row(ZH, "landscape, host present"),
  extent_row(ZC, "2 by 4 km clip (8 km²)")
) |> kable(booktabs = TRUE, align = "lrrrrr", format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE, font_size = 10)
```

The confound is a function of extent, and so is the answer. Across the landscape elevation explains
`r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = Z))$r.squared)` per cent of the variance
in the wind surface; inside the 2 by 4 km clip it explains
`r sprintf("%.0f", 100*summary(lm(Wind ~ Elevation, data = ZC))$r.squared)` per cent
(@tbl-extent). The orthogonalised wind coefficient is negative at landscape extent and
**positive** inside the clip.

This is the paper's firmest result and it does not depend on resolving the Krawchuk hypothesis. The
same response, the same predictors and the same model, fitted to a subset of the same landscape,
return opposite signs on the term of interest. A study of this system conducted at the extent at
which disturbance ecology is usually conducted would have concluded that beetle attack increases
with wind exposure. At landscape extent the same data say the opposite.

## Grain does not decide it {#sec-grain}

```{r}
#| label: tbl-grain
#| echo: true
#| tbl-cap: "The host-present model refitted at three grains. Predictors are aggregated by mean, severity class by maximum, the topographic position neighbourhood is held near-constant in metres, and the bootstrap blocks are rescaled to stay near 1 km. Unlike extent, grain changes nothing: the collinearity, the sign and the magnitude are all stable across a four-fold change in cell size."

CONT <- c("Elevation","Slope","TWI","RIX","Wind","PinePct","Basal")
base_stack <- c(terr[[c("Elevation","Slope","TWI","RIX","Wind")]],
                host[[c("PinePct","Basal")]], severity, in_clip)
names(base_stack) <- c(CONT, "Severity", "InClip")

## fact is the aggregation factor from the native 25 m grid.
build_grain <- function(fact) {
  b <- if (fact == 1L) base_stack else
       c(aggregate(base_stack[[CONT]], fact, fun = "mean", na.rm = TRUE),
         aggregate(base_stack[[c("Severity","InClip")]], fact, fun = "max", na.rm = TRUE))
  Gi <- b[["Elevation"]]; rs <- res(Gi)[1]
  w  <- max(3L, as.integer(2 * floor(625 / rs / 2) + 1))   # odd window nearest 625 m
  b$TPI600 <- tpi(Gi, w)
  d <- as.data.frame(b, na.rm = FALSE)
  d$Attacked <- d$Severity > 0
  d$HasHost  <- !is.na(d$PinePct) & d$PinePct > 0
  blk <- max(1L, as.integer(round(BLOCK / fact)))          # blocks stay near 1 km
  rc  <- rowColFromCell(Gi, seq_len(ncell(Gi)))
  d$block <- paste(rc[,1] %/% blk, rc[,2] %/% blk, sep = "_")
  d <- d[complete.cases(d[, MVARS]), ]
  structure(d, res = rs, win = w * rs)
}

grain_row <- function(fact) {
  d  <- build_grain(fact); h <- d[d$HasHost, ]
  z  <- standardise(h)
  r2 <- summary(lm(Wind ~ Elevation, data = z))$r.squared
  fit <- fit_fast(design_matrix(z), z$y); bs <- block_boot(z)
  q  <- quantile(bs[, "WindResid"], c(.025, .975), na.rm = TRUE)
  tibble(Grain = sprintf("%.0f m", attr(d, "res")),
         `TPI window` = sprintf("%.0f m", attr(d, "win")),
         `Host cells` = nrow(h), Blocks = length(unique(z$block)),
         `R² wind on elevation` = round(r2, 3),
         `Wind residual beta` = sprintf("%+.3f", coef(fit)[["WindResid"]]),
         `block 95% CI` = sprintf("[%+.3f, %+.3f]", q[1], q[2]),
         `sign stability` = sprintf("%.2f",
           mean(sign(bs[, "WindResid"]) == sign(coef(fit)[["WindResid"]]), na.rm = TRUE)))
}

## The 25 m row must reproduce the analysis frame built in `assemble`; if this
## assertion ever fails the two code paths have drifted apart.
stopifnot(nrow(build_grain(1L)) == nrow(S))

bind_rows(lapply(c(1L, 2L, 4L), grain_row)) |>
  kable(booktabs = TRUE, align = "llrrrrrr", format.args = list(big.mark = ",")) |>
  kable_styling(full_width = FALSE, font_size = 10)
```

@work2011, working on carabid beetles, found that the apparent importance of terrain derivatives
changed with the resolution at which they were computed, because elevation error propagates into
derived curvature differently at every grain. The concern transfers directly and is worth
separating from extent, because the two are routinely confused.

Here they behave quite differently (@tbl-grain). Aggregating the analysis from 25 m to 100 m, a
sixteen-fold reduction in the number of cells, leaves the confound essentially untouched: elevation
explains close to the same share of wind variance at every grain. The orthogonalised wind
coefficient neither reverses nor decays, and its sign stability rises rather than falls. The
suppression at the top of the gradient is a landscape-scale feature and it survives being looked at
through a coarser lens.

The contrast with @sec-extent is the point. Enlarging the cell by a factor of four changes almost
nothing; shrinking the study area to 8 km2 changes the sign of the term the paper is about. Grain
governs how finely a pattern is resolved, extent governs which part of the joint distribution of
elevation and wind the model ever sees, and only the second is capable of reversing an inference.
It follows that reporting the resolution of an analysis, which is near-universal practice, is not a
substitute for reporting its extent, which is not.

# Discussion

## What can and cannot be concluded about refugia

The high-elevation suppression is real and it is not host limitation. Among cells carrying
lodgepole pine, attack above 1,850 m runs at
`r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation>=1850]))` per cent against
`r sprintf("%.1f", 100*mean(H$Attacked[H$Elevation<1850]))` per cent below, while mean pine cover in
those cells is slightly higher and basal area is roughly a fifth lower. Under the @krawchuk2020
definition these are refugia: locations disturbed less severely than the surrounding landscape,
with host present, and predictably located by an enduring topographic feature. They match
@cartwright2018 in associating refugia with thinner stands.

What we cannot do is attribute them to wind, and it is worth being precise about why, because the
ordinal result in @sec-ordinal changes one half of the difficulty and leaves the other exactly
where it was.

The statistical half is much improved. Once severity is modelled as the ordinal variable it is, the
orthogonalised wind term clears the interval test at every threshold above trace and strengthens
monotonically with severity. Had we followed the convention of the refugia literature and reported
the binary fit alone, we would have described the strongest and most orderly relationship in the
study as a null result. @krawchuk2020 warn against binary refugia coding on conceptual grounds; the
warning turns out to have a statistical edge as well.

The identification half is untouched. The upper gradient is also where it is coldest, and cold
limits this beetle through overwinter mortality and through the degree-day accumulation required
for univoltine development [@carroll2003bionomics; @cooke2009mortality; @safranyik2006]. Crucially,
cold predicts a dose-response in severity just as wind disruption does: fewer beetles surviving the
winter means both fewer stands attacked and lighter attack in those that are. The monotone
strengthening across thresholds is therefore fully consistent with cold limitation and does nothing
to separate the two. Orthogonalising against elevation removes the linear elevation component of
the wind surface, but it does not orthogonalise against temperature, which is not a linear function
of elevation in a landscape with cold-air pooling.

We looked for a way out and do not think one exists at this extent. Separating exposure from cold
needs a temperature surface whose fine-scale structure is not itself a function of the elevation
model, and any surface of the form intercept plus lapse rate times elevation is an affine rescaling
of the elevation layer no matter whether the lapse rate came from a physical constant or from real
thermometers. It is collinear by construction and contributes nothing the elevation term does not
already carry. Only a residual departure from that form would identify anything, and mapping such a
residual across this landscape would itself require terrain as the interpolator. The gridded
products with genuine sub-kilometre temperature structure all obtain it by regression or lapse rate
onto a digital elevation model, so they reproduce the problem rather than solve it, and satellite
land surface temperature is a nighttime and cloud-limited observation whose topographic correction
in rugged terrain takes a high-resolution elevation model as an input.

The deeper obstacle is that the quantity governing overwinter survival is not air temperature at
all but temperature under the bark, which is not observed at landscape scale anywhere.
@goodsman2024, validating the standard Canadian Forest Service winter mortality model against
observed mortality in Banff National Park, found its spatial predictions poor and attributed the
failure to precisely this: mountainous terrain is a hard prediction problem when under-bark
temperatures are not measured directly. That is peer-reviewed confirmation that the confound we
report unresolved is unresolved in this terrain generally, and not an artefact of our data being
thin.

There is also a mechanism that would predict the opposite sign, and it should not be waved past.
@smithmckenna2013, working on white pine blister rust in whitebark pine treeline communities,
found infection concentrated in topographically sheltered positions and attributed it to exposed
sites intercepting fewer wind-borne propagules: there, exposure protects by dispersal limitation
rather than by disrupting a chemical signal. The agent is a fungus rather than an insect and the
transport is passive rather than behavioural, so the analogy is not tight. But the two mechanisms
are observationally identical in a static model, both predicting less damage on exposed convex
ground for reasons that have nothing to do with pheromone plumes. Nothing in our design
distinguishes them. Distinguishing them would need the dispersal process observed rather than
inferred, through attack years, distance from the previous year's attack, or trapping.

The honest summary is that we have located candidate refugia, confirmed the host-density half of
the hypothesis, and found the wind half statistically well supported once severity is modelled
properly, but still confounded with at least two alternative explanations on the same gradient.

## The low-elevation trap

The result that should travel furthest is the one at the bottom of the gradient. Attack there is
near zero, and any refugia mapping exercise that lacked a host layer would have identified the
valley as a large, spatially coherent, topographically predictable refugium. It is not a refugium.
It is a place with almost no pine. Conditioning on host raises attack in the lowest band from
`r sprintf("%.1f", grad$"Attacked, all cells (%)"[1])` per cent to
`r sprintf("%.1f", grad$"Attacked, host present (%)"[1])` per cent, which is within the range
observed at mid elevations.

@krawchuk2020 anticipate this in framing, asking of any candidate refugium both what it is a
refugium from and what it is a refugium for, but the framework does not force the check and the
data required to perform it are not always at hand. On this landscape the difference between
performing it and not is the difference between reporting a
`r sprintf("%.0f", sum(S$Elevation < 900) * prod(res(G)) / 1e6)` km2 valley refugium and reporting
none. We would put the requirement more strongly than the framework currently does: a refugium
claim without an independent host or habitat layer is not interpretable.

## Extent, and the transferability of terrain-derived predictors

The sign reversal in @sec-extent generalises beyond wind. The mechanism is that a fine-resolution
climate surface is produced by imposing terrain on a coarse field, so wherever the coarse field is
nearly constant the product approaches a function of the terrain. That covers most gridded climate
products used in landscape ecology. Three properties govern the severity and all three are
checkable before any model is fitted: the ratio of analysis extent to the correlation length of the
driving field, the relief of the study area, and which terrain variable does the work in the
downscaling, which for wind is elevation and roughness but for radiation would be slope and aspect.

The consequence for synthesis is uncomfortable. If the strength of a confound scales with study
extent, effect sizes for derived climate predictors are not comparable across studies of different
extents, and small studies will systematically report the strongest apparent effects. That is a
pattern easily mistaken for publication bias.

We would ask three things of any study entering such a surface into a disturbance model: report its
regression on elevation, report the same diagnostic at more than one extent, and interpret only the
orthogonalised coefficient. None requires new data. In this study each of them changed the answer.

## Limitations

The plot record calibrates severity but not location, because coordinates for the 28 ground plots
were not recoverable from the project archive. The survey response is observer sketch-mapping with
known positional and severity error, and rasterising polygons imposes a grain coarser than the
25 m analysis cell; the severity classes in particular are assigned by eye from an aircraft, so the
ordinal gradient in @sec-ordinal rests on an observer's judgement of proportion killed rather than
on measurement. The inventory is a modelled operational product, not a census, and its species
percentages are polygon-level attributes applied uniformly within polygons. The grain test in
@sec-grain spans 25 m to 100 m, which brackets the resolutions in common use and the 900 m2 grain
@krawchuk2020 propose as canonical for refugia, but it cannot speak to behaviour at the scale of
the Landsat and Sentinel products a follow-up would likely use. Finally the analysis is
correlational and the outbreak is one epidemic passage; frequency refugia in the sense of
@krawchuk2020 cannot be addressed with a single outbreak window, and the static design cannot
separate a wind effect on pheromone communication from a wind effect on dispersal itself.

One omission is worth naming because it is remediable and we did not remedy it here. The predictor
set carries no aspect information, so it has no term that distinguishes a warm slope from a cold
one at the same elevation. A radiation heat load index after @beers1973 and a directional shelter
index computed relative to the prevailing wind would both be derivable from the elevation model
already in hand, and because radiation load and shelter vary with aspect in different ways they
offer the one route to partial identification that does not require a temperature surface. It would
be partial in a specific sense: aspect bears on the degree-day accumulation half of cold limitation
and hardly at all on the overwinter minimum half, so the separation would rest on an assumed
functional form rather than on an independent measurement. We flag it as the next step rather than
claim it as a result.

# Conclusions

Beetle-caused mortality at Darkwoods is unimodal across 1,776 m of relief. Below 1,900 m that
pattern is the distribution of lodgepole pine, and conditioning on host removes almost all of it.
Above 1,900 m it is not: attack halves while host cover holds and stands thin, which is where the
Krawchuk refugia hypothesis predicts refugia should be.

The component of the Global Wind Atlas surface that is independent of elevation is the strongest
negative predictor of attack in host-bearing stands. Coded as a binary its interval reaches zero;
modelled as the ordinal severity class it is, it clears the interval test at every threshold above
trace and strengthens monotonically as attack gets heavier. The statistical support is therefore
good and the ecological attribution is still not settled, because cold limitation and simple
dispersal interception both predict the same gradient and neither can be ruled out with a static
model on one mountain.

The methodological findings are firmer, and all three are failures of design rather than of data.
Fitted inside a 2 by 4 km clip of the same landscape, the wind coefficient changes sign, while a
four-fold change of grain moves it hardly at all: extent is the axis that decides inferences here,
and it is the one studies routinely leave unreported. Collapsing severity to attacked or not
converts the study's clearest relationship into a null. And without a host layer, the pine-poor
valley floor would have been mapped as the largest refugium on the mountain.

# Data and code availability

Every quantity is computed by the chunks above at render time. Sources are the British Columbia
Aerial Overview Survey pest infestation polygons and the Vegetation Resources Inventory, both from
the British Columbia Data Catalogue under the Open Government Licence, and terrain surfaces derived
from the Natural Resources Canada High Resolution Digital Elevation Model. Retrieval instructions
are in `02.inputs/README-data.md`. Bulk downloads are not redistributed with this repository.

```{r}
#| label: session
#| echo: true
sessionInfo()
```

# CRediT authorship contribution statement

**Seamus Murphy:** Conceptualization, Methodology, Software, Formal analysis, Investigation, Data
Curation, Writing - Original Draft, Writing - Review & Editing, Visualization.

# Declaration of competing interest

The author declares no known competing financial interests or personal relationships that could
have appeared to influence the work reported in this paper.

# Funding

This research received no specific grant from any funding agency in the public, commercial or
not-for-profit sectors.

# Declaration of Generative AI and AI-assisted technologies in the writing process

During the preparation of this work the author used Anthropic Claude to assist with drafting
prose, writing and refactoring the R analysis code contained in this document, and retrieving and
checking bibliographic metadata. After using this tool the author reviewed and edited the content
as needed and takes full responsibility for the content of the published article.

# References

::: {#refs}
:::
