---
title: "Testing the wind disruption hypothesis for beetle refugia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
    affiliation: "Independent Researcher, Nelson, British Columbia, Canada"
abstract: >
  DRAFT PLACEHOLDER. Unstructured abstract, 300 words maximum, containing the
  major keywords. Not yet written: see the Word budget chunk below before
  drafting, and condense from the abstract of beetle-topography-wind-study.qmd.
keywords:
  - Dendroctonus ponderosae
  - Pinus contorta
  - elevational gradient
  - aerial overview survey
  - quadratic mean diameter
  - scale dependence
date: today
format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: false
execute:
  eval: true
  echo: true
  warning: false
  message: false
  error: false
  comment: NA
engine: knitr
citeproc: true
bibliography: ../04.references/references-beetle.bib
csl: ../04.references/apa-6th-edition.csl
---

<!--
TARGET VENUE: Journal of Applied Entomology (Wiley), Original Article.
Rules this file is built to satisfy, all quoted from
04.references/Submission Guidelines - Journal of Applied Entomology.pdf:

  Word limit   "6000 limit (without references)"
  Abstract     "an abstract of no more than 300 words containing the major keywords"
  Keywords     "six keywords. These keywords should not include words in the title"
  Structure    "Introduction, Materials and Methods, Results, Discussion (each
               separately). A Conclusions section is optional."
  Running head "A short running title of less than 40 characters"
  References   "prepared according to the Publication Manual of the American
               Psychological Association (6th edition)"
  CRediT       "Please insert author contribution in the manuscript above references."
  Data         "A link to your data must be included in your Data Availability
               Statement." Data on request or as supplementary files is NOT accepted.
  COI          "The Conflict of Interest statement should be included as part of the
               main text file."
  Footnotes    "Footnotes to the text are not allowed."
  Species      Common name followed by scientific name on first use in title,
               abstract and text.
  Tables       "supplied as editable files, not pasted as images"; footnote symbols
               in the order daggar, double-daggar, section, pilcrow; asterisks
               reserved for P-values; SD or SEM identified in the headings.
  Figures      "supplied as separate files"; colour free online, print colour
               charged; line figures preferably black and white.
  Fig legends  "supplied as a complete list in the text".
  Files        Title page, main text, figures, tables and supporting information
               are SEPARATE files. Provision of all of these is mandatory at
               submission. Splitting this render into those files is a build step,
               not something Quarto does on its own.
  Charges      "There are no charges / fees for publishing pages except for printed
               colour figures."

STRUCTURAL DEVIATION, flagged deliberately: the standing project rule is that a
manuscript carries a numbered literature review as its own section before the
Introduction. JAE mandates the four-section body above and has no slot for one.
The review therefore moves to Supporting Information as Appendix S1, keeps its
numbered findings, and is cited by number from the Introduction. Substance kept,
placement changed. Raise this if the trade is unacceptable.

Running title (goes on the separate title page file, not here):
  "Wind and beetle refugia in BC"  -- 29 characters, under the 40 allowed.
-->

```{r}
#| label: setup
#| echo: false
#| output: false

pkgs <- c("terra","knitr","kableExtra","ggplot2","dplyr","patchwork","tibble",
          "sf","ggspatial","tidyterra","ggrepel","rnaturalearth","scales","parallel")
invisible(lapply(pkgs, function(p) library(p, character.only = TRUE)))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE,
                      comment = NA, dpi = 500, fig.align = "center",
                      cache.path = "_cache-short/", autodep = TRUE, cache.lazy = FALSE)
knitr::dep_auto()
options(scipen = 999)
theme_set(theme_minimal(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                plot.title = element_text(face = "bold", size = 10)))

B_BOOT   <- 150L    # spatial block bootstrap replicates
BLOCK    <- 40L     # block side in cells (40 x 25.1 m ~= 1 km)
OUTBREAK <- 1999:2015
ALPINE   <- 1900
```

```{r}
#| label: paths
#| echo: false

DATA_ROOT <- Sys.getenv("DARKWOODS_DATA",
  "/Users/seamus/repos/publications-pending/Darkwoods-Disturbance-Paper/3.SpatialData")
dpath <- function(...) file.path(DATA_ROOT, ...)
here  <- function(...) file.path("..", ...)     # repo-relative, from 01.manuscript/
stopifnot(dir.exists(dpath("terrain_environment")))

G <- rast(dpath("terrain_environment", "Elevation.utm.tif"))
GRID_RELIEF <- diff(as.vector(minmax(G)))
```

```{r}
#| label: word-budget
#| echo: true

## The 6000-word limit is the one constraint this file exists to satisfy, so it is
## measured at render time from this file's own source rather than trusted to a
## word processor after the fact. The parent manuscript carried {{WC_TOTAL}}
## placeholders that never substituted and silently reported nothing for months;
## this chunk is the fix. It counts the four mandated body sections only, which is
## what "without references" means here, and stops the render when the body is over
## budget so an over-length draft cannot reach a submission package unnoticed.

jae_wordcount <- function(path = "manuscript-short.qmd",
                          from = "# Introduction",
                          to   = "# Conflict of Interest") {
  src <- paste(readLines(path, warn = FALSE), collapse = "\n")
  src <- sub("^---.*?\n---\n", "", src)              # YAML
  src <- gsub("<!--.*?-->", "", src)                 # HTML comments
  src <- gsub("```\\{r[^\n]*\n.*?\n```", "", src)    # code chunks
  src <- gsub("`r [^`]*`", "NUM", src)               # inline code -> one word
  i <- regexpr(from, src, fixed = TRUE)
  j <- regexpr(to,   src, fixed = TRUE)
  if (i < 0) stop("word-budget: start marker not found: ", from)
  body <- if (j > i) substr(src, i, j - 1) else substr(src, i, nchar(src))
  body <- gsub("^#+ .*$", "", body)                  # headings are not body prose
  length(strsplit(trimws(body), "\\s+")[[1]])
}

WC_BODY  <- jae_wordcount()
WC_LIMIT <- 6000L
cat(sprintf("Main text (Introduction to Discussion, references excluded): %d / %d words\n",
            WC_BODY, WC_LIMIT))
if (WC_BODY > WC_LIMIT)
  stop(sprintf("Over the JAE Original Article limit by %d words.", WC_BODY - WC_LIMIT))
```

# Introduction

DRAFT. Budget roughly 900 words. Condense from the parent manuscript's
Introduction and fold in the numbered findings of the literature review, which
now sits in Appendix S1 and is cited here by finding number. On first use the
common name is followed by the scientific name: mountain pine beetle
(*Dendroctonus ponderosae* Hopkins), lodgepole pine (*Pinus contorta* Douglas ex
Loudon). State the objectives last.

# Materials and Methods

DRAFT. Budget roughly 1,500 words. Port from the parent's Materials and Methods,
dropping the map furniture prose and keeping the study area, the response, the
host attributes, the terrain and wind surfaces, the climate records, the
orthogonalisation, and the block bootstrap.

# Results

DRAFT. Budget roughly 1,800 words. Retain the elevational distribution, the stem
diameter effects, the wind disruption test, the nonlinear respecification that
collapses the wind term, and the winter temperature comparison. The extent and
grain sensitivity analyses move to Appendix S1 and are reported here in one
sentence each.

# Discussion

DRAFT. Budget roughly 1,600 words. Keep the specification-artefact argument, the
measurement mismatch between an annual mean at 10 m over open ground and
instantaneous below-canopy speeds, the thermal alternative, and the
host-absence confounding.

# Conclusions

DRAFT. Budget roughly 200 words. Optional under JAE rules; retained because the
paper's claim is a negative result and a negative result needs a plain closing
statement.

# Conflict of Interest

The author declares no conflict of interest.

# Author Contribution

**Seamus Murphy:** Conceptualization; Data curation; Formal analysis;
Investigation; Methodology; Software; Validation; Visualization; Writing -
original draft; Writing - review and editing.

# Data Availability Statement

PLACEHOLDER, BLOCKING. JAE requires a resolvable link with a DOI, and states that
data on request or supplied as supplementary files is not accepted. A repository
deposit must exist before submission. Dryad archiving charges are paid by Wiley
for papers published in this journal, and a "Private for Peer Review" link should
be quoted here while the paper is under review.

# References

::: {#refs}
:::

# Tables

Tables are listed here complete with legend and footnotes, then exported as an
editable file for submission. Footnote symbols run in the order dagger, double
dagger, section, pilcrow; asterisks are reserved for P-values; SD and SEM are
named in the column headings.

# Figure Legends

The complete list of legends belongs here in the main text file, while the
figures themselves are uploaded as separate files.

# Supporting Information

Appendix S1 carries the literature review with its numbered findings and search
protocol, the extent and grain sensitivity analyses, and the supporting tables
and figures. Supplied as separate files, hosted without editing or typesetting.
