---
title: "Terrain shelter and flight-period wind act on mountain pine beetle attack at two scales"
subtitle: "Landing surfaces, plume disruption and refugia from *Dendroctonus ponderosae* in the Selkirk Mountains of British Columbia"
author:
  - name: Seamus Murphy
    orcid: 0000-0002-1792-0351
    email: seamusrobertmurphy@gmail.com
    corresponding: true
    affiliations:
      - name: Forest Carbon Verification, TÜV SÜD
        city: Comox
        region: British Columbia
        country: Canada
keywords:
  - mountain pine beetle
  - forest disturbance
  - insect outbreaks
  - Landsat time series
  - refugia
  - British Columbia

date: today
keep-md: true

format:
  docx:
    prefer-html: true
    reference-doc: ../04.references/style-formal.docx
    highlight-style: pygments
    number-sections: true
  html:
    page-layout: full
    toc: true
    number-sections: true
    embed-resources: true
    toc-title: ""
    toc-location: right
    toc-depth: 3
    theme:
      - ../04.references/style.scss
      - cosmo
    code-fold: false
    code-tools: true
    code-link: true
    grid:
      body-width: 1100px
      margin-width: 220px

execute:
  eval: true
  echo: false
  warning: false
  message: false
  error: false
  comment: NA

always_allow_html: true
engine: knitr
citeproc: true
editor: visual
bibliography: ../04.references/references-beetle.bib
# Journal of Pest Science cites "by name and year in parentheses", (Thompson 1990), and
# lists references in the Springer Basic author-date form with full DOI links. The style
# file is the Zotero repository's springer-basic-author-date.csl, fetched 2026-09-01. The
# journal was Journal of Applied Entomology until 2026-09-01, which wanted APA 6th; that
# file stays in 04.references but nothing reads it.
csl: ../04.references/springer-basic-author-date.csl
df-print: kable
---

```{r}
#| label: host-structure-switch
#| echo: false
#| output: false

## This draft is the annualized refit. The shared preamble reads this switch and fits
## every model on the annual VRI snapshots instead of the 2025 composite, so the abstract
## and the Results report the same models. Remove it and this file silently becomes a
## copy of the archived static-composite draft, with two extra sections that disagree
## with the rest of it.
USE_VRI_TIMESERIES <- TRUE

## FIELD-PLOT VARIANT, 2026-09-02. Reads 02.inputs/beetle-fieldplots, where the annual
## classifier is trained on the parent study's 28 ground plots, labelled by tertiles of
## killed pine basal area, plus derived undisturbed plots, and predicted over the study
## perimeter with water masked. The sixteen-day response is unchanged, because it is a
## threshold on NDMI and never used the classifier. The term guard reports rather than
## stops, so a term the prose quotes that this fit does not contain prints as blank.
BEETLE_ROOT <- "beetle-fieldplots"
GUARD_DIAGNOSTIC <- TRUE
```

{{< include _sections/_pipeline.qmd >}}

{{< include _sections/_preamble.qmd >}}

```{r}
#| label: vri-refit
#| echo: false
#| output: false

## The two comparisons this draft exists to report, read from the files the scripts wrote
## rather than retyped. 55-build-vri-timeseries-table.R builds the annualized tables and
## 56-refit-vri-timeseries.R fits both models; 52- and 53- do the same for the wind window.
VRIS <- read.csv(file.path(BC, "model-data", "vri_refit_comparison.csv"))
VSRC <- read.csv(file.path(BC, "model-data", "vri_year_source.csv"))
FWS  <- read.csv(file.path(BC, "model-data", "flight_window_sensitivity.csv"))

vr <- function(m, col) VRIS[[col]][VRIS$model == m]
b  <- function(x) sprintf("%+.4f", x)
pp <- function(x) if (x < 0.001) "< 0.001" else sprintf("= %.3f", x)

## The field-plot classifier's own diagnostics, read from the files the variant wrote.
CLS_PL <- read.csv(file.path(BC, "plot-locations", "fieldplots_with_undisturbed.csv"))
CLS_LOO <- read.csv(file.path(BC, "red-stage-darkwoods", "classifier_loo.csv"))
CLS_BR <- quantile(CLS_PL$pi_mpb_killed[CLS_PL$source == "field"], c(1/3, 2/3))
CLS_NU <- sum(CLS_PL$source == "derived"); CLS_N <- CLS_LOO$n
CLS_ACC <- CLS_LOO$accuracy; CLS_KAPPA <- CLS_LOO$kappa

VRI_YEARS <- nrow(VSRC)
VRI_SUB   <- VSRC$year[VSRC$substituted]
VRI_SUBN  <- length(VRI_SUB)

## The share of the density by terrain-exposure interaction that flight-window radiation
## takes, read from the two fits rather than typed, because the fraction is quoted three
## times and the composite draft's "two fifths" had outlived the numbers it described.
RAD_SHARE <- 100 * (1 - abs(co3$beta[co3$term == I_GEO]) / abs(cnr$beta[cnr$term == I_GEO]))
sh <- function(v, col) sprintf("%+.3f", shift[[col]][shift$term == v])
```

# Abstract {.unnumbered}

```{r}
#| label: abstract-wordcount
#| include: false

## The abstract is the last thing measured, because Journal of Pest Science asks for
## "an abstract of 150 to 250 words" and an abstract over the cap is returned unread.
## The count is taken from the rendered text, not from the source, so inline
## expressions are counted as the words they become.
ABS_LIMIT <- 250
```

Disturbance refugia from mountain pine beetle (*Dendroctonus ponderosae* Hopkins) outbreaks have been hypothesised by @krawchuk2020 to form where tree defences remain effective, on shaded ground that spares trees water stress or in thin stands with few large hosts where wind may disrupt the aggregation pheromone, and none has been fitted spatially. All three were tested across `r fmt(AREA_HA)` ha of the Selkirk Mountains, British Columbia, using `r EP_EPOCHS` sixteen-day Landsat epochs over `r spellnum(EP_YEARS)` outbreak years with annualized forest inventory, geomorphometric variables and a terrain-resolved MicroMet wind field from hourly station records. Wind disruption was supported conditionally, in that attack fell where a thin stand and strong flight-period wind coincided, the density-by-wind interaction `r be1(EINT[1])` for stem density (`r pe1T(EINT[1])`) and `r be1(EINT[2])` for standing volume (`r pe1T(EINT[2])`), both holding against within-season contagion. Terrain acted differently. Leeward ground had more attack as a main effect, `r b3("wind_shelter")`, independent of stand density, and open gentle ground above shallow valleys more still, sky view `r b3("sky_view")`, which is the pattern expected if wind-borne beetles, which land without discriminating hosts, settle where the flow slows. Attack responded to host size as a threshold, peaking at `r qb("attack_pc", "25-30")` per cent in the 25 to 30 cm quadratic mean diameter class. The shade hypothesis failed, since north-facing slopes showed more attack (`r b3("northness")`). Refugia are therefore proposed to lie on warm, windward ground in thin stands during windy flight periods, and the terrain ruggedness signal of the companion study resolves into shelter and openness.

# Keywords {.unnumbered}

mountain pine beetle, forest disturbance, insect outbreaks, Landsat time series, refugia, British Columbia

```{=html}
<!-- Journal of Pest Science asks for "4 to 6 keywords which can be used for indexing
     purposes" and sets no rule against title words. Quarto's docx writer puts the YAML
     `keywords` field into the file's core properties only, where no editor sees it, so
     they are printed here as well. The two lists must be kept identical. -->
```


# Introduction

Mountain pine beetle (*Dendroctonus ponderosae* Hopkins) has killed more lodgepole pine (*Pinus contorta* Douglas ex Loudon) in British Columbia than any other cause on record [@taylor2003; @sambaraju2021]. Mortality is not distributed evenly across a landscape, and the stands that survive supply the structure and seed from which the next forest develops. Such stands are termed disturbance refugia, places buffered from disturbance over time [@krawchuk2020]. Locating a refugium does not explain it, and explanation requires a mechanism linking survival to a measurable property of the site, such as terrain, soil or stand structure [@cartwright2018].

@krawchuk2020 proposed such a mechanism, that refugia could occur "in areas with cooler temperatures (eg from topographic shading) that protect trees from water stress; in areas with lower host density, allowing for greater wind disruption of beetle pheromone communication and more vigorous tree growth and chemical defenses; and in areas with fewer large-diameter host trees" (p. 239). These are three testable claims. Topographic shading reduces attack by relieving water stress on cool ground. Low host density reduces attack by admitting the wind that disperses the aggregation pheromone. A scarcity of large-diameter hosts reduces attack by limiting brood production. Two of the three act through terrain, and none has been fitted spatially. Of the prior work cited here, only @murphy2026 enters a wind term, in a companion study on this same landscape whose response is conifer regeneration rather than beetle attack.

That companion study is the origin of this one. @murphy2026 fitted the density of conifer seedlings after the 2015 Mt Midgeley fire against seed source distance, fire severity, the red and grey stages of the beetle outbreak, aspect, wind and terrain ruggedness, on the grounds that topography acts as "both disturbance modifier and regeneration filter", and terrain ruggedness survived its variable selection into the final set of eight covariates, entering the model for all seedlings at +0.626 (p < 0.001), the largest positive terrain term in that study. A ruggedness coefficient records that relief matters without saying which property of the relief the organism responds to. The beetle outbreak that paper entered as a covariate is the response here, and the single ruggedness index is replaced by the terrain quantities the beetle's biology names, exposure to the prevailing wind, openness to the sky, position on the slope and depth of the valley, so that the relief signal the parent study found can be read component by component.

The wind mechanism is a claim about conditions within the canopy rather than in open air. @cartwright2018, the only prior study designed to model the controls on an insect refugium, located refugia in stands of low basal area, reasoning that "thinner stands also increase wind penetration, helping to disperse beetle pheromones and disrupt chemical communications needed to coordinate attacks." @powell2014 state the converse as a condition for outbreak, that "higher local host density, which minimizes pheromone plume dispersion, reduces wind, and promotes successful switching to nearby hosts, positively influences outbreak propensity." A model that removes stand density and then reads a terrain coefficient as a wind effect has eliminated the pathway it set out to test.

Terrain acts on attack through more than airflow, because flight is itself confined to a temperature window, between 19 and 41 degrees C and mostly between 22 and 32, and most flights occur "on bright sunny days, and peak flight is in the early to mid afternoon" [@safranyik2006chap1; @mccambridge1971]. Radiation during the flight window is therefore a different quantity from radiation accumulated across the season, and it is the seasonal quantity that the shading pathway concerns. Light, temperature and wind are therefore less three competing explanations than one bundle of correlated conditions, because stand density governs "tree vigour and within-stand microclimate, which in turn influence success of bark beetle dispersal, host selection, attack or brood development" [@safranyik2006chap1] [@bartos1989]. Cool sites also slow development enough to push the beetle onto a two-year cycle [@sambaraju2021], so shade and vigour both predict less attack on cool ground by routes this design cannot separate.

Two further constraints determine the interval over which a wind term can meaningfully be measured, and both are temporal rather than spatial, the first operating within the day. @gray1972 recorded emergence in *D. ponderosae* against ambient temperature and found that emergence "began when temperatures rose above 20' C" and "ceased in the afternoon at about the same threshold temperature", with "Peak emergence occurred between 11 a. m. and 2 p. m." and "the sharp decline in activity which occurred when ambient temperatures exceeded 30' C". Flight is therefore limited at both ends and confined to a few hours of the afternoon, so a daily or monthly mean wind speed averages across many hours in which no beetle is flying.

The second constraint operates across the season and is a property of the population rather than of the individual, because mass attack is a threshold phenomenon. @howe2022 conclude that "a combination of stand-level spatial aggregation, behavioral shifts, and higher quality of attainable hosts defines a critical threshold beyond which continual population growth becomes self-driving", and @cooke2025 define the irruption threshold as "the population density at which endemic populations may transition towards the epidemic state". @carroll2004bionomics place the same argument on the thermal side, requiring "thermal environments conducive to overwintering survival and with sufficient heat accumulation to maintain a synchronous univoltine life cycle". Two consequences follow for the present design. Attack in one year is not independent of attack in the year before, which is why previous-year and neighbourhood pressure enter the models. An environmental variable acting on this system does so by regulating that threshold rather than by adding to attack [@cooke2025]. A wind effect on mass attack through the plume is therefore expected as an interaction with host density, and not as a main effect.

A third consideration concerns where a dispersing beetle comes down, and it changes what a terrain main effect can mean. @hynum1980 monitored landing on lodgepole pine with landing traps and found that the beetle "is not attracted to lodgepole pine" before the first gallery is started and that beetles "were unable to distinguish between hosts, dead hosts and nonhosts during landing", so where a beetle lands is decided by its transport rather than by the tree beneath it. Within a stand that transport is short and wind-directed, in that @safranyik1989 predicted the directional distribution of dispersing beetles from hourly wind direction, and @safranyik1992 recaptured most released beetles near 3 m above the ground and estimated that "only 0.2 % of the marked beetles dispersed above the stand canopy". Above the canopy it is long, because @jackson2008 observed beetles by radar to more than 800 m above the forest, and @chen2017 describe long-distance dispersal as three phases, "the ascent, transport, and descent", whose mechanisms "are poorly known", with unstable air under the canopy during emergence that "would help loft beetles above the forest canopy". A beetle descending from that transport arrives as a wind-borne particle, and @giroday2011 set out what follows, that landscape features "provide impactive surfaces for interception of insects", that a heavier airborne object is more likely to "impact a surface rather than be swept around the obstacle into lee-ward eddies", and that settlement rises "in areas where wind speed is reduced". Where the flow slows is the lee of ridges and sheltered ground, which is the pattern the wind shelter index of @plattner2004 was built to predict for snow.

Deposition and plume disruption therefore make different predictions, and the difference is what this design can test. Deposition acts before any host is chosen and predicts a main effect of terrain shelter that does not depend on stand density. Plume disruption acts on an aggregation already under way and predicts an interaction between stand density and wind, with no requirement that shelter act alone. A terrain coefficient read without this distinction can be assigned to the wrong mechanism.

No available product reports the quantity the plume hypothesis concerns, an instantaneous below-canopy wind speed at flight height during the flight period. Gridded climatologies report a long-run mean at 10 m over open ground and are downscaled over a digital elevation model, which makes them in part a transform of the terrain variables offered alongside them [@badger2014; @davis2023]. Wind is therefore taken from station observations of each 16-day period, modified over the terrain with the MicroMet model of @liston2006, and the response is measured at Landsat's own 16-day repeat, which is the finest interval the sensor and the flight period share.

@tbl-hypotheses sets out the direction expected of each attribute and the reasoning behind it. This study addresses five questions. Does stand density predict moderate-to-high disturbance, as the pheromone mechanism requires? Does topographic shading predict it, against the competing microclimate and adaptive-seasonality arguments? Does terrain shape predict it once density and radiation are already in the model? Is the density effect conditional on wind, which is the form plume disruption takes? Does terrain shelter act as a main effect, which is what deposition of wind-borne beetles predicts, or only through stand density, which is what plume disruption predicts?

```{r}
#| label: tbl-hypotheses
#| echo: false
#| tbl-cap: "Landscape attributes entered in this study, the direction expected of each, and the reasoning behind that expectation."

tibble::tribble(
  ~Attribute, ~Expected, ~Rationale,
  "Stand basal area, volume, crown closure, stems", "positive",
  "Denser canopy holds the pheromone plume together; thinner stands admit wind that disperses it [@krawchuk2020; @cartwright2018; @powell2014].",
  "Quadratic mean diameter", "positive above 25 cm",
  "Trees under 25 cm are beetle sinks, over 25 cm are sources [@carroll2004bionomics].",
  "Lodgepole pine cover and basal area", "positive",
  "Attack cannot occur where the host is absent; a cell without pine is not a refugium [@cartwright2018].",
  "Flight-period direct radiation", "positive",
  "Flight is confined to 19 to 41 degrees C and peaks on bright afternoons, so sunlit slopes are reachable [@safranyik2006chap1; @mccambridge1971].",
  "Growing-season direct radiation", "negative",
  "Shading cools, relieving water stress and sustaining defence, and cool sites also push the beetle toward a two-year cycle [@krawchuk2020; @sambaraju2021].",
  "Terrain exposure to wind", "negative",
  "Wind disrupts the pheromone plume, so exposed ground should have less attack [@krawchuk2020].",
  "Terrain shelter and sky view", "positive, as a main effect",
  "Beetles land without discriminating hosts after wind-borne transport, so sheltered, open ground should receive more landings whatever the stand [@hynum1980; @jackson2008; @chen2017].",
  "Terrain ruggedness", "uncertain",
  "The strongest terrain term in the companion study on this landscape, but fitted there against regeneration rather than attack [@murphy2026].",
  "Convergence, wetness, valley depth, slope position", "positive in convergent terrain",
  "Infested groups gather in draws and gullies, and deep snow insulates overwintering brood [@safranyik2006chap1].",
  "Flight-period wind speed", "negative",
  "The mechanism acts during dispersal, and no study specifies the interval, so it is taken at the flight period [@krawchuk2020; @jones2019].",
  "Elevation", "uncertain",
  "A composite of temperature, snowpack, season length and host distribution that this design cannot separate [@sambaraju2021]."
) |>
  save_tbl("table-1-hypotheses") |>
  kable(booktabs = TRUE, align = "lll", row.names = FALSE)
```

# Materials and methods

## Study area

@tbl-inventory sets out the datasets this study combines and the resolution of each. The asymmetry it shows is the one on which the analysis depends, in that the response varies every sixteen days, the inventory once a year, the station wind every hour, and the terrain not at all.

```{r}
#| label: tbl-inventory
#| echo: false
#| tbl-cap: "The datasets this study combines, with the structure and resolution of each. Every count is read from the files at render time. Spatial resolution is the grid the variable is analysed on; temporal resolution is the interval at which it varies. Note that the response varies every 16 days, the inventory once a year and the terrain not at all, which is the asymmetry the wind analysis depends on."

data_inventory(BC, SA) |>
  save_tbl("table-2-inventory") |>
  kable(booktabs = TRUE, align = "llllllr", row.names = FALSE)
```

The study area is `r fmt(AREA_HA)` ha of the Selkirk Mountains in southeastern British Columbia, `r fmt(N_CELL)` cells at 30 m, spanning `r fmt(ELEV_R[1])` to `r fmt(ELEV_R[2])` m, which is `r fmt(diff(ELEV_R))` m of relief.

The coordinate reference system is EPSG:`r EPSG`, `r CRSNAME`, which is the parent study's grid and was adopted so that results are directly comparable to it.

The perimeter is anchored on the 2015 Mt Midgeley fire, `r fmt(BURN_HA)` ha, which is the parent study's site. That study reports its site as spanning "elevations from 830 to 1744 m a.s.l."; the digital elevation model used here returns `r fmt(burn_e[1])` to `r fmt(burn_e[2])` m inside the same perimeter, which reproduces the statement. The present study extends beyond that boundary, because 480 ha does not contain the stand-density contrast the pheromone mechanism requires. The extension is constrained rather than arbitrary, in that the perimeter is the burn buffered 5 km and then cut to the elevation band the parent study's site occupies, so that the added ground is ecologically comparable.

```{r}
#| label: fig-study-area
#| echo: false
#| fig-cap: "Landscape, terrain and stand surfaces across the study area, all EPSG:3153 at 30 m over Esri World Shaded Relief. (a) elevation; (b) terrain ruggedness index; (c) windward-leeward index at the prevailing bearing; (d) flight-window direct radiation, the thermal limit on flight; (e) growing-season direct radiation, the shading pathway; (f) the MicroMet wind weighting factor at the prevailing bearing; (g) stand basal area; (h) quadratic mean diameter, whose 25 cm source-sink threshold falls near the midpoint of the scale; (i) live stems per hectare. The white outline is the study perimeter and the red outline the 2015 Mt Midgeley burn. Contours are at 200 m."
#| fig-height: 8.4
#| fig-width: 7.5

geo   <- rast(file.path(SA, "geomorphometry_context.tif"))
vri   <- rast(file.path(SA, "vri_context.tif"))
elevc <- rast(file.path(SA, "dem_context.tif"))
ctx   <- map_context(per, burn, SA)
## The unmasked weighting from 49-context-covariates.R, not the stored per-bin stack,
## which is masked to the perimeter and would leave this the one panel with a hole in it.
mmb   <- rast(file.path(SA, "micromet_weight_context.tif"))

m <- function(r, title, pal = "viridis")
  academic_map(r, title, ctx, pal, base_size = 7)

(m(elevc, "(a) Elevation (m)") |
 m(geo[["tri"]], "(b) Terrain ruggedness index", "magma") |
 m(geo[["wind_effect"]], "(c) Windward-leeward index", "cividis")) /
(m(geo[["solar_flight_direct"]], "(d) Flight-window radiation (kWh/m2)", "inferno") |
 m(geo[["solar_season_direct"]], "(e) Growing-season radiation (kWh/m2)", "inferno") |
 m(mmb, "(f) MicroMet wind weighting", "mako")) /
(m(vri[["BASAL_AREA"]], "(g) Stand basal area (m2/ha)", "mako") |
 m(vri[["QUAD_DIAM_125"]], "(h) Quadratic mean diameter (cm)", "viridis") |
 m(vri[["VRI_LIVE_STEMS_PER_HA"]], "(i) Live stems per hectare", "viridis"))
```

## Beetle disturbance

The response is annual moderate-to-high beetle disturbance, classified from Landsat imagery by a support vector machine. Training used the 28 ground plots of @murphy2026, in which the basal area of pine killed by the beetle was measured on 20 m plots centred within 60 by 60 m quadrants of four Landsat cells, so that no plot straddles a cell edge, following the plot-to-pixel protocols of @carlson2017 and @lentile2006. Plots were labelled low, moderate or high by tertiles of killed basal area, at `r sprintf('%.1f', CLS_BR[1])` and `r sprintf('%.1f', CLS_BR[2])` square metres per hectare, with undisturbed forest as the unaffected class. The classifier separates moderate or high from the rest on the changes in NDVI, NBR and tasselled-cap wetness against 2005, taken in the year of each plot's deepest NDMI decline. Accuracy is leave-one-out, because a held-out quarter of `r CLS_N` plots is too small to trust, and was `r sprintf('%.3f', CLS_ACC)`, kappa `r sprintf('%.3f', CLS_KAPPA)`. Annual maps were predicted over the perimeter with water masked, each year on its own. It covers `r spellnum(length(YRS))` outbreak years, `r paste(range(YRS), collapse = " to ")`, excluding 2012, the one year covered only by Landsat 7, which has flown with its scan-line corrector off since May 2003. Annual prevalence inside the perimeter runs `r sprintf("%.1f", 100*min(prev_yr$prev))` to `r sprintf("%.1f", 100*max(prev_yr$prev))` per cent, pooled `r sprintf("%.1f", 100*PREV)` per cent over `r fmt(N_ROWS)` cell-years. @fig-first-attack shows where the outbreak arrived first and how much of the perimeter it reached each year.

Years are never merged into one layer, because merging destroys the year-to-year variation this study measures and because the merged layer over the wider grid reached 73 per cent of the landscape, a figure not credible for a beetle outbreak. The provincial aerial overview survey is not used as a response and supplies no training label, and is retained only as a visual check.

```{r}
#| label: tbl-prevalence
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by year inside the study perimeter."

prev_yr |>
  transmute(Year = year, Cells = fmt(n),
            `Moderate-to-high (%)` = sprintf("%.1f", 100*prev)) |>
  save_tbl("table-3-prevalence") |>
  kable(booktabs = TRUE, align = "lrr", row.names = FALSE)
```

```{r}
#| label: fig-first-attack
#| echo: false
#| fig-cap: "Spread of moderate-to-high beetle disturbance across the study perimeter. (a) The first year in which each cell entered the moderate-to-high class, over shaded relief with the perimeter in white and the 2015 Mt Midgeley burn in red; cells never classed as attacked show the relief alone. This is a display product built from the eight annual maps, which are fitted separately and never merged for analysis. (b) The share of perimeter cells classed moderate-to-high in each year."
#| fig-height: 5.4
#| fig-width: 7.5

msk <- rast(file.path(SA, "perimeter_mask.tif"))
ann <- rast(lapply(YRS, function(y) {
  r <- rast(file.path(BC, "red-stage-darkwoods", sprintf("modhigh_%d.tif", y)))
  project(ifel(as.numeric(r) == 2, 1, 0), msk, method = "near")
}))
first <- app(ann, function(v) { i <- which(v == 1)[1]; if (is.na(i)) NA_real_ else YRS[i] })
first <- mask(first, msk); names(first) <- "first_year"
ctx <- map_context(per, burn, SA)
pa <- academic_map(first, "(a) First year classed moderate-to-high", ctx, "viridis", base_size = 8) +
  scale_fill_viridis_c(breaks = YRS, labels = YRS, na.value = "transparent", name = NULL)
pb <- ggplot(prev_yr, aes(factor(year), 100 * prev)) +
  geom_col(fill = "grey35", width = 0.75) +
  labs(x = NULL, y = "Moderate-to-high (% of perimeter cells)", title = "(b) Attack by year") +
  theme_bw(base_size = 8) + theme(panel.grid.minor = element_blank())
pa + pb + plot_layout(widths = c(2.2, 1))
```

## Stand structure

Stand structure comes from the provincial Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved by web feature service over the perimeter. Six attributes cover the mechanism and the stem-size threshold, namely total basal area, crown closure, live stems per hectare, quadratic mean diameter over stems 12.5 cm and up, stand age, and susceptible pine basal area, formed as total basal area times the pine share of cover.

Every model is fitted on annualized stand structure. Each cell-year takes the inventory snapshot the province published for that year, depleted for harvest and projected for growth to it, so the host terms vary in time as well as in space.

`r if (VRI_SUBN) sprintf("One year is substituted. The %d delivery lists basal area and live stems as columns and fills neither, 0 and 3 per cent of polygons against 70 to 97 per cent in every other year, so the preceding year is carried forward rather than averaged. The inventory is itself a projection, so carrying forward reproduces a stand structure the province published, where averaging would invent one it did not.", VRI_SUB[1])`

Cell-years at or above the `r QMD_THRESH` cm source-sink threshold make up `r sprintf("%.1f", PCT_ABOVE_QMD)` per cent of the sample.

```{r}
#| label: tbl-vri
#| echo: false
#| tbl-cap: "Stand structure across the study perimeter, from the Vegetation Resources Inventory. n is cell-years. SD is the standard deviation of the landscape; SE is the standard error of the mean and is small by construction at this n, so it should not be read as precision about any one cell. Skewness and kurtosis are the bias-corrected third and fourth standardised moments; kurtosis is excess, so 0 is Gaussian and positive is heavy-tailed."

describe_vars(d, VRI_V, VRI_LABELS) |>
  describe_table() |>
  save_tbl("table-4-stand-structure") |>
  kable(booktabs = TRUE, align = "lrrrrrrrrr", row.names = FALSE)
```

The inventory is a projected operational product, not a census, and its polygons here have reference years spanning several decades. A polygon interpreted from late-outbreak photography describes a stand the beetle had already attacked, so that basal area and pine cover are post-attack over part of the study window. No unattacked vintage exists, and the limitation is therefore reported rather than corrected, and returned to in the Discussion.

## Geomorphometry

Terrain is described by `r length(GEO_V)` geomorphometric surfaces computed with SAGA GIS over the full reprojected elevation model and clipped afterwards, so that a search radius near the boundary still falls on measured ground. The set decomposes the single ruggedness index of @murphy2026 into the quantities the biology names, and ruggedness itself is kept as a candidate so that the decomposition is tested against it rather than assumed.

Radiation is computed twice because two mechanisms require different quantities. Flight-window insolation is direct and diffuse radiation over 1 July to 15 August restricted to 12:00 to 17:00, the hours identified as the flight peak [@safranyik2006chap1], while growing-season insolation is the whole-day total from 1 May to 30 September, which is the shading quantity the first mechanism of @krawchuk2020 concerns. A single annual heat index cannot separate the two. Flight-window direct radiation spans a `r sprintf("%.0f", max(d$solar_flight_direct)/min(d$solar_flight_direct))`-fold range here against `r sprintf("%.1f", max(d$solar_season_total)/min(d$solar_season_total))`-fold for the season total.

Exposure enters as the windward-leeward index and effective air flow height at the prevailing bearing, the omnidirectional wind exposition index, the wind shelter index of @plattner2004, which is "the maximum gradient within a given radius in upwind direction", and topographic openness. Shape enters as terrain ruggedness, vector ruggedness, topographic position and its multi-scale form, convergence, slope, curvature and geomorphon class. Landform enters as topographic wetness, valley depth, height above the valley floor, normalised height and mid-slope position, because infested groups are reported in draws and gullies and deep snow insulates overwintering brood [@safranyik2006chap1]. Aspect enters as its northward and eastward components with the heat load index of @mccune2002.

## Terrain-resolved wind {#sec-micromet}

```{r}
#| label: micromet-load
#| echo: false
#| output: false

mmsum <- read.csv(file.path(BC, "covariates", "wind-micromet", "micromet_summary.csv"))
mmW   <- rast(file.path(BC, "covariates", "wind-micromet", "wind_weight_by_direction.tif"))
MM_WW <- range(as.vector(minmax(mmW)))
```

Station wind interpolated from four to seven valley stations is nearly flat within a year, so wind is additionally computed as a field varying in space, using the MicroMet model of @liston2006. The wind component of that model is seven closed-form equations and was implemented directly from the source paper, whose equation numbers are given below.

Terrain slope $\beta$ and slope azimuth $\xi$ come from the elevation model, equations 12 and 13 of @liston2006. Curvature $\Omega_c$ is a cell's elevation minus the mean of the two opposite cells one curvature length scale away, taken on four direction lines and averaged (their equation 14); the length scale is estimated as the first lag at which elevation autocorrelation falls below 0.5, which is 600 m here. Writing $\theta$ for the wind bearing, the slope in the wind direction is

$$\Omega_s = \beta \cos(\theta - \xi)$$ {#eq-slope}

with $\Omega_s$ and $\Omega_c$ each scaled to $[-0.5,\,0.5]$. The terrain weighting factor applied to the observed speed $W$ is

$$W_w = 1 + \gamma_s \Omega_s + \gamma_c \Omega_c, \qquad \gamma_s = \gamma_c = 0.5$$ {#eq-weight}

giving the terrain-modified speed

$$W_t = W_w W$$ {#eq-speed}

and a diversion of the wind direction of

$$\theta_d = -\tfrac{1}{2}\, \Omega_s \sin\!\left[2(\xi - \theta)\right].$$ {#eq-divert}

By @eq-weight, $W_w$ depends on direction and not on speed, so it is computed once for each of 16 direction bins and each hourly observation is multiplied by the surface for its own bin, so that nothing is averaged before the terrain acts on it. Over all bins $W_w$ runs `r sprintf("%.2f", MM_WW[1])` to `r sprintf("%.2f", MM_WW[2])`. Speed and direction are combined as vector components, because averaging degrees across the 360 to 0 discontinuity is not meaningful. The resulting field varies `r sprintf("%.1f", min(mmsum$spatial_range))` to `r sprintf("%.1f", max(mmsum$spatial_range))` km/h across the grid within a year and is not a repackaged terrain index, correlating `r cc("mm_flight_mean", "wind_effect")` with the windward-leeward index and `r cc("mm_flight_mean", "solar_flight_direct")` with flight-window radiation. Its strongest association is with elevation, `r cc("mm_flight_mean", "elevation")`.

## Flight-window wind

The flight window is 1 July to 15 August and the hours 12:00 to 17:00, and neither bound was chosen from these data. The dates are the flight period @safranyik2006chap1 give for this region, and the hours follow their "peak flight is in the early to mid afternoon" together with the 11:00 to 14:00 emergence peak of @gray1972, taken here to the later side because emergence precedes the flight it initiates. Because the window is fixed a priori it can be checked against the landscape's own climate, and @fig-flight-window does that. Across `r fmt(CLIM_N)` hourly station records from May to September of the nine study years, `r sprintf("%.1f", 100*CLIM_IN)` per cent of afternoon hours inside the window fall within the 19 to 41 degrees C flight range, against `r sprintf("%.1f", 100*CLIM_OUT)` per cent outside it, and the afternoon mean is `r sprintf("%.1f", CLIM_T_IN)` against `r sprintf("%.1f", CLIM_T_OUT)` degrees C. The window therefore roughly doubles the share of hours in which flight is thermally possible. The same figure shows the same daily pattern in wind. Mean speed peaks in the same hours as temperature, at `r sprintf("%.1f", CLIM_W_MAX)` km/h in mid-afternoon against `r sprintf("%.1f", CLIM_W_MIN)` km/h at dawn, so the hours the beetle can fly are also the windiest of the day, a coincidence of timing the mechanism depends on.

Station wind is Environment and Climate Change Canada hourly records, never reduced before the terrain acts on them.

Hourly readings give detail rather than independent observations, because summarised over a whole season every hour of wind collapses into one number per metric, a coefficient then rests on as many values as there are seasons, and its standard error looks small without being believable. Stand density varies from cell to cell while a season's wind does not, so the response is measured over each sixteen-day epoch, wind varies within a season as well as between seasons, and the interaction the mechanism predicts is identified.

```{r}
#| label: fig-flight-window
#| echo: false
#| fig-cap: "The flight window against the landscape's own climate, from hourly Environment and Climate Change Canada station records for May to September of the nine study years. The window, 1 July to 15 August, is the shaded band; it is taken from the bionomics and is not fitted to these data. (a) Mean afternoon temperature by day of year, with the 19 to 41 degrees C flight range and the 22 to 32 degrees C peak band marked. (b) The share of afternoon hours falling inside the flight range, day by day. (c) The diurnal curve, mean temperature and the share of all hours inside the flight range, by hour, with the 12:00 to 17:00 restriction shaded. (d) Mean wind speed by hour, on the same axis, showing that the hours the beetle can fly are also the windiest of the day."
#| fig-height: 6.2
#| fig-width: 7.5

wd <- range(unlist(lapply(unique(clim_d$year), win_doy)))
band <- annotate("rect", xmin = wd[1], xmax = wd[2], ymin = -Inf, ymax = Inf,
                 fill = "#fdae61", alpha = 0.18)
th <- theme_bw(base_size = 8) + theme(panel.grid.minor = element_blank())

pa <- ggplot(clim_d, aes(doy, pm_temp)) + band +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = T_GATE[1], ymax = T_GATE[2],
           fill = "#4575b4", alpha = 0.10) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = T_PEAK[1], ymax = T_PEAK[2],
           fill = "#4575b4", alpha = 0.14) +
  geom_point(aes(colour = factor(year)), size = 0.5, alpha = 0.55) +
  geom_smooth(se = FALSE, colour = "black", linewidth = 0.5, method = "loess", span = 0.3) +
  scale_colour_viridis_d(name = "Year", option = "turbo") +
  labs(x = "Day of year", y = "Afternoon temperature (C)",
       title = "(a) Afternoon temperature and the flight range") + th

pb <- ggplot(clim_d, aes(doy, 100 * pm_in_gate)) + band +
  geom_point(aes(colour = factor(year)), size = 0.5, alpha = 0.55, show.legend = FALSE) +
  geom_smooth(se = FALSE, colour = "black", linewidth = 0.5, method = "loess", span = 0.3) +
  scale_colour_viridis_d(option = "turbo") +
  labs(x = "Day of year", y = "Afternoon hours in range (%)",
       title = "(b) Share of afternoon hours flight is possible") + th

pc <- ggplot(clim_x, aes(hour)) +
  annotate("rect", xmin = FW_HR[1], xmax = FW_HR[2], ymin = -Inf, ymax = Inf,
           fill = "#fdae61", alpha = 0.22) +
  geom_hline(yintercept = T_GATE[1], linetype = 2, linewidth = 0.3, colour = "grey40") +
  geom_line(aes(y = temp), linewidth = 0.7, colour = "#d73027") +
  geom_line(aes(y = 100 * in_gate / 3), linewidth = 0.6, colour = "#4575b4") +
  scale_y_continuous("Temperature (C)",
    sec.axis = sec_axis(~ . * 3, name = "Hours in range (%)")) +
  labs(x = "Hour of day", title = "(c) Diurnal cycle of temperature and the flight range") + th

pd <- ggplot(clim_x, aes(hour, wind)) +
  annotate("rect", xmin = FW_HR[1], xmax = FW_HR[2], ymin = -Inf, ymax = Inf,
           fill = "#fdae61", alpha = 0.22) +
  geom_line(linewidth = 0.7, colour = "#1a9850") +
  labs(x = "Hour of day", y = "Wind speed (km/h)",
       title = "(d) Wind peaks in the same hours") + th

(pa | pb) / (pc | pd)
```

## Variable selection {#sec-selection}

The candidate set is `r N_CAND` variables in six groups, one group for each pathway the Introduction names. The groups are stand density, host quality, terrain exposure to wind, which holds the shelter and openness indices the deposition question turns on, terrain shape, landscape context, and flight-window wind. Selection proceeds in four stages and is required to retain at least one variable from each pathway, so that a collinearity filter cannot silently delete a hypothesis the Introduction established.

Fitting uses a class-balanced sample of `r fmt(N_BAL)` cell-years, `r fmt(4000)` of each class per year. Landscape prevalence is about `r sprintf("%.0f", 100*PREV)` per cent, and an unbalanced fit at that prevalence reports the intercept rather than the covariates.

## Models

Every model is a logistic regression of moderate-to-high disturbance $y_{it}$ in cell $i$ and year $t$ on standardised covariates,

$$\operatorname{logit}\Pr(y_{it}=1) = \alpha + \mathbf{x}_{it}^{\top}\boldsymbol{\beta}
+ \gamma_{g(i)} + \sum_{k} \delta_k\, z_{k,it}$$ {#eq-model}

where $\gamma_{g(i)}$ is the effect of the geomorphon landform class $g$ of cell $i$ and the $z_{k,it}$ are the interaction terms. Coefficients are therefore log-odds per standard deviation. Four annual models are fitted, each adding one claim to the last, so that every term's contribution is visible as a change in fit rather than asserted, and the sixteen-day models described under flight-window wind carry the interaction in time.

M0 fits the shading and host-size mechanisms together with landform context, that is host size, topographic shading, and the terrain that governs where cold air and snow collect. M1 adds stand density, which the first question concerns. M2 adds terrain shape, terrain exposure to wind and flight-window radiation, which the second and third questions concern, and its shelter main effect is the first half of the fifth. M3 adds the three interactions the fourth and fifth questions require, stand density by terrain exposure, by flight-window radiation and by flight-window wind. The fifth question is answered by reading the shelter main effect in M2 and M3 against the density by shelter interaction in M3, because deposition predicts the first without the second and plume disruption predicts the second whether or not the first is present. The sixteen-day models answer the fourth question in the domain where wind varies within a season.

# Results {#sec-results}

```{r}
#| label: tbl-aic
#| echo: false
#| tbl-cap: "Model comparison, panel A of the model table. Each model adds one pathway to the previous. AIC ranks on likelihood and a parameter penalty; the remaining columns are predictive error on the fitted probabilities. RMSE is the root mean squared error and is the square root of the Brier score; RMSE (%) expresses it against the prevalence of the response. Brier skill is the improvement over predicting the prevalence for every cell, where 0 is no better than the base rate. MAPE and Theil's U are not reported, because both divide by the observed value, which is zero for the majority class of a binary response, so both are undefined without discarding that class."

fit_metrics_table(
  fit_metrics(M0, "M0 host size + shading + landform"),
  fit_metrics(M1, "M1 + stand density"),
  fit_metrics(M2, "M2 + terrain and flight radiation"),
  fit_metrics(M3, "M3 + interactions")) |>
  save_tbl("table-5-model-comparison") |>
  kable(booktabs = TRUE, align = "lrrrrrrrr", row.names = FALSE)
```

```{r}
#| label: tbl-m3
#| echo: false
#| tbl-cap: "Full model M3, panel B of the model table, continuous terms ordered by absolute effect. Coefficients are log-odds per standard deviation on a class-balanced sample, so the intercept is not landscape prevalence. SE is the standard error of the coefficient. The geomorphon landform classes are also in this model and are not reported here."

co3 |>
  filter(term != "(Intercept)", !grepl("^geomorphon", term)) |>
  arrange(desc(abs(beta))) |>
  transmute(Term = pretty_terms(term), Beta = sprintf("%+.3f", beta),
            SE = sprintf("%.3f", se), z = sprintf("%.2f", z), p = sapply(p, pv)) |>
  save_tbl("table-6-full-model") |>
  kable(booktabs = TRUE, align = "lrrrr", row.names = FALSE)
```

Each pathway improved the fit. AIC fell `r fmt(aic$AIC[1] - aic$AIC[2])` when stand density entered, a further `r fmt(aic$AIC[2] - aic$AIC[3])` with terrain and flight-window radiation, and a further `r fmt(aic$AIC[3] - aic$AIC[4])` with the interactions (@tbl-aic).

Stand basal area was the strongest density term, entering at `r b3("BASAL_AREA")` log-odds per standard deviation, an odds ratio of `r or3("BASAL_AREA")`, so that a stand one standard deviation above the mean in basal area had `r or3pc("BASAL_AREA")` per cent higher odds of moderate-to-high disturbance (`r p3T("BASAL_AREA")`). Live stems per hectare took the same sign at `r b3("VRI_LIVE_STEMS_PER_HA")` (`r p3T("VRI_LIVE_STEMS_PER_HA")`) and crown closure the opposite at `r b3("CROWN_CLOSURE")` (`r p3T("CROWN_CLOSURE")`).

Basal area and live stems both rose with attack, the direction a plume-holding canopy implies, while crown closure fell. Crown closure is also the one measure the inventory caps, recorded to a ceiling of 60 per cent, so that its coefficient describes a truncated variable and should not be read as the others are.

Radiation entered the model once, as direct radiation during the flight window at `r b3("solar_flight_direct")`, which represents the thermal limit on flying. Growing-season radiation did not survive selection, so northness stands in for the shading pathway, the two correlating `r cc("solar_season_direct", "northness")`. Northness was `r b3("northness")`, so that shaded, north-facing ground had more attack rather than less, against the prediction.

Terrain predicted attack independently of both stand structure and radiation, and it did so through shelter and openness rather than through ruggedness, because neither terrain ruggedness nor the windward-leeward index survived selection into the full model. The wind shelter index as computed here rises on slopes facing the prevailing bearing, correlating `r sprintf("%+.2f", cor(d$wind_shelter, cos((d$aspect - WDIR) * pi / 180), method = "spearman", use = "complete.obs"))` with that orientation, and it entered at `r b3("wind_shelter")` (`r p3T("wind_shelter")`), so leeward ground had more attack and windward ground less. Sky view factor, which is high on gentle, open, upper ground, was `r b3("sky_view")` (`r p3T("sky_view")`), height above the valley floor `r b3("height_valley_floor")` and valley depth `r b3("valley_depth")`, so attack sat on open, sheltered ground standing above shallow valleys rather than in deep ones. Elevation was `r b3("elevation")`, and the vector ruggedness measure `r b3("vrm")` was not distinguishable from zero (`r p3T("vrm")`).

Fitted without flight-window radiation, stand density interacted with terrain exposure at `r bnr(I_GEO)`, which would ordinarily be read as plume disruption acting through terrain. With radiation in the model that interaction shrank to `r b3(I_GEO)` and was no longer distinguishable from zero (`r p3T(I_GEO)`), while density interacted with radiation at `r b3(paste0(DENS[1], ":", FSUN[1]))`. On a range whose prevailing bearing is `r sprintf("%.0f", WDIR)` degrees the windward slopes face west, and these are also the slopes taking afternoon sun, so that `r sprintf("%.0f", RAD_SHARE)` per cent of the apparent interaction was radiation and what remained was too small to support a claim on its own. The shelter main effect did not move between the two fits, so the fifth question resolved one way. Shelter acted alone, as deposition predicts, and not through stand density, as plume disruption would require.

```{r}
#| label: fig-interaction
#| echo: false
#| fig-cap: "The refugia mechanism as fitted. (a) Predicted probability of moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem density, all other terms held at their means; the lines cross, so wind raises attack in thin stands and lowers it in dense ones, which is the interaction. (b) The same for standing volume, where the interaction is a fifth the size and attenuates the dense-stand slope without reversing it, so the lines converge rather than cross over the observed range. (c) Coefficients of the 16-day model, with and without the within-season spread term; the shift of a term between the two fits is what that term absorbs. (d) Epoch prevalence against epoch mean wind, one point per epoch, years distinguished by both shape and shade so the panel remains legible in black and white."
#| fig-height: 6

nd_base <- be[1, , drop = FALSE]
for (v in EV) nd_base[[v]] <- 0
pred_panel <- function(dv, lab) {
  qs <- quantile(be[[dv]], c(0.1, 0.9))
  gr <- expand.grid(w = seq(min(be$ep_wind_mean), max(be$ep_wind_mean), length.out = 60),
                    d = qs)
  nd <- nd_base[rep(1, nrow(gr)), ]
  nd$ep_wind_mean <- gr$w; nd[[dv]] <- gr$d
  nd$geomorphon <- factor(levels(be$geomorphon)[1], levels = levels(be$geomorphon))
  gr$p <- predict(E1, newdata = nd, type = "response")
  gr$Density <- factor(gr$d, labels = c("low (10th pct)", "high (90th pct)"))
  ggplot(gr, aes(w, p, colour = Density)) + geom_line(linewidth = 0.8) +
    scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
    labs(x = "Epoch wind (SD units)", y = "P(moderate-to-high)", title = lab) +
    theme_minimal(base_size = 9)
}
pa <- pred_panel("VRI_LIVE_STEMS_PER_HA", "(a) Stem density")
pb <- pred_panel("LIVE_STAND_VOLUME_125", "(b) Standing volume")

cf <- bind_rows(
  ce1 |> mutate(model = "E1 all epochs"),
  ce2 |> mutate(model = "E2 with lag")) |>
  filter(!grepl("^geomorphon|Intercept", term)) |>
  group_by(term) |> filter(any(abs(beta) > 0.05)) |> ungroup()
## pretty_terms() here for the same reason the tables use it: panel (c) shipped raw column
## names, ep_lag_nbr90 and QUAD_DIAM_125, beside a document that labels every other term.
cf$label <- pretty_terms(cf$term)
pc <- ggplot(cf, aes(reorder(label, beta), beta, colour = model)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_pointrange(aes(ymin = beta - 1.96*se, ymax = beta + 1.96*se),
                  position = position_dodge(width = 0.5), size = 0.25) +
  coord_flip() + scale_colour_manual(values = c("#1b9e77", "#7570b3")) +
  labs(x = NULL, y = "Log-odds per SD", title = "(c) 16-day model coefficients") +
  theme_minimal(base_size = 8) + theme(legend.position = "bottom")

pd <- epsum |>
  inner_join(epwind, by = c("year","epoch")) |>
  ggplot(aes(mean_kmh, prevalence)) +
  geom_point(aes(colour = factor(year), shape = factor(year)), size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.5) +
  scale_colour_viridis_d(option = "viridis", end = 0.9) +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 4, 18)) +
  ## Both scales must carry the same name or ggplot draws two legends and titles the
  ## shape one "factor(year)", which is the raw expression rather than a label.
  labs(x = "Epoch mean wind (km/h)", y = "Epoch prevalence",
       colour = "Year", shape = "Year", title = "(d) Epochs") +
  theme_minimal(base_size = 8)

(pa | pb) / (pc | pd)
```

## Density and wind

Stand density interacted negatively with terrain-resolved wind, which is the form the pheromone mechanism predicts. Attack fell where a thin stand and strong wind coincided, the interaction `r be1(EINT[1])` for stem density (`r pe1T(EINT[1])`) and `r be1(EINT[2])` for standing volume (`r pe1T(EINT[2])`). Expressed as odds, each standard deviation of the epoch wind regime multiplied the contribution of stem density by `r ore1(EINT[1])`, a reduction of `r ore1pcabs(EINT[1])` per cent. Both interactions held after the previous epoch of the same season was entered, so neither is the outbreak's own spread appearing as a wind coefficient.

Wind alone gave no protection, its main effect being `r be1("ep_wind_mean")`, which is marginally more attack rather than less, and the density terms were positive, so that a thin stand in still air showed no reduction either. The mechanism was present only where the two coincided.

## Previous-year pressure {#sec-autologistic}

Attack recurred where it had already occurred, by a margin large enough to change how every other coefficient should be read. A cell attacked in one year was between `r sprintf("%.0f", min(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))` and `r sprintf("%.0f", max(read.csv(file.path(BC, "model-data", "lag1_persistence.csv"))$odds_ratio))` times more likely to be attacked in the next. Entering persistence and 90 m neighbourhood spread separately, following the autologistic design used for this province, gave `r b4("lag_self")` and `r b4("lag_nbr90")`, raising discrimination from `r sprintf("%.3f", AUC3l)` to `r sprintf("%.3f", AUC4)`.

The informative result was which environmental terms held after that term entered. Sky view moved from `r sh("sky_view", "beta_env")` to `r sh("sky_view", "beta_auto")` and the wind shelter index from `r sh("wind_shelter", "beta_env")` to `r sh("wind_shelter", "beta_auto")`, both describing conditions a cell has whether or not the beetle was ever present. The landform terms were less stable, valley depth falling from `r sh("valley_depth", "beta_env")` to `r sh("valley_depth", "beta_auto")` while mid-slope position moved from `r sh("midslope_position", "beta_env")` to `r sh("midslope_position", "beta_auto")`. Part of what the landform terms measured was therefore where the outbreak had already been, which is the same failure the terrain-wind index showed against radiation.

## Host as diameter {#sec-diameter}

```{r}
#| label: tbl-qmd
#| echo: false
#| tbl-cap: "Moderate-to-high beetle disturbance by quadratic mean diameter class, on the balanced sample. The 25 cm boundary is the source-sink threshold of the species' bionomics. Intervals are Wilson score intervals on the class proportion, which is why they are asymmetric in the smallest class. Note that 30 m cells in a spreading outbreak are not independent, so the tests reported beneath this table are anti-conservative."

qmd_bin |>
  diameter_table() |>
  save_tbl("table-7-diameter-class") |>
  kable(booktabs = TRUE, align = "lrrrr", row.names = FALSE)
```

Coding host as diameter changed the picture that cover alone gives. Attack peaked in the 25 to 30 cm class at `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent and fell above it, to `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "30-40"])` per cent at 30 to 40 cm and `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == ">40"])` per cent above 40 (@tbl-qmd). The step across the 25 cm source-sink boundary is from `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "20-25"])` to `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent.

In the full model diameter was `r b3("QUAD_DIAM_125")` per standard deviation (z = `r z3("QUAD_DIAM_125")`), negative because the relationship is not monotone, since a linear term fitted through a humped response returns the slope of its falling limb, which is the larger part of the range. The class table rather than the coefficient is therefore the result here, and the interpretation stands, because a stand's cover says nothing about whether its stems can produce brood.

The pattern was not an artefact of the class boundaries, in that across all six classes attack depended on diameter class, $\chi^2$ = `r sprintf("%.1f", QT$X2)` on `r QT$df` degrees of freedom, p `r pt(QT$chi$p.value)`, with Cramer's V = `r sprintf("%.3f", QT$V)`. The step across the 25 cm source-sink boundary specifically, from the 20 to 25 class to the 25 to 30 class, is `r sprintf("%+.1f", QT$step_diff)` percentage points, 95 per cent confidence interval `r sprintf("%.1f to %.1f", -QT$step_ci[2], -QT$step_ci[1])`, p `r pt(QT$step$p.value)`. With `r fmt(sum(qmd_bin$n))` cells a chi-square is significant on trivial differences, which is why the effect size is quoted beside it, and 30 m cells in a spreading outbreak are not independent, so both p-values are anti-conservative.

The smallest class gives little evidence, with `r fmt(qmd_bin$n[qmd_bin$bin == "<15"])` cells against `r fmt(qmd_bin$n[qmd_bin$bin == "25-30"])` in the modal class, and a stand below 15 cm quadratic mean diameter in an inventory whose vintage postdates the outbreak is more likely a stand from which the beetle had already removed the large stems than a stand attacked at that size.

## The wind window {#sec-window}

The wind variable is the epoch wind regime, the mean over every hourly station observation in the sixteen days an epoch covers, each hour modified over the terrain by its own bearing. It is not restricted to the flight window, and that choice follows from how the insect disperses. @jackson2008 tracked mountain pine beetle on weather radar and confirmed the returns by aerial capture, finding beetles "at altitudes up to more than 800 m above the forest canopy" and estimating that those in flight above the canopy "may move 30 to 110 km$\cdot$day$^{-1}$", at a mean density of 4,950 and a maximum of 18,600 beetles per hectare. @ainslie2010 subsequently reproduced that above-canopy dispersion with an atmospheric model. An insect carried in the boundary layer at those heights and over those distances is exposed to the atmospheric regime of the period rather than to the wind measured within a stand between noon and five, and the variable is named accordingly.

@tbl-window refits the 16-day model under four definitions of the window. These are every hour, the 12:00 to 17:00 hours the flight peak occupies, the 11:00 to 18:00 hours that fall inside the thermal limits on this landscape more than half the time, and only those hours whose observed temperature fell inside the 19 to 41 degrees C range itself. The stem-density interaction was negative and distinguishable from zero under all four definitions, so the mechanism does not depend on the choice, although it was largest over the full regime and the standing-volume interaction was the term sensitive to that choice.

```{r}
#| label: tbl-window
#| echo: false
#| tbl-cap: "The 16-day model refitted under four definitions of the wind window. The epoch wind regime, every hour, is the definition used throughout this paper. Coefficients are log-odds per standard deviation. The thermal-limits row is fitted on fewer cell-epochs because two epochs had too few qualifying hours, so its AIC is not comparable with the other three."

FWS |>
  transmute(
    Definition = c("Epoch wind regime (all hours)", "12:00-17:00", "11:00-18:00",
                   "Inside the 19-41 C range")[match(definition,
                   c("all24","h12_17","h11_18","gate"))],
    `Cell-epochs` = formatC(n, format = "d", big.mark = ","),
    `Wind (km/h)` = sprintf("%.2f", wind_mean),
    `Stems x wind` = sprintf("%+.4f", stems_x_wind),
    `p` = sapply(stems_p, function(x) if (x < 0.001) "< 0.001" else sprintf("%.3f", x)),
    `Volume x wind` = sprintf("%+.4f", volume_x_wind),
    `p ` = sapply(volume_p, function(x) if (x < 0.001) "< 0.001" else sprintf("%.3f", x)),
    AUC = sprintf("%.3f", AUC)) |>
  save_tbl("table-8-flight-window") |>
  kable(booktabs = TRUE, align = "lrrrrrrr", row.names = FALSE)
```

# Discussion

## Two scales

The five questions resolved into one pattern. Terrain acted on attack as a main effect of shelter and openness that did not depend on stand density, and wind acted on attack as an interaction with stand density that appeared only where wind varied in time. Deposition predicts the first and plume disruption the second, and neither mechanism produced the other's signature. The one place the two could have been confused, the density by shelter interaction, fell by `r sprintf("%.0f", RAD_SHARE)` per cent when flight-window radiation entered and could not then be distinguished from zero, while the shelter main effect did not move, so a terrain index alone would have supported the plume mechanism on evidence that was partly sunlight.

This is unlikely to be peculiar to the present dataset. A shelter index and an afternoon radiation surface are both functions of slope and aspect, and on a range whose prevailing flight-window bearing is `r sprintf("%.0f", WDIR)` degrees the slopes that meet the wind are the west-facing ones, which are also the slopes taking afternoon sun during the flight peak. Any study entering a terrain wind index without a radiation term on the same clock will attribute radiation to wind, and any study reading a terrain main effect as plume disruption has assigned a deposition signature to the wrong mechanism.

## What was supported

Of the three mechanisms @krawchuk2020 propose, two were supported and one failed. Low host density admitting the wind that disperses the aggregation pheromone was supported in the conditional form the mechanism specifies rather than as a main effect. The claim was never that dense stands are attacked more, but that a dense canopy holds a plume together and that the advantage this confers on the beetle should diminish as wind speed rises, and both halves of that prediction are present. A scarcity of large-diameter hosts was supported as a threshold rather than as a gradient, attack peaking at `r sprintf("%.1f", 100*qmd_bin$attack[qmd_bin$bin == "25-30"])` per cent in the 25 to 30 cm class, at the source-sink boundary of @carroll2004bionomics, and falling away on both sides.

Topographic shading was not supported, and it fails on a surrogate rather than on the quantity the mechanism names. Growing-season radiation did not survive variable selection, so northness stands in for the pathway, the two correlating `r cc("solar_season_direct", "northness")`. Northness was `r b3("northness")`, so that shaded ground had more attack where the mechanism requires less. Northness tests aspect, and aspect measures more than shade, which makes this the least secure of the three verdicts, and the water-stress pathway @krawchuk2020 describe is not measured here at all.

The fifth question, which @krawchuk2020 did not pose, was answered as well, in that terrain shelter acted as a main effect and not through density, the deposition signature the companion study's ruggedness coefficient was recording without being able to name.

## The wind claim

Stand density interacts with terrain-resolved wind at `r be1(EINT[1])` for stem density and `r be1(EINT[2])` for standing volume, both negative, at p `r pe1t(EINT[1])` and p `r pe1t(EINT[2])` respectively. A dense stand therefore confers less advantage on the beetle where and when wind speed is higher, which is what @krawchuk2020 predict. The result held after the previous epoch of the same season was entered, so it is not the outbreak's own spread appearing as a wind coefficient.

Two limits remain attached to the claim. The interaction is small, roughly a tenth of the stand basal area main effect, and the wind field, although it varies in space, remains terrain-modified station data rather than measurement on the ridge. What is established is therefore that the modelled wind field behaves as the mechanism requires, not that the air itself did.

## Landing zones

Read with the dispersal literature, the terrain results describe where beetles come down rather than where they choose to attack. @hynum1980 showed that landing is not host-directed, @jackson2008 that part of the population travels far above the canopy, and @chen2017 that the descent phase of that transport is the least understood. A beetle descending through slowing air behaves as any wind-borne particle does and settles where the flow decelerates, in the lee of ridges and on sheltered slopes, which is what the shelter index measures and why the same index predicts where snow accumulates [@plattner2004]. Three results fit that reading and none contradicts it. Leeward ground had more attack as a main effect, `r b3("wind_shelter")`, and the effect did not detectably depend on stand density. Open, gently sloping ground above shallow valleys had more attack, sky view `r b3("sky_view")`, height above the valley floor `r b3("height_valley_floor")` and valley depth `r b3("valley_depth")`. The density by wind interaction that plume disruption predicts appeared only in the sixteen-day models, where wind varies in time, and not through terrain, where it varies in space. Terrain shelter therefore reads as a landing zone and the sixteen-day wind as plume disruption, two mechanisms at two scales rather than one.

The one prior study to read terrain this way found the opposite face of the same frame, in that @giroday2011 associated the highest infestation intensities in the Peace River region with southwest-facing open slopes and mid-slope ridges, the surfaces that meet a westerly wind, after first establishment in canyons and valleys, whereas here the association was leeward and deep valleys held the least attack. Impaction on faces that meet the wind and settlement where it slows are both deposition, and which one a landscape shows may depend on how much of its relief a descending beetle clears.

This is an interpretation rather than an observation, because no beetle was tracked to the ground here. Sky view factor also sets diffuse radiation, so part of its coefficient may be sunlight, and deep valleys are also the coldest cells on the landscape, so the same pattern would follow from slower development as readily as from deposition [@sambaraju2021]. The component reading is nonetheless what the parent study could not give, because a ruggedness index records that relief matters without saying how, and decomposed here into shelter, openness and valley position the relief signal resolves into a landing surface.

## What stays confounded

Elevation is among the largest terms in the model at `r b3("elevation")`, but it is not a single quantity, combining temperature, snowpack, growing-season length and the distribution of lodgepole pine in a way this design cannot separate. Reporting the elevation coefficient as a result would therefore be reporting a composite.

The same caution applies in reverse to every terrain term in the model, since radiation absorbed `r sprintf("%.0f", RAD_SHARE)` per cent of one terrain interaction and there is no guarantee that some further unmeasured environmental surface would not absorb the remainder, or another term entirely.

Three limits sit outside the model and bound every result in it. The inventory postdates part of the outbreak, because polygons interpreted after the beetle passed describe the stand it left, and total basal area discriminates only weakly on its own, a univariate AUC of `r sprintf("%.3f", uni$auc[uni$term == "BASAL_AREA"])`. The response is classified rather than observed, from `r CLS_N` plots inside one burn, so the reported accuracy measures how well the classifier reproduces those plots and not agreement with ground mortality elsewhere. Every conclusion rests on one mountain range across `r spellnum(EP_YEARS)` years.

## Conclusions

Terrain and wind acted on mountain pine beetle attack by two mechanisms at two scales. Across `r EP_EPOCHS` sixteen-day epochs, stand density interacted negatively with terrain-resolved wind, `r be1(EINT[1])` for stem density and `r be1(EINT[2])` for standing volume, and both held after a within-season spread term was entered, so a dense canopy conferred less advantage on the beetle where and when wind speed was higher, which is what plume disruption requires. Across the landscape, terrain shelter acted as a main effect and not through stand density, leeward ground having more attack at `r b3("wind_shelter")` whatever the stand, and open gentle ground above shallow valleys more still, which is what deposition of wind-borne beetles requires. Sheltered open ground therefore behaves as a landing zone, and the ruggedness signal the companion study found on this landscape resolves into that surface.

Of the three mechanisms @krawchuk2020 name, two were supported. Host density was supported through stand basal area at `r b3(DENS[1])`, with live stems agreeing in sign and crown closure alone running counter, and large-diameter host as a threshold at 25 to 30 cm rather than as a gradient, at the source-sink boundary of @carroll2004bionomics. Topographic shading was not supported, failing on its surrogate northness at `r b3("northness")`.

Two measurement results extend beyond this landscape. A terrain wind index is in part a measurement of incident radiation, because entering flight-window radiation moved the density by shelter interaction from `r bnr(I_GEO)` to `r b3(I_GEO)`, so a study entering such an index without a radiation term on the same clock will attribute radiation to wind. A landform variable in a spreading outbreak also records in part where the outbreak has already been.

For pest management the result means that refugia cannot be mapped from terrain alone, and that the map has two layers. A stand's exposure to the prevailing wind sets how many beetles arrive, and its density during the windy weeks of the flight period sets how many of those succeed, so thinning can be expected to lower attack on windward ground during windy flight periods and to do little on the sheltered lee slopes where beetles come down, and only by the modest margin the interaction measures.

# Author contributions {.unnumbered}

SM conceived and designed the study, assembled the data, performed the analysis and wrote the manuscript.

# Acknowledgements {.unnumbered}

This work used no external funding. Beetle disturbance was classified from Landsat Collection 2 Level-2 surface reflectance distributed by the United States Geological Survey; stand structure from the British Columbia Data Catalogue; terrain from Natural Resources Canada; and wind from Environment and Climate Change Canada. The author thanks those agencies for maintaining the open archives the study rests on.

# References {.unnumbered}

::: {#refs}
:::

# Statements and declarations {.unnumbered}

## Funding

The author declares that no funds, grants, or other support were received during the preparation of this manuscript.

## Competing interests

The author has no relevant financial or non-financial interests to disclose.

## Ethics approval

This study used only publicly archived remote sensing, inventory and meteorological data. No animals, human participants or protected material were involved, so no ethics approval was required.

## Data availability

Every dataset used here is public and none was collected by the author. Beetle disturbance is classified from Landsat Collection 2 Level-2 surface reflectance. Stand structure is the provincial Vegetation Resources Inventory layer `VEG_COMP_LYR_R1_POLY`, retrieved from the British Columbia Data Catalogue web feature service by `02.inputs/beetle/34-fetch-vri.py`. Terrain derives from the Natural Resources Canada High Resolution Digital Elevation Model [@nrcan2017], with geomorphometry computed by SAGA GIS in `37-geomorphometry.R`. Wind is Environment and Climate Change Canada hourly station data, retrieved by `30-wind-hourly-metrics.R` and `36-wind-direction.R`.

All derived data and the complete analysis code that reproduce every number, table and figure in this article are available at <https://github.com/seamusrobertmurphy/beetle-topography-and-wind-study>.

```{=html}
<!-- Journal of Pest Science requires a data availability statement and "strongly
     encourages" a repository deposit; it does not mandate one for data of this kind.
     The statement therefore names the public repository alone. Seamus ruled on
     2026-09-01 that no Dryad deposit is needed unless the journal states otherwise. -->
```

# Figure legends {.unnumbered}

**Figure 1.** Landscape, terrain and stand surfaces across the study area, all EPSG:3153 at 30 m over Esri World Shaded Relief, which is reproduced under Esri's attribution requirement. (a) elevation; (b) terrain ruggedness index; (c) windward-leeward index at the prevailing bearing; (d) flight-window direct radiation, the thermal limit on flight; (e) growing-season direct radiation, the shading pathway; (f) the MicroMet wind weighting factor at the prevailing bearing; (g) stand basal area; (h) quadratic mean diameter, whose 25 cm source-sink threshold falls near the midpoint of the scale; (i) live stems per hectare. The white outline is the study perimeter and the red outline the 2015 Mt Midgeley burn. Kootenay Lake and the Kootenay River are from the British Columbia Freshwater Atlas. Contours are at 200 m. The surfaces are drawn unclipped, so they extend beyond the perimeter within which the models are fitted.

**Figure 2.** Spread of moderate-to-high beetle disturbance across the study perimeter. (a) The first year in which each cell entered the moderate-to-high class, over shaded relief with the perimeter in white and the 2015 Mt Midgeley burn in red; cells never classed as attacked show the relief alone. This is a display product built from the eight annual maps, which are fitted separately and never merged for analysis. (b) The share of perimeter cells classed moderate-to-high in each year.

**Figure 3.** The flight window against the landscape's own climate, from hourly Environment and Climate Change Canada station records for May to September of the nine study years. The window, 1 July to 15 August, is the shaded band; it is taken from the bionomics and is not fitted to these data. (a) Mean afternoon temperature by day of year, with the 19 to 41 degrees C flight range and the 22 to 32 degrees C peak band marked. (b) The share of afternoon hours falling inside the flight range, day by day. (c) The diurnal curve, mean temperature and the share of all hours inside the flight range, by hour, with the 12:00 to 17:00 restriction shaded. (d) Mean wind speed by hour, on the same axis, showing that the hours the beetle can fly are also the windiest of the day.

**Figure 4.** The refugia mechanism as fitted. (a) Predicted probability of moderate-to-high disturbance against terrain-resolved epoch wind, at the 10th and 90th percentiles of stem density, all other terms held at their means; the lines cross, so wind raises attack in thin stands and lowers it in dense ones, which is the interaction. (b) The same for standing volume, where the interaction is a fifth the size and attenuates the dense-stand slope without reversing it, so the lines converge rather than cross over the observed range. (c) Coefficients of the 16-day model, with and without the within-season spread term. (d) Epoch prevalence against epoch mean wind, one point per epoch, years distinguished by both shape and shade so the panel remains legible in black and white.

```{r}
#| label: session
#| echo: false

sessionInfo()
```

```{r}
#| label: save-figures
#| echo: false
#| output: false

## Figures are copied to 03.outputs/PNG as the render finishes, so the outputs folder
## holds the current figures beside the current tables and neither can drift from the
## article. save_tbl() does the same for tables as each one prints. This must sit after
## the last figure chunk, because knitr writes a figure file only when its chunk runs.
##
## Quarto writes figures to <stem>_files/figure-<format>/ during a render and replays them
## from .quarto/_freeze/ when a chunk is frozen, so both are searched.

.pngdir <- c(
  file.path(here::here("01.manuscript"), "Manuscript-copy_files", "figure-docx"),
  file.path(here::here(".quarto", "_freeze", "01.manuscript", "Manuscript-copy"), "figure-docx"))
.out <- here::here("03.outputs", "PNG-beetle-fieldplots")
dir.create(.out, recursive = TRUE, showWarnings = FALSE)
for (.d in .pngdir) {
  if (!dir.exists(.d)) next
  .f <- list.files(.d, pattern = "\\.png$", full.names = TRUE)
  if (length(.f)) file.copy(.f, .out, overwrite = TRUE)
}
rm(.pngdir, .out, .d, .f)
```
